---
title: "Prepare_genotype_or_vcf"
output: html_document
---

# Note for the fastlmm code:
1. Use --double-id \ because strains in the PLINK matrix share the same FID and IID (e.g., SACE_GAL   SACE_GAL) for all other than SNPs
2. Don't use --double-id \ because the strain name is split at the underscore in the plink matrix (e.g., SACE    GAL) for SNPs - don't


# 1. SNPs - 
The file I finally used was sent to me by Gauthier, this was the snp file used in the Pan transcriptome paper
I don't need to filter for any strain names, code will automatically consider only the strains present in the paper

``` file used

'SNPs': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/SNPs/SNPs_no_change_name_and_no_strain_filtering_only_MAF15_filtered/full969Matrix.SNPs.Biallelic.Var.DP10.99percNonMissing.MAF15.NoSubTelo20kb.plink'

}
```

## Only MAF filtering without any strain filtering - no change of name and no strain filtering

```{}
plink --bfile full969Matrix.SNPs.Biallelic.Var.DP10.99percNonMissing.MAF5.NoSubTelo20kb.plink \
      --maf 0.15 \
      --make-bed \
      --out /projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/SNPs/SNPs_no_change_name_and_no_strain_filtering_only_MAF15_filtered/full969Matrix.SNPs.Biallelic.Var.DP10.99percNonMissing.MAF15.NoSubTelo20kb.plink


```


# 2. CNVs - again this CNV file was used in the Pan transcriptome paper

``` file used 

        'CNVs': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/CNVs/CNVs_no_change_name_and_no_strain_filtering_only_MAF15_filtered/CNV_969strains_gain_maf15_N400_20221121.plink',

```

## Only MAF filtering without any strain filtering - no change of name and no strain filtering

```{}
plink --bfile CNV_969strains_gain_maf5_N400_20221121.plink \
      --maf 0.15 \
      --make-bed \
      --out /projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/CNVs/CNVs_no_change_name_and_no_strain_filtering_only_MAF15_filtered/CNV_969strains_gain_maf15_N400_20221121.plink
```




# 3. Indels - The vcf file was sent by Anne, it contained only 96 strains of our interest


## step 1. make a plink file with everything, bad idea to filter anything here because the way I named strains could be different than the way vcf file has named and you don't want it to filter additional things.. -- geno 0.01 -- filtering missing genotype rate greater than 1%.. keeps only the variant where at least 99% samples have a called genotype. 

```
plink --vcf 96strains.vcf \
       --make-bed \
       --allow-extra-chr \
       --maf 0.15 \
       --geno 0.01 \
       --out ./SNPs_indels/96strains.snps.indels.plink
```
```{r}
#first load in the .fam SNPs and Indels file and convert the .fam file where if the family_ID matches Individual_ID you don't change anything but if they don't match then you rename both the rows such that you join family and Individual IDs i.e., if family_ID is SACE and individual_ID is YCY you make the rows to be SACE_YCY in both family and individual IDs. 

fam_file <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SNPs_indels/96strains.snps.indels.plink.fam"
library(data.table)

# Load .fam file (6 columns: FID, IID, PID, MID, SEX, PHENO)
fam <- fread(fam_file, header = FALSE, col.names = c("FID", "IID", "PID", "MID", "SEX", "PHENO"))

# Rename FID and IID where they don't match
fam[FID != IID, `:=`(FID = paste(FID, IID, sep = "_"),
                      IID = paste(FID, IID, sep = "_"))]

#convert AVI_1a to AVI
if ("AVI_1a" %in% fam$FID) {
  fam[FID == "AVI_1a", `:=`(FID = "AVI", IID = "AVI")]
  cat("AVI_1a converted to AVI\n")
} else {
  cat("AVI_1a not found, no changes made\n")
}

#first check if 96 strains exists in this file
genotypes_96 <- read.table(file = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/sample_used_two_column.txt")
colnames(genotypes_96) <- c("familyID", "IndividualID")

# Check which 96 strains are in the fam file
in_fam <- genotypes_96$IndividualID %in% fam$IID

cat("Strains found in fam file:", sum(in_fam), "\n")
cat("Strains NOT found in fam file:", sum(!in_fam), "\n")

# See which ones are missing
missing_strains <- genotypes_96$IndividualID[!in_fam]
print(missing_strains)

##copy bim and bed file into this folder 
# Write out the modified .fam file
fwrite(fam, file = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SNPs_indels/modified_snp_indel_plink_for_96strains/96strains.snps.indels.plink.fam", sep = " ", col.names = FALSE)


#step 2 update the .bim file, first check if the chromsome is in the integer format if not convert to integer.
# Load the .bim file
bim_file <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SNPs_indels/modified_snp_indel_plink_for_96strains/96strains.snps.indels.plink.bim"
bim <- read.table(bim_file, header=FALSE, stringsAsFactors=FALSE)
colnames(bim) <- c("CHR", "SNP", "cM", "BP", "A1", "A2")
head(bim)

# Create SNP ID as chr_pos_ref_alt (ref=allele2, alt=allele1)
bim$SNP <- paste(bim$CHR, bim$BP, bim$A2, bim$A1, sep = "_")

# Check chromosome format
cat("Current CHR values (unique):\n")
print(unique(bim$CHR))
cat("CHR class:", class(bim$CHR), "\n")

# Remove "Chromosome" prefix and convert to integer
bim$CHR <- as.integer(gsub("chromosome", "", bim$CHR))

# Verify
cat("Unique CHR values after conversion:\n")
print(unique(bim$CHR))
cat("Any NAs introduced:", sum(is.na(bim$CHR)), "\n")

head(bim)

# Write out the modified .fam file
fwrite(bim, file = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SNPs_indels/modified_snp_indel_plink_for_96strains/96strains.snps.indels.plink.bim", sep = " ", col.names = FALSE)
```


``` file used

'Indels': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SNPs_indels/modified_snp_indel_plink_for_96strains/Indels_only_96/indels.96samples',
       
```

##Step 2
### Make it snp only. Can't do everything in the first step because write-snplist can only be used on a plink file and not on the vcf file as an input
```{}
plink --bfile 96strains.snps.indels.plink \
       --make-bed \
       --snps-only \
       --write-snplist \
       --out ./SNPs_only_96/snps.96samples

```

### Make it indels only
``` {}
plink --bfile 96strains.snps.indels.plink \
      --exclude ./SNPs_only/snps.85samples.snplist \
      --make-bed \
      --allow-extra-chr \
      --out ./Indels_only_96/indels.96samples
```


#4. SVs 


``` file used

'SVs': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SVs_from_SV.1087/SV.1087samples_updated_fam_files/SV.1087samples.SVs',
   
```

### Step 1. 
```
##SVs are filtered in the same way, previously the SVs were used as provided in this file: /projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SV.1087Samples.vcf

###Step 1. create plink file from SV vcf file -- REMEMBER I have to filter for 0.15 MAF

plink --vcf SV.1087Samples.vcf \
       --make-bed \
       --allow-extra-chr \
       --maf 0.15 \
       --out ./SVs_from_SV.1087/SV.1087samples.SVs

```
```{r}
library(data.table)

# =============================================================================
# STEP 1: Update .fam file
# =============================================================================

fam_file <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SVs_from_SV.1087/SV.1087samples.SVs.fam"

fam <- fread(fam_file, header = FALSE, col.names = c("FID", "IID", "PID", "MID", "SEX", "PHENO"))

# Rename FID and IID where they don't match by joining them
fam[FID != IID, `:=`(FID = paste(FID, IID, sep = "_"),
                      IID = paste(FID, IID, sep = "_"))]

# Convert AVI_1a to AVI
if ("AVI_1a" %in% fam$FID) {
  fam[FID == "AVI_1a", `:=`(FID = "AVI", IID = "AVI")]
  cat("AVI_1a converted to AVI\n")
} else {
  cat("AVI_1a not found, no changes made\n")
}

# Check if 96 strains exist in this file
genotypes_96 <- read.table(file = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/sample_used_two_column.txt")
colnames(genotypes_96) <- c("familyID", "IndividualID")

in_fam <- genotypes_96$IndividualID %in% fam$IID
cat("Strains found in fam file:", sum(in_fam), "\n")
cat("Strains NOT found in fam file:", sum(!in_fam), "\n")
missing_strains <- genotypes_96$IndividualID[!in_fam]
print(missing_strains)

fwrite(fam, 
       file = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SVs_from_SV.1087/SV.1087samples_updated_fam_files/SV.1087samples.SVs.fam", 
       sep = "\t", col.names = FALSE)
```

```{r}
# =============================================================================
# STEP 2: Update .bim file
# =============================================================================

bim_file <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SVs_from_SV.1087/SV.1087samples.SVs.bim"

bim <- read.table(bim_file, header = FALSE, stringsAsFactors = FALSE)
colnames(bim) <- c("CHR", "SNP", "cM", "BP", "A1", "A2")

# Check chromosome format
cat("Current CHR values (unique):\n")
print(unique(bim$CHR))
cat("CHR class:", class(bim$CHR), "\n")

# Remove "chromosome" prefix and convert to integer
bim$CHR <- as.integer(gsub("chromosome", "", bim$CHR))
cat("Unique CHR values after conversion:\n")
print(unique(bim$CHR))
cat("Any NAs introduced:", sum(is.na(bim$CHR)), "\n")

# Replace allele sequences with generic ref/alt indicators
# 1 = reference allele, 2 = alternate allele
# The original sequences are too long for bed_reader to parse,
# and FaST-LMM does not use allele identity for the association test
bim$A1 <- 1  # reference
bim$A2 <- 2  # alternate

cat("A1/A2 columns replaced with 1 (ref) and 2 (alt)\n")
head(bim)

# Save updated BIM with new filename to distinguish from original
fwrite(bim, 
       file = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SVs_from_SV.1087/SV.1087samples_updated_fam_files/SV.1087samples.SVs.bim", 
       sep = "\t", col.names = FALSE)

cat("Done. BIM written with A1=1 (ref), A2=2 (alt)\n")
```

I don't need to filter and can use the same file
