---
title: "1.cell_cycle_assignment.Rmd"
author: "Akriti Agrawal"
date: "2025-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For finding eQTLs in cell cycle phase there are a few steps
1. Assignment of cell cycle phase which we do on normalized data
2. Then we use this assignment on raw seurat object for pseudobulking. Pseudobulking is always on raw data but before that as a separate step we do need to generate normalized seurat object to have it assigned with cell cycle phases 



# Load packages
```{r}
# Install BiocManager if you don't have it
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# # Install the organism database
# BiocManager::install("org.Sc.sgd.db")
library(org.Sc.sgd.db)
library(Seurat)
library(ggplot2)
library(dplyr)

# Be explicit if needed
select <- dplyr::select
filter <- dplyr::filter
```

#Load the demuxlet output
```{r}
data <- read.delim("../../demuxlet/merged_demuxlet_output_YPD.tsv")
head(data)
str(data)

data$BEST.LLK <- as.numeric(data$BEST.LLK)
data$NUM.READS <- as.numeric(data$NUM.READS)
data$NUM.SNPS <- as.numeric(data$NUM.SNPS)
data$NEXT.LLK <- as.numeric(data$NEXT.LLK)
data$DIFF.LLK.BEST.NEXT <- as.numeric(data$DIFF.LLK.BEST.NEXT)
data$BEST.POSTERIOR <- as.numeric(data$BEST.POSTERIOR)
data$SNG.POSTERIOR <- as.numeric(data$SNG.POSTERIOR)
data$SNG.BEST.LLK <- as.numeric(data$SNG.BEST.LLK)
data$SNG.NEXT.LLK <- as.numeric(data$SNG.NEXT.LLK)
data$SNG.ONLY.POSTERIOR <- as.numeric(data$SNG.ONLY.POSTERIOR)
data$DBL.BEST.LLK <- as.numeric(data$DBL.BEST.LLK)
data$DIFF.LLK.SNG.DBL <- as.numeric(data$DIFF.LLK.SNG.DBL)
data <- na.omit(data)
```

```{r}
expression_data <- Read10X(data.dir = "../../alignment/output_of_alignment_YPD1_with_clean_gtf/run_count/outs/filtered_feature_bc_matrix/")
Na_va_data <- CreateSeuratObject(counts = expression_data, project = "natural variants", min.cells = 3, min.features = 200)
head(Na_va_data@meta.data)
print(dim(Na_va_data))

GetAssayData(Na_va_data, slot = "counts")
```
# Add the cell identities
Add the cell identities assigned by demuxlet to the metadata of the seurat object Na_va_data.
```{r}
# Add all the columns of the data dataframe to the metadata of the Seurat object
#Match the rows of demuxlet assignment "data" with the seurat metadata colum "BARCODE". because both have common barcodes
metadata_to_add <- data[match(colnames(Na_va_data), data$BARCODE), ]
#Add the matching demuxlet identities to the metadata
Na_va_data <- AddMetaData(Na_va_data, metadata = metadata_to_add)
```

## Perfrom qc from seurat pipeline. 
### Step.1 : Filter the cells with low or suspiciously high ncount_RNA
```{r}
# Filter cells based on nCount_RNA and nFeature_RNA
#Na_va_data_subsetted <- subset(Na_va_data, subset = nFeature_RNA > 200) Don't need it
```


###  Step.2 : Exclude the entries forKANMX and NATMX. 

We do this because DGY3 was part of a separate pool and it formed a different cluster from all the other strains. We removed KANMX and NATMX because these barcoded antibiotic markers were added in the DGY3 to identify them. 
```{r}

# Check the dimensions of the filtered Seurat object
print(dim(Na_va_data))

# Optionally, check the first few rows of the metadata to confirm the filtering
head(Na_va_data@meta.data)

# Remove specific DGY3
Na_va_data_subsetted <- subset(Na_va_data, subset = SNG.BEST.GUESS != "DGY3")

# Check the dimensions of the filtered Seurat object
print(dim(Na_va_data_subsetted))

```

### Step 3. Remove entries containing AMB and DBL droplet.type. 

We'll be using the demuxlet identities for cluster identification and we want to be sure to keep only the unique identities
```{r}
# Subset Na_va_data to include only entries where DROPLET.TYPE is "AMB" or "DBL"
Na_va_data_subsetted <- subset(Na_va_data_subsetted, !(subset = DROPLET.TYPE %in% c("AMB", "DBL")))

# Check the dimensions of the filtered Seurat object
print(dim(Na_va_data_subsetted))

# Optionally, check the first few rows of the metadata to confirm the filtering
head(Na_va_data_subsetted@meta.data)

#Optional - you can also set identities of the cluster using Idents
Na_va_data_subsetted <- SetIdent(Na_va_data_subsetted, value = Na_va_data_subsetted$SNG.BEST.GUESS)

saveRDS(Na_va_data_subsetted, file = "../data/Na_va_data_subsetted.rds")
```

### Step 4. Normalization - 

Aims to adjust for differences in cellular sequencing depth, due to fluctuations in cellular RNA content, efficiency in lysis and reverse transcription and stochastix samling during next-gen sequencing. It also aims to do variance stabilization - to address the relationship between gene abundance and gene variance. And to ensure both lowly and highly expressed genes can contribute to the cellular states

```{r}
###Normalization using log normalization and scaling of the data
Normalized_Na_va_data <- NormalizeData(Na_va_data_subsetted)
Normalized_Na_va_data <- FindVariableFeatures(Normalized_Na_va_data, selection.method = "vst", nfeatures = 2000)
# Scale the data
all.genes <- rownames(Normalized_Na_va_data)
Normalized_Na_va_data <- ScaleData(Normalized_Na_va_data, features = all.genes)
```


### Step 5. - PCA

```{r}
# Perform linear dimensional reduction (PCA)
Normalized_Na_va_data <- RunPCA(Normalized_Na_va_data, features = VariableFeatures(object = Normalized_Na_va_data))

# Save the normalized Seurat object
saveRDS(Normalized_Na_va_data, file = "../data/Normalized_Na_va_data.rds")
```


## Cell cycle gene assingment


### Step 1. Clustering - Cell clustering is on the basis of PCA scores -- each PC representing a metafeature that combines information across a correlated featue set

1. FindNeighbors -- constructs KNN graph based on the euclidean distance in PCA space, refine the edge weights between any two cells based on shared overlap in their local neighboorhood (jaccard similarity -- if the two cells have same number of similar features they are more similar then the other)

2. Find clusters -- Actuall clustering -- Louvain algorithm -- to iteratively group cells together, with the goal of optimizing the standard modularity function.

```{r}
Normalized_Na_va_data_clustered_for_Cell_cycle <- FindNeighbors(Normalized_Na_va_data, dims = 1:5)
Normalized_Na_va_data_clustered_for_Cell_cycle  <- FindClusters(Normalized_Na_va_data_clustered_for_Cell_cycle , resolution = 0.5)
```

#Plot the UMAP
```{r}
Normalized_Na_va_data_clustered_for_Cell_cycle <- RunUMAP(Normalized_Na_va_data_clustered_for_Cell_cycle, dims = 1:5)
DimPlot(Normalized_Na_va_data_clustered_for_Cell_cycle, reduction = "umap")
```

### Step 2. Find differentially expressed markers
For every cluster compared to all remaining cells, report only the positive ones

```{r}
Normalized_Na_va_data_clustered_for_Cell_cycle.markers <- FindAllMarkers(Normalized_Na_va_data_clustered_for_Cell_cycle, only.pos = TRUE)

Normalized_Na_va_data_clustered_for_Cell_cycle.markers %>% 
  group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 0.2 & p_val_adj < 0.05)

```

### Step 3. Add common name and ORF column to the list
```{r}
add_gene_annotation_columns <- function(markers_df) {
  # Extract all gene names
  gene_names <- markers_df$gene
  
  # Initialize new columns
  markers_df$orf_id <- NA
  markers_df$common_name <- NA
  
  # Identify which genes are likely already ORFs (systematic names)
  is_orf <- grepl("^Y[A-Z]{2}\\d", gene_names)
  
  # Set ORF IDs for genes already in ORF format
  markers_df$orf_id[is_orf] <- gene_names[is_orf]
  
  # Set common names for genes not in ORF format (already common names)
  markers_df$common_name[!is_orf] <- gene_names[!is_orf]
  
  # Look up common names for ORFs
  if(sum(is_orf) > 0) {
    orf_common_mapping <- tryCatch({
      AnnotationDbi::select(org.Sc.sgd.db, 
                            keys = gene_names[is_orf],
                            columns = "COMMON", 
                            keytype = "ORF")
    }, error = function(e) {
      warning("Error mapping ORFs to common names: ", e$message)
      return(NULL)
    })
    
    # Apply the mapping to our dataframe
    if(!is.null(orf_common_mapping)) {
      for(i in which(is_orf)) {
        orf <- gene_names[i]
        common <- orf_common_mapping$COMMON[orf_common_mapping$ORF == orf]
        if(length(common) > 0 && !is.na(common[1])) {
          markers_df$common_name[i] <- common[1]
        }
      }
    }
  }
  
  # Look up ORFs for common names
  if(sum(!is_orf) > 0) {
    common_orf_mapping <- tryCatch({
      AnnotationDbi::select(org.Sc.sgd.db, 
                            keys = gene_names[!is_orf],
                            columns = "ORF", 
                            keytype = "COMMON")
    }, error = function(e) {
      warning("Error mapping common names to ORFs: ", e$message)
      return(NULL)
    })
    
    # Apply the mapping to our dataframe
    if(!is.null(common_orf_mapping)) {
      for(i in which(!is_orf)) {
        common <- gene_names[i]
        orf <- common_orf_mapping$ORF[common_orf_mapping$COMMON == common]
        if(length(orf) > 0 && !is.na(orf[1])) {
          markers_df$orf_id[i] <- orf[1]
        }
      }
    }
  }
  
  # Report mapping statistics
  cat(sprintf("Gene annotation statistics:\n"))
  cat(sprintf("  - Total genes: %d\n", length(gene_names)))
  cat(sprintf("  - Original ORFs: %d\n", sum(is_orf)))
  cat(sprintf("  - Original common names: %d\n", sum(!is_orf)))
  cat(sprintf("  - Genes with resolved common names: %d\n", sum(!is.na(markers_df$common_name))))
  cat(sprintf("  - Genes with resolved ORF IDs: %d\n", sum(!is.na(markers_df$orf_id))))
  
  return(markers_df)
}

# Apply the function to your dataframe
Normalized_Na_va_data_clustered_for_Cell_cycle.markers <- add_gene_annotation_columns(Normalized_Na_va_data_clustered_for_Cell_cycle.markers)

write.table(Normalized_Na_va_data_clustered_for_Cell_cycle.markers, file = "../data/Normalized_Na_va_data_clustered_for_Cell_cycle.markers.txt")

# Normalized_Na_va_data_clustered_for_Cell_cycle.markers <- read.table("../data/Normalized_Na_va_data_clustered_for_Cell_cycle.markers.txt")

str(Normalized_Na_va_data_clustered_for_Cell_cycle.markers)
head(Normalized_Na_va_data_clustered_for_Cell_cycle.markers)
```

### Step 4. Annotate cell cycle using cell cycle annotation -- as annotate by Chris jackson
```{r}
cell_cycle_annotation <- read.delim("../../../cell_cycle_annotation.tsv")
str(cell_cycle_annotation)
table(cell_cycle_annotation$Group)
```


####Step 4.1 Define cell cycle groups to keep
```{r}
# Define the cell cycle groups to keep
cell_cycle_groups <- c("M/G1", "G1", "S", "G2", "M")

# Filter the table to include only cell cycle groups
cell_cycle_annotation <- cell_cycle_annotation[cell_cycle_annotation$Group %in% cell_cycle_groups, ]

# Print some information about the filtering
cat("Filtered table has", nrow(cell_cycle_annotation), "rows\n")
cat("Retained groups:", paste(unique(cell_cycle_annotation$Group), collapse = ", "), "\n")

cell_cycle_annotation$Group <- gsub("M/G1", "M-G1", cell_cycle_annotation$Group)

# If you want to write the filtered data back to a file
write.table(cell_cycle_annotation, file = "../data/CellCycle_Genes.tsv", sep = "\t", 
          row.names = FALSE, quote = FALSE)

cell_cycle_annotation <- read.table("../data/CellCycle_Genes.tsv", header = T)
str(cell_cycle_annotation)
table(cell_cycle_annotation$Group)
```

####Step 4.2 Perform gene-to-group mapping (which gene belongs to which cell cycle phase)

```{r}
# Make a copy of your current data frame
annotated_df <- Normalized_Na_va_data_clustered_for_Cell_cycle.markers

# Add a new column for cell cycle group
annotated_df$cell_cycle_group <- NA

# Annotate each gene with its cell cycle group
for (i in 1:nrow(annotated_df)) {
  # Get the current gene's ORF ID
  current_orf <- annotated_df$orf_id[i]
  
  # Find matching entry in cell cycle annotation
  match_idx <- which(cell_cycle_annotation$Gene == current_orf)
  
  # If a match is found, add the cell cycle group
  if (length(match_idx) > 0) {
    annotated_df$cell_cycle_group[i] <- cell_cycle_annotation$Group[match_idx[1]]
  }
}

# Summarize the data
table(annotated_df$cell_cycle_group, useNA = "ifany")

# You can also create a table showing how many genes from each cell cycle group 
# appear in each cluster
if ("cluster" %in% colnames(annotated_df)) {
  cell_cycle_by_cluster <- table(annotated_df$cluster, annotated_df$cell_cycle_group, useNA = "ifany")
  print(cell_cycle_by_cluster)
}

# Save the annotated data if needed
# write.table(annotated_df, file = "annotated_markers.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

####Step 4.3 Perform cell-to-group assignment (which cell cycle phase each cell is in)

For single-cell RNA-seq data, assigning cells to cell cycle phases involves:
1. Considering the combined expression of all genes in each cell cycle group
2. Determining which cell cycle program is most active in each cell


In Seurat objects, the data is organized in assays (like "RNA", "integrated", etc.), and within each assay, there are different "slots" or "layers" that contain different types of data:

"counts": The raw count data (unmodified, original counts)
"data": The normalized data (after normalization)
"scale.data": The scaled data (after scaling)

When we talk about "each phase might have a different baseline level of expression," we're addressing an important biological reality in gene expression data.
Different cell cycle phases (G1, S, G2, M, etc.) involve different biological processes, and the genes associated with each phase may have inherently different expression levels. For example:

Some phases might have genes that are naturally expressed at higher levels even in their "baseline" state
Some phases might have more genes associated with them than others
Some phases might have genes that are more widely expressed across many cell types

Without standardizing the scores, you might incorrectly assign cells to phases that simply have higher overall expression. For instance, if G1 genes are generally expressed at higher levels than S phase genes, cells might be incorrectly classified as G1 even when they're actually in S phase.

Z-scoring solves this problem by:

Subtracting the mean (centering the distribution around zero)
Dividing by the standard deviation (making the spread comparable)

After Z-scoring, what matters is not the absolute expression level but how much a cell's expression of a particular phase's genes deviates from the average across all cells. This allows for fair comparison between phases with different baseline expression characteristics.
For example, if a cell has:

A G1 score that's 1.5 standard deviations above the mean G1 score
An S score that's 2.0 standard deviations above the mean S score

The cell would be assigned to S phase because its expression of S phase genes is more unusual/distinctive relative to other cells, even if the raw G1 expression values might have been higher.

```{r}
# Create a list to store cell cycle gene sets
cell_cycle_gene_sets <- list()

# Populate the gene sets with genes from each category
for(group in cell_cycle_groups) {
  cell_cycle_gene_sets[[group]] <- cell_cycle_annotation$Gene[cell_cycle_annotation$Group == group]
}

# Initialize matrix to store scores
scores_matrix <- matrix(0, ncol = length(cell_cycle_gene_sets), nrow = ncol(Normalized_Na_va_data_clustered_for_Cell_cycle))
colnames(scores_matrix) <- names(cell_cycle_gene_sets)

# Get normalized expression matrix
expr_matrix <- GetAssayData(Normalized_Na_va_data_clustered_for_Cell_cycle, layer = "data")

# Calculate total counts per cell
total_counts <- colSums(expr_matrix)

# For each cell cycle group, calculate the score:
#This double loop is the core scoring calculation:
#For each cell cycle phase (outer loop)
#Get the list of genes for this phase
#Find which genes actually exist in our dataset
#For each cell (inner loop)
#Sum the expression of all genes in this phase
#Divide by the total expression in the cell (normalization)
#Store the result in the scores matrix
#This gives a normalized score representing what fraction of each cell's total expression comes from genes in each cell cycle #phase.

for(i in 1:length(cell_cycle_gene_sets)) {
  group <- names(cell_cycle_gene_sets)[i]
  genes <- cell_cycle_gene_sets[[group]]
  
  # Find genes present in the dataset
  genes_in_data <- intersect(genes, rownames(expr_matrix))
  
  if(length(genes_in_data) > 0) {
    # For each cell
    for(cell_idx in 1:ncol(expr_matrix)) {
      # Sum expression of genes in this cell cycle group
      cell_expression <- sum(expr_matrix[genes_in_data, cell_idx])
      
      # Normalize by total expression in this cell
      scores_matrix[cell_idx, i] <- cell_expression / total_counts[cell_idx]
    }
  }
}

#This standardizes each column of scores (each cell cycle phase):
#Calculate the mean and standard deviation of scores for each phase
#Convert each score to a z-score: (value - mean) / standard deviation
#Z-scoring makes the scores comparable across different phases, since each phase might have a different baseline level of expression.
# Z-score each column (for each cell cycle group)

for(i in 1:ncol(scores_matrix)) {
  col_mean <- mean(scores_matrix[, i])
  col_sd <- sd(scores_matrix[, i])
  scores_matrix[, i] <- (scores_matrix[, i] - col_mean) / col_sd
}


# Find the maximum score for each cell
cell_phases <- colnames(scores_matrix)[max.col(scores_matrix)]

#This finds which cell cycle phase has the highest z-score for each cell:
#max.col(scores_matrix) returns the column index with the maximum value for each row
#colnames(scores_matrix)[...] converts these indices to the corresponding phase names


```

####Step 4.4 Add cell cycle phases to seurat meta object
```{r}
# Add cell cycle phase to Seurat object metadata
Normalized_Na_va_data_clustered_for_Cell_cycle$cell_cycle_phase <- cell_phases
Normalized_Na_va_data_clustered_for_Cell_cycle$cell_cycle_phase <- factor(
  Normalized_Na_va_data_clustered_for_Cell_cycle$cell_cycle_phase, 
  levels = cell_cycle_groups
)

# Add individual scores to metadata
for(group in names(cell_cycle_gene_sets)) {
  col_name <- paste0("cc_", group)
  Normalized_Na_va_data_clustered_for_Cell_cycle[[col_name]] <- scores_matrix[, group]
}

```


####Step 4.5 #Save the data with cell cycle assignment 

```{r}
saveRDS(Normalized_Na_va_data_clustered_for_Cell_cycle, "../data/seurat_with_cell_cycle.rds")
#Normalized_Na_va_data_clustered_for_Cell_cycle <- readRDS("../data/seurat_with_cell_cycle.rds")
```


Visualize the cluster in dimplot
```{r}
# Show distribution of cells across cell cycle phases
print(table(Normalized_Na_va_data_clustered_for_Cell_cycle$cell_cycle_phase))

# Visualize cell cycle phases on a dimensionality reduction plot and save as PDF
if("umap" %in% names(Normalized_Na_va_data_clustered_for_Cell_cycle@reductions)) {
  # Open PDF device
  pdf("../results/cell_cycle_phase_plot.pdf", width = 10, height = 8)
  
   # Create the plot
cell_cycle_plot <- DimPlot(Normalized_Na_va_data_clustered_for_Cell_cycle, 
              reduction = "umap", 
              group.by = "cell_cycle_phase",
              pt.size = 0.5,
              label = TRUE,
              label.size = 5,
              repel = TRUE) +
        ggtitle("Cell Cycle Phase Classification")

# Get the same limits as your first plot
x_limits <- range(Normalized_Na_va_data_clustered_for_Cell_cycle@reductions$umap@cell.embeddings[, 1])
y_limits <- range(Normalized_Na_va_data_clustered_for_Cell_cycle@reductions$umap@cell.embeddings[, 2])
max_limit <- max(abs(x_limits), abs(y_limits))

# Apply the same limits
cell_cycle_plot <- cell_cycle_plot + 
  coord_cartesian(xlim = c(-max_limit, max_limit), ylim = c(-max_limit, max_limit))

# Now print the plot
print(cell_cycle_plot)
  
  # Close the PDF device
  dev.off()
  
  # Also display the plot in the R session
  print(DimPlot(Normalized_Na_va_data_clustered_for_Cell_cycle, 
                reduction = "umap", 
                group.by = "cell_cycle_phase"))
  
  cat("Plot saved as 'cell_cycle_phase_plot.pdf'\n")
}

# # You can also save the cell counts to a separate file
# write.csv(table(Normalized_Na_va_data_clustered_for_Cell_cycle$cell_cycle_phase),
#           "../results/seurat_outputs/cell_cycle_counts.csv")


```


