---
title: "Comprehensive_eQTL_analysis"
output: html_document
---

# ============================================================================
# UNIFIED eQTL ANALYSIS: ALL HITS (SHARED + NOT SHARED FROM BOTH STUDIES)
# 
# Purpose: Create one master dataframe with all eQTLs from both studies
#          Analyze -log10(p-value) and threshold proximity together
#          Create scatter plot: Reference -log10(p-value) vs Your -log10(p-value)
# ============================================================================
```{r}
library(tidyverse)
library(ggplot2)
library(gridExtra)
```


```{r}
all_signif_snps_0.05 <- read.table("../../results/all_signif_SNP_71_with_threshold.txt", 
                           header = TRUE, 
                           sep = "\t", 
                           stringsAsFactors = FALSE)

# Convert dots to dashes in Pheno column
all_signif_snps_0.05 <- all_signif_snps_0.05 %>%
  mutate(Pheno = gsub("\\.", "-", Pheno))

#remove the Source_Phenotype that starts with X and a number
all_signif_snps_0.05_cleaned <- all_signif_snps_0.05 %>%
  filter(!grepl("^[X0-9]", Pheno))

# Show summary
cat("Total associations:", nrow(all_signif_snps_0.05_cleaned), "\n")
cat("Unique genes:", length(unique(all_signif_snps_0.05_cleaned$Pheno)), "\n")


all_signif_snps_0.05_cleaned$Chr_ChrPos_Pheno <- paste0(all_signif_snps_0.05_cleaned$Chr, "_", 
                                                    all_signif_snps_0.05_cleaned$ChrPos, "_", 
                                                    all_signif_snps_0.05_cleaned$Pheno)


```

```{r}

pan_transcriptome <- read.csv("/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/pan_transcriptome/data/downloaded_data_from_pan_transcriptome_paper/GWAS_combined_lgcCorr_ldPruned_noBonferroni_20221207.tab", header=TRUE)
# #create a new column where we concatenate ChrPos and Pheno and call it ChrPos_Pheno
pan_transcriptome$Chr_ChrPos_Pheno <- paste0(pan_transcriptome$Chr, "_",
                                                    pan_transcriptome$ChrPos, "_",
                                                    pan_transcriptome$Pheno)
#filter out the snp eQTLs detected
snp_pan_transcriptome <- pan_transcriptome[pan_transcriptome$subtype == "SNP", ]
snp_pan_transcriptome

```

# ============================================================================
# STEP 1: CREATE UNIFIED DATAFRAME
# ============================================================================
```{r}

# Prepare YOUR data
your_data_prep <- all_signif_snps_0.05_cleaned %>%
  mutate(
    Your_neglog10p = -log10(PValue),
    Your_thr_neglog10p = -log10(thr_perm),
    Your_distance_above_threshold = Your_neglog10p - Your_thr_neglog10p,
    Chr_ChrPos_Pheno = paste0(Chr, "_", ChrPos, "_", Pheno)
  ) %>%
  dplyr::select(Chr_ChrPos_Pheno, Your_neglog10p, Your_thr_neglog10p, 
                Your_distance_above_threshold, Pheno, Chr, ChrPos, 
                PValue, thr_perm, Your_SnpWeight = SnpWeight)  # <-- rename on select

# Prepare REFERENCE data
ref_data_prep <- snp_pan_transcriptome %>%
  mutate(
    Ref_neglog10p = -log10(PValue),
    Ref_thr_neglog10p = -log10(thr_perm),
    Ref_distance_above_threshold = Ref_neglog10p - Ref_thr_neglog10p,
    Chr_ChrPos_Pheno = paste0(Chr, "_", ChrPos, "_", Pheno)
  ) %>%
  dplyr::select(Chr_ChrPos_Pheno, Ref_neglog10p, Ref_thr_neglog10p, 
                Ref_distance_above_threshold, Pheno, Chr, ChrPos, 
                PValue, thr_perm, Ref_SnpWeight = SnpWeight)  # <-- rename on select


```
# ============================================================================
# STEP 2: CREATE COMPREHENSIVE UNIFIED DATAFRAME
# ============================================================================
```{r}
# Step 1: Mark which eQTLs are in your study
your_ids <- your_data_prep$Chr_ChrPos_Pheno
ref_ids <- ref_data_prep$Chr_ChrPos_Pheno

# Step 2: Start with ALL eQTLs from reference
all_eqtls <- ref_data_prep %>%
  mutate(
    In_your_study = Chr_ChrPos_Pheno %in% your_ids,
    Category = ifelse(In_your_study, "SHARED", "REFERENCE-ONLY")
  )

# Step 3: Add YOUR data for SHARED eQTLs via left_join
all_eqtls <- all_eqtls %>%
  left_join(
    your_data_prep %>% dplyr::select(Chr_ChrPos_Pheno, Your_neglog10p, Your_thr_neglog10p, 
                                     Your_distance_above_threshold, Your_SnpWeight),  # <-- add here
    by = "Chr_ChrPos_Pheno"
  )

# Step 4: Create YOUR-ONLY eQTLs (in your data but NOT in reference)
your_only_ids <- setdiff(your_ids, ref_ids)
your_only_data <- your_data_prep %>%
  filter(Chr_ChrPos_Pheno %in% your_only_ids) %>%
  mutate(
    Category = "YOUR-ONLY",
    In_your_study = TRUE,
    Ref_neglog10p = NA,
    Ref_thr_neglog10p = NA,
    Ref_distance_above_threshold = NA,
    Ref_SnpWeight = NA                  # <-- add here
  ) %>%
  dplyr::select(Chr_ChrPos_Pheno, Ref_neglog10p, Ref_thr_neglog10p, Ref_distance_above_threshold,
                Ref_SnpWeight,                    # <-- add here
                Your_neglog10p, Your_thr_neglog10p, Your_distance_above_threshold,
                Your_SnpWeight,                   # <-- add here
                Pheno, Chr, ChrPos, PValue, thr_perm, Category, In_your_study)
# Step 5: Combine all three categories
unified_data <- bind_rows(all_eqtls, your_only_data)

cat("Total eQTLs in unified dataset: ", nrow(unified_data), "\n\n")

cat("Categories:\n")
cat("  SHARED (in both studies): ", sum(unified_data$Category == "SHARED"), "\n")
cat("  REFERENCE-ONLY (in reference, not you): ", sum(unified_data$Category == "REFERENCE-ONLY"), "\n")
cat("  YOUR-ONLY (in you, not reference): ", sum(unified_data$Category == "YOUR-ONLY"), "\n\n")
```
```{r}
library(ggplot2)
library(dplyr)
library(scales)

# ── 1. Build summary dataframe ────────────────────────────────────────────────
unified_summary <- data.frame(
  Category = c("SHARED", "Caudel_et_al_Only", "My_analysis_Only"),
  Count    = c(
    sum(unified_data$Category == "SHARED"),
    sum(unified_data$Category == "REFERENCE-ONLY"),
    sum(unified_data$Category == "YOUR-ONLY")
  )
) %>%
  mutate(Category = factor(Category,
                           levels = c("SHARED", "Caudel_et_al_Only", "My_analysis_Only")))

# ── 2. Compute totals for title ───────────────────────────────────────────────
n_reference  <- sum(unified_data$Category %in% c("SHARED", "REFERENCE-ONLY"))
n_reproduced <- sum(unified_data$Category %in% c("SHARED", "YOUR-ONLY"))
n_unified    <- nrow(unified_data)

# ── 3. Colors ─────────────────────────────────────────────────────────────────
category_colors <- c(
  "SHARED"             = "#4DAC26",   # publication-friendly green (replaces base "green")
  "Caudel_et_al_Only"  = "#8B2500",
  "My_analysis_Only"   = "#2171B5"
)

# ── 4. Clean x-axis labels (avoids overlap without rotation) ──────────────────
x_labels <- c(
  "SHARED"             = "Shared",
  "Caudel_et_al_Only"  = "Caudal\net al. only",
  "My_analysis_Only"   = "This study\nonly"
)

# ── 5. Plot ───────────────────────────────────────────────────────────────────
p_unified <- ggplot(unified_summary, aes(x = Category, y = Count, fill = Category)) +
  geom_bar(stat = "identity", width = 0.55,
           color = "white", linewidth = 0.4) +
  geom_text(aes(label = scales::comma(Count)),
            vjust = -0.6, size = 4.5,
            color = "black", fontface = "bold") +
  scale_fill_manual(values = category_colors) +
  scale_x_discrete(labels = x_labels) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    title   = paste0("Unified SNP-eQTL dataset (n = ", scales::comma(n_unified), ")"),
    x       = NULL,
    y       = "Number of SNP-eQTLs",
    caption = "Shared: eQTLs recovered in both this study and Caudal et al."
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title         = element_text(size = 13, face = "bold", color = "black",
                                      hjust = 0.5, margin = margin(b = 4)),
    axis.text.x        = element_text(size = 12, color = "black", lineheight = 1.1),
    axis.text.y        = element_text(size = 12, color = "black"),
    axis.title.y       = element_text(size = 13, color = "black", face = "bold",
                                      margin = margin(r = 10)),
    plot.caption       = element_text(size = 9, color = "grey30", hjust = 0,
                                      margin = margin(t = 8)),
    legend.position    = "none",
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.4),
    panel.grid.minor.y = element_blank(),
    axis.line          = element_line(color = "black", linewidth = 0.5),
    axis.ticks         = element_line(color = "black", linewidth = 0.5),
    plot.margin        = margin(14, 18, 12, 14)
  )

p_unified

# ── 6. Save as TIFF ───────────────────────────────────────────────────────────
ggsave(
  filename    = "../../results/publication/unified_snp_eqtl_71_sc_bulk_paper.tiff",
  plot        = p_unified,
  device      = "tiff",
  width       = 4.5,
  height      = 4.5,
  units       = "in",
  dpi         = 300,
  compression = "lzw"
)

message("Saved: unified_snp_eqtl.tiff")
```
# ============================================================================
# STEP 3: SUMMARY STATISTICS BY CATEGORY
# ============================================================================
```{r}
cat("STEP 3: Summary Statistics by Category\n")
cat("──────────────────────────────────────\n\n")

# SHARED eQTLs
shared_df <- unified_data %>% filter(Category == "SHARED")
cat("SHARED eQTLs (n=", nrow(shared_df), "):\n")
cat("  Your -log10(p-value) - Mean: ", round(mean(shared_df$Your_neglog10p, na.rm=T), 3),
    " | Median: ", round(median(shared_df$Your_neglog10p, na.rm=T), 3),
    " | SD: ", round(sd(shared_df$Your_neglog10p, na.rm=T), 3), "\n")
cat("  Reference -log10(p-value) - Mean: ", round(mean(shared_df$Ref_neglog10p, na.rm=T), 3),
    " | Median: ", round(median(shared_df$Ref_neglog10p, na.rm=T), 3),
    " | SD: ", round(sd(shared_df$Ref_neglog10p, na.rm=T), 3), "\n")
cat("  Your distance above threshold - Mean: ", round(mean(shared_df$Your_distance_above_threshold, na.rm=T), 3), "\n")
cat("  Reference distance above threshold - Mean: ", round(mean(shared_df$Ref_distance_above_threshold, na.rm=T), 3), "\n\n")

# REFERENCE-ONLY eQTLs
ref_only_df <- unified_data %>% filter(Category == "REFERENCE-ONLY")
cat("REFERENCE-ONLY eQTLs (n=", nrow(ref_only_df), "):\n")
cat("  Reference -log10(p-value) - Mean: ", round(mean(ref_only_df$Ref_neglog10p, na.rm=T), 3),
    " | Median: ", round(median(ref_only_df$Ref_neglog10p, na.rm=T), 3),
    " | SD: ", round(sd(ref_only_df$Ref_neglog10p, na.rm=T), 3), "\n")
cat("  Reference distance above threshold - Mean: ", round(mean(ref_only_df$Ref_distance_above_threshold, na.rm=T), 3), "\n\n")

# YOUR-ONLY eQTLs
your_only_df <- unified_data %>% filter(Category == "YOUR-ONLY")
cat("YOUR-ONLY eQTLs (n=", nrow(your_only_df), "):\n")
cat("  Your -log10(p-value) - Mean: ", round(mean(your_only_df$Your_neglog10p, na.rm=T), 3),
    " | Median: ", round(median(your_only_df$Your_neglog10p, na.rm=T), 3),
    " | SD: ", round(sd(your_only_df$Your_neglog10p, na.rm=T), 3), "\n")
cat("  Your distance above threshold - Mean: ", round(mean(your_only_df$Your_distance_above_threshold, na.rm=T), 3), "\n\n")

```
# ============================================================================
# STEP 4: -log10(p-value) CORRELATION ANALYSIS (SHARED ONLY)
# ============================================================================
```{r}

# Remove NAs for correlation
shared_clean <- shared_df %>%
  filter(!is.na(Your_neglog10p) & !is.na(Ref_neglog10p))

cor_neglog10p_shared <- cor(shared_clean$Your_neglog10p, shared_clean$Ref_neglog10p, method="pearson")
cor_test_shared <- cor.test(shared_clean$Your_neglog10p, shared_clean$Ref_neglog10p, method="pearson")

cat("SHARED eQTLs -log10(p-value) Correlation:\n")
cat("  Pearson r: ", round(cor_neglog10p_shared, 4), "\n")
cat("  p-value: ", format(cor_test_shared$p.value, scientific=TRUE), "\n")
cat("  95% CI: [", round(cor_test_shared$conf.int[1], 4), ", ", 
    round(cor_test_shared$conf.int[2], 4), "]\n\n")

if (cor_neglog10p_shared > 0.90) {
  cat("✓✓✓ NEARLY PERFECT correlation\n")
  cat("    Your and reference detect SAME signals with same strength\n\n")
} else if (cor_neglog10p_shared > 0.80) {
  cat("✓✓ VERY STRONG correlation\n")
  cat("    Your and reference detect SAME signals\n\n")
} else if (cor_neglog10p_shared > 0.70) {
  cat("✓ STRONG correlation\n")
  cat("    Your and reference detect mostly same signals\n\n")
} else {
  cat("Moderate correlation\n")
  cat("    Some differences in signal strength\n\n")
}
```
# ============================================================================
# STEP 5: THRESHOLD PROXIMITY - ALL CATEGORIES
# ============================================================================
```{r}
cat("Distance above threshold (how far from failing):\n\n")

cat("SHARED:\n")
cat("  Your distance - Mean: ", round(mean(shared_df$Your_distance_above_threshold, na.rm=T), 3),
    " | Median: ", round(median(shared_df$Your_distance_above_threshold, na.rm=T), 3), "\n")
cat("  Ref distance - Mean: ", round(mean(shared_df$Ref_distance_above_threshold, na.rm=T), 3),
    " | Median: ", round(median(shared_df$Ref_distance_above_threshold, na.rm=T), 3), "\n\n")

cat("REFERENCE-ONLY:\n")
cat("  Ref distance - Mean: ", round(mean(ref_only_df$Ref_distance_above_threshold, na.rm=T), 3),
    " | Median: ", round(median(ref_only_df$Ref_distance_above_threshold, na.rm=T), 3), "\n\n")

cat("YOUR-ONLY:\n")
cat("  Your distance - Mean: ", round(mean(your_only_df$Your_distance_above_threshold, na.rm=T), 3),
    " | Median: ", round(median(your_only_df$Your_distance_above_threshold, na.rm=T), 3), "\n\n")
```

# ============================================================================
# STEP 7: COMPARISON TABLE
# ============================================================================
```{r}
comparison_table <- data.frame(
  Category = c("SHARED", "REFERENCE-ONLY", "YOUR-ONLY"),
  Count = c(nrow(shared_df), nrow(ref_only_df), nrow(your_only_df)),
  Your_Mean_neglog10p = c(round(mean(shared_df$Your_neglog10p, na.rm=T), 3),
                    "N/A",
                    round(mean(your_only_df$Your_neglog10p, na.rm=T), 3)),
  Ref_Mean_neglog10p = c(round(mean(shared_df$Ref_neglog10p, na.rm=T), 3),
                   round(mean(ref_only_df$Ref_neglog10p, na.rm=T), 3),
                   "N/A"),
  Your_Mean_Distance = c(round(mean(shared_df$Your_distance_above_threshold, na.rm=T), 3),
                         "N/A",
                         round(mean(your_only_df$Your_distance_above_threshold, na.rm=T), 3)),
  Ref_Mean_Distance = c(round(mean(shared_df$Ref_distance_above_threshold, na.rm=T), 3),
                        round(mean(ref_only_df$Ref_distance_above_threshold, na.rm=T), 3),
                        "N/A")
)

print(comparison_table)
cat("\n")

```
```{r}
library(ggplot2)
library(tidyverse)

# ── Color scheme ──────────────────────────────────────────────────────────────
category_colors <- c(
  "SHARED"             = "green3",
  "Caudel_et_al_Only"  = "skyblue",
  "My_analysis_Only"   = "orange"
)

clean_labels <- c(
  "SHARED"             = "Shared",
  "Caudel_et_al_Only"  = "Caudal et al. only",
  "My_analysis_Only"   = "This study only"
)

# ── Prepare data ──────────────────────────────────────────────────────────────
shared_df    <- unified_data %>% filter(Category == "SHARED")
ref_only_df  <- unified_data %>% filter(Category == "REFERENCE-ONLY")
your_only_df <- unified_data %>% filter(Category == "YOUR-ONLY")

# ── Median annotation dataframes ─────────────────────────────────────────────
means_1a <- data.frame(
  Source     = c("SHARED", "Caudel_et_al_Only", "My_analysis_Only"),
  median_val = c(
    median(shared_df$Your_neglog10p,    na.rm = TRUE),
    median(ref_only_df$Ref_neglog10p,   na.rm = TRUE),
    median(your_only_df$Your_neglog10p, na.rm = TRUE)
  )
)

means_1b <- data.frame(
  Source     = c("SHARED", "Caudel_et_al_Only", "My_analysis_Only"),
  median_val = c(
    median(shared_df$Ref_neglog10p,     na.rm = TRUE),
    median(ref_only_df$Ref_neglog10p,   na.rm = TRUE),
    median(your_only_df$Your_neglog10p, na.rm = TRUE)
  )
)

# ── Shared theme ──────────────────────────────────────────────────────────────
shared_theme <- theme_minimal(base_size = 16) +
  theme(
    plot.title      = element_blank(),
    plot.subtitle   = element_blank(),
    axis.title.x    = element_text(size = 14, color = "grey20", margin = margin(t = 8)),
    axis.title.y    = element_text(size = 14, color = "grey20", margin = margin(r = 8)),
    axis.text.x     = element_text(size = 13, color = "grey20"),
    axis.text.y     = element_text(size = 13, color = "grey20"),
    legend.position = "right",
    legend.title    = element_text(size = 13, face = "bold"),
    legend.text     = element_text(size = 12),
    legend.key.size = unit(0.6, "cm"),
    plot.margin     = margin(20, 16, 12, 12)
  )

# ── Function to build base plot ───────────────────────────────────────────────
build_density_plot <- function(plot_data, means_df) {
  ggplot(plot_data, aes(x = neglog10p_value, fill = Source, color = Source)) +
    geom_density(alpha = 0.4, linewidth = 0.8) +
    geom_vline(data = means_df,
               aes(xintercept = median_val, color = Source),
               linetype = "dashed", linewidth = 1, alpha = 0.9) +
    geom_text(data = means_df,
              aes(x     = median_val,
                  y     = Inf,
                  label = paste0(clean_labels[Source], "\nmedian=", round(median_val, 2)),
                  color = Source,
                  vjust = c(1.3, 3.8, 6.3)),
              hjust       = -0.07,
              size        = 4,
              fontface    = "bold",
              show.legend = FALSE) +
    scale_fill_manual(values = category_colors, labels = clean_labels) +
    scale_color_manual(values = category_colors, labels = clean_labels) +
    labs(
      x     = expression(-log[10](italic(p)-value)),
      y     = "Density",
      fill  = "Category",
      color = "Category"
    ) +
    shared_theme
}

# ── Plot data ─────────────────────────────────────────────────────────────────
plot_data_a <- bind_rows(
  shared_df    %>% filter(!is.na(Your_neglog10p)) %>%
    mutate(neglog10p_value = Your_neglog10p, Source = "SHARED"),
  ref_only_df  %>% filter(!is.na(Ref_neglog10p))  %>%
    mutate(neglog10p_value = Ref_neglog10p,  Source = "Caudel_et_al_Only"),
  your_only_df %>% filter(!is.na(Your_neglog10p)) %>%
    mutate(neglog10p_value = Your_neglog10p, Source = "My_analysis_Only")
) %>% mutate(Source = factor(Source, levels = c("SHARED", "Caudel_et_al_Only", "My_analysis_Only")))

plot_data_b <- bind_rows(
  shared_df    %>% filter(!is.na(Ref_neglog10p))  %>%
    mutate(neglog10p_value = Ref_neglog10p,  Source = "SHARED"),
  ref_only_df  %>% filter(!is.na(Ref_neglog10p))  %>%
    mutate(neglog10p_value = Ref_neglog10p,  Source = "Caudel_et_al_Only"),
  your_only_df %>% filter(!is.na(Your_neglog10p)) %>%
    mutate(neglog10p_value = Your_neglog10p, Source = "My_analysis_Only")
) %>% mutate(Source = factor(Source, levels = c("SHARED", "Caudel_et_al_Only", "My_analysis_Only")))

# ============================================================================
# FIGURE 1A & 1B
# ============================================================================
p_dist1a <- build_density_plot(plot_data_a, means_1a)
p_dist1b <- build_density_plot(plot_data_b, means_1b)

print(p_dist1a)
print(p_dist1b)

# ============================================================================
# SAVE AS TIFF
# ============================================================================
plots_to_save <- list(
  "plot_dist1a" = p_dist1a,
  "plot_dist1b" = p_dist1b
)

for (nm in names(plots_to_save)) {
  ggsave(
    filename    = paste0("../../results/publication/", nm, ".tiff"),
    plot        = plots_to_save[[nm]],
    device      = "tiff",
    width       = 11,
    height      = 6,
    units       = "in",
    dpi         = 300,
    compression = "lzw"
  )
  message("Saved: ", nm, ".tiff")
}
```
```{r}

```


```{r}
library(ggplot2)
library(tidyverse)
library(scales)

# ── Color scheme ──────────────────────────────────────────────────────────────
category_colors <- c(
  "SHARED"      = "green3",
  "BulkRNAseq"  = "skyblue",
  "scRNAseq"    = "orange"
)

clean_labels <- c(
  "SHARED"      = "Shared",
  "BulkRNAseq"  = "Bulk RNA-seq only",
  "scRNAseq"    = "scRNA-seq only"
)

# ── Prepare data ──────────────────────────────────────────────────────────────
unified_renamed <- unified_data %>%
  mutate(Category = recode(Category,
                           "REFERENCE-ONLY" = "BulkRNAseq",
                           "YOUR-ONLY"      = "scRNAseq"))

# ============================================================================
# PLOT 1: BOXPLOT + JITTER
# ============================================================================
plot1_data <- unified_renamed %>%
  mutate(
    neglog10p = case_when(
      Category == "SHARED"      ~ (Ref_neglog10p + Your_neglog10p) / 2,
      Category == "BulkRNAseq"  ~ Ref_neglog10p,
      Category == "scRNAseq"    ~ Your_neglog10p
    ),
    Category = factor(Category, levels = c("BulkRNAseq", "SHARED", "scRNAseq"))
  ) %>%
  filter(!is.na(neglog10p))

summary_stats <- plot1_data %>%
  group_by(Category) %>%
  summarise(
    n      = n(),
    median = median(neglog10p),
    mean   = mean(neglog10p),
    .groups = "drop"
  )

p1 <- ggplot(plot1_data,
             aes(x = Category, y = neglog10p, color = Category, fill = Category)) +
  geom_boxplot(
    alpha         = 0.15,
    width         = 0.5,
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0.2,
    alpha = 0.35,
    size  = 1.5
  ) +
  stat_summary(
    fun   = mean,
    geom  = "point",
    shape = 18,
    size  = 5,
    color = "white"
  ) +
  geom_text(
    data        = summary_stats,
    aes(x       = Category,
        y       = -0.5,
        label   = paste0("n=", scales::comma(n), "\nmedian=", round(median, 1))),
    inherit.aes = FALSE,
    size        = 4,
    color       = "black",
    fontface    = "bold",
    vjust       = 1
  ) +
  scale_color_manual(values = category_colors, labels = clean_labels) +
  scale_fill_manual(values  = category_colors, labels = clean_labels) +
  scale_x_discrete(labels = clean_labels) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.08))) +
  labs(
    x       = NULL,
    y       = expression(-log[10](italic(p)-value)),
    caption = "Shared: average \u2212log\u2081\u2080(p) of both studies.  Bulk RNA-seq / scRNA-seq only: single study value."
  ) +
  theme_minimal(base_size = 16) +
  theme(
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    plot.title         = element_blank(),
    plot.subtitle      = element_blank(),
    plot.caption       = element_text(size = 10, color = "grey30"),
    axis.text.x        = element_text(size = 14, color = "black", lineheight = 1.1),
    axis.text.y        = element_text(size = 14, color = "black"),
    axis.title.y       = element_text(size = 15, color = "black", face = "bold",
                                      margin = margin(r = 8)),
    plot.margin        = margin(12, 16, 12, 12)
  )

print(p1)

ggsave(
  filename    = "../../results/publication/plot1_signal_strength_by_category.tiff",
  plot        = p1,
  device      = "tiff",
  width       = 9,
  height      = 7,
  units       = "in",
  dpi         = 300,
  compression = "lzw"
)
message("Saved: plot1_signal_strength_by_category.tiff")

# ============================================================================
# PLOT 2: SCATTER — scRNA-seq vs. Bulk RNA-seq -log10(p)
# ============================================================================
plot2_data <- unified_renamed %>%
  mutate(
    x_val    = if_else(is.na(Ref_neglog10p),  0, Ref_neglog10p),
    y_val    = if_else(is.na(Your_neglog10p), 0, Your_neglog10p),
    Category = factor(Category, levels = c("BulkRNAseq", "scRNAseq", "SHARED"))
  )

axis_limit <- max(c(plot2_data$x_val, plot2_data$y_val), na.rm = TRUE) * 1.05
n_shared   <- sum(plot2_data$Category == "SHARED")
n_scrna    <- sum(plot2_data$Category == "scRNAseq")
n_bulk     <- sum(plot2_data$Category == "BulkRNAseq")

p2 <- ggplot(plot2_data,
             aes(x = x_val, y = y_val, color = Category)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey60", linewidth = 0.7) +
  geom_point(
    data  = plot2_data %>% filter(Category != "SHARED"),
    alpha = 0.4,
    size  = 1.8
  ) +
  geom_point(
    data  = plot2_data %>% filter(Category == "SHARED"),
    alpha = 0.6,
    size  = 2
  ) +
  scale_color_manual(
    values = category_colors,
    labels = c(
      "SHARED"      = paste0("Shared (n=", scales::comma(n_shared), ")"),
      "BulkRNAseq"  = paste0("Bulk RNA-seq only (n=", scales::comma(n_bulk), ")"),
      "scRNAseq"    = paste0("scRNA-seq only (n=", scales::comma(n_scrna), ")")
    )
  ) +
  coord_fixed(ratio = 1,
              xlim  = c(0, axis_limit),
              ylim  = c(0, axis_limit)) +
  labs(
    x     = expression(bold("Bulk RNA-seq") ~ -log[10](italic(p)-value)),
    y     = expression(bold("scRNA-seq") ~ -log[10](italic(p)-value)),
    color = NULL
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title      = element_blank(),
    plot.subtitle   = element_blank(),
    axis.text.x     = element_text(size = 14, color = "black"),
    axis.text.y     = element_text(size = 14, color = "black"),
    axis.title.x    = element_text(size = 15, color = "black", face = "bold",
                                   margin = margin(t = 8)),
    axis.title.y    = element_text(size = 15, color = "black", face = "bold",
                                   margin = margin(r = 8)),
    legend.position = "bottom",
    legend.title    = element_blank(),
    legend.text     = element_text(size = 12, color = "black"),
    legend.key.size = unit(0.6, "cm"),
    plot.margin     = margin(12, 16, 12, 12)
  )

print(p2)

ggsave(
  filename    = "../../results/publication/plot2_scatter_scRNAseq_vs_bulk.tiff",
  plot        = p2,
  device      = "tiff",
  width       = 9,
  height      = 8,
  units       = "in",
  dpi         = 300,
  compression = "lzw"
)
message("Saved: plot2_scatter_scRNAseq_vs_bulk.tiff")
```
```{r}
library(ggplot2)
library(tidyverse)
library(scales)

# ── Color scheme ──────────────────────────────────────────────────────────────
category_colors <- c(
  "SHARED"      = "green3",
  "BulkRNAseq"  = "skyblue",
  "scRNAseq"    = "orange"
)

# ============================================================================
# PLOT 3: SCATTER — scRNA-seq vs. Bulk RNA-seq Effect Size (SnpWeight)
# ============================================================================
plot3_data <- unified_renamed %>%
  mutate(
    x_val    = if_else(is.na(Ref_SnpWeight),  0, Ref_SnpWeight),
    y_val    = if_else(is.na(Your_SnpWeight), 0, Your_SnpWeight),
    Category = factor(Category,
                      levels = c("BulkRNAseq", "scRNAseq", "SHARED"))  # match unified_renamed
  )

axis_abs_max <- max(abs(c(plot3_data$x_val, plot3_data$y_val)), na.rm = TRUE) * 1.05
n_shared <- sum(plot3_data$Category == "SHARED")
n_scrna  <- sum(plot3_data$Category == "scRNAseq")
n_bulk   <- sum(plot3_data$Category == "BulkRNAseq")

p3 <- ggplot(plot3_data,
             aes(x = x_val, y = y_val, color = Category)) +
  geom_abline(slope =  1, intercept = 0,
              linetype = "dashed", color = "grey60", linewidth = 0.7) +
  geom_abline(slope = -1, intercept = 0,
              linetype = "dotted", color = "grey80", linewidth = 0.5) +
  geom_hline(yintercept = 0, color = "grey85", linewidth = 0.4) +
  geom_vline(xintercept = 0, color = "grey85", linewidth = 0.4) +
  geom_point(
    data  = plot3_data %>% filter(Category != "SHARED"),
    alpha = 0.4, size = 1.8
  ) +
  geom_point(
    data  = plot3_data %>% filter(Category == "SHARED"),
    alpha = 0.6, size = 2
  ) +
  scale_color_manual(
    values = category_colors,
    labels = c(
      "SHARED"      = paste0("Shared (n=", scales::comma(n_shared), ")"),
      "BulkRNAseq"  = paste0("Bulk RNA-seq only (n=", scales::comma(n_bulk), ")"),
      "scRNAseq"    = paste0("scRNA-seq only (n=", scales::comma(n_scrna), ")")
    )
  ) +
  coord_fixed(ratio = 1,
              xlim  = c(-axis_abs_max, axis_abs_max),
              ylim  = c(-axis_abs_max, axis_abs_max)) +
  labs(
    x     = expression(bold("Bulk RNA-seq  SnpWeight  ") * (beta)),
    y     = expression(bold("scRNA-seq  SnpWeight  ") * (beta)),
    color = NULL
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title      = element_blank(),
    plot.subtitle   = element_blank(),
    axis.text.x     = element_text(size = 14, color = "black"),
    axis.text.y     = element_text(size = 14, color = "black"),
    axis.title.x    = element_text(size = 15, color = "black", face = "bold",
                                   margin = margin(t = 8)),
    axis.title.y    = element_text(size = 15, color = "black", face = "bold",
                                   margin = margin(r = 8)),
    legend.position = "bottom",
    legend.title    = element_blank(),
    legend.text     = element_text(size = 12, color = "black"),
    legend.key.size = unit(0.6, "cm"),
    plot.margin     = margin(12, 16, 12, 12)
  )

print(p3)

ggsave(
  filename    = "../../results/publication/plot3_scatter_effect_size_snpweight_sc_bulk.tiff",
  plot        = p3,
  device      = "tiff",
  width       = 9,
  height      = 8,
  units       = "in",
  dpi         = 300,
  compression = "lzw"
)
message("Saved: plot3_scatter_effect_size_snpweight_sc_bulk_downsampled.tiff")

```
