---
title: "Panel B — Single-cell raw-data validation of cis-eQTLs (all variant types)"
author: "Akriti Agrawal"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

```{r libs}
library(dplyr)
library(ggplot2)
library(Seurat)
library(data.table)
library(BEDMatrix)
library(MASS)
library(gridExtra)
```

# 1. Configuration

Inputs are read from `RES`; new outputs (verification CSVs, plots) are written to `OUT`.
Each variant type points to its own PLINK genotype matrix. The eQTL annotations come
from the single combined file (split by `Genetic_variants`); to use per-type files
instead, set `annot` per entry and set `USE_COMBINED <- FALSE`.

```{r config}
RES <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/Agrawal_single_cell_eqtl_cell_cycle/Figure.3/Fig.3-Panel-A-C-D-cis-and-trans-eqtl-YPD-variant-type/results/"
OUT <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/Agrawal_single_cell_eqtl_cell_cycle/Figure.3/Fig.3-Panel-A-C-D-cis-and-trans-eqtl-YPD-variant-type/results/"
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)

# shared single-cell object (cells -> strain via $SNG.BEST.GUESS)
SEURAT_FILE <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/gene_expression_analysis/output_from_pseudobulk_fastlmm/Na_va_data_subsetted.rds"

# single combined annotation file (has Genetic_variants + cis_trans columns)
USE_COMBINED <- TRUE
COMBINED_ANNOT <- file.path(OUT, "all_clumped_eqtl_annotated_YPD.1.txt")

# how many cis eQTLs to VERIFY per type (ranked by FastLMM P), and how many to PLOT
N_VERIFY <- 60      # candidate pool per type (bounds knit time)
N_TOP    <- 5       # best examples plotted/saved per type

# per-type genotype matrices (+ optional per-type annotation file if USE_COMBINED = FALSE)
variant_cfg <- list(
  SNPs = list(
    plink = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/SNPs/SNPs_no_change_name_and_no_strain_filtering_only_MAF15_filtered/full969Matrix.SNPs.Biallelic.Var.DP10.99percNonMissing.MAF15.NoSubTelo20kb.plink",
    annot = file.path(RES, "snp_data_annotated_cis_trans.txt"),
    allele_mode = "parse"),
  CNVs = list(
    plink = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/CNVs/CNVs_no_change_name_and_no_strain_filtering_only_MAF15_filtered/CNV_969strains_gain_maf15_N400_20221121.plink",
    annot = file.path(RES, "cnv_data_annotated_cis_trans.txt"),
    allele_mode = "dosage"),
  InDels = list(
    plink = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SNPs_indels/modified_snp_indel_plink_for_96strains/Indels_only_96/indels.96samples",
    annot = file.path(RES, "indel_data_annotated_cis_trans.txt"),
    allele_mode = "parse"),
  SVs = list(
    plink = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SVs_from_SV.1087/SV.1087samples_updated_fam_files/SV.1087samples.SVs",
    annot = file.path(RES, "sv_data_annotated_cis_trans.txt"),
    allele_mode = "dosage")
)

variant_order  <- c("SNPs", "InDels", "SVs", "CNVs")
variant_colors <- c(SNPs = "#4472C4", InDels = "#FFC000", SVs = "#C41E3A", CNVs = "#70AD47")
```

# 2. Load the single-cell object (once)

```{r load-seurat}
seurat_object <- readRDS(SEURAT_FILE)
cat("Seurat object:", ncol(seurat_object), "cells,", nrow(seurat_object), "genes\n")

# combined annotation, split by variant type; keep CIS only
if (USE_COMBINED) {
  all_annot <- fread(COMBINED_ANNOT)
  stopifnot("Genetic_variants" %in% names(all_annot), "cis_trans" %in% names(all_annot))
  cat("Combined annotation rows:", nrow(all_annot), "\n")
  print(all_annot[, .N, by = .(Genetic_variants, cis_trans)])
}
```

# 3. Helper functions

Column auto-detection makes the verifier work regardless of whether the gene column
is `Pheno` (combined schema) or `Source_Phenotype` (per-type schema).

```{r detectors}
detect_snp_column <- function(data) {
  for (col in c("SNP", "SNP_ID", "snp_id", "snp", "variant_id", "variant", "rsid"))
    if (col %in% colnames(data)) { cat("SNP column:", col, "\n"); return(col) }
  stop("No SNP column. Have: ", paste(colnames(data), collapse = ", "))
}
detect_gene_column <- function(data) {
  for (col in c("Pheno", "Source_Phenotype", "Gene", "gene", "gene_id", "Gene_ID", "target_gene"))
    if (col %in% colnames(data)) { cat("Gene column:", col, "\n"); return(col) }
  stop("No gene column. Have: ", paste(colnames(data), collapse = ", "))
}
detect_gene_common_name_column <- function(data) {
  for (col in c("gene_common_name", "Gene_Common_Name", "common_name", "symbol", "gene_symbol"))
    if (col %in% colnames(data)) { cat("Gene common-name column:", col, "\n"); return(col) }
  NULL
}
```

# verified = how many cis-eQTLs entered the verification loop. 
#Successful = of those, how many actually completed the negative-binomial GLM and produced a verification β (i.e. Status == "Success").
A common reason for eQTL to not be successful will be insufficient data per strain
```{r verify-fn}
# Verify a set of eQTLs against raw single-cell counts (negative-binomial GLM).
verify_eqtls_all <- function(fastlmm_results, seurat_object, bim_data, fam_data,
                             bed_matrix, n_eqtls = NULL, save_file = NULL) {
  snp_col  <- detect_snp_column(fastlmm_results)
  gene_col <- detect_gene_column(fastlmm_results)
  gene_common_col <- detect_gene_common_name_column(fastlmm_results)

  available_snps  <- intersect(fastlmm_results[[snp_col]], bim_data$snp_id)
  cat("Available variants:", length(available_snps), "of",
      length(unique(fastlmm_results[[snp_col]])), "\n")
  eqtls_to_verify <- fastlmm_results %>% filter(.data[[snp_col]] %in% available_snps)

  if (!is.null(n_eqtls)) {
    eqtls_to_verify <- eqtls_to_verify %>% arrange(PValue) %>% head(n_eqtls)
    cat("Verifying top", nrow(eqtls_to_verify), "by FastLMM P\n")
  } else {
    cat("Verifying all", nrow(eqtls_to_verify), "eQTLs\n")
  }

  results <- data.frame(
    SNP_ID = character(nrow(eqtls_to_verify)), Gene_ID = character(nrow(eqtls_to_verify)),
    Gene_Common_Name = character(nrow(eqtls_to_verify)), Resolved_Gene_ID = character(nrow(eqtls_to_verify)),
    Status = character(nrow(eqtls_to_verify)), N_Cells = integer(nrow(eqtls_to_verify)),
    Verification_Beta = numeric(nrow(eqtls_to_verify)), Verification_P = numeric(nrow(eqtls_to_verify)),
    FastLMM_Beta = numeric(nrow(eqtls_to_verify)), FastLMM_P = numeric(nrow(eqtls_to_verify)),
    Effect_Agreement = logical(nrow(eqtls_to_verify)), R_Squared = numeric(nrow(eqtls_to_verify)),
    stringsAsFactors = FALSE)

  cell_sample_map      <- seurat_object$SNG.BEST.GUESS
  cells_with_genotypes <- names(cell_sample_map)[cell_sample_map %in% fam_data$iid]
  cat("Cells with genotype:", length(cells_with_genotypes), "\n")

  successful <- 0
  for (i in seq_len(nrow(eqtls_to_verify))) {
    snp_id  <- eqtls_to_verify[[snp_col]][i]
    gene_id <- eqtls_to_verify[[gene_col]][i]
    gene_common <- if (!is.null(gene_common_col)) eqtls_to_verify[[gene_common_col]][i] else ""

    results$SNP_ID[i] <- snp_id; results$Gene_ID[i] <- gene_id
    results$Gene_Common_Name[i] <- ifelse(is.na(gene_common), "", gene_common)
    results$FastLMM_Beta[i] <- eqtls_to_verify$SnpWeight[i]
    results$FastLMM_P[i]    <- eqtls_to_verify$PValue[i]

    resolved_gene <- if (gene_id %in% rownames(seurat_object)) gene_id else
      if (!is.na(gene_common) && gene_common != "" && gene_common %in% rownames(seurat_object)) gene_common else NULL
    if (is.null(resolved_gene)) { results$Status[i] <- "Gene not found"; next }
    results$Resolved_Gene_ID[i] <- resolved_gene

    snp_idx <- which(bim_data$snp_id == snp_id)[1]
    if (is.na(snp_idx)) { results$Status[i] <- "SNP not found"; next }

    genotypes        <- bed_matrix[, snp_idx]
    plink_sample_map <- setNames(genotypes, fam_data$iid)
    cell_genotypes   <- plink_sample_map[cell_sample_map[cells_with_genotypes]]
    names(cell_genotypes) <- cells_with_genotypes
    cell_genotypes   <- cell_genotypes[!is.na(cell_genotypes)]
    if (length(cell_genotypes) < 20 || length(unique(cell_genotypes)) < 2) {
      results$Status[i] <- "Insufficient data"; next }

    gene_expr <- GetAssayData(seurat_object, slot = "counts")[resolved_gene, names(cell_genotypes)]
    plot_data <- data.frame(expression = as.numeric(gene_expr), genotype = cell_genotypes)
    plot_data <- plot_data[complete.cases(plot_data), ]
    results$N_Cells[i] <- nrow(plot_data)

    tryCatch({
      model <- glm.nb(expression ~ genotype, data = plot_data)
      null_model <- glm.nb(expression ~ 1, data = plot_data)
      cs <- summary(model)$coefficients
      results$Verification_Beta[i] <- cs[2, 1]
      results$Verification_P[i]    <- cs[2, 4]
      results$Effect_Agreement[i]  <- sign(cs[2, 1]) == sign(results$FastLMM_Beta[i])
      results$R_Squared[i]         <- as.numeric(1 - logLik(model) / logLik(null_model))
      results$Status[i] <- "Success"; successful <- successful + 1
    }, error = function(e) { results$Status[i] <- "Model failed" })
  }

  cat("Success:", successful, "/", nrow(results),
      "(", round(100 * successful / max(nrow(results), 1), 1), "%)\n")
  sd_ok <- results[results$Status == "Success", ]
  if (nrow(sd_ok) > 1)
    cat("Beta correlation:", round(cor(sd_ok$Verification_Beta, sd_ok$FastLMM_Beta), 3), "\n")
  if (!is.null(save_file)) write.csv(results, save_file, row.names = FALSE)
  results
}
```

```{r plot-fn}
# Genotype vs single-cell expression plot. `variant_type` sets genotype axis labels
# (REF/ALT parsed from the ID for SNP/InDel; dosage for CNV/SV).
create_enhanced_verification_plot <- function(plot_data, result, association, variant_type) {

  unique_genotypes <- sort(unique(plot_data$genotype))
  if (variant_type %in% c("SNPs", "InDels")) {
    snp_parts <- strsplit(result$snp_id, "_")[[1]]
    if (length(snp_parts) >= 4) { ref_allele <- snp_parts[3]; alt_allele <- snp_parts[4] }
    else { ref_allele <- "Ref"; alt_allele <- "Alt" }
    lab3 <- c(paste0(ref_allele, "/", ref_allele),
              paste0(ref_allele, "/", alt_allele),
              paste0(alt_allele, "/", alt_allele))
    genotype_labels <- lab3[unique_genotypes + 1]
    x_title <- "Genotype"
  } else {
    genotype_labels <- paste0(unique_genotypes)   # dosage (0/1/2 copies)
    x_title <- paste0(variant_type, " dosage")
  }
  plot_data$genotype_label <- factor(plot_data$genotype, levels = unique_genotypes,
                                     labels = genotype_labels)

  genotype_stats <- plot_data %>% group_by(genotype_label) %>%
    summarise(n_cells = n(), n_samples = length(unique(sample)),
              mean_expr = mean(expression), se_expr = sd(expression) / sqrt(n()),
              .groups = "drop")

  fold_change  <- exp(result$verification_beta)
  fastlmm_beta <- if ("SnpWeight" %in% names(association)) association$SnpWeight else NA
  fastlmm_pval <- if ("PValue"   %in% names(association)) association$PValue   else NA
  fastlmm_log2fc      <- if (!is.na(fastlmm_beta)) fastlmm_beta / log(2) else NA
  fastlmm_fold_change <- if (!is.na(fastlmm_beta)) exp(fastlmm_beta)      else NA
  effect_agreement    <- if (!is.na(fastlmm_beta)) sign(result$verification_beta) == sign(fastlmm_beta) else NA

  shared_labs <- labs(
    title = paste0(variant_type, " eQTL: ", result$snp_id, " \u2192 ", result$resolved_gene_id,
                   ifelse(result$resolved_gene_id != result$gene_id,
                          paste0(" (", result$gene_id, ")"), "")),
    subtitle = paste0("Verification: \u03b2=", round(result$verification_beta, 3),
      ", Log2FC=", round(result$verification_log2fc, 2), ", FC=", round(fold_change, 2),
      ", P=", format(result$verification_pval, digits = 3),
      ifelse(!is.na(fastlmm_beta),
        paste0("\nFastLMM: \u03b2=", round(fastlmm_beta, 3), ", Log2FC=", round(fastlmm_log2fc, 2),
               ", FC=", round(fastlmm_fold_change, 2), ", P=", format(fastlmm_pval, digits = 3),
               " | Agreement: ", ifelse(is.na(effect_agreement), "Unknown",
                                        ifelse(effect_agreement, "\u2713", "\u2717"))), "")),
    x = x_title,
    caption = paste0("N=", nrow(plot_data), " cells from ",
                     length(unique(plot_data$sample)), " samples | red line shows genotype trend"))

  shared_theme <- theme_minimal(base_size = 12) +
    theme(legend.position = "none",
          plot.title = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 10)),
          plot.subtitle = element_text(size = 12, hjust = 0.5, margin = margin(b = 15), lineheight = 1.2),
          plot.caption = element_text(size = 10, color = "black", hjust = 0.5),
          axis.title.x = element_text(size = 16, margin = margin(t = 10)),
          axis.title.y = element_text(size = 16, margin = margin(r = 10)),
          axis.text = element_text(size = 12),
          panel.grid.major = element_line(color = "grey85", size = 0.4),
          panel.grid.minor = element_blank(),
          panel.background = element_rect(fill = "white", color = NA),
          plot.background  = element_rect(fill = "white", color = NA))

  shared_x_scale <- scale_x_discrete(labels = function(x) {
    sr <- genotype_stats[genotype_stats$genotype_label == x, ]
    if (nrow(sr) > 0) paste0(x, "\nN = ", sr$n_cells, "\n(", sr$n_samples, " samples)") else x })
  shared_fill <- scale_fill_manual(values = c("#66C2A5", "#FC8D62", "#8DA0CB"))

  if (median(plot_data$expression) == 0) {
    # zero-inflated gene: mean +/- SE bar plot
    p <- ggplot(genotype_stats, aes(genotype_label, mean_expr, fill = genotype_label)) +
      geom_bar(stat = "identity", width = 0.4, alpha = 0.8) +
      geom_errorbar(aes(ymin = mean_expr - se_expr, ymax = mean_expr + se_expr),
                    width = 0.15, size = 0.8) +
      geom_line(aes(as.numeric(genotype_label), mean_expr, group = 1),
                color = "red", size = 1.5, alpha = 0.8) +
      geom_point(aes(as.numeric(genotype_label), mean_expr),
                 color = "red", size = 3, shape = 16) +
      shared_fill + shared_labs + labs(y = "Mean expression (UMI counts)") +
      shared_theme + shared_x_scale + scale_y_continuous(expand = expansion(mult = c(0.05, 0.2)))
  } else {
    y_upper <- max(quantile(plot_data$expression, 0.95) * 1.2, 2)
    p <- ggplot(plot_data, aes(genotype_label, expression)) +
      geom_boxplot(aes(fill = genotype_label), width = 0.4, alpha = 0.8, outlier.shape = NA) +
      stat_summary(fun = mean, geom = "point", size = 2.5, color = "black", shape = 18) +
      geom_line(data = genotype_stats, aes(as.numeric(genotype_label), mean_expr, group = 1),
                color = "red", size = 1.5, alpha = 0.8, inherit.aes = FALSE) +
      geom_point(data = genotype_stats, aes(as.numeric(genotype_label), mean_expr),
                 color = "red", size = 3, shape = 16, inherit.aes = FALSE) +
      shared_fill + shared_labs + labs(y = "Gene expression (UMI counts)") +
      shared_theme + shared_x_scale +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
      coord_cartesian(ylim = c(0, y_upper))
  }
  p
}
```

```{r topn-fn}
# From a verified results table, plot the top-N successes by |verification beta|.
plot_top_eqtls <- function(results_table, variant_type, n_plots,
                           seurat_object, bim_data, fam_data, bed_matrix,
                           out_dir, save_plots = TRUE) {
  results_table$SNP_ID <- as.character(results_table$SNP_ID)
  top <- results_table %>% filter(Status == "Success") %>%
    mutate(Abs_Beta = abs(Verification_Beta)) %>%
    arrange(desc(Abs_Beta)) %>% head(n_plots)
  cat("Top", nrow(top), variant_type, "cis-eQTLs by |verified beta|\n")
  if (nrow(top) == 0) return(list())

  cell_sample_map      <- seurat_object$SNG.BEST.GUESS
  cells_with_genotypes <- names(cell_sample_map)[cell_sample_map %in% fam_data$iid]
  plots_list <- list()

  for (i in seq_len(nrow(top))) {
    row <- top[i, ]
    snp_id        <- as.character(row$SNP_ID)
    gene_id       <- as.character(row$Gene_ID)
    resolved_gene <- as.character(row$Resolved_Gene_ID)

    snp_idx <- which(bim_data$snp_id == snp_id)[1]
    if (is.na(snp_idx)) { cat("skip (not in bim):", snp_id, "\n"); next }

    genotypes        <- bed_matrix[, snp_idx]
    plink_sample_map <- setNames(genotypes, fam_data$iid)
    cell_genotypes   <- plink_sample_map[cell_sample_map[cells_with_genotypes]]
    names(cell_genotypes) <- cells_with_genotypes
    cell_genotypes   <- cell_genotypes[!is.na(cell_genotypes)]

    gene_expr <- GetAssayData(seurat_object, slot = "counts")[resolved_gene, names(cell_genotypes)]
    plot_data <- data.frame(cell = names(cell_genotypes),
                            expression = as.numeric(gene_expr),
                            genotype = cell_genotypes,
                            sample = cell_sample_map[names(cell_genotypes)])
    plot_data <- plot_data[complete.cases(plot_data), ]

    result <- list(snp_id = snp_id, gene_id = gene_id, resolved_gene_id = resolved_gene,
                   verification_beta = row$Verification_Beta,
                   verification_pval = row$Verification_P,
                   verification_log2fc = row$Verification_Beta / log(2))
    association <- data.frame(SnpWeight = row$FastLMM_Beta, PValue = row$FastLMM_P)

    p <- create_enhanced_verification_plot(plot_data, result, association, variant_type)
    plots_list[[i]] <- p
    print(p)   # embed in the compiled HTML

    if (save_plots) {
  fn <- file.path(out_dir, sprintf("PanelB_%s_top%02d_%s_%s.tiff", variant_type, i,
                    gsub("[^A-Za-z0-9]", "_", snp_id),
                    gsub("[^A-Za-z0-9]", "_", resolved_gene)))
  ggsave(fn, p, width = 12, height = 8, units = "in", dpi = 300, 
         compression = "lzw", bg = "white")
}
  plots_list
}
```

# 4. Run per variant type (cis only, top 5)

Each type: load its PLINK matrix, take the cis eQTLs, verify the top `N_VERIFY`
against single-cell counts, then plot and save the best `N_TOP` by verified effect size.

```{r run-all, results='asis', fig.width=12, fig.height=8}
all_summaries <- list()

for (vt in variant_order) {
  cat("\n\n## ", vt, "\n\n", sep = "")
  cfg <- variant_cfg[[vt]]

  # genotype matrix
  bed_matrix <- BEDMatrix(cfg$plink)
  bim_data   <- fread(paste0(cfg$plink, ".bim"),
                      col.names = c("chr","snp_id","gen_dist","pos","allele1","allele2"))
  fam_data   <- fread(paste0(cfg$plink, ".fam"),
                      col.names = c("fid","iid","father","mother","sex","phenotype"))

  # cis eQTLs for this type
  if (USE_COMBINED) {
    ann <- all_annot[Genetic_variants == vt]
  } else {
    ann <- fread(cfg$annot)
  }
  cis <- as.data.frame(ann[ann$cis_trans == "Cis", ])
  cat("cis eQTLs:", nrow(cis), " | PLINK strains:", nrow(fam_data),
      " | variants:", nrow(bim_data), "\n\n")
  if (nrow(cis) == 0) { cat("_No cis eQTLs — skipped._\n"); next }

  res <- verify_eqtls_all(cis, seurat_object, bim_data, fam_data, bed_matrix,
                          n_eqtls = N_VERIFY,
                          save_file = file.path(OUT, paste0("PanelB_verify_", vt, "_cis.csv")))

  plot_top_eqtls(res, vt, N_TOP, seurat_object, bim_data, fam_data, bed_matrix,
                 out_dir = OUT, save_plots = TRUE)

  ok <- res[res$Status == "Success", ]
  all_summaries[[vt]] <- data.frame(
    Variant_type   = vt,
    cis_eQTLs      = nrow(cis),
    verified       = nrow(res),
    successful     = nrow(ok),
    effect_agree   = if (nrow(ok)) sum(ok$Effect_Agreement, na.rm = TRUE) else 0,
    pct_agree      = if (nrow(ok)) round(100 * mean(ok$Effect_Agreement, na.rm = TRUE), 1) else NA,
    beta_cor       = if (nrow(ok) > 1) round(cor(ok$Verification_Beta, ok$FastLMM_Beta), 3) else NA)
}
```

# 5. Cross-type validation summary

```{r summary-table}
summary_tbl <- bind_rows(all_summaries)
print(summary_tbl)
write.csv(summary_tbl, file.path(OUT, "PanelB_validation_summary_all_types.csv"), row.names = FALSE)
```


