---
title: "Figure 1A — Phylogenetic distribution of the 96 sequenced strains"
author: "Akriti | Gresham Lab, NYU"
output: html_document
---

Phylogenetic tree of the 1011 *S. cerevisiae* collection (Peter et al. 2018),
with the 96 strains used in this study highlighted as red tip points. Branches are
coloured by clade assignment. Blue text indicates the 96 sequenced strains.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(ggplot2)
library(ggtree)
library(viridis)
library(dplyr)
library(grid)  # for additional text annotation options
```

```{r load}
# tree1011  : phylo object for the 1011 collection
# cladesdf  : per-strain metadata (STD_name, clades)
load("../data/1011_tree_assets.RData")
```

```{r strains}
strains_96 <- c(
  "BNV", "BQM", "ALS", "CAE", "CGS", "AQA", "CDT", "BRK", "BHM", "AMP",
  "BHK", "BTT", "CNK", "CNH", "BVG", "BBQ", "BFP", "BTG", "CLB", "BFH",
  "BGD", "BGF", "BGE", "BDN", "CBN", "CFF", "SACE_YAG", "SACE_YCF", "BDB", "BHG",
  "SACE_MAA", "AFE", "ATS", "ATR", "AAP", "BDL", "CPI", "CPK", "CPN", "BDQ",
  "BNC", "BMV", "BDP", "BCK", "CBL", "AKI", "SACE_YCE", "CDE", "SACE_YBJ", "ANN",
  "CQD", "CQF", "APQ", "AKH", "BAF", "BAM", "BAQ", "BAG", "BAL", "CEG",
  "CCQ", "CCR", "CCS", "SACE_YCR", "BAK", "BAI", "CCN", "CDL", "BMA", "AKM",
  "BMB", "BMC", "SACE_YCY", "BAN", "BAP", "CCD", "CCG", "CDF", "CDH", "CDI",
  "AVI", "ANH", "ANG", "SACE_YCU", "CCV", "SACE_YCM", "SACE_YDE", "CLI", "CLN", "CMI",
  "CLP", "CGL", "CGN", "BHQ", "BEF", "SACE_YCK"
)
stopifnot(length(strains_96) == 96)
cladesdf <- cladesdf %>%
  mutate(highlight = ifelse(STD_name %in% strains_96, "In subset", "Not in subset"))
# sanity check: all 96 present in the tree metadata
setdiff(strains_96, cladesdf$STD_name)
```

```{r colours}
# clade colours from viridis; subset tips red, all other tips invisible
cols <- c(viridis(length(unique(cladesdf$clades))),
          "In subset"     = "red",
          "Not in subset" = "transparent")
names(cols) <- c(unique(cladesdf$clades), "In subset", "Not in subset")
```

```{r}
fig1A <- ggtree(tree1011,
                layout     = "ape",
                open.angle = 10,
                size       = 0.25,
                aes(color = clades)) %<+% cladesdf +
  geom_tippoint(aes(color = highlight), size = 2) +
  scale_color_manual(values = cols) +
  theme_void() +
  theme(legend.position = "none",
        plot.background = element_rect(fill = "transparent", colour = NA))

fig1A
```


```{r figure, fig.width = 14, fig.height = 9}
fig1A <- ggtree(tree1011,
                layout      = "ape",
                open.angle  = 10,
                size        = 0.2,
                aes(color = clades)) %<+% cladesdf +
  geom_tippoint(aes(color = highlight), size = 2) +
  scale_color_manual(
    values = cols,
    breaks = unique(cladesdf$clades),
    name   = "Clade"
  ) +
  guides(color = guide_legend(
    override.aes = list(
      size = 3,
      linewidth = 1
    ),
    ncol = 1
  )) +
  theme(
    legend.position      = "right",
    legend.text          = element_text(size = 18),
    legend.title         = element_text(size = 18, face = "bold"),
    legend.background    = element_rect(fill = "white", color = "black", linewidth = 0.3),
    legend.key           = element_rect(fill = "white", color = NA),
    legend.key.height    = unit(0.6, "cm"),
    legend.key.width     = unit(0.6, "cm"),
    legend.margin        = margin(8, 8, 8, 8),
    legend.spacing.y     = unit(0.3, "cm"),
    panel.grid.major     = element_blank(),
    panel.grid.minor     = element_blank(),
    panel.background     = element_rect(fill = "transparent", colour = NA),
    plot.background      = element_rect(fill = "transparent", colour = NA)
  )

fig1A

```
Figure 1A. The 96 sequenced strains are broadly distributed across major phylogenetic clades of S. cerevisiae.

Phylogenetic tree of the 1011 S. cerevisiae collection (Peter et al. 2018), with branches colored by clade assignment. Red dots indicate the 96 strains selected for this study. The distribution of red dots across all 26 major clades demonstrates that strain selection encompasses the phylogenetic diversity of the species rather than being biased toward any single clade. This broad representation ensures that eQTL analyses capture genetic variation across diverse genetic backgrounds and evolutionary history of S. cerevisiae.
```{r save}
ggsave("../results/tree_with_96_strain.pdf", plot = fig1A,
       width = 14, height = 9, dpi = 300,
       device = "pdf", useDingbats = FALSE)

ggsave("../results/tree_with_96_strain.tiff", plot = fig1A,
       width = 14, height = 9, dpi = 300,
       device = "tiff", compression = "lzw")

ggsave("../results/tree_with_96_strain.png", plot = fig1A,
       width = 14, height = 9, dpi = 300,
       device = "png", bg = "white")
```


```{r}
library(ggtree); library(ape); library(dplyr); library(viridis)

# For each clade: MRCA node, and what fraction of its descendants actually belong
clade_nodes <- cladesdf %>%
  filter(!is.na(clades)) %>%
  group_by(clades) %>%
  filter(n() >= 3) %>%                      # 1-2 tips can't form a useful blob
  summarise(tips = list(STD_name), n_tips = n(), .groups = "drop") %>%
  rowwise() %>%
  mutate(
    node = tryCatch(getMRCA(tree1011, match(tips, tree1011$tip.label)),
                    error = function(e) NA_integer_),
    n_desc = if (!is.na(node))
      length(geiger::tips(tree1011, node)) else NA_integer_,
    purity = n_tips / n_desc
  ) %>%
  ungroup()

clade_nodes %>%
  select(clades, n_tips, n_desc, purity) %>%
  arrange(purity) %>%
  print(n = Inf)
```

```{r}
library(ggtree); library(ape); library(dplyr); library(viridis)

# ── Which clades are monophyletic enough to encircle? ───────────────────────
clade_nodes <- cladesdf %>%
  filter(!is.na(clades)) %>%
  group_by(clades) %>%
  filter(n() >= 3) %>%
  summarise(tips = list(STD_name), n_tips = n(), .groups = "drop") %>%
  rowwise() %>%
  mutate(
    node   = tryCatch(getMRCA(tree1011, match(tips, tree1011$tip.label)),
                      error = function(e) NA_integer_),
    n_desc = if (!is.na(node))
               length(phangorn::Descendants(tree1011, node, "tips")[[1]])
             else NA_integer_,
    purity = n_tips / n_desc
  ) %>%
  ungroup()

PURITY_MIN <- 0.5
keep <- clade_nodes %>%
  filter(!is.na(node), purity >= PURITY_MIN) %>%
  arrange(desc(n_desc))          # largest first so smaller blobs draw on top

cat("Highlighting", nrow(keep), "of", nrow(clade_nodes), "clades\n")
cat("Dropped (not monophyletic):",
    paste(setdiff(clade_nodes$clades, keep$clades), collapse = ", "), "\n")

# ── Draw ────────────────────────────────────────────────────────────────────
clade_cols <- setNames(viridis(nrow(keep), option = "turbo"), keep$clades)

p <- ggtree(tree1011, layout = "ape", open.angle = 10,
            size = 0.15, colour = "grey30") %<+% cladesdf

for (i in seq_len(nrow(keep))) {
  p <- p + geom_hilight(node = keep$node[i], type = "encircle",
                        fill = clade_cols[keep$clades[i]],
                        alpha = 0.25, expand = 0.01)
}

fig1A <- p +
  geom_tippoint(aes(subset = label %in% strains_96),
                colour = "red", size = 1.8) +
  theme_void() +
  theme(legend.position = "none",
        plot.background = element_rect(fill = "white", colour = NA))

fig1A
```
```{r}
library(phangorn)

# All internal nodes and their descendant tips, computed once
int_nodes <- (Ntip(tree1011) + 1):(Ntip(tree1011) + Nnode(tree1011))
desc_list <- Descendants(tree1011, int_nodes, "tips")
desc_labs <- lapply(desc_list, function(i) tree1011$tip.label[i])

# For one clade: the largest node that is >= purity_min pure for that clade
core_node <- function(clade_tips, purity_min = 0.75, min_tips = 3) {
  n_in  <- vapply(desc_labs, function(d) sum(d %in% clade_tips), integer(1))
  n_tot <- lengths(desc_labs)
  ok    <- n_in >= min_tips & (n_in / n_tot) >= purity_min
  if (!any(ok)) return(tibble(node = NA_integer_, n_in = 0L,
                              n_tot = 0L, purity = NA_real_))
  best <- which(ok)[which.max(n_in[ok])]          # most clade tips captured
  tibble(node   = int_nodes[best],
         n_in   = n_in[best],
         n_tot  = n_tot[best],
         purity = n_in[best] / n_tot[best])
}

clade_cores <- cladesdf %>%
  filter(!is.na(clades)) %>%
  group_by(clades) %>%
  filter(n() >= 3) %>%
  summarise(tips = list(STD_name), n_clade = n(), .groups = "drop") %>%
  rowwise() %>%
  mutate(core_node(tips)) %>%
  ungroup() %>%
  mutate(coverage = n_in / n_clade)               # fraction of the clade shaded

clade_cores %>%
  select(clades, n_clade, n_in, n_tot, purity, coverage) %>%
  arrange(coverage) %>%
  print(n = Inf)

COVERAGE_MIN <- 0.4     # don't shade a clade if the core is a small fraction of it

keep <- clade_cores %>%
  filter(!is.na(node), coverage >= COVERAGE_MIN) %>%
  arrange(desc(n_tot))                    # big blobs first, small on top

cat("Shading", nrow(keep), "clades\n")
cat("Median coverage:", round(median(keep$coverage), 2), "\n")
cat("Not shaded:", paste(setdiff(clade_cores$clades, keep$clades),
                         collapse = ", "), "\n")

clade_cols <- setNames(viridis(nrow(keep), option = "turbo"), keep$clades)

p <- ggtree(tree1011, layout = "ape", open.angle = 10,
            size = 0.15, colour = "grey30") %<+% cladesdf

for (i in seq_len(nrow(keep))) {
  p <- p + geom_hilight(node = keep$node[i], type = "encircle",
                        fill = clade_cols[keep$clades[i]],
                        alpha = 0.25, expand = 0.01)
}

fig1A <- p +
  geom_tippoint(aes(subset = label %in% strains_96),
                colour = "red", size = 1.8) +
  theme_void() +
  theme(legend.position = "none",
        plot.background = element_rect(fill = "white", colour = NA))

fig1A

ggsave("../results/tree_with_96_strain_with_clade_blob.tiff", plot = fig1A,
       width = 14, height = 9, dpi = 300,
       device = "tiff", compression = "lzw")

```


