---
title: "Supplementary Figures 4-6 — strain and gene expression reproducibility"
output: html_document
---

- **S4** Strain expression is reproducible between replicates
- **S5** Reproducibility improves with cells assigned per strain (A: mean cells, B: min cells)
- **S6** Per-gene expression correlation across strains

YPD and fluconazole only — sc 2% glucose has no replicate experiment.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE, dpi = 300)
```

```{r libraries}
library(tidyverse)
library(DESeq2)
library(patchwork)
```

# ── Configuration ──────────────────────────────────────────────
# N_PERM / set.seed(42) removed — those only fed the permutation-null
# test in the old S4 (same-strain vs different-strain), which is dropped.

```{r config}
BASE <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all"

conditions <- list(
  YPD = list(
    label = "YPD",
    pb1   = file.path(BASE, "YPD.1/fastlmm/data/pheno_data/pseudobulk_reliable_transcripts_YPD.txt"),
    pb2   = file.path(BASE, "YPD.2/fastlmm/data/pheno_data/pseudobulk_reliable_transcripts_YPD.txt"),
    rds1  = file.path(BASE, "YPD.1/gene_expression_analysis/output_from_pseudobulk_fastlmm/Na_va_data_subsetted.rds"),
    rds2  = file.path(BASE, "YPD.2/gene_expression_analysis/Na_va_data_subsetted.rds")
  ),
  Fluconazole = list(
    label = "Fluconazole",
    pb1   = file.path(BASE, "Flu.1/fastlmm/data/pheno_data/pseudobulk_reliable_transcripts_Flu1.txt"),
    pb2   = file.path(BASE, "Flu.2/fastlmm/data/pheno_data/pseudobulk_reliable_transcripts_Flu2.txt"),
    rds1  = file.path(BASE, "Flu.1/gene_expression_analysis/output_from_pseudobulk_fastlmm/Na_va_data_subsetted.rds"),
    rds2  = file.path(BASE, "Flu.2/gene_expression_analysis/output_from_pseudobulk_fastlmm/Na_va_data_subsetted.rds")
  )
)

condition_colors <- c("YPD" = "#2c5f8a", "Fluconazole" = "#c1554d")
lvl <- names(conditions)

OUT_DIR <- "../results"
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

```{r theme}
pub_theme <- theme_classic(base_size = 9) +
  theme(
    axis.title       = element_text(size = 9, face = "bold", color = "black"),
    axis.text        = element_text(size = 8, color = "black"),
    axis.line        = element_line(linewidth = 0.4, color = "black"),
    axis.ticks       = element_line(linewidth = 0.35, color = "black"),
    strip.text       = element_text(size = 9, face = "bold"),
    strip.background = element_blank(),
    panel.grid       = element_blank(),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

# Helpers

```{r helpers}
read_pseudobulk <- function(path) {
  if (!file.exists(path)) stop("File not found: ", path)
  as.matrix(read.csv(path, sep = "\t", row.names = 1, header = TRUE,
                     check.names = FALSE))
}

deseq_normalize <- function(mat, label = "") {
  m   <- round(as.matrix(mat))
  dds <- DESeqDataSetFromMatrix(
    countData = m,
    colData   = data.frame(row.names = colnames(m), condition = factor(colnames(m))),
    design    = ~ 1
  )
  dds <- tryCatch(estimateSizeFactors(dds),
                  error = function(e) estimateSizeFactors(dds, type = "poscounts"))
  counts(dds, normalized = TRUE)
}

# Per-strain QC from the Seurat object. Metadata only — the object is dropped
# immediately so four of these don't sit in memory at once.
strain_qc <- function(rds_path, replicate) {
  obj <- readRDS(rds_path)
  md  <- obj@meta.data
  rm(obj); gc(verbose = FALSE)

  md %>%
    group_by(strain = SNG.BEST.GUESS) %>%
    summarise(n_cells    = n(),
              median_umi = median(nCount_RNA),
              .groups    = "drop") %>%
    rename_with(~ paste0(.x, "_", replicate), -strain)
}
```

# ── Compute ────────────────────────────────────────────────────
# Removed: cross_mat (full n x n strain pairwise Spearman), diff_r,
# self_rank/n_best, permutation null (N_PERM draws) — that whole block
# only existed to drive the old density-plot S4 (same-strain vs
# different-strain), which duplicates Fig 1 Panel D. Per-strain
# reproducibility (replicate_r) now comes from a single O(n) pass
# instead of an O(n^2) matrix, and feeds both the new S4 scatter labels
# and the existing S5 QC plot — no duplicate correlation computation.

```{r compute}
analyse_condition <- function(cfg) {

  x <- read_pseudobulk(cfg$pb1)
  y <- read_pseudobulk(cfg$pb2)

  genes   <- intersect(rownames(x), rownames(y))
  strains <- intersect(colnames(x), colnames(y))
  cat("\n---", cfg$label, "---\n")
  cat("Shared strains:", length(strains), "| shared genes:", length(genes), "\n")

  x <- x[genes, strains, drop = FALSE]
  y <- y[genes, strains, drop = FALSE]

  # Per-strain replicate correlation (Spearman is rank-based, so raw
  # counts and log1p(counts) give the identical r — computed once here).
  replicate_r <- vapply(strains, function(s) {
    cor(x[, s], y[, s], method = "spearman")
  }, numeric(1))

  # --- S4: per-strain, per-gene scatter (one facet per strain) --------------
  scatter_df <- map_dfr(strains, function(s) {
    tibble(condition = cfg$label, strain = s, gene = genes,
           val1 = x[, s] + 1, val2 = y[, s] + 1, r = replicate_r[s])
  }) %>%
    mutate(facet_label = paste0(strain, "  (r = ", round(r, 3), ")"))

  # --- S5: per-strain QC ------------------------------------------------------
  qc <- full_join(strain_qc(cfg$rds1, "r1"), strain_qc(cfg$rds2, "r2"),
                  by = "strain") %>%
    mutate(mean_cells = (n_cells_r1 + n_cells_r2) / 2,
           min_cells  = pmin(n_cells_r1, n_cells_r2))

  strain_df <- tibble(condition = cfg$label, strain = strains, r = replicate_r) %>%
    left_join(qc, by = "strain")

  cat("Strains with QC data:", sum(!is.na(strain_df$mean_cells)),
      "of", nrow(strain_df), "\n")

  # --- S6: per-gene Spearman across strains (DESeq2-normalized) --------------
  xn <- deseq_normalize(x)
  yn <- deseq_normalize(y)

  gene_r <- vapply(seq_len(nrow(xn)), function(i) {
    suppressWarnings(cor(xn[i, ], yn[i, ], method = "spearman", use = "complete.obs"))
  }, numeric(1))

  gene_df <- tibble(condition = cfg$label, gene = rownames(xn), r = gene_r) %>%
    mutate(cor_category = case_when(
      r >= 0.7            ~ "high (\u22650.7)",
      r >= 0.4 & r < 0.7  ~ "moderate (0.4\u20130.7)",
      r <  0.4            ~ "low (<0.4)",
      TRUE                ~ NA_character_))

  list(
    scatter_df = scatter_df,
    strain_df  = strain_df,
    gene_df    = gene_df,
    stats      = tibble(condition          = cfg$label,
                        n_strains          = length(strains),
                        n_genes            = length(genes),
                        median_replicate_r = median(replicate_r),
                        median_gene_r      = median(gene_r, na.rm = TRUE),
                        pct_gene_r_0.4     = 100 * mean(gene_r > 0.4, na.rm = TRUE))
  )
}

res <- lapply(conditions, analyse_condition)

fct <- function(d) mutate(d, condition = factor(condition, levels = lvl))
scatter_df_all <- fct(bind_rows(lapply(res, `[[`, "scatter_df")))
strain_df_all  <- fct(bind_rows(lapply(res, `[[`, "strain_df")))
gene_df_all    <- fct(bind_rows(lapply(res, `[[`, "gene_df")))
stats_all      <- fct(bind_rows(lapply(res, `[[`, "stats")))

knitr::kable(stats_all, digits = 3)
```

# ── Supplementary Figure 4 ───────────────────────────────────────
# Per-strain replicate scatter grid (raw counts, log-log, one facet per
# strain, r annotated in each facet title) — generalized from the
# original Flu.1-vs-Flu.2-only code to run on both conditions. Replaces
# the old same-strain-vs-different-strain density plot, which duplicated
# Fig 1 Panel D; this version instead gives a reviewer the strain-by-strain
# detail behind Panel D's aggregate claim.

```{r figS4, fig.width=7, fig.height=3.2}
make_strain_scatter_fig <- function(df, cond_label) {
  d <- df %>%
    filter(condition == cond_label) %>%
    mutate(facet_label = factor(facet_label, levels = unique(facet_label[order(-r)])))

  ggplot(d, aes(x = val1, y = val2)) +
    geom_point(alpha = 0.08, size = 0.15, color = condition_colors[[cond_label]]) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", color = "grey40", linewidth = 0.3) +
    scale_x_log10() +
    scale_y_log10() +
    facet_wrap(~ facet_label, ncol = 9) +
    labs(
      title = paste0("Per-strain replicate concordance: ", cond_label),
      x     = paste0(cond_label, " replicate 1 counts (log10, +1)"),
      y     = paste0(cond_label, " replicate 2 counts (log10, +1)")
    ) +
    theme_classic(base_size = 10) +
    theme(
      plot.title    = element_text(size = 18, face = "bold", hjust = 0.5),
      axis.title    = element_text(size = 16, face = "bold"),
      axis.text     = element_text(size = 16, color = "black"),
      axis.text.x   = element_text(size = 16, color = "black",
                                   angle = 90, hjust = 1, vjust = 0.5),
      strip.text    = element_text(size = 10, face = "bold"),
      panel.spacing = unit(0.6, "lines")
    )
}

figS4_YPD <- make_strain_scatter_fig(scatter_df_all, "YPD")
figS4_Flu <- make_strain_scatter_fig(scatter_df_all, "Fluconazole")

ggsave(file.path(OUT_DIR, "FigS4_YPD_per_strain_scatter.pdf"), figS4_YPD,
       width = 40, height = 35, limitsize = FALSE)
ggsave(file.path(OUT_DIR, "FigS4_Flu_per_strain_scatter.pdf"), figS4_Flu,
       width = 40, height = 35, limitsize = FALSE)
```

# Supplementary Figure 5

```{r figS5, fig.width=7, fig.height=6}
s5_long <- strain_df_all %>%
  select(condition, strain, r, mean_cells, min_cells) %>%
  pivot_longer(c(mean_cells, min_cells),
               names_to = "metric", values_to = "cells") %>%
  mutate(metric = recode(metric,
                         mean_cells = "Mean cells per strain",
                         min_cells  = "Min cells per strain")) %>%
  filter(!is.na(cells))

s5_labels <- s5_long %>%
  group_by(condition, metric) %>%
  summarise(rho = cor(cells, r, method = "spearman", use = "complete.obs"),
            .groups = "drop") %>%
  mutate(lab = sprintf("r = %.3f", rho))

figS5 <- ggplot(s5_long, aes(x = cells, y = r, color = condition)) +
  geom_point(size = 1.4, alpha = 0.75) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.6, color = "grey25") +
  geom_text(data = s5_labels, aes(x = Inf, y = -Inf, label = lab),
            inherit.aes = FALSE, hjust = 1.2, vjust = -0.8,
            size = 2.6, fontface = "bold") +
  scale_color_manual(values = condition_colors, guide = "none") +
  facet_grid(condition ~ metric, scales = "free") +
  labs(x = "Cells assigned per strain",
       y = "Spearman correlation between replicates") +
  pub_theme

figS5

ggsave(file.path(OUT_DIR, "FigS5_reproducibility_vs_cells.pdf"), figS5,
       width = 7, height = 6, units = "in")
```

# Supplementary Figure 6

```{r figS6, fig.width=7, fig.height=3.2}
cat_colors <- c("high (\u22650.7)"        = "#2166ac",
                "moderate (0.4\u20130.7)" = "#92c5de",
                "low (<0.4)"              = "#d1e5f0")

s6_labels <- stats_all %>%
  mutate(lab = sprintf("median r = %.2f\n%.0f%% above 0.4",
                       median_gene_r, pct_gene_r_0.4))

figS6 <- ggplot(filter(gene_df_all, !is.na(cor_category)),
                aes(x = r, fill = cor_category)) +
  geom_histogram(bins = 50, color = "white", linewidth = 0.2) +
  geom_vline(xintercept = c(0.4, 0.7), linetype = "dashed",
             color = "grey30", linewidth = 0.4) +
  geom_text(data = s6_labels, aes(x = -Inf, y = Inf, label = lab),
            inherit.aes = FALSE, hjust = -0.1, vjust = 1.2, size = 2.5) +
  scale_fill_manual(values = cat_colors, name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ condition, nrow = 1) +
  labs(x = "Per-gene Spearman correlation across strains",
       y = "Number of genes") +
  pub_theme +
  theme(legend.position = "bottom")

figS6

ggsave(file.path(OUT_DIR, "FigS6_per_gene_correlation.pdf"), figS6,
       width = 7, height = 3.2, units = "in")
```

# Source data

```{r source_data}
write.csv(stats_all,     file.path(OUT_DIR, "FigS4-6_summary_stats.csv"),        row.names = FALSE)
write.csv(strain_df_all, file.path(OUT_DIR, "FigS4-5_per_strain.csv"),           row.names = FALSE)
write.csv(gene_df_all,   file.path(OUT_DIR, "FigS6_per_gene_correlations.csv"),  row.names = FALSE)
```

```{r session}
sessionInfo()
```
