---
title: "Cross-condition eQTL comparison: YPD, SC-2% glucose, Fluconazole"
author: "Akriti Agrawal"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 6)
```

```{r libs}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(ggrepel)
library(scales)
library(ggvenn)
library(clusterProfiler)
library(org.Sc.sgd.db)

# resolve dplyr masking
bind_rows <- dplyr::bind_rows
count     <- dplyr::count
select    <- dplyr::select
filter    <- dplyr::filter

OUT <- "../results"
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)

cond_colors <- c(YPD = "#4CAF50", SC2Glu = "#FF9800", Fluconazole = "#9C27B0")
cond_levels <- c("YPD", "SC2Glu", "Fluconazole")
variant_levels <- c("SNPs", "InDels", "SVs", "CNVs")
```

# 1. Load inputs

Gene matrices (for phenotype/strain counts) and the annotated clumped eQTL tables
for each condition.

```{r load}
gm_path <- function(cond_dir)
  file.path("/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all",
            cond_dir, "fastlmm/data/pheno_data/gene_matrix.txt")

gene_matrix_YPD.1        <- read.table(gm_path("YPD.1"), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
gene_matrix_sc2glu       <- read.table(gm_path("sc2glu.2"), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
gene_matrix_fluconazole  <- read.table("/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/data/pheno_data/gene_matrix.txt",
                                        header = TRUE, sep = "\t", stringsAsFactors = FALSE)

all_clumped_YPD.1       <- read.table("/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results/all_clumped_eqtl_annotated_YPD.1.txt",
                                       sep = "\t", header = TRUE)
all_clumped_sc2glu      <- read_tsv("/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/all_clumped_eqtl_annotated_sc2glu.txt",
                                     show_col_types = FALSE) %>% as.data.frame()
all_clumped_fluconazole <- read.table("/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/all_clumped_eqtl_annotated_fluc.txt",
                                       sep = "\t", header = TRUE)
```

# 2. Between-strain expression variance (power QC)

**Why this comes first.** eQTL detection depends on a gene's expression varying
across strains — a gene with near-constant expression carries no signal for any
variant to explain, so it can never yield an eQTL regardless of genetics. Before
comparing eQTL yields across conditions, we check whether the three conditions
offer *equal variance to map on*. If one condition has systematically lower
between-strain variance, its lower eQTL count is partly a data/power property,
not necessarily "less genetic regulation."

For genes and strains shared between conditions we compute per-gene variance
across strains and compare the distributions (paired Wilcoxon; density, boxplot,
per-gene scatter). Conclusion carried downstream: between-strain variance is
highest in YPD, intermediate in SC-2% glucose, lowest in fluconazole — so
absolute eQTL counts are not directly comparable, and the comparison leans on
proportions and effect-size concordance rather than raw counts.

```{r var-matrices}
library(tibble)

# first column = gene ID (matches Section 4's ncol-1 strain count), whatever it is named
to_mat <- function(gm) {
  m <- as.matrix(gm[, -1, drop = FALSE]); rownames(m) <- gm[[1]]; m
}
ypd_mat         <- to_mat(gene_matrix_YPD.1)
sc2glu_mat      <- to_mat(gene_matrix_sc2glu)
fluconazole_mat <- to_mat(gene_matrix_fluconazole)

cat("YPD matrix:",         nrow(ypd_mat),         "genes x", ncol(ypd_mat),         "strains\n")
cat("SC 2% glucose:",      nrow(sc2glu_mat),      "genes x", ncol(sc2glu_mat),      "strains\n")
cat("Fluconazole matrix:", nrow(fluconazole_mat), "genes x", ncol(fluconazole_mat), "strains\n")

# per-gene variance across strains, on any gene x strain subset
var_across <- function(mat, genes, strains) apply(mat[genes, strains, drop = FALSE], 1, var, na.rm = TRUE)
```

```{r var-3way}
# --- genes/strains shared by all three conditions ---
g3 <- Reduce(intersect, list(rownames(ypd_mat), rownames(sc2glu_mat), rownames(fluconazole_mat)))
s3 <- Reduce(intersect, list(colnames(ypd_mat), colnames(sc2glu_mat), colnames(fluconazole_mat)))
cat("All-3 shared:", length(g3), "genes x", length(s3), "strains\n")

var3 <- data.frame(
  gene        = g3,
  YPD         = var_across(ypd_mat,         g3, s3),
  SC2Glu      = var_across(sc2glu_mat,      g3, s3),
  Fluconazole = var_across(fluconazole_mat, g3, s3))

var3_long <- var3 %>%
  pivot_longer(c(YPD, SC2Glu, Fluconazole), names_to = "condition", values_to = "variance") %>%
  mutate(condition = factor(condition, levels = cond_levels)) %>%
  filter(!is.na(variance))

med3 <- var3_long %>% group_by(condition) %>%
  summarise(median_val = median(variance, na.rm = TRUE), .groups = "drop") %>%
  mutate(median_log = log2(median_val + 1))
cat("Median variance (all-3 shared):\n"); print(med3[, c("condition","median_val")])

# paired Wilcoxon among the 3-way shared set
cat("\nPaired Wilcoxon (two-sided), all-3 shared:\n")
cat("  YPD vs SC2Glu:      p =", format(wilcox.test(var3$YPD, var3$SC2Glu, paired = TRUE)$p.value, digits = 3), "\n")
cat("  YPD vs Fluconazole: p =", format(wilcox.test(var3$YPD, var3$Fluconazole, paired = TRUE)$p.value, digits = 3), "\n")
cat("  SC2Glu vs Fluc:     p =", format(wilcox.test(var3$SC2Glu, var3$Fluconazole, paired = TRUE)$p.value, digits = 3), "\n")

# which condition has the highest variance per gene
winner <- var3 %>% mutate(highest = case_when(
  YPD >= SC2Glu & YPD >= Fluconazole ~ "YPD",
  SC2Glu >= YPD & SC2Glu >= Fluconazole ~ "SC2Glu",
  TRUE ~ "Fluconazole")) %>%
  count(highest) %>% mutate(pct = round(100 * n / sum(n), 1)) %>% arrange(desc(n))
cat("\nGene wins highest variance:\n"); print(winner)
```

```{r var-density, fig.width=8, fig.height=5}
cond_labs <- c(YPD = "YPD", SC2Glu = "SC 2% glucose", Fluconazole = "Fluconazole")

figV1 <- ggplot(var3_long, aes(log2(variance + 1), fill = condition, color = condition)) +
  geom_density(alpha = 0.4, linewidth = 0.8) +
  geom_vline(data = med3, aes(xintercept = median_log, color = condition),
             linetype = "dashed", linewidth = 1) +
  geom_text(data = med3,
            aes(x = median_log, y = Inf, color = condition,
                label = paste0(cond_labs[as.character(condition)], "\nmedian=", round(median_val, 2))),
            vjust = c(1.3, 3.8, 6.3), hjust = -0.07, size = 3.8, fontface = "bold", show.legend = FALSE) +
  scale_fill_manual(values = cond_colors, labels = cond_labs) +
  scale_color_manual(values = cond_colors, labels = cond_labs) +
  labs(title = "Gene expression variance across strains",
       subtitle = paste0("Shared ", length(g3), " genes x ", length(s3), " strains"),
       x = "log2(variance + 1)", y = "Density", fill = "Condition", color = "Condition") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
        axis.title = element_text(face = "bold"))
print(figV1)
ggsave(file.path(OUT, "figV1_variance_density.tiff"), figV1,
       width = 8, height = 5, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

```{r var-box, fig.width=6, fig.height=5}
figV2 <- ggplot(var3_long, aes(condition, log2(variance + 1), fill = condition)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
  scale_fill_manual(values = cond_colors, guide = "none") +
  scale_x_discrete(labels = cond_labs) +
  labs(title = "Gene expression variance across strains",
       subtitle = paste0("Shared ", length(g3), " genes x ", length(s3), " strains"),
       x = NULL, y = "log2(variance + 1)") +
  theme_classic(base_size = 13) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
        axis.title = element_text(face = "bold"))
print(figV2)
ggsave(file.path(OUT, "figV2_variance_boxplot.tiff"), figV2,
       width = 6, height = 5, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

```{r var-scatter, fig.width=6, fig.height=5}
# pairwise per-gene scatter, each on its own shared gene/strain set (n differs by pair)
scatter_pair <- function(mA, mB, labA, labB) {
  g <- intersect(rownames(mA), rownames(mB)); s <- intersect(colnames(mA), colnames(mB))
  d <- data.frame(vA = var_across(mA, g, s), vB = var_across(mB, g, s))
  p <- format(wilcox.test(d$vA, d$vB, paired = TRUE)$p.value, digits = 3)
  ggplot(d, aes(log2(vB + 1), log2(vA + 1))) +
    geom_point(alpha = 0.2, size = 0.8, color = "grey40") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
    annotate("text", x = Inf, y = -Inf, label = paste0("Wilcoxon p = ", p),
             hjust = 1.1, vjust = -0.5, size = 4, fontface = "bold") +
    labs(title = paste0("Variance: ", labA, " vs ", labB),
         subtitle = paste0(length(s), " shared strains"),
         x = paste0("log2(variance + 1) - ", labB),
         y = paste0("log2(variance + 1) - ", labA)) +
    theme_classic(base_size = 13) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
          axis.title = element_text(face = "bold"))
}
print(scatter_pair(ypd_mat,    fluconazole_mat, "YPD",           "Fluconazole"))
print(scatter_pair(ypd_mat,    sc2glu_mat,      "YPD",           "SC 2% glucose"))
print(scatter_pair(sc2glu_mat, fluconazole_mat, "SC 2% glucose", "Fluconazole"))
```

# 3. Combine once

All downstream chunks reuse `combined`. `CT` normalizes the cis/trans column
(the tables carry either `cis_trans` or `Location`).

```{r combine}
all_clumped_YPD.1$condition       <- "YPD"
all_clumped_sc2glu$condition      <- "SC2Glu"
all_clumped_fluconazole$condition <- "Fluconazole"

get_cistrans <- function(df) if ("cis_trans" %in% names(df)) df$cis_trans else df$Location
signed_col   <- if ("SnpWeight" %in% names(all_clumped_YPD.1)) "SnpWeight" else "EffectSize"

combined <- bind_rows(all_clumped_YPD.1, all_clumped_sc2glu, all_clumped_fluconazole)
combined$CT        <- get_cistrans(combined)
combined$condition <- factor(combined$condition, levels = cond_levels)

# eGene sets and SNP-gene pair sets, reused throughout
genes_YPD    <- unique(all_clumped_YPD.1$Pheno)
genes_sc2glu <- unique(all_clumped_sc2glu$Pheno)
genes_flu    <- unique(all_clumped_fluconazole$Pheno)

ypd_pairs    <- unique(paste(all_clumped_YPD.1$SNP, all_clumped_YPD.1$Pheno, sep = "_"))
sc2glu_pairs <- unique(paste(all_clumped_sc2glu$SNP, all_clumped_sc2glu$Pheno, sep = "_"))
flu_pairs    <- unique(paste(all_clumped_fluconazole$SNP, all_clumped_fluconazole$Pheno, sep = "_"))
```

# 4. eQTL detection and experimental scale (Supplementary)

The single count-based figure, kept as **supplementary**. Absolute eQTL numbers
track strain count and between-strain variance (Section 2), so a drop under
fluconazole is expected on power grounds and is *not* interpreted as biology.
Everything downstream uses proportions and effect-size concordance instead.

```{r scale, fig.width=7, fig.height=4.5}
pheno_strain_summary <- data.frame(
  condition    = cond_levels,
  n_phenotypes = c(nrow(gene_matrix_YPD.1), nrow(gene_matrix_sc2glu), nrow(gene_matrix_fluconazole)),
  n_strains    = c(ncol(gene_matrix_YPD.1) - 1, ncol(gene_matrix_sc2glu) - 1, ncol(gene_matrix_fluconazole) - 1),
  n_eQTLs      = c(nrow(all_clumped_YPD.1), nrow(all_clumped_sc2gflu), nrow(all_clumped_fluconazole)))
print(pheno_strain_summary)

figS_scale <- pheno_strain_summary %>%
  pivot_longer(c(n_phenotypes, n_strains, n_eQTLs), names_to = "metric", values_to = "count") %>%
  mutate(condition = factor(condition, levels = cond_levels),
         metric = factor(metric, levels = c("n_phenotypes","n_strains","n_eQTLs"),
                         labels = c("Phenotypes (genes)","Strains","eQTLs detected"))) %>%
  ggplot(aes(metric, count, fill = condition)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = count), position = position_dodge(0.7), vjust = -0.4, size = 3.3, fontface = "bold") +
  scale_fill_manual(values = cond_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = "Count", fill = "Condition",
       title = "Experimental scale and eQTL detection (supplementary)") +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 30, hjust = 1))
print(figS_scale)
ggsave(file.path(OUT, "figS_scale_eqtl_detection.tiff"), figS_scale,
       width = 7, height = 4.5, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

**Supplementary:** genomic distribution of eQTLs per condition and variant type
(eQTL count per chromosome). Descriptive only — shows where detected eQTLs fall
along the genome in each condition; goes in the supplement alongside the scale
figure above.

```{r geno-dist-supp, fig.width=10, fig.height=6}
# ---- SUPPLEMENTARY figure ----
roman_lv <- c("I","II","III","IV","V","VI","VII","VIII",
              "IX","X","XI","XII","XIII","XIV","XV","XVI")
norm_chr_lab <- function(x){
  x <- as.character(x)
  x <- gsub("^chr(omosome)?[_ ]?", "", x, ignore.case = TRUE)
  num <- suppressWarnings(as.integer(x))
  ifelse(!is.na(num) & num %in% 1:16, roman_lv[num],
         ifelse(x %in% roman_lv, x, NA_character_))
}
vchr_col <- intersect(c("Chr","var_chr","SNP_Chr","chromosome","var_chr_roman"),
                      names(combined))[1]

geno_ok <- tryCatch({
  gd <- combined %>%
    mutate(chr = factor(norm_chr_lab(.data[[vchr_col]]), levels = roman_lv),
           Genetic_variants = factor(Genetic_variants, levels = variant_levels),
           condition = factor(condition, levels = cond_levels)) %>%
    filter(!is.na(chr)) %>%
    count(condition, Genetic_variants, chr)

  variant_colors <- c(SNPs="#4472C4", InDels="#FFC000", SVs="#C41E3A", CNVs="#70AD47")
  figS_geno <- ggplot(gd, aes(chr, n, fill = Genetic_variants)) +
    geom_col(position = position_dodge2(preserve = "single", padding = 0.15),
             width = 0.8, color = "grey30", linewidth = 0.15) +
    facet_grid(condition ~ ., scales = "free_y") +
    scale_fill_manual(values = variant_colors, name = "Variant type") +
    labs(x = "Chromosome", y = "eQTLs detected",
         title = "Genomic distribution of eQTLs by condition and variant type (supplementary)") +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          strip.text.y = element_text(face = "bold", angle = 0),
          panel.grid.minor = element_blank(),
          legend.position = "top",
          axis.text.x = element_text(size = 8))
  print(figS_geno)
  ggsave(file.path(OUT, "figS_genomic_distribution_by_condition_type.tiff"), figS_geno,
         width = 10, height = 6, dpi = 600, units = "in", device = "tiff", compression = "lzw")
  TRUE
}, error = function(e){ cat("Genomic distribution skipped:", conditionMessage(e),
   "\n(no recognizable variant chromosome column in `combined`)\n"); FALSE })
```

# 5. Effect size by variant type and location

Effect size is split by **variant type** (facet columns: SNPs, InDels, SVs,
CNVs) and **location** (x-axis: Cis, Trans), with the **three conditions shown
side by side as coloured violins** within each cell. This makes the
cross-condition comparison the primary read: for a given variant type and
location, is |β| the same in YPD, SC-2%-glucose, and fluconazole? The cis/trans
split still carries the secondary message — SVs/CNVs look small because they are
almost all trans, and trans < cis within every type. A per-cell Kruskal-Wallis
tests whether the three conditions differ in |β|, printed where at least two
conditions have ≥3 eQTLs.


What the test is. Within one panel (say SNPs) at one location (say Cis), you have three groups of |β| values — the YPD cis-SNP effect sizes, the SC2Glu cis-SNP effect sizes, the fluconazole cis-SNP effect sizes. The Kruskal-Wallis test asks a single question: are these three distributions of effect size the same, or does at least one condition differ? It's the non-parametric (rank-based) version of a one-way ANOVA — it doesn't assume the effect sizes are normally distributed, which they aren't.

So in the SNPs panel:

the left KW p-value = "do YPD, SC2Glu, fluconazole differ in cis-SNP effect size?"
the right KW p-value = "do they differ in trans-SNP effect size?"

```{r effect-size, fig.width=12, fig.height=6}
combined$abs_beta <- abs(combined[[signed_col]])

es <- combined %>%
  filter(CT %in% c("Cis", "Trans"), !is.na(abs_beta)) %>%
  mutate(Genetic_variants = factor(Genetic_variants, levels = variant_levels),
         condition        = factor(condition, levels = cond_levels),
         CT               = factor(CT, levels = c("Cis", "Trans")))

# per-condition eQTL counts (shown under each violin)
n_lab <- es %>% count(Genetic_variants, CT, condition, name = "n") %>%
  mutate(condition = factor(condition, levels = cond_levels))

# Within each variant type x location cell, compare the THREE conditions
# (Kruskal-Wallis) where each condition has >= 3 eQTLs.
p_lab <- es %>% group_by(Genetic_variants, CT) %>%
  summarise(p = {
    ok <- table(condition[abs_beta >= 0])
    if (sum(ok >= 3) >= 2 && n_distinct(condition[!is.na(abs_beta)]) >= 2)
      tryCatch(format.pval(kruskal.test(abs_beta ~ condition)$p.value, digits = 2),
               error = function(e) NA_character_)
    else NA_character_
  }, .groups = "drop") %>%
  filter(!is.na(p)) %>% mutate(label = paste0("KW p = ", p))

dodge <- position_dodge(width = 0.8)
figES <- ggplot(es, aes(CT, abs_beta, fill = condition)) +
  geom_violin(trim = FALSE, alpha = 0.6, scale = "width", position = dodge,
              width = 0.75, linewidth = 0.3) +
  geom_boxplot(aes(group = interaction(CT, condition)),
               width = 0.12, fill = "white", outlier.shape = NA, position = dodge,
               linewidth = 0.3) +
  facet_grid(. ~ Genetic_variants) +
  geom_text(data = p_lab, aes(x = as.integer(factor(CT, levels = c("Cis","Trans"))),
                              y = Inf, label = label),
            inherit.aes = FALSE, vjust = 1.4, size = 2.7) +
  geom_text(data = n_lab, aes(x = CT, y = 0, label = paste0("n=", n),
                              color = condition, group = condition),
            position = dodge, vjust = 1.5, size = 2.5, fontface = "bold",
            show.legend = FALSE) +
  scale_fill_manual(values = cond_colors, name = "Condition") +
  scale_color_manual(values = cond_colors, guide = "none") +
  coord_cartesian(ylim = c(0, quantile(es$abs_beta, 0.99, na.rm = TRUE) * 1.15),
                  clip = "off") +
  labs(x = "Variant location", y = "Absolute effect size (|\u03b2|)",
       title = "Effect size by variant type and location: conditions side by side") +
  theme_bw(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        strip.text = element_text(face = "bold"),
        legend.position = "top",
        panel.grid.minor = element_blank(),
        plot.margin = margin(t = 10, r = 15, b = 25, l = 10))
print(figES)
ggsave(file.path(OUT, "fig_effect_size_by_type_and_location.tiff"), figES,
       width = 12, height = 6, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

The cis/trans composition itself (few cis for SV/CNV) is reported as counts —
this is the number the effect-size story rests on:

```{r cis-trans-numbers}
cat("Cis/trans counts by variant type and condition:\n")
print(with(combined, table(Genetic_variants, CT, condition)))
```

# 6. Stress-vs-baseline eQTL overlap (two pairwise Venns)

SNP-gene pairs are the sharing currency (shared *regulation*, not just a shared
gene). The design is deliberately **two pairwise comparisons, both centred on
fluconazole**: YPD vs Fluconazole and SC-2%-glucose vs Fluconazole. YPD and
SC2Glu are near-identical baselines, so each pairwise Venn is an independent
read of the same question — *what does fluconazole change relative to normal
growth?* Agreement between the two panels replicates the drug signature across
two baselines. The YPD-vs-SC2Glu overlap (the two baselines against each other)
is reported only as a sanity check in the supplement — it should be large.

```{r overlap-pairs, fig.width=12, fig.height=6}
# pairwise-shared SNP-gene pair sets (reused downstream)
shared_ypd_flu <- intersect(ypd_pairs, flu_pairs)
shared_sc2_flu <- intersect(sc2glu_pairs, flu_pairs)
shared_ypd_sc2 <- intersect(ypd_pairs, sc2glu_pairs)
shared_all3    <- Reduce(intersect, list(ypd_pairs, flu_pairs, sc2glu_pairs))
shared_any     <- unique(c(shared_ypd_flu, shared_sc2_flu, shared_ypd_sc2))

cat("SNP-gene pair overlap:\n")
cat("  YPD-Flu shared:   ", length(shared_ypd_flu),
    "| YPD only:", length(setdiff(ypd_pairs, flu_pairs)),
    "| Flu only:", length(setdiff(flu_pairs, ypd_pairs)), "\n")
cat("  SC2Glu-Flu shared:", length(shared_sc2_flu),
    "| SC2Glu only:", length(setdiff(sc2glu_pairs, flu_pairs)),
    "| Flu only:", length(setdiff(flu_pairs, sc2glu_pairs)), "\n")
cat("  YPD-SC2Glu shared (baseline sanity):", length(shared_ypd_sc2), "\n")

p_venn_yf <- ggvenn(list(YPD = ypd_pairs, Fluconazole = flu_pairs),
       fill_color = unname(cond_colors[c("YPD","Fluconazole")]),
       set_name_size = 5, text_size = 4.5, show_percentage = FALSE) +
  ggtitle("YPD vs Fluconazole") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))

p_venn_sf <- ggvenn(list(SC2Glu = sc2glu_pairs, Fluconazole = flu_pairs),
       fill_color = unname(cond_colors[c("SC2Glu","Fluconazole")]),
       set_name_size = 5, text_size = 4.5, show_percentage = FALSE) +
  ggtitle("SC-2%-glucose vs Fluconazole") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))

print(p_venn_yf); print(p_venn_sf)
ggsave(file.path(OUT, "fig_venn_YPD_vs_Flu.tiff"), p_venn_yf, width = 6, height = 6,
       dpi = 600, units = "in", device = "tiff", compression = "lzw")
ggsave(file.path(OUT, "fig_venn_SC2Glu_vs_Flu.tiff"), p_venn_sf, width = 6, height = 6,
       dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

```{r overlap-egene-supp, fig.width=12, fig.height=6}
total_genes_ref <- read.csv("/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv") %>%
  filter(gene_biotype == "protein_coding") %>% pull(gene_id) %>% unique() %>% length()

run_fisher_pair <- function(setA, setB, nameA, nameB, total) {
  overlap <- intersect(setA, setB)
  mat <- matrix(c(length(overlap), length(setdiff(setA, setB)),
                  length(setdiff(setB, setA)), total - length(union(setA, setB))),
                nrow = 2, byrow = TRUE)
  res <- fisher.test(mat)
  cat(sprintf("%s vs %s: overlap=%d  OR=%.2f  p=%.3g\n",
              nameA, nameB, length(overlap), res$estimate, res$p.value))
  res
}
cat("=== eGene overlap Fisher tests (supplement) ===\n")
invisible(run_fisher_pair(genes_YPD, genes_flu,    "YPD", "Flu",    total_genes_ref))
invisible(run_fisher_pair(genes_sc2glu, genes_flu, "SC2Glu", "Flu", total_genes_ref))
invisible(run_fisher_pair(genes_YPD, genes_sc2glu, "YPD", "SC2Glu (baseline)", total_genes_ref))

pe_yf <- ggvenn(list(YPD = genes_YPD, Fluconazole = genes_flu),
       fill_color = unname(cond_colors[c("YPD","Fluconazole")]),
       set_name_size = 5, text_size = 4.5, show_percentage = FALSE) +
  ggtitle("eGene overlap: YPD vs Fluconazole (supp)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13))
pe_sf <- ggvenn(list(SC2Glu = genes_sc2glu, Fluconazole = genes_flu),
       fill_color = unname(cond_colors[c("SC2Glu","Fluconazole")]),
       set_name_size = 5, text_size = 4.5, show_percentage = FALSE) +
  ggtitle("eGene overlap: SC2Glu vs Fluconazole (supp)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13))
print(pe_yf); print(pe_sf)
```

# 7. MAIN FIGURE A — condition-sharing of eQTLs by variant type

Complementary to the Venn: the Venn gives pooled shared/unique counts; this
breaks the *same* sharing down by variant type, asking whether structural
variants (SV/CNV) are more condition-labile than SNPs. Keep only if that
variant-type-lability contrast is a point in the paper.

```{r figA-sharing, fig.width=7, fig.height=5}
pair_tab <- combined %>%
  mutate(pair = paste(SNP, Pheno, sep = "_")) %>%
  distinct(pair, condition, Genetic_variants)

share_prop <- pair_tab %>%
  group_by(pair, Genetic_variants) %>%
  summarise(n_cond = n_distinct(condition), .groups = "drop") %>%
  mutate(sharing = case_when(n_cond == 3 ~ "All three",
                             n_cond == 2 ~ "Two conditions",
                             TRUE        ~ "Condition-specific")) %>%
  count(Genetic_variants, sharing) %>%
  group_by(Genetic_variants) %>% mutate(prop = n / sum(n)) %>% ungroup() %>%
  mutate(sharing = factor(sharing, levels = c("Condition-specific","Two conditions","All three")),
         Genetic_variants = factor(Genetic_variants, levels = variant_levels))

figA <- ggplot(share_prop, aes(Genetic_variants, prop, fill = sharing)) +
  geom_col(width = 0.72, color = "black", linewidth = 0.3) +
  geom_text(aes(label = ifelse(prop > 0.04, percent(prop, 1), "")),
            position = position_stack(vjust = 0.5), size = 3.2, fontface = "bold") +
  scale_fill_manual(values = c("Condition-specific" = "#B0BEC5",
                               "Two conditions" = "#FFB74D", "All three" = "#455A64"), name = NULL) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "Fraction of eQTLs", title = "Condition-sharing of eQTLs by variant type") +
  theme_classic(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "top", axis.text.x = element_text(face = "bold"))
print(figA)
ggsave(file.path(OUT, "figA_sharing_by_variant_type.tiff"), figA,
       width = 7, height = 5, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

# 8. MAIN FIGURE B — effect-size concordance of shared eQTLs (all pairwise)

For eQTLs (SNP-gene pairs) detected in **any** two conditions, do the effect
sizes agree, and are there sign flips (regulatory reversal)? One panel per
condition-pair; N and sign-flip count reported per panel.

```{r figB-concordance, fig.width=10, fig.height=4.5}
mk <- function(df) df %>% transmute(pair = paste(SNP, Pheno, sep = "_"),
                                    beta = .data[[signed_col]], Genetic_variants)
build_pair <- function(dfA, dfB, labA, labB) {
  inner_join(mk(dfA), mk(dfB), by = c("pair","Genetic_variants"), suffix = c("_A","_B")) %>%
    mutate(comparison = paste0(labA, " vs ", labB))
}
shared_beta <- bind_rows(
  build_pair(all_clumped_YPD.1,  all_clumped_fluconazole, "YPD",    "Flu"),
  build_pair(all_clumped_sc2glu, all_clumped_fluconazole, "SC2Glu", "Flu"),
  build_pair(all_clumped_YPD.1,  all_clumped_sc2glu,      "YPD",    "SC2Glu")) %>%
  mutate(sign_flip = sign(beta_A) != sign(beta_B),
         comparison = factor(comparison, levels = c("YPD vs Flu","SC2Glu vs Flu","YPD vs SC2Glu")))

concordance_summary <- shared_beta %>% group_by(comparison) %>%
  summarise(n_shared = n(), n_sign_flip = sum(sign_flip),
            pct_flip = round(100 * mean(sign_flip), 1),
            pearson_r = round(cor(beta_A, beta_B), 3), .groups = "drop")
cat("Effect-size concordance per condition-pair:\n"); print(concordance_summary)

figB <- ggplot(shared_beta, aes(beta_A, beta_B, color = sign_flip)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.6, size = 1.6) +
  geom_text(data = concordance_summary, inherit.aes = FALSE,
            aes(x = -Inf, y = Inf, label = paste0("n=", n_shared, ", r=", pearson_r,
                                                   "\nflips=", n_sign_flip, " (", pct_flip, "%)")),
            hjust = -0.08, vjust = 1.2, size = 3.2, fontface = "bold") +
  facet_wrap(~ comparison, ncol = 3) +
  scale_color_manual(values = c("FALSE" = "#455A64", "TRUE" = "#C41E3A"),
                     labels = c("Concordant", "Sign flip"), name = NULL) +
  labs(x = "Effect size (condition 1)", y = "Effect size (condition 2)",
       title = "Effect-size concordance of shared eQTLs") +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "top",
        strip.background = element_rect(fill = "grey90", color = NA),
        strip.text = element_text(face = "bold"))
print(figB)
ggsave(file.path(OUT, "figB_effect_concordance_all_pairs.tiff"), figB,
       width = 10, height = 4.5, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```




# 10. GO enrichment — two stress-vs-baseline comparisons

Two comparative GO dotplots, one per baseline, each with three eGene categories:

- **Plot 1 (YPD vs Fluconazole):** YPD-only, Flu-only, Shared YPD-Flu
- **Plot 2 (SC2Glu vs Fluconazole):** SC2Glu-only, Flu-only, Shared SC2Glu-Flu

Same rationale as the Venns — keep the stress against one baseline at a time,
so the x-axis reads as *lost under stress / gained under stress / robust to
stress* rather than a six-way grid. If the **Flu-only** column shows the same
enriched terms in both plots, the fluconazole response is reproducible across
baselines. Over-representation vs the protein-coding universe (`enrichGO`, ORF
keys, `ont="ALL"`, BH, p < 0.05, q < 0.1); categories with <10 genes are skipped.

```{r go-setup}
library(clusterProfiler)
library(org.Sc.sgd.db)

tested_universe <- read.csv("/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv") %>%
  filter(gene_biotype == "protein_coding") %>% pull(gene_id) %>% unique()

run_enrichGO <- function(gene_list, label, universe, ontology = "ALL") {
  g <- intersect(gene_list, universe)
  cat("\n", label, "| genes in universe:", length(g), "\n")
  if (length(g) < 10) { cat("  too few genes (<10), skipping\n"); return(NULL) }
  res <- enrichGO(gene = g, universe = universe, OrgDb = org.Sc.sgd.db, keyType = "ORF",
                  ont = ontology, pAdjustMethod = "BH", pvalueCutoff = 0.05, qvalueCutoff = 0.1)
  if (is.null(res) || nrow(res) == 0) { cat("  no significant terms\n"); return(NULL) }
  cat("  significant terms:", nrow(res), "\n"); res
}

# build one comparative dotplot from a named list of gene sets
plot_go_pair <- function(go_lists, title, outfile) {
  cat("\n===", title, "=== eGene category sizes:\n"); print(sapply(go_lists, length))
  go_res <- lapply(names(go_lists), function(g)
              run_enrichGO(go_lists[[g]], g, tested_universe)) %>% setNames(names(go_lists))
  go_res <- go_res[!sapply(go_res, is.null)]
  if (length(go_res) < 2) {
    cat("  Fewer than 2 categories with significant terms — dotplot skipped.\n"); return(invisible(NULL))
  }
  SHOW_N <- 10   # terms per category
  cg <- merge_result(go_res)
  cg@compareClusterResult$Cluster <- factor(cg@compareClusterResult$Cluster, levels = names(go_lists))
  # wide wrap: keep most terms on ONE line; only very long ones break
  cg@compareClusterResult$Description <- stringr::str_wrap(cg@compareClusterResult$Description, width = 60)

  p <- dotplot(cg, showCategory = SHOW_N) + ggtitle(title) + theme_bw(base_size = 13) +
    theme(plot.title  = element_text(hjust = 0.5, face = "bold", size = 15, margin = margin(b = 10)),
          axis.text.y = element_text(size = 8.5, color = "black", lineheight = 0.85),
          axis.text.x = element_text(size = 11, face = "bold", color = "black", angle = 30, hjust = 1),
          axis.title.x = element_text(size = 12, face = "bold"), axis.title.y = element_blank(),
          legend.title = element_text(size = 11, face = "bold"), legend.text = element_text(size = 10))
  print(p)
  # height scales with the number of distinct terms actually shown, so rows never overlap
  n_terms <- length(unique(p$data$Description))
  ggsave(file.path(OUT, outfile), p, width = 10, height = max(7, 0.42 * n_terms + 2),
         units = "in", dpi = 300, device = "tiff", compression = "lzw", limitsize = FALSE)
  invisible(p)
}
```

```{r go-YPD-vs-Flu, fig.width=10, fig.height=10}
plot_go_pair(list(
  "YPD-only"       = setdiff(genes_YPD, genes_flu),
  "Flu-only"       = setdiff(genes_flu, genes_YPD),
  "Shared YPD-Flu" = intersect(genes_YPD, genes_flu)),
  "GO enrichment: YPD vs Fluconazole", "GO_comparison_YPD_vs_Flu.tiff")
```

```{r go-SC2Glu-vs-Flu, fig.width=10, fig.height=10}
plot_go_pair(list(
  "SC2Glu-only"       = setdiff(genes_sc2glu, genes_flu),
  "Flu-only"          = setdiff(genes_flu, genes_sc2glu),
  "Shared SC2Glu-Flu" = intersect(genes_sc2glu, genes_flu)),
  "GO enrichment: SC-2%-glucose vs Fluconazole", "GO_comparison_SC2Glu_vs_Flu.tiff")
```

# 11. Fluconazole-specific eQTLs — candidate drug-response loci

Fluconazole-specific SNP-gene pairs (present in fluconazole, in neither control),
split by whether the target gene is also a baseline eGene, then cross-referenced
against the linear-regression over-expressed gene list.

```{r flu-specific}
genes_baseline_any <- union(genes_YPD, genes_sc2glu)
flu_specific_pairs <- setdiff(flu_pairs, union(ypd_pairs, sc2glu_pairs))
all_clumped_flu_specific <- all_clumped_fluconazole %>%
  filter(paste(SNP, Pheno, sep = "_") %in% flu_specific_pairs)

genes_shared_gene_new_variant <- all_clumped_flu_specific %>%
  filter(Pheno %in% genes_baseline_any) %>% pull(Pheno) %>% unique()
genes_fully_novel <- all_clumped_flu_specific %>%
  filter(!Pheno %in% genes_baseline_any) %>% pull(Pheno) %>% unique()
cat("Flu-specific variant, baseline gene:", length(genes_shared_gene_new_variant), "\n")
cat("Fully novel (gene + variant):", length(genes_fully_novel), "\n")

lr_path <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/significant_genes_E2_fluc_from_linear_Regression.txt"
if (file.exists(lr_path)) {
  overexp_genes_lr <- unique(read.table(lr_path, header = TRUE, stringsAsFactors = FALSE)$ORF)
  cat("\nOverlap with LR over-expressed genes:\n")
  cat("  baseline-gene set:", length(intersect(overexp_genes_lr, genes_shared_gene_new_variant)), "\n")
  cat("  novel set:", length(intersect(overexp_genes_lr, genes_fully_novel)), "\n")
}
```

# 12. Session info

```{r session}
sessionInfo()
```

