---
title: "eQTL Concordance and GO Enrichment Analysis Across Conditions"
author: "Akriti Agrawal"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(ggplot2)
library(gridExtra)
library(clusterProfiler)
library(org.Sc.sgd.db)
library(enrichplot)
library(stringr)

# resolve dplyr masking
bind_rows <- dplyr::bind_rows
count     <- dplyr::count
select    <- dplyr::select
filter    <- dplyr::filter
```

# STEP 1: Load and Prepare Data

```{r load-data}
# ── YPD.1 ────────────────────────────────────────────────────
all_snps_eqtl_YPD1 <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results/all_signif_SNP_YPD.1_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

all_CNV_eqtl_YPD1 <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results/all_signif_CNV_YPD.1_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

all_indel_eqtl_YPD1 <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results/all_signif_Indels_YPD.1_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

all_SV_eqtl_YPD1 <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results/all_signif_SVs_YPD.1_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

# ── fluconazole ────────────────────────────────────────────────────
all_snps_eqtl_FLU <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/all_signif_SNP_flu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

all_CNV_eqtl_FLU <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/all_signif_CNV_flu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

all_indel_eqtl_FLU <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/all_signif_Indels_flu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

all_SV_eqtl_FLU <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/all_signif_SVs_flu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

# ── sc2glu ────────────────────────────────────────────────────
all_snps_eqtl_SC2GLU <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/all_signif_SNP_sc2glu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

all_CNV_eqtl_SC2GLU <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/all_signif_CNV_sc2glu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

all_indel_eqtl_SC2GLU <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/all_signif_Indels_sc2glu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

all_SV_eqtl_SC2GLU <- read.table(
  "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/all_signif_SVs_sc2glu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt",
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
)

cat("✓ Data loaded for all variant types and conditions\n\n")
```

# STEP 2: Directory Definitions and Output Paths

```{r setup-directories}
DIRS_flu <- list(
  SNPs   = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/results_SNPs_968_with_double_id_final_filtered_for_ncount_less_than_5/",
  CNVs   = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/results_CNVs_with_double_id_final_filtered_for_ncount_less_than_5",
  InDels = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/results_Indels_with_double_id_final_filtered_for_ncount_less_than_5",
  SVs    = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/results_SVs_with_double_id_final_filtered_for_ncount_less_than_5"
)

DIRS_YPD1 <- list(
  SNPs   = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results/results_SNPs_968_with_double_id_final_filtered_for_ncount_less_than_5/",
  CNVs   = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results/results_CNVs_with_double_id_final_filtered_for_ncount_less_than_5",
  InDels = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results/results_Indels_with_double_id_final_filtered_for_ncount_less_than_5",
  SVs    = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results/results_SVs_with_double_id_final_filtered_for_ncount_less_than_5"
)

DIRS_sc2glu <- list(
  SNPs   = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/results_SNPs_968_with_double_id_final_filtered_for_ncount_less_than_5_run2/",
  CNVs   = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/results_CNVs_with_double_id_final_filtered_for_ncount_less_than_5",
  InDels = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/results_Indels_with_double_id_final_filtered_for_ncount_less_than_5",
  SVs    = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/results_SVs_with_double_id_final_filtered_for_ncount_less_than_5"
)

OUTPUT_DIR <- "../results/"
if (!dir.exists(OUTPUT_DIR)) {
  dir.create(OUTPUT_DIR, recursive = TRUE)
}

cat("✓ Directory mappings defined\n")
cat("✓ Output directory ready:", OUTPUT_DIR, "\n\n")
```

# STEP 3: Combine All Variant Types into Single "Significant" Datasets

```{r combine-variants}
flu_signif_combined <- bind_rows(
  all_snps_eqtl_FLU %>% mutate(Variant_Type = "SNPs"),
  all_CNV_eqtl_FLU %>% mutate(Variant_Type = "CNVs"),
  all_indel_eqtl_FLU %>% mutate(Variant_Type = "InDels"),
  all_SV_eqtl_FLU %>% mutate(Variant_Type = "SVs")
)

ypd1_signif_combined <- bind_rows(
  all_snps_eqtl_YPD1 %>% mutate(Variant_Type = "SNPs"),
  all_CNV_eqtl_YPD1 %>% mutate(Variant_Type = "CNVs"),
  all_indel_eqtl_YPD1 %>% mutate(Variant_Type = "InDels"),
  all_SV_eqtl_YPD1 %>% mutate(Variant_Type = "SVs")
)

sc2glu_signif_combined <- bind_rows(
  all_snps_eqtl_SC2GLU %>% mutate(Variant_Type = "SNPs"),
  all_CNV_eqtl_SC2GLU %>% mutate(Variant_Type = "CNVs"),
  all_indel_eqtl_SC2GLU %>% mutate(Variant_Type = "InDels"),
  all_SV_eqtl_SC2GLU %>% mutate(Variant_Type = "SVs")
)

cat("Fluconazole significant eQTLs (all variants combined):", nrow(flu_signif_combined), "\n")
cat("  SNPs:  ", nrow(all_snps_eqtl_FLU), "\n")
cat("  CNVs:  ", nrow(all_CNV_eqtl_FLU), "\n")
cat("  InDels:", nrow(all_indel_eqtl_FLU), "\n")
cat("  SVs:   ", nrow(all_SV_eqtl_FLU), "\n\n")

cat("YPD.1 significant eQTLs (all variants combined):", nrow(ypd1_signif_combined), "\n\n")

cat("SC2Glu significant eQTLs (all variants combined):", nrow(sc2glu_signif_combined), "\n\n")
```

# STEP 4: Generalized Retrieval Function for All Comparisons

```{r retrieval-function}
# retrieve_concordance_data_combined <- function(signif_data, cond1_name, cond2_name, 
#                                                 all_dirs_cond1, all_dirs_cond2) {
#   
#   cat(sprintf("Retrieving: %s (significant) → %s (lookup)\n", cond1_name, cond2_name))
#   
#   cond1_genes <- unique(signif_data$Source_Phenotype)
#   cat(sprintf("  Genes to process: %d\n", length(cond1_genes)))
#   
#   cond2_lookup <- lapply(seq_along(cond1_genes), function(i) {
#     gene <- cond1_genes[i]
#     if (i %% 50 == 0) cat(sprintf("    [%d/%d]\n", i, length(cond1_genes)))
#     
#     gene_variants <- signif_data %>% filter(Source_Phenotype == gene)
#     if (nrow(gene_variants) == 0) return(NULL)
#     
#     var_types <- unique(gene_variants$Variant_Type)
#     
#     results_for_gene <- lapply(var_types, function(vtype) {
#       dir_cond2 <- all_dirs_cond2[[vtype]]
#       path_cond2 <- file.path(dir_cond2, gene, paste0(gene, ".first_assoc.txt"))
#       
#       if (!file.exists(path_cond2)) return(NULL)
#       
#       cond1_vars <- gene_variants %>% 
#         filter(Variant_Type == vtype) %>% 
#         select(SNP, Beta_Cond1 = SnpWeight, P_Cond1 = PValue, Variant_Type)
#       
#       if (nrow(cond1_vars) == 0) return(NULL)
#       
#       cond2 <- tryCatch({
#         read.table(path_cond2, header = TRUE, sep = "\t", 
#                    stringsAsFactors = FALSE, quote = "") %>%
#           select(SNP, Beta_Cond2 = SnpWeight, P_Cond2 = PValue)
#       }, error = function(e) NULL)
#       
#       if (is.null(cond2)) return(NULL)
#       
#       inner_join(cond1_vars, cond2, by = "SNP")
#       
#     }) %>% bind_rows()
#     
#     if (nrow(results_for_gene) == 0) return(NULL)
#     results_for_gene %>% mutate(Gene = gene)
#     
#   }) %>% bind_rows()
#   
#   final_data <- cond2_lookup %>%
#     mutate(Condition1 = cond1_name, Condition2 = cond2_name)
#   
#   cat(sprintf("  Retrieved: %d variant×gene pairs\n", nrow(final_data)))
#   cat(sprintf("  Breakdown by variant type:\n"))
#   for (vtype in c("SNPs", "CNVs", "InDels", "SVs")) {
#     n_type <- nrow(filter(final_data, Variant_Type == vtype))
#     if (n_type > 0) cat(sprintf("    %s: %d\n", vtype, n_type))
#   }
#   cat("\n")
#   
#   return(final_data)
# }
# 
# cat("✓ Retrieval function defined\n\n")
```

# STEP 5: Data Retrieval

```{r retrieve-concordance}
# cat("╔════════════════════════════════════════════╗\n")
# cat("║  RETRIEVING CONCORDANCE DATA               ║\n")
# cat("╚════════════════════════════════════════════╝\n\n")
# 
# conc_flu_ypd1 <- retrieve_concordance_data_combined(
#   flu_signif_combined, "Fluconazole", "YPD.1", DIRS_flu, DIRS_YPD1
# )
# 
# conc_flu_sc2glu <- retrieve_concordance_data_combined(
#   flu_signif_combined, "Fluconazole", "sc2glu", DIRS_flu, DIRS_sc2glu
# )
# 
# cat("✓ Concordance data retrieved\n\n")
```

# STEP 6: Save to RDS and Reload

```{r save-and-reload-rds}
# saveRDS(conc_flu_ypd1, file.path(OUTPUT_DIR, "conc_flu_ypd1.rds"))
# saveRDS(conc_flu_sc2glu, file.path(OUTPUT_DIR, "conc_flu_sc2glu.rds"))
# 
# cat("✓ Saved concordance data to RDS files\n\n")

# Reload from RDS
conc_flu_ypd1 <- readRDS(file.path(OUTPUT_DIR, "conc_flu_ypd1.rds"))
conc_flu_sc2glu <- readRDS(file.path(OUTPUT_DIR, "conc_flu_sc2glu.rds"))

cat("✓ Reloaded from RDS\n")
cat("  conc_flu_ypd1:   ", nrow(conc_flu_ypd1), "rows\n")
cat("  conc_flu_sc2glu: ", nrow(conc_flu_sc2glu), "rows\n\n")
```

# STEP 7: Concordance Statistics Calculation

```{r concordance-stats-detailed}
calculate_concordance_stats <- function(concordance_data, flu_total, comparison_name, 
                                        cond1_name = "Fluconazole", cond2_name = "Condition2") {
  
  cat(sprintf("\n%s (all variants)\n", comparison_name))
  cat(strrep("=", 70), "\n\n")
  
  n_retrieved <- nrow(concordance_data)
  n_not_found <- flu_total - n_retrieved
  
  directions <- sign(concordance_data$Beta_Cond1) == sign(concordance_data$Beta_Cond2)
  n_concordant <- sum(directions, na.rm = TRUE)
  n_discordant <- n_retrieved - n_concordant
  obs_sign_conc <- n_concordant / n_retrieved
  
  # Observed effect size correlations
  obs_r <- cor(concordance_data$Beta_Cond1, concordance_data$Beta_Cond2, method = "pearson")
  obs_rho <- cor(concordance_data$Beta_Cond1, concordance_data$Beta_Cond2, method = "spearman")
  
  # Binomial test
  binom_result <- binom.test(n_concordant, n_retrieved, p = 0.5, 
                              alternative = "greater")
  
  # Permutation tests
  cat("Running permutation tests (n=5000)...\n")
  n_perms <- 5000
  
  perm_sign <- replicate(n_perms, {
    perm_indices <- sample(1:n_retrieved, n_retrieved, replace = TRUE)
    perm_beta2 <- concordance_data$Beta_Cond2[perm_indices]
    sum(sign(concordance_data$Beta_Cond1) == sign(perm_beta2), na.rm = TRUE) / n_retrieved
  })
  
  perm_r <- replicate(n_perms, {
    perm_indices <- sample(1:n_retrieved, n_retrieved, replace = TRUE)
    perm_beta2 <- concordance_data$Beta_Cond2[perm_indices]
    cor(concordance_data$Beta_Cond1, perm_beta2, method = "pearson")
  })
  
  perm_p_sign <- sum(perm_sign >= obs_sign_conc) / n_perms
  perm_p_r <- sum(perm_r >= obs_r) / n_perms
  
  # Print results
  cat(sprintf("\n╔════════════════════════════════════════════╗\n"))
  cat(sprintf("║  eQTL SHARING BREAKDOWN                    ║\n"))
  cat(sprintf("╚════════════════════════════════════════════╝\n\n"))
  
  cat(sprintf("Total %s significant eQTLs: %d\n", cond1_name, flu_total))
  cat(sprintf("  NOT found in %s: %d (%.1f%%)\n", cond2_name, n_not_found, 
              100 * n_not_found / flu_total))
  cat(sprintf("  FOUND in %s: %d (%.1f%%)\n", cond2_name, n_retrieved, 
              100 * n_retrieved / flu_total))
  cat(sprintf("    ├─ Concordant (same direction): %d (%.1f%% of found)\n", 
              n_concordant, 100 * obs_sign_conc))
  cat(sprintf("    └─ Discordant (opposite direction): %d (%.1f%% of found)\n\n", 
              n_discordant, 100 * (1 - obs_sign_conc)))
  
  cat(sprintf("\nEffect size correlations (n=%d):\n", n_retrieved))
  cat(sprintf("  Pearson r:   %.4f\n", obs_r))
  cat(sprintf("  Spearman ρ:  %.4f\n\n", obs_rho))
  
  cat("Concordance by variant type:\n")
  for (vtype in c("SNPs", "CNVs", "InDels", "SVs")) {
    subset <- concordance_data %>% filter(Variant_Type == vtype)
    if (nrow(subset) > 0) {
      n_conc_vt <- sum(sign(subset$Beta_Cond1) == sign(subset$Beta_Cond2), na.rm = TRUE)
      pct_conc <- 100 * n_conc_vt / nrow(subset)
      cat(sprintf("  %s (n=%d): %.1f%% concordant\n", vtype, nrow(subset), pct_conc))
    }
  }
  
  cat("\nStatistical tests:\n")
  cat(sprintf("  Binomial test (H0: concordance = 0.50)\n"))
  cat(sprintf("    p-value: %.3e\n", binom_result$p.value))
  cat(sprintf("  Permutation test (n = %d)\n", n_perms))
  cat(sprintf("    Sign concordance p-value: %.4f\n", perm_p_sign))
  cat(sprintf("    Pearson r p-value:        %.4f\n\n", perm_p_r))
  
  cat("✓ Statistics calculated\n\n")
  
  return(list(
    data = concordance_data,
    flu_total = flu_total,
    n_not_found = n_not_found,
    n_retrieved = n_retrieved,
    n_concordant = n_concordant,
    n_discordant = n_discordant,
    obs_sign_conc = obs_sign_conc,
    obs_r = obs_r,
    obs_rho = obs_rho,
    perm_sign = perm_sign,
    perm_r = perm_r,
    perm_p_sign = perm_p_sign,
    perm_p_r = perm_p_r,
    binom_result = binom_result,
    n_total = n_retrieved
  ))
}

conc_flu_ypd1_stats <- calculate_concordance_stats(
  conc_flu_ypd1, 
  flu_total = nrow(flu_signif_combined),
  comparison_name = "Fluconazole vs YPD.1",
  cond1_name = "Fluconazole",
  cond2_name = "YPD.1"
)

conc_flu_sc2glu_stats <- calculate_concordance_stats(
  conc_flu_sc2glu,
  flu_total = nrow(flu_signif_combined),
  comparison_name = "Fluconazole vs SC2Glu",
  cond1_name = "Fluconazole",
  cond2_name = "SC2Glu"
)
```

# STEP 8: Effect Size Plots

```{r effect-size-plots}
make_concordance_plot_by_variant <- function(conc_data, stats, comparison_label) {
  
  variant_colors <- c(
    "SNPs" = "#3498DB",
    "CNVs" = "#E67E22",
    "InDels" = "#9B59B6",
    "SVs" = "#E74C3C"
  )
  
  plot_data <- conc_data %>%
    mutate(
      direction = ifelse(sign(Beta_Cond1) == sign(Beta_Cond2), 
                        "Concordant", "Discordant")
    )
  
  # Get data range for positioning labels
  x_range <- range(plot_data$Beta_Cond1, na.rm = TRUE)
  x_width <- x_range[2] - x_range[1]
  
  # Calculate label positions on regression lines at different x-positions
  label_positions <- data.frame(
    Variant_Type = c("SNPs", "CNVs", "InDels", "SVs"),
    x_pos = c(
      x_range[1] + 0.15 * x_width,
      x_range[1] + 0.35 * x_width,
      x_range[1] + 0.55 * x_width,
      x_range[1] + 0.75 * x_width
    ),
    y_pos = numeric(4),
    stringsAsFactors = FALSE
  )
  
  # Fit regression for each variant type
  for(i in 1:nrow(label_positions)) {
    vtype <- label_positions$Variant_Type[i]
    subset <- plot_data %>% filter(Variant_Type == vtype)
    
    if(nrow(subset) > 1) {
      fit <- lm(Beta_Cond2 ~ Beta_Cond1, data = subset)
      y_val <- predict(fit, newdata = data.frame(Beta_Cond1 = label_positions$x_pos[i]))
      label_positions$y_pos[i] <- y_val
    }
  }
  
  # Create the plot
  p <- ggplot(plot_data, aes(x = Beta_Cond1, y = Beta_Cond2)) +
    
    # Points colored by direction
    geom_point(aes(color = direction), size = 2, alpha = 0.3, show.legend = TRUE) +
    
    # Regression lines by variant type
    geom_smooth(data = plot_data %>% filter(Variant_Type == "SNPs"),
                method = "lm", se = FALSE, color = variant_colors["SNPs"], 
                linewidth = 1.2, linetype = "solid", show.legend = FALSE) +
    geom_smooth(data = plot_data %>% filter(Variant_Type == "CNVs"),
                method = "lm", se = FALSE, color = variant_colors["CNVs"], 
                linewidth = 1.2, linetype = "solid", show.legend = FALSE) +
    geom_smooth(data = plot_data %>% filter(Variant_Type == "InDels"),
                method = "lm", se = FALSE, color = variant_colors["InDels"], 
                linewidth = 1.2, linetype = "solid", show.legend = FALSE) +
    geom_smooth(data = plot_data %>% filter(Variant_Type == "SVs"),
                method = "lm", se = FALSE, color = variant_colors["SVs"], 
                linewidth = 1.2, linetype = "solid", show.legend = FALSE) +
    
    # 1:1 reference line
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", 
                color = "grey50", linewidth = 0.6, alpha = 0.7) +
    
    # Reference lines at 0
    geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3, alpha = 0.5) +
    geom_vline(xintercept = 0, color = "grey70", linewidth = 0.3, alpha = 0.5) +
    
    # Manual color scale for direction
    scale_color_manual(
      values = c("Concordant" = "#27AE60", "Discordant" = "#E74C3C"),
      name = "Direction"
    ) +
    
    # Add labels directly on regression lines - colored by variant type
    geom_text(data = label_positions,
              aes(x = x_pos, y = y_pos, label = Variant_Type, color = Variant_Type),
              size = 4.2, fontface = "bold", show.legend = FALSE,
              bbox = list(boxcolour = "white", alpha = 0.7, pad = unit(1, "mm"))) +
    
    # Unified color scale
    scale_color_manual(
      values = c("Concordant" = "#27AE60", "Discordant" = "#E74C3C",
                 "SNPs" = variant_colors["SNPs"], "CNVs" = variant_colors["CNVs"],
                 "InDels" = variant_colors["InDels"], "SVs" = variant_colors["SVs"]),
      name = "Direction"
    ) +
    
    # Statistics annotation
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.2,
             label = sprintf("Pearson r = %.3f\nSpearman ρ = %.3f\nSign concordance = %.1f%%\nn = %d",
                             stats$obs_r, stats$obs_rho, 100 * stats$obs_sign_conc, stats$n_total),
             size = 3.6, fontface = "bold",
             bbox = list(boxcolour = "white", alpha = 0.85, size = 0.5)) +
    
    labs(title = comparison_label,
         x = "Effect size β (Condition 1)",
         y = "Effect size β (Condition 2)",
         subtitle = "Points: colored by concordance | Lines & labels: colored by variant type | Dashed: 1:1 reference") +
    
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
      plot.subtitle = element_text(size = 9, hjust = 0.5, color = "grey60"),
      axis.title = element_text(face = "bold"),
      legend.position = "right",
      panel.grid.major = element_line(color = "grey95", linewidth = 0.2)
    )
  
  return(p)
}

p_flu_ypd1 <- make_concordance_plot_by_variant(
  conc_flu_ypd1, conc_flu_ypd1_stats, "Fluconazole vs YPD.1"
)

p_flu_sc2glu <- make_concordance_plot_by_variant(
  conc_flu_sc2glu, conc_flu_sc2glu_stats, "Fluconazole vs SC2Glu"
)

combined_scatter <- gridExtra::grid.arrange(p_flu_ypd1, p_flu_sc2glu, nrow = 1)
print(combined_scatter)

ggsave(file.path(OUTPUT_DIR, "Fig_effect_size_concordance.tiff"),
       combined_scatter, width = 14, height = 6, dpi = 600, compression = "lzw")

cat("╔════════════════════════════════════════════╗\n")
cat("║  FIGURE LEGEND - Concordance Plot         ║\n")
cat("╚════════════════════════════════════════════╝\n\n")
cat("Points (colored by concordance direction):\n")
cat("  • Green points  = Concordant (same direction)\n")
cat("  • Red points    = Discordant (opposite direction)\n\n")
cat("Regression lines & labels (colored by variant type):\n")
cat("  • Blue line & SNPs label      = SNPs\n")
cat("  • Orange line & CNVs label    = CNVs\n")
cat("  • Purple line & InDels label  = InDels\n")
cat("  • Red line & SVs label        = SVs\n\n")
cat("Reference lines:\n")
cat("  • Dashed grey line = Perfect 1:1 correlation\n\n")
cat("✓ Saved: Fig_effect_size_concordance.tiff\n\n")
```

# STEP 9: Null Distribution Plots

```{r null-distribution-plots}
make_null_plot_combined <- function(stats, comparison_name) {
  
  perm_df <- data.frame(
    sign_conc = stats$perm_sign,
    pearson_r = stats$perm_r
  )
  
  # Panel A: Sign concordance
  p_null_sign <- ggplot(perm_df, aes(x = sign_conc)) +
    geom_histogram(bins = 50, fill = "grey70", color = "white", linewidth = 0.2) +
    geom_vline(xintercept = stats$obs_sign_conc, color = "#E74C3C", linewidth = 1.2) +
    geom_vline(xintercept = 0.5, color = "black", linewidth = 0.5, linetype = "dashed") +
    annotate("text", x = Inf, y = Inf,
             label = sprintf("Observed = %.3f\np-value = %.2e",
                             stats$obs_sign_conc, 
                             stats$binom_result$p.value),
             hjust = 1.05, vjust = 1.4, size = 3.5, fontface = "bold",
             bbox = list(boxcolour = "white", alpha = 0.8, size = 0.5)) +
    labs(title = "Sign Concordance Distribution",
         x = "Sign concordance (permutation)", y = "Frequency") +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(face = "bold", size = 12))
  
  # Panel B: Pearson r
  p_null_r <- ggplot(perm_df, aes(x = pearson_r)) +
    geom_histogram(bins = 50, fill = "grey70", color = "white", linewidth = 0.2) +
    geom_vline(xintercept = stats$obs_r, color = "#27AE60", linewidth = 1.2) +
    geom_vline(xintercept = 0, color = "black", linewidth = 0.5, linetype = "dashed") +
    annotate("text", x = Inf, y = Inf,
             label = sprintf("Observed = %.3f\np-value = %.4f",
                             stats$obs_r, 
                             stats$perm_p_r),
             hjust = 1.05, vjust = 1.4, size = 3.5, fontface = "bold",
             bbox = list(boxcolour = "white", alpha = 0.8, size = 0.5)) +
    labs(title = "Pearson r Distribution",
         x = "Pearson r (permutation)", y = "Frequency") +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(face = "bold", size = 12))
  
  gridExtra::grid.arrange(p_null_sign, p_null_r, nrow = 1)
}

p_null_flu_ypd1 <- make_null_plot_combined(conc_flu_ypd1_stats, "Fluconazole vs YPD.1")
p_null_flu_sc2glu <- make_null_plot_combined(conc_flu_sc2glu_stats, "Fluconazole vs SC2Glu")

print(p_null_flu_ypd1)
ggsave(file.path(OUTPUT_DIR, "Fig_null_distribution_Flu_vs_YPD1.tiff"),
       p_null_flu_ypd1, width = 12, height = 5, dpi = 600, compression = "lzw")

print(p_null_flu_sc2glu)
ggsave(file.path(OUTPUT_DIR, "Fig_null_distribution_Flu_vs_sc2glu.tiff"),
       p_null_flu_sc2glu, width = 12, height = 5, dpi = 600, compression = "lzw")

cat("✓ Saved null distribution plots\n\n")
```

---

# Analysis: Categorization, Concordance, and GO Enrichment

```{r categorize-eqtls}
categorize_eqtls <- function(signif_data_cond1, concordance_data, cond1_name, cond2_name) {
  
  cat(sprintf("\n%s vs %s\n", cond1_name, cond2_name))
  cat(strrep("=", 70), "\n\n")
  
  all_sig_cond1 <- signif_data_cond1 %>%
    select(SNP, Gene = Source_Phenotype, Beta_Cond1 = SnpWeight, 
           Variant_Type, P_Cond1 = PValue) %>%
    mutate(pair_id = paste0(SNP, "_", Gene))
  
  found_in_cond2 <- concordance_data %>%
    mutate(pair_id = paste0(SNP, "_", Gene)) %>%
    select(pair_id, Beta_Cond2, P_Cond2)
  
  categorized <- all_sig_cond1 %>%
    left_join(found_in_cond2, by = "pair_id") %>%
    mutate(
      Found_in_both = !is.na(Beta_Cond2),
      Direction_Match = ifelse(Found_in_both, sign(Beta_Cond1) == sign(Beta_Cond2), NA),
      Category = case_when(
        !Found_in_both ~ sprintf("%s-only", cond1_name),
        Found_in_both & Direction_Match ~ "Shared-Concordant",
        Found_in_both & !Direction_Match ~ "Shared-Discordant"
      )
    )
  
  cat(sprintf("TOTAL %s eQTLs: %d\n\n", cond1_name, nrow(all_sig_cond1)))
  cat(sprintf("%s-ONLY: %d (%.1f%%)\n", cond1_name, sum(!categorized$Found_in_both), 
              100 * sum(!categorized$Found_in_both) / nrow(all_sig_cond1)))
  cat(sprintf("Shared-Concordant: %d (%.1f%% of shared)\n",
              sum(categorized$Category == "Shared-Concordant", na.rm = TRUE),
              100 * sum(categorized$Category == "Shared-Concordant", na.rm = TRUE) / sum(categorized$Found_in_both)))
  cat(sprintf("Shared-Discordant: %d (%.1f%% of shared)\n\n",
              sum(categorized$Category == "Shared-Discordant", na.rm = TRUE),
              100 * sum(categorized$Category == "Shared-Discordant", na.rm = TRUE) / sum(categorized$Found_in_both)))
  
  return(categorized)
}

eqtl_categories_flu_ypd <- categorize_eqtls(flu_signif_combined, conc_flu_ypd1, 
                                             "Fluconazole", "YPD.1")
eqtl_categories_flu_sc2 <- categorize_eqtls(flu_signif_combined, conc_flu_sc2glu,
                                             "Fluconazole", "SC2Glu")
```

```{r extract-genes}
extract_category_genes <- function(categorized_data, cond1_name, cond2_name) {
  
  genes_cond1_only <- categorized_data %>%
    filter(Category == sprintf("%s-only", cond1_name)) %>%
    pull(Gene) %>% unique()
  
  genes_shared_concordant <- categorized_data %>%
    filter(Category == "Shared-Concordant") %>%
    pull(Gene) %>% unique()
  
  genes_shared_discordant <- categorized_data %>%
    filter(Category == "Shared-Discordant") %>%
    pull(Gene) %>% unique()
  
  genes_cond2_only <- setdiff(
    categorized_data %>% filter(!is.na(Beta_Cond2)) %>% pull(Gene) %>% unique(),
    categorized_data$Gene %>% unique()
  )
  
  cat(sprintf("%s-specific: %d | %s-specific: %d | Concordant: %d | Discordant: %d\n\n",
              cond1_name, length(genes_cond1_only), cond2_name, length(genes_cond2_only),
              length(genes_shared_concordant), length(genes_shared_discordant)))
  
  return(list(cond1_only = genes_cond1_only, cond2_only = genes_cond2_only,
              shared_concordant = genes_shared_concordant, shared_discordant = genes_shared_discordant))
}

gene_sets_flu_ypd <- extract_category_genes(eqtl_categories_flu_ypd, "Fluconazole", "YPD.1")
gene_sets_flu_sc2 <- extract_category_genes(eqtl_categories_flu_sc2, "Fluconazole", "SC2Glu")
```

```{r plot-sharing}
plot_eqtl_sharing <- function(categorized_data, title) {
  
  summary_df <- categorized_data %>%
    group_by(Category) %>%
    summarise(N = n(), .groups = "drop") %>%
    mutate(Pct = round(100 * N / sum(N), 1),
           Category = factor(Category, levels = c("Fluconazole-only", "Shared-Concordant", 
                                                   "Shared-Discordant")))
  
  colors <- c("Fluconazole-only" = "#E74C3C", "Shared-Concordant" = "#27AE60", 
              "Shared-Discordant" = "#F39C12")
  
  ggplot(summary_df, aes(x = Category, y = N, fill = Category)) +
    geom_col(width = 0.6, alpha = 0.8) +
    geom_text(aes(label = sprintf("n=%d\n%.1f%%", N, Pct)), vjust = -0.3, size = 4, fontface = "bold") +
    scale_fill_manual(values = colors) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(title = title, y = "Number of eQTLs", x = NULL) +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          axis.text.x = element_text(angle = 30, hjust = 1, face = "bold"),
          legend.position = "none")
}

p1 <- plot_eqtl_sharing(eqtl_categories_flu_ypd, "Fluconazole vs YPD.1")
p2 <- plot_eqtl_sharing(eqtl_categories_flu_sc2, "Fluconazole vs SC2Glu")

combined_sharing <- gridExtra::grid.arrange(p1, p2, nrow = 1)
print(combined_sharing)

ggsave(file.path(OUTPUT_DIR, "Fig_eQTL_sharing.tiff"), combined_sharing, width = 12, height = 5, dpi = 600)
cat("✓ Saved: Fig_eQTL_sharing.tiff\n\n")
```

```{r summary-table}
cat("\n╔════════════════════════════════════════════════════════════════╗\n")
cat("║  VERIFYING STATISTICS OBJECTS                                 ║\n")
cat("╚════════════════════════════════════════════════════════════════╝\n\n")

cat("conc_flu_ypd1_stats fields:\n")
print(names(conc_flu_ypd1_stats))
cat("\nconc_flu_sc2glu_stats fields:\n")
print(names(conc_flu_sc2glu_stats))
cat("\n")

summary_table_detailed <- data.frame(
  Comparison = c("Flu vs YPD.1", "Flu vs SC2Glu"),
  Total_Flu_eQTLs = c(
    conc_flu_ypd1_stats$flu_total,
    conc_flu_sc2glu_stats$flu_total
  ),
  Flu_Only = c(
    conc_flu_ypd1_stats$n_not_found,
    conc_flu_sc2glu_stats$n_not_found
  ),
  Flu_Only_Pct = c(
    sprintf("%.1f%%", 100 * conc_flu_ypd1_stats$n_not_found / conc_flu_ypd1_stats$flu_total),
    sprintf("%.1f%%", 100 * conc_flu_sc2glu_stats$n_not_found / conc_flu_sc2glu_stats$flu_total)
  ),
  Found_in_Other = c(
    conc_flu_ypd1_stats$n_retrieved,
    conc_flu_sc2glu_stats$n_retrieved
  ),
  Found_Pct = c(
    sprintf("%.1f%%", 100 * conc_flu_ypd1_stats$n_retrieved / conc_flu_ypd1_stats$flu_total),
    sprintf("%.1f%%", 100 * conc_flu_sc2glu_stats$n_retrieved / conc_flu_sc2glu_stats$flu_total)
  ),
  Concordant = c(
    conc_flu_ypd1_stats$n_concordant,
    conc_flu_sc2glu_stats$n_concordant
  ),
  Concordant_Pct = c(
    sprintf("%.1f%%", 100 * conc_flu_ypd1_stats$obs_sign_conc),
    sprintf("%.1f%%", 100 * conc_flu_sc2glu_stats$obs_sign_conc)
  ),
  Discordant = c(
    conc_flu_ypd1_stats$n_discordant,
    conc_flu_sc2glu_stats$n_discordant
  ),
  Discordant_Pct = c(
    sprintf("%.1f%%", 100 * (1 - conc_flu_ypd1_stats$obs_sign_conc)),
    sprintf("%.1f%%", 100 * (1 - conc_flu_sc2glu_stats$obs_sign_conc))
  ),
  Pearson_r = c(
    sprintf("%.4f", conc_flu_ypd1_stats$obs_r),
    sprintf("%.4f", conc_flu_sc2glu_stats$obs_r)
  ),
  Spearman_rho = c(
    sprintf("%.4f", conc_flu_ypd1_stats$obs_rho),
    sprintf("%.4f", conc_flu_sc2glu_stats$obs_rho)
  ),
  stringsAsFactors = FALSE
)

cat("╔════════════════════════════════════════════════════════════════╗\n")
cat("║  COMPREHENSIVE ANALYSIS SUMMARY                              ║\n")
cat("╚════════════════════════════════════════════════════════════════╝\n\n")

cat("eQTL SHARING AND CONCORDANCE SUMMARY\n")
cat(strrep("=", 140), "\n\n")
print(summary_table_detailed, row.names = FALSE, digits = 4)

cat("\n\nINTERPRETATION:\n")
cat("────────────────────────────────────────────────────────────────\n")
cat("• Total_Flu_eQTLs: Total significant eQTLs in Fluconazole\n")
cat("• Flu_Only: Fluconazole eQTLs NOT found in the other condition\n")
cat("• Found_in_Other: Fluconazole eQTLs that ARE detected in the other condition\n")
cat("• Concordant: Of found eQTLs, those with same direction of effect\n")
cat("• Discordant: Of found eQTLs, those with opposite direction of effect\n")
cat("• Pearson r / Spearman ρ: Correlation of effect sizes (only for found eQTLs)\n\n")

cat("\nQUICK REFERENCE:\n")
cat("────────────────────────────────────────────────────────────────\n\n")

cat("Fluconazole vs YPD.1:\n")
cat(sprintf("  Total Flu eQTLs: %d\n", conc_flu_ypd1_stats$flu_total))
cat(sprintf("  Flu-only (not found in YPD.1): %d (%.1f%%)\n", 
            conc_flu_ypd1_stats$n_not_found,
            100 * conc_flu_ypd1_stats$n_not_found / conc_flu_ypd1_stats$flu_total))
cat(sprintf("  Found in YPD.1: %d (%.1f%%)\n", 
            conc_flu_ypd1_stats$n_retrieved,
            100 * conc_flu_ypd1_stats$n_retrieved / conc_flu_ypd1_stats$flu_total))
cat(sprintf("    ├─ Concordant (same direction): %d (%.1f%% of found)\n", 
            conc_flu_ypd1_stats$n_concordant,
            100 * conc_flu_ypd1_stats$obs_sign_conc))
cat(sprintf("    └─ Discordant (opposite direction): %d (%.1f%% of found)\n", 
            conc_flu_ypd1_stats$n_discordant,
            100 * (1 - conc_flu_ypd1_stats$obs_sign_conc)))
cat(sprintf("  Effect size correlation: r = %.4f, ρ = %.4f\n\n", 
            conc_flu_ypd1_stats$obs_r, conc_flu_ypd1_stats$obs_rho))

cat("Fluconazole vs SC2Glu:\n")
cat(sprintf("  Total Flu eQTLs: %d\n", conc_flu_sc2glu_stats$flu_total))
cat(sprintf("  Flu-only (not found in SC2Glu): %d (%.1f%%)\n", 
            conc_flu_sc2glu_stats$n_not_found,
            100 * conc_flu_sc2glu_stats$n_not_found / conc_flu_sc2glu_stats$flu_total))
cat(sprintf("  Found in SC2Glu: %d (%.1f%%)\n", 
            conc_flu_sc2glu_stats$n_retrieved,
            100 * conc_flu_sc2glu_stats$n_retrieved / conc_flu_sc2glu_stats$flu_total))
cat(sprintf("    ├─ Concordant (same direction): %d (%.1f%% of found)\n", 
            conc_flu_sc2glu_stats$n_concordant,
            100 * conc_flu_sc2glu_stats$obs_sign_conc))
cat(sprintf("    └─ Discordant (opposite direction): %d (%.1f%% of found)\n", 
            conc_flu_sc2glu_stats$n_discordant,
            100 * (1 - conc_flu_sc2glu_stats$obs_sign_conc)))
cat(sprintf("  Effect size correlation: r = %.4f, ρ = %.4f\n\n", 
            conc_flu_sc2glu_stats$obs_r, conc_flu_sc2glu_stats$obs_rho))

cat("\n✓ Analysis complete\n")
```

```{r variant-breakdown}
# ============================================================================
# Variant type breakdown
# ============================================================================

variant_breakdown <- rbind(
  conc_flu_ypd1_stats$data %>%
    group_by(Variant_Type) %>%
    summarise(
      Comparison = "Flu vs YPD.1",
      N_eQTLs = n(),
      Concordant = sum(sign(Beta_Cond1) == sign(Beta_Cond2)),
      Concordant_Pct = round((sum(sign(Beta_Cond1) == sign(Beta_Cond2)) / n()) * 100, 2),
      Pearson_r = round(cor(Beta_Cond1, Beta_Cond2, method = "pearson"), 4),
      Spearman_rho = round(cor(Beta_Cond1, Beta_Cond2, method = "spearman"), 4),
      .groups = "drop"
    ),
  conc_flu_sc2glu_stats$data %>%
    group_by(Variant_Type) %>%
    summarise(
      Comparison = "Flu vs sc2glu",
      N_eQTLs = n(),
      Concordant = sum(sign(Beta_Cond1) == sign(Beta_Cond2)),
      Concordant_Pct = round((sum(sign(Beta_Cond1) == sign(Beta_Cond2)) / n()) * 100, 2),
      Pearson_r = round(cor(Beta_Cond1, Beta_Cond2, method = "pearson"), 4),
      Spearman_rho = round(cor(Beta_Cond1, Beta_Cond2, method = "spearman"), 4),
      .groups = "drop"
    )
)

cat("\n")
cat("════════════════════════════════════════════════════════════\n")
cat("VARIANT TYPE BREAKDOWN\n")
cat("════════════════════════════════════════════════════════════\n\n")
print(variant_breakdown)
```

```{r identify-condition-specific}
# ============================================================================
# Identify YPD.1-only and SC2Glu-only eQTLs (simplified)
# ============================================================================

cat("╔════════════════════════════════════════════════════════════════╗\n")
cat("║  IDENTIFYING CONDITION-SPECIFIC eQTLs                         ║\n")
cat("╚════════════════════════════════════════════════════════════════╝\n\n")

# YPD.1-specific eQTLs
cat("YPD.1-specific eQTLs:\n")
cat("─────────────────────────────────────────\n\n")

ypd1_genes <- unique(ypd1_signif_combined$Source_Phenotype)
flu_genes <- unique(flu_signif_combined$Source_Phenotype)

ypd1_only_genes <- setdiff(ypd1_genes, flu_genes)
ypd1_shared_genes <- intersect(ypd1_genes, flu_genes)

cat(sprintf("Total YPD.1 significant eQTLs: %d\n", nrow(ypd1_signif_combined)))
cat(sprintf("  Found in Fluconazole: %d (%.1f%%)\n", 
            nrow(ypd1_signif_combined %>% filter(Source_Phenotype %in% flu_genes)),
            100 * nrow(ypd1_signif_combined %>% filter(Source_Phenotype %in% flu_genes)) / nrow(ypd1_signif_combined)))
cat(sprintf("  YPD.1-only (not in Flu): %d (%.1f%%)\n", 
            nrow(ypd1_signif_combined %>% filter(Source_Phenotype %in% ypd1_only_genes)),
            100 * nrow(ypd1_signif_combined %>% filter(Source_Phenotype %in% ypd1_only_genes)) / nrow(ypd1_signif_combined)))
cat(sprintf("  Genes: %d YPD.1-only, %d shared\n\n", 
            length(ypd1_only_genes), length(ypd1_shared_genes)))

# SC2Glu-specific eQTLs
cat("SC2Glu-specific eQTLs:\n")
cat("─────────────────────────────────────────\n\n")

sc2glu_genes <- unique(sc2glu_signif_combined$Source_Phenotype)

sc2glu_only_genes <- setdiff(sc2glu_genes, flu_genes)
sc2glu_shared_genes <- intersect(sc2glu_genes, flu_genes)

cat(sprintf("Total SC2Glu significant eQTLs: %d\n", nrow(sc2glu_signif_combined)))
cat(sprintf("  Found in Fluconazole: %d (%.1f%%)\n", 
            nrow(sc2glu_signif_combined %>% filter(Source_Phenotype %in% flu_genes)),
            100 * nrow(sc2glu_signif_combined %>% filter(Source_Phenotype %in% flu_genes)) / nrow(sc2glu_signif_combined)))
cat(sprintf("  SC2Glu-only (not in Flu): %d (%.1f%%)\n", 
            nrow(sc2glu_signif_combined %>% filter(Source_Phenotype %in% sc2glu_only_genes)),
            100 * nrow(sc2glu_signif_combined %>% filter(Source_Phenotype %in% sc2glu_only_genes)) / nrow(sc2glu_signif_combined)))
cat(sprintf("  Genes: %d SC2Glu-only, %d shared\n\n", 
            length(sc2glu_only_genes), length(sc2glu_shared_genes)))

# Save gene lists
ypd1_only_eqtls <- ypd1_signif_combined %>% filter(Source_Phenotype %in% ypd1_only_genes)
sc2glu_only_eqtls <- sc2glu_signif_combined %>% filter(Source_Phenotype %in% sc2glu_only_genes)

write.csv(ypd1_only_eqtls, file.path(OUTPUT_DIR, "YPD1_only_eQTLs.csv"), row.names = FALSE)
write.csv(sc2glu_only_eqtls, file.path(OUTPUT_DIR, "SC2Glu_only_eQTLs.csv"), row.names = FALSE)

cat("✓ Condition-specific eQTL analysis complete\n")
cat(sprintf("✓ Saved: YPD1_only_eQTLs.csv (%d eQTLs)\n", nrow(ypd1_only_eqtls)))
cat(sprintf("✓ Saved: SC2Glu_only_eQTLs.csv (%d eQTLs)\n\n", nrow(sc2glu_only_eqtls)))
```

```{r go-prep}
# ============================================================================
# CHUNK 12A: PREPARE DATA FOR GO ENRICHMENT
# ============================================================================

prepare_concordant_genes <- function(conc_stats_data, comparison_name,
                                      min_effect_size_corr = 0.5) {
  
  cat("\n╔════════════════════════════════════════════╗\n")
  cat(sprintf("║  GO PREP: %s\n", comparison_name))
  cat("╚════════════════════════════════════════════╝\n\n")
  
  data <- conc_stats_data
  
  concordant_vars <- data %>%
    filter(
      sign(Beta_Cond1) == sign(Beta_Cond2),
      abs(Beta_Cond1) > 0 & abs(Beta_Cond2) > 0
    )
  
  gene_concordance <- concordant_vars %>%
    group_by(Gene) %>%
    summarise(
      N_variants = n(),
      Mean_abs_beta_cond1 = mean(abs(Beta_Cond1)),
      Mean_abs_beta_cond2 = mean(abs(Beta_Cond2)),
      Pearson_r = cor(Beta_Cond1, Beta_Cond2, method = "pearson"),
      N_concordant = sum(sign(Beta_Cond1) == sign(Beta_Cond2)),
      Pct_concordant = (N_concordant / n()) * 100,
      .groups = "drop"
    ) %>%
    mutate(
      Has_high_correlation = abs(Pearson_r) > min_effect_size_corr
    )
  
  genes_high_concordance <- gene_concordance %>%
    filter(Has_high_correlation) %>%
    pull(Gene) %>%
    unique()
  
  genes_low_concordance <- gene_concordance %>%
    filter(!Has_high_correlation) %>%
    pull(Gene) %>%
    unique()
  
  cat(sprintf("Total genes tested: %d\n", n_distinct(data$Gene)))
  cat(sprintf("Genes with high concordance (|r| > %.2f): %d\n", 
              min_effect_size_corr, length(genes_high_concordance)))
  cat(sprintf("Genes with low concordance: %d\n", length(genes_low_concordance)))
  cat(sprintf("\nVariant breakdown:\n"))
  cat(sprintf("  Total variants in data: %d\n", nrow(data)))
  cat(sprintf("  Concordant variants (same sign): %d\n", nrow(concordant_vars)))
  cat(sprintf("  Discordant variants: %d\n\n", nrow(data) - nrow(concordant_vars)))
  
  return(list(
    comparison_name = comparison_name,
    genes_high_concordance = genes_high_concordance,
    genes_low_concordance = genes_low_concordance,
    gene_concordance_stats = gene_concordance,
    all_concordant_data = concordant_vars
  ))
}

go_prep_flu_ypd1 <- prepare_concordant_genes(
  conc_flu_ypd1_stats$data, 
  "Fluconazole vs YPD.1",
  min_effect_size_corr = 0.5
)

go_prep_flu_sc2glu <- prepare_concordant_genes(
  conc_flu_sc2glu_stats$data, 
  "Fluconazole vs sc2glu",
  min_effect_size_corr = 0.5
)

cat("✓ GO preparation complete\n\n")
```

```{r go-enrichment}
# ============================================================================
# CHUNK 12B: GO ENRICHMENT - Three-way comparison
# ============================================================================

library(clusterProfiler)
library(org.Sc.sgd.db)

# Define tested universe
tested_universe <- read.csv(
  "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"
) %>%
  filter(gene_biotype == "protein_coding") %>% 
  pull(gene_id) %>% 
  unique()

cat(sprintf("Tested universe: %d protein-coding genes\n\n", length(tested_universe)))

# Prepare gene categories
prepare_three_way_categories <- function(concordance_data, cond1_name, cond2_name,
                                         all_signif_cond1, all_signif_cond2,
                                         min_effect_size_corr = 0.5) {
  
  cat(sprintf("\n╔════════════════════════════════════════════╗\n"))
  cat(sprintf("║  Preparing categories: %s vs %s\n", cond1_name, cond2_name))
  cat(sprintf("╚════════════════════════════════════════════╝\n\n"))
  
  genes_cond1 <- unique(all_signif_cond1$Source_Phenotype)
  genes_cond2 <- unique(all_signif_cond2$Source_Phenotype)
  
  genes_shared <- intersect(genes_cond1, genes_cond2)
  
  gene_corr <- concordance_data %>%
    filter(Gene %in% genes_shared) %>%
    group_by(Gene) %>%
    summarise(
      Pearson_r = cor(Beta_Cond1, Beta_Cond2, method = "pearson"),
      .groups = "drop"
    )
  
  genes_high_conc_shared <- gene_corr %>%
    filter(abs(Pearson_r) > min_effect_size_corr) %>%
    pull(Gene) %>%
    unique()
  
  genes_cond1_only <- setdiff(genes_cond1, genes_cond2)
  genes_cond2_only <- setdiff(genes_cond2, genes_cond1)
  
  cat(sprintf("%s total genes: %d\n", cond1_name, length(genes_cond1)))
  cat(sprintf("%s total genes: %d\n\n", cond2_name, length(genes_cond2)))
  
  cat(sprintf("Shared genes (in both): %d\n", length(genes_shared)))
  cat(sprintf("  → High concordance (|r| > %.2f): %d\n", 
              min_effect_size_corr, length(genes_high_conc_shared)))
  cat(sprintf("  → Low concordance: %d\n\n", 
              length(genes_shared) - length(genes_high_conc_shared)))
  
  cat(sprintf("%s-only genes: %d\n", cond1_name, length(genes_cond1_only)))
  cat(sprintf("%s-only genes: %d\n\n", cond2_name, length(genes_cond2_only)))
  
  list(
    cond1_only = genes_cond1_only,
    high_conc_shared = genes_high_conc_shared,
    cond2_only = genes_cond2_only,
    all_shared = genes_shared,
    cond1_name = cond1_name,
    cond2_name = cond2_name
  )
}

cat("\n═══════════════════════════════════════════════════\n")
cat("PREPARING GENE CATEGORIES\n")
cat("═══════════════════════════════════════════════════\n")

go_categories_flu_ypd1 <- prepare_three_way_categories(
  conc_flu_ypd1,
  cond1_name = "Fluconazole",
  cond2_name = "YPD.1",
  all_signif_cond1 = flu_signif_combined,
  all_signif_cond2 = ypd1_signif_combined,
  min_effect_size_corr = 0.5
)

go_categories_flu_sc2glu <- prepare_three_way_categories(
  conc_flu_sc2glu,
  cond1_name = "Fluconazole",
  cond2_name = "sc2glu",
  all_signif_cond1 = flu_signif_combined,
  all_signif_cond2 = sc2glu_signif_combined,
  min_effect_size_corr = 0.5
)

cat("\n✓ Gene categories prepared\n\n")

# GO enrichment function
run_enrichGO <- function(gene_list, label, universe, ontology = "ALL") {
  
  g <- intersect(gene_list, universe)
  cat(sprintf("  %s | genes in universe: %d\n", label, length(g)))
  
  if (length(g) < 10) { 
    cat("    → too few genes (<10), skipping\n")
    return(NULL) 
  }
  
  res <- enrichGO(
    gene = g, 
    universe = universe, 
    OrgDb = org.Sc.sgd.db, 
    keyType = "ORF",
    ont = ontology, 
    pAdjustMethod = "BH", 
    pvalueCutoff = 0.05, 
    qvalueCutoff = 0.1
  )
  
  if (is.null(res) || nrow(res) == 0) { 
    cat("    → no significant terms\n")
    return(NULL) 
  }
  
  cat(sprintf("    → significant terms: %d\n", nrow(res)))
  return(res)
}

# Plot comparison function
plot_go_three_way <- function(gene_categories, tested_universe, title, outfile) {
  
  cat(sprintf("\n═══ %s ═══\n", title))
  
  # Build the three categories with SIMPLIFIED labels
  go_lists <- list(
    gene_categories$cond2_only,
    gene_categories$high_conc_shared,
    gene_categories$cond1_only
  )
  
  names(go_lists) <- c(
    sprintf("%s-only", gene_categories$cond2_name),
    "Shared\n(high concordance)",
    sprintf("%s-only", gene_categories$cond1_name)
  )
  
  cat("eGene category sizes:\n")
  print(sapply(go_lists, length))
  
  # Run enrichment for each category
  cat("\nRunning GO enrichment:\n")
  go_res <- lapply(names(go_lists), function(g) {
    run_enrichGO(go_lists[[g]], g, tested_universe, ont = "ALL")
  }) %>% setNames(names(go_lists))
  
  # Filter out NULL results
  go_res <- go_res[!sapply(go_res, is.null)]
  
  if (length(go_res) < 2) {
    cat("\n⚠️  Fewer than 2 categories with significant terms — plot skipped.\n\n")
    return(invisible(NULL))
  }
  
  cat(sprintf("\n✓ %d categories with significant GO terms\n", length(go_res)))
  
  SHOW_N <- 15
  
  # Merge results for comparison
  cg <- merge_result(go_res)
  cg@compareClusterResult$Cluster <- factor(
    cg@compareClusterResult$Cluster, 
    levels = names(go_lists)
  )
  
  # Wrap descriptions
  cg@compareClusterResult$Description <- str_wrap(
    cg@compareClusterResult$Description, width = 60
  )
  
  # Create dotplot
  p <- dotplot(cg, showCategory = SHOW_N) + 
    ggtitle(title) + 
    theme_bw(base_size = 12) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14, 
                                margin = margin(b = 10)),
      axis.text.y = element_text(size = 8, color = "black", lineheight = 0.85),
      axis.text.x = element_text(size = 10, face = "bold", color = "black", 
                                 angle = 30, hjust = 1),
      axis.title.x = element_text(size = 11, face = "bold"),
      axis.title.y = element_blank(),
      legend.title = element_text(size = 10, face = "bold"),
      legend.text = element_text(size = 9)
    )
  
  print(p)
  
  # Calculate height based on number of terms
  n_terms <- length(unique(p$data$Description))
  height <- max(7, 0.35 * n_terms + 2)
  
  ggsave(
    file.path(OUTPUT_DIR, outfile), p, 
    width = 12, height = height,
    units = "in", dpi = 300, device = "tiff", 
    compression = "lzw", limitsize = FALSE
  )
  
  cat(sprintf("✓ Saved: %s\n\n", outfile))
  invisible(p)
}

# ============================================================================
# GO: Fluconazole vs YPD.1 (Three-way)
# ============================================================================

cat("\n╔════════════════════════════════════════════╗\n")
cat("║  GO ENRICHMENT: Fluconazole vs YPD.1      ║\n")
cat("║  (YPD-only | High Concordance | Flu-only) ║\n")
cat("╚════════════════════════════════════════════╝\n")

p_go_flu_ypd1 <- plot_go_three_way(
  go_categories_flu_ypd1,
  tested_universe,
  "GO Enrichment: Fluconazole vs YPD.1",
  "GO_Flu_vs_YPD1_three_way.tiff"
)

# ============================================================================
# GO: Fluconazole vs sc2glu (Three-way)
# ============================================================================

cat("╔════════════════════════════════════════════╗\n")
cat("║  GO ENRICHMENT: Fluconazole vs sc2glu     ║\n")
cat("║ (SC2Glu-only | High Concordance | Flu-only)║\n")
cat("╚════════════════════════════════════════════╝\n")

p_go_flu_sc2glu <- plot_go_three_way(
  go_categories_flu_sc2glu,
  tested_universe,
  "GO Enrichment: Fluconazole vs sc2glu",
  "GO_Flu_vs_sc2glu_three_way.tiff"
)

cat("\n✓ GO enrichment analysis complete\n\n")

# ============================================================================
# Summary table of gene categories
# ============================================================================

go_summary <- data.frame(
  Comparison = c("Flu vs YPD.1", "Flu vs YPD.1", "Flu vs YPD.1",
                 "Flu vs sc2glu", "Flu vs sc2glu", "Flu vs sc2glu"),
  Category = c("YPD.1-only", "High Concordance", "Fluconazole-only",
               "sc2glu-only", "High Concordance", "Fluconazole-only"),
  N_Genes = c(
    length(go_categories_flu_ypd1$cond2_only),
    length(go_categories_flu_ypd1$high_conc_shared),
    length(go_categories_flu_ypd1$cond1_only),
    length(go_categories_flu_sc2glu$cond2_only),
    length(go_categories_flu_sc2glu$high_conc_shared),
    length(go_categories_flu_sc2glu$cond1_only)
  ),
  Interpretation = c(
    "Condition-specific", "Shared across conditions", "Condition-specific",
    "Condition-specific", "Shared across conditions", "Condition-specific"
  )
)

cat("════════════════════════════════════════════════════════════\n")
cat("GO ENRICHMENT GENE CATEGORIES SUMMARY\n")
cat("════════════════════════════════════════════════════════════\n\n")
print(go_summary)

write.csv(go_summary,
          file.path(OUTPUT_DIR, "GO_gene_categories_summary.csv"),
          row.names = FALSE)

cat("\n✓ Saved: GO_gene_categories_summary.csv\n")
```

```{r go-concordance-summary}
# ============================================================================
# CHUNK 12C: SUMMARY - Concordant genes by category
# ============================================================================

concordance_gene_summary <- rbind(
  data.frame(
    Comparison = "Flu vs YPD.1",
    Category = c("High Concordance", "Low Concordance"),
    N_Genes = c(
      length(go_prep_flu_ypd1$genes_high_concordance),
      length(go_prep_flu_ypd1$genes_low_concordance)
    ),
    Mean_Effect_Size_Corr = c(
      mean(filter(go_prep_flu_ypd1$gene_concordance_stats, 
                  Has_high_correlation)$Pearson_r, na.rm = TRUE),
      mean(filter(go_prep_flu_ypd1$gene_concordance_stats, 
                  !Has_high_correlation)$Pearson_r, na.rm = TRUE)
    )
  ),
  data.frame(
    Comparison = "Flu vs sc2glu",
    Category = c("High Concordance", "Low Concordance"),
    N_Genes = c(
      length(go_prep_flu_sc2glu$genes_high_concordance),
      length(go_prep_flu_sc2glu$genes_low_concordance)
    ),
    Mean_Effect_Size_Corr = c(
      mean(filter(go_prep_flu_sc2glu$gene_concordance_stats, 
                  Has_high_correlation)$Pearson_r, na.rm = TRUE),
      mean(filter(go_prep_flu_sc2glu$gene_concordance_stats, 
                  !Has_high_correlation)$Pearson_r, na.rm = TRUE)
    )
  )
)

cat("\n")
cat("════════════════════════════════════════════════════════════\n")
cat("CONCORDANT GENES SUMMARY\n")
cat("════════════════════════════════════════════════════════════\n\n")
print(concordance_gene_summary)

write.csv(concordance_gene_summary,
          file.path(OUTPUT_DIR, "Concordant_Genes_Summary.csv"),
          row.names = FALSE)

cat("\n✓ Saved: Concordant_Genes_Summary.csv\n\n")

# Save the detailed concordance statistics per gene
write.csv(go_prep_flu_ypd1$gene_concordance_stats,
          file.path(OUTPUT_DIR, "Gene_Concordance_Stats_Flu_vs_YPD1.csv"),
          row.names = FALSE)

write.csv(go_prep_flu_sc2glu$gene_concordance_stats,
          file.path(OUTPUT_DIR, "Gene_Concordance_Stats_Flu_vs_sc2glu.csv"),
          row.names = FALSE)

cat("✓ Saved: Gene_Concordance_Stats_*.csv\n")
```

---

**All steps complete!** This is your final, fully integrated `.Rmd` file with all changes included.