---
title: "Fig.3-supplementary-YPD1-vs-YPD2-eQTL-reproducibility"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# eQTL replicate reproducibility: YPD.1 vs YPD.2

This supplementary analysis checks how reproducible the FastLMM eQTL calls are
between the two independent YPD replicate datasets (YPD.1, YPD.2), across
SNPs, CNVs, and InDels. It reports (1) overlap of significant eQTL calls at
the association/SNP/gene level, (2) concordance of p-values and effect sizes
for shared eQTLs, and (3) a permutation-based test of whether effect-size
sign concordance exceeds chance.

```{r libs}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(scales)
  library(patchwork)
})

select <- dplyr::select
filter <- dplyr::filter

BASE <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all"
OUT  <- "../results"
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)

variant_colors <- c("SNP" = "#4472C4", "CNV" = "#70AD47", "Indel" = "#FFC000")
variant_levels <- c("SNP", "CNV", "Indel", "All")

dataset_colors <- c("YPD.1" = "#E76F51", "YPD.2" = "#2196ac")

category_colors <- c(
  "SHARED"     = "#4DAC26",
  "YPD1-ONLY"  = "#E76F51",
  "YPD2-ONLY"  = "#2196ac"
)
clean_labels <- c(
  "SHARED"    = "Shared",
  "YPD1-ONLY" = "YPD.1 only",
  "YPD2-ONLY" = "YPD.2 only"
)

pub_theme <- theme_classic(base_size = 14) +
  theme(
    axis.text.x        = element_text(size = 12, color = "black"),
    axis.text.y        = element_text(size = 12, color = "black"),
    axis.title.y       = element_text(size = 13, face = "bold", margin = margin(r = 10)),
    axis.title.x       = element_blank(),
    legend.position    = "none",
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.4),
    axis.line          = element_line(color = "black", linewidth = 0.5),
    axis.ticks         = element_line(color = "black", linewidth = 0.5),
    strip.text         = element_text(size = 12, face = "bold"),
    plot.margin        = margin(14, 18, 12, 14)
  )

save_tiff <- function(plot, filename, width = 10, height = 4.5) {
  ggsave(
    filename    = file.path(OUT, filename),
    plot        = plot,
    device      = "tiff",
    width       = width,
    height      = height,
    units       = "in",
    dpi         = 600,
    compression = "lzw"
  )
  message("Saved: ", filename)
}
```

## 1. Load significant eQTL calls (YPD.1 and YPD.2)

```{r load-signif}
read_signif <- function(cond, vt_file) {
  read.table(
    file.path(BASE, cond, "fastlmm/results", vt_file),
    header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = ""
  )
}

all_snps_eqtl_YPD1  <- read_signif("YPD.1", "all_signif_SNP_YPD.1_with_double_id_with_threshold.txt")
all_CNV_eqtl_YPD1   <- read_signif("YPD.1", "all_signif_CNV_YPD.1_with_double_id_with_threshold.txt")
all_indel_eqtl_YPD1 <- read_signif("YPD.1", "all_signif_Indels_YPD.1_with_double_id_with_threshold.txt")

all_snps_eqtl_YPD2  <- read_signif("YPD.2", "all_signif_SNP_YPD.1_with_double_id_with_threshold.txt")
all_CNV_eqtl_YPD2   <- read_signif("YPD.2", "all_signif_CNV_YPD.1_with_double_id_with_threshold.txt")
all_indel_eqtl_YPD2 <- read_signif("YPD.2", "all_signif_Indels_YPD.1_with_double_id_with_threshold.txt")

add_key <- function(df) {
  df$Chr_ChrPos_Pheno <- paste0(df$Chr, "_", df$ChrPos, "_", df$Pheno)
  df
}

all_snps_eqtl_YPD1  <- add_key(all_snps_eqtl_YPD1)
all_CNV_eqtl_YPD1   <- add_key(all_CNV_eqtl_YPD1)
all_indel_eqtl_YPD1 <- add_key(all_indel_eqtl_YPD1)
all_snps_eqtl_YPD2  <- add_key(all_snps_eqtl_YPD2)
all_CNV_eqtl_YPD2   <- add_key(all_CNV_eqtl_YPD2)
all_indel_eqtl_YPD2 <- add_key(all_indel_eqtl_YPD2)

all_eqtl_YPD1 <- bind_rows(
  mutate(all_snps_eqtl_YPD1,  variant_type = "SNP"),
  mutate(all_CNV_eqtl_YPD1,   variant_type = "CNV"),
  mutate(all_indel_eqtl_YPD1, variant_type = "Indel")
)
all_eqtl_YPD2 <- bind_rows(
  mutate(all_snps_eqtl_YPD2,  variant_type = "SNP"),
  mutate(all_CNV_eqtl_YPD2,   variant_type = "CNV"),
  mutate(all_indel_eqtl_YPD2, variant_type = "Indel")
)

cat("YPD.1 total eQTLs (all variant types):", nrow(all_eqtl_YPD1), "\n")
cat("YPD.2 total eQTLs (all variant types):", nrow(all_eqtl_YPD2), "\n")
```

## 2. Overlap at association / SNP / gene level

```{r compare-fn}
compare_eqtls <- function(df1, df2, label1 = "YPD.1", label2 = "YPD.2", variant = "All") {
  assocs1 <- unique(df1$Chr_ChrPos_Pheno); assocs2 <- unique(df2$Chr_ChrPos_Pheno)
  common_assocs <- intersect(assocs1, assocs2)

  snps1 <- unique(df1$SNP); snps2 <- unique(df2$SNP)
  common_snps <- intersect(snps1, snps2)

  genes1 <- unique(df1$Pheno); genes2 <- unique(df2$Pheno)
  common_genes <- intersect(genes1, genes2)

  invisible(list(
    variant = variant,
    assocs1 = assocs1, assocs2 = assocs2, common_assocs = common_assocs,
    snps1 = snps1, snps2 = snps2, common_snps = common_snps,
    genes1 = genes1, genes2 = genes2, common_genes = common_genes
  ))
}

res_snp   <- compare_eqtls(all_snps_eqtl_YPD1,  all_snps_eqtl_YPD2,  variant = "SNP")
res_cnv   <- compare_eqtls(all_CNV_eqtl_YPD1,   all_CNV_eqtl_YPD2,   variant = "CNV")
res_indel <- compare_eqtls(all_indel_eqtl_YPD1, all_indel_eqtl_YPD2, variant = "Indel")
res_all   <- compare_eqtls(all_eqtl_YPD1,        all_eqtl_YPD2,       variant = "All")

make_summary_row <- function(res, level = "association") {
  if (level == "association") { v1 <- res$assocs1; v2 <- res$assocs2; common <- res$common_assocs }
  else if (level == "SNP")    { v1 <- res$snps1;   v2 <- res$snps2;   common <- res$common_snps }
  else                        { v1 <- res$genes1;  v2 <- res$genes2;  common <- res$common_genes }

  data.frame(
    Variant     = res$variant, Level = level,
    YPD1_n      = length(v1), YPD2_n = length(v2), Shared_n = length(common),
    Pct_of_YPD1 = round(100 * length(common) / length(v1), 1),
    Pct_of_YPD2 = round(100 * length(common) / length(v2), 1)
  )
}

summary_table <- bind_rows(
  make_summary_row(res_snp,   "association"), make_summary_row(res_snp,   "SNP"), make_summary_row(res_snp,   "gene"),
  make_summary_row(res_cnv,   "association"), make_summary_row(res_cnv,   "SNP"), make_summary_row(res_cnv,   "gene"),
  make_summary_row(res_indel, "association"), make_summary_row(res_indel, "SNP"), make_summary_row(res_indel, "gene"),
  make_summary_row(res_all,   "association"), make_summary_row(res_all,   "SNP"), make_summary_row(res_all,   "gene")
)

print(summary_table, row.names = FALSE)
write.csv(summary_table, file.path(OUT, "YPD1_vs_YPD2_overlap_summary_table.csv"), row.names = FALSE)
```

## 3. MAIN: total eQTL counts per variant type (dodged, faceted)

```{r fig-counts}
count_df <- data.frame(
  Dataset      = rep(c("YPD.1", "YPD.2"), 4),
  Variant_type = c("SNP", "SNP", "CNV", "CNV", "Indel", "Indel", "All", "All"),
  Count        = c(
    nrow(all_snps_eqtl_YPD1),  nrow(all_snps_eqtl_YPD2),
    nrow(all_CNV_eqtl_YPD1),   nrow(all_CNV_eqtl_YPD2),
    nrow(all_indel_eqtl_YPD1), nrow(all_indel_eqtl_YPD2),
    nrow(all_eqtl_YPD1),       nrow(all_eqtl_YPD2)
  )
) %>%
  mutate(
    Variant_type = factor(Variant_type, levels = variant_levels),
    Dataset      = factor(Dataset, levels = c("YPD.1", "YPD.2"))
  )

fig_counts <- ggplot(count_df, aes(x = Dataset, y = Count, fill = Dataset)) +
  geom_col(width = 0.6, color = "white", linewidth = 0.3, position = position_dodge(width = 0.7)) +
  geom_text(aes(label = scales::comma(Count)), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = dataset_colors) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ Variant_type, scales = "free_y", nrow = 1) +
  labs(y = "Number of significant eQTLs") +
  pub_theme

print(fig_counts)
save_tiff(fig_counts, "figS_YPD1_vs_YPD2_eqtl_counts_by_variant.tiff", width = 10, height = 4.5)
```

## 4. MAIN: shared vs unique eQTL associations per variant type

```{r fig-overlap}
overlap_df <- bind_rows(
  data.frame(Variant = "SNP",   Category = c("YPD.1 only", "YPD.2 only", "Shared"),
             Count = c(length(setdiff(res_snp$assocs1, res_snp$assocs2)),
                       length(setdiff(res_snp$assocs2, res_snp$assocs1)),
                       length(res_snp$common_assocs))),
  data.frame(Variant = "CNV",   Category = c("YPD.1 only", "YPD.2 only", "Shared"),
             Count = c(length(setdiff(res_cnv$assocs1, res_cnv$assocs2)),
                       length(setdiff(res_cnv$assocs2, res_cnv$assocs1)),
                       length(res_cnv$common_assocs))),
  data.frame(Variant = "Indel", Category = c("YPD.1 only", "YPD.2 only", "Shared"),
             Count = c(length(setdiff(res_indel$assocs1, res_indel$assocs2)),
                       length(setdiff(res_indel$assocs2, res_indel$assocs1)),
                       length(res_indel$common_assocs))),
  data.frame(Variant = "All",   Category = c("YPD.1 only", "YPD.2 only", "Shared"),
             Count = c(length(setdiff(res_all$assocs1, res_all$assocs2)),
                       length(setdiff(res_all$assocs2, res_all$assocs1)),
                       length(res_all$common_assocs)))
) %>%
  mutate(
    Variant  = factor(Variant,  levels = variant_levels),
    Category = factor(Category, levels = c("YPD.1 only", "YPD.2 only", "Shared"))
  )

overlap_colors <- c("YPD.1 only" = "#E76F51", "YPD.2 only" = "#2196ac", "Shared" = "#4CAF50")

fig_overlap <- ggplot(overlap_df, aes(x = Category, y = Count, fill = Category)) +
  geom_col(width = 0.6, color = "white", linewidth = 0.3) +
  geom_text(aes(label = scales::comma(Count)), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = overlap_colors) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ Variant, scales = "free_y", nrow = 1) +
  labs(y = "Number of associations") +
  pub_theme +
  theme(axis.text.x = element_text(size = 11, color = "black", angle = 35, hjust = 1),
        plot.margin = margin(14, 18, 20, 14))

print(fig_overlap)
save_tiff(fig_overlap, "figS_YPD1_vs_YPD2_overlap_by_variant.tiff", width = 10, height = 4.5)
```

## 5. Per-eQTL categorization (shared / YPD.1-only / YPD.2-only)

```{r build-unified}
build_unified <- function(df1, df2) {
  d1 <- df1 %>%
    mutate(YPD1_neglog10p = -log10(PValue), YPD1_thr_neglog10p = -log10(thr_perm),
           YPD1_distance_above_thr = YPD1_neglog10p - YPD1_thr_neglog10p) %>%
    select(Chr_ChrPos_Pheno, Pheno, Chr, ChrPos,
           YPD1_neglog10p, YPD1_thr_neglog10p, YPD1_distance_above_thr,
           YPD1_SnpWeight = SnpWeight)

  d2 <- df2 %>%
    mutate(YPD2_neglog10p = -log10(PValue), YPD2_thr_neglog10p = -log10(thr_perm),
           YPD2_distance_above_thr = YPD2_neglog10p - YPD2_thr_neglog10p) %>%
    select(Chr_ChrPos_Pheno, Pheno, Chr, ChrPos,
           YPD2_neglog10p, YPD2_thr_neglog10p, YPD2_distance_above_thr,
           YPD2_SnpWeight = SnpWeight)

  ids1 <- d1$Chr_ChrPos_Pheno; ids2 <- d2$Chr_ChrPos_Pheno

  unified <- d2 %>%
    mutate(In_YPD1 = Chr_ChrPos_Pheno %in% ids1,
           Category = ifelse(In_YPD1, "SHARED", "YPD2-ONLY")) %>%
    left_join(d1 %>% select(Chr_ChrPos_Pheno, YPD1_neglog10p, YPD1_thr_neglog10p,
                             YPD1_distance_above_thr, YPD1_SnpWeight),
              by = "Chr_ChrPos_Pheno")

  ypd1_only <- d1 %>%
    filter(!Chr_ChrPos_Pheno %in% ids2) %>%
    mutate(Category = "YPD1-ONLY", In_YPD1 = TRUE,
           YPD2_neglog10p = NA, YPD2_thr_neglog10p = NA,
           YPD2_distance_above_thr = NA, YPD2_SnpWeight = NA) %>%
    select(Chr_ChrPos_Pheno, Pheno, Chr, ChrPos,
           YPD2_neglog10p, YPD2_thr_neglog10p, YPD2_distance_above_thr, YPD2_SnpWeight,
           YPD1_neglog10p, YPD1_thr_neglog10p, YPD1_distance_above_thr, YPD1_SnpWeight,
           Category, In_YPD1)

  bind_rows(unified, ypd1_only)
}

unified_snp   <- build_unified(all_snps_eqtl_YPD1,  all_snps_eqtl_YPD2)
unified_cnv   <- build_unified(all_CNV_eqtl_YPD1,   all_CNV_eqtl_YPD2)
unified_indel <- build_unified(all_indel_eqtl_YPD1, all_indel_eqtl_YPD2)
unified_all   <- build_unified(all_eqtl_YPD1,        all_eqtl_YPD2)

for (nm in c("SNP", "CNV", "Indel", "All")) {
  ud <- get(paste0("unified_", tolower(nm)))
  cat(nm, "— Total:", nrow(ud),
      "| Shared:", sum(ud$Category == "SHARED"),
      "| YPD1-only:", sum(ud$Category == "YPD1-ONLY"),
      "| YPD2-only:", sum(ud$Category == "YPD2-ONLY"), "\n")
}
```

## 6. Signal strength by category (boxplot + jitter)

```{r fig-boxplot}
make_boxplot <- function(ud, variant_label) {
  plot_data <- ud %>%
    mutate(
      neglog10p = case_when(
        Category == "SHARED"     ~ (YPD1_neglog10p + YPD2_neglog10p) / 2,
        Category == "YPD1-ONLY"  ~ YPD1_neglog10p,
        Category == "YPD2-ONLY"  ~ YPD2_neglog10p
      ),
      Category = factor(Category, levels = c("YPD1-ONLY", "YPD2-ONLY", "SHARED")),
      variant_label = variant_label   # <-- ADD THIS LINE
    ) %>%
    filter(!is.na(neglog10p))

  summary_stats <- plot_data %>%
    group_by(Category) %>%
    summarise(n = n(), median = median(neglog10p), .groups = "drop")
  
  # Get shared y-axis limits across all unified datasets

  all_neglog10p <- c(
  unified_snp$YPD1_neglog10p,   unified_snp$YPD2_neglog10p,
  unified_cnv$YPD1_neglog10p,   unified_cnv$YPD2_neglog10p,
  unified_indel$YPD1_neglog10p, unified_indel$YPD2_neglog10p,
  unified_all$YPD1_neglog10p,   unified_all$YPD2_neglog10p
)

y_max <- max(all_neglog10p, na.rm = TRUE) * 1.05
y_min <- min(all_neglog10p, na.rm = TRUE)

  ggplot(plot_data, aes(x = Category, y = neglog10p, color = Category, fill = Category)) +
    geom_boxplot(alpha = 0.15, width = 0.5, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.3, size = 1.5) +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "white") +
    geom_text(data = summary_stats,
              aes(x = Category, y = -0.5,
                  label = paste0("n=", scales::comma(n), "\nmedian=", round(median, 1))),
              inherit.aes = FALSE, size = 3.8, color = "black", fontface = "bold", vjust = 1) +
    scale_color_manual(values = category_colors, labels = clean_labels) +
    scale_fill_manual(values = category_colors, labels = clean_labels) +
    scale_x_discrete(labels = clean_labels) +
    scale_y_continuous(expand = expansion(mult = c(0.15, 0.08))) +
    labs(x = NULL, y = expression(-log[10](italic(p)-value))) +
    facet_wrap(~variant_label) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position    = "none",
      panel.grid.major.x = element_blank(),
      strip.text         = element_text(size = 13, face = "bold"),
      axis.text.x        = element_text(size = 12, color = "black", angle = 35, hjust = 1),
      axis.text.y        = element_text(size = 12, color = "black"),
      axis.title.y       = element_text(size = 13, face = "bold", margin = margin(r = 8)),
      plot.margin        = margin(12, 16, 20, 12)
    )
}

box_list <- list(
  make_boxplot(unified_snp,   "SNP"),
  make_boxplot(unified_cnv,   "CNV"),
  make_boxplot(unified_indel, "Indel"),
  make_boxplot(unified_all,   "All")
)

# Apply shared y limits to all panels
box_list <- lapply(box_list, function(p) {
  p + scale_y_continuous(limits = c(y_min, y_max),
                         expand = expansion(mult = c(0.15, 0.08)))
})

fig_boxplots <- wrap_plots(box_list, nrow = 1)
print(fig_boxplots)
save_tiff(fig_boxplots, "figS_YPD1_vs_YPD2_boxplot_neglog10p_by_variant.tiff", width = 14, height = 5)
```

## 7. MAIN: effect-size (SnpWeight) and p-value concordance for shared eQTLs

```{r fig-scatter-pval}
make_scatter_pval <- function(ud, variant_label) {
  plot_data <- ud %>%
    mutate(
      x_val    = if_else(is.na(YPD2_neglog10p), 0, YPD2_neglog10p),
      y_val    = if_else(is.na(YPD1_neglog10p), 0, YPD1_neglog10p),
      Category = factor(Category, levels = c("YPD1-ONLY", "YPD2-ONLY", "SHARED"))
    )
  ax_lim   <- max(c(plot_data$x_val, plot_data$y_val), na.rm = TRUE) * 1.05
  n_shared <- sum(plot_data$Category == "SHARED")
  n_ypd1   <- sum(plot_data$Category == "YPD1-ONLY")
  n_ypd2   <- sum(plot_data$Category == "YPD2-ONLY")

  ggplot(plot_data, aes(x = x_val, y = y_val, color = Category)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.7) +
    geom_point(data = plot_data %>% filter(Category != "SHARED"), alpha = 0.4, size = 1.8) +
    geom_point(data = plot_data %>% filter(Category == "SHARED"), alpha = 0.6, size = 2) +
    scale_color_manual(values = category_colors,
      labels = c("YPD1-ONLY" = paste0("YPD.1 only (n=", comma(n_ypd1), ")"),
                 "YPD2-ONLY" = paste0("YPD.2 only (n=", comma(n_ypd2), ")"),
                 "SHARED"    = paste0("Shared (n=", comma(n_shared), ")"))) +
    coord_fixed(ratio = 1, xlim = c(0, ax_lim), ylim = c(0, ax_lim)) +
    labs(title = variant_label,
         x = expression(bold("YPD.2") ~ -log[10](italic(p)-value)),
         y = expression(bold("YPD.1") ~ -log[10](italic(p)-value)), color = NULL) +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
          axis.text = element_text(size = 11, color = "black"),
          legend.position = "bottom", legend.text = element_text(size = 9),
          plot.margin = margin(10, 12, 10, 10))
}

scatter_pval_list <- list(
  make_scatter_pval(unified_snp, "SNP"), make_scatter_pval(unified_cnv, "CNV"),
  make_scatter_pval(unified_indel, "Indel"), make_scatter_pval(unified_all, "All")
)
fig_scatter_pval <- wrap_plots(scatter_pval_list, nrow = 1)
print(fig_scatter_pval)
save_tiff(fig_scatter_pval, "figS_YPD1_vs_YPD2_scatter_pval_by_variant.tiff", width = 15, height = 4.5)
```

```{r fig-scatter-effect}
make_scatter_effect <- function(ud, variant_label) {
  plot_data <- ud %>%
    mutate(
      x_val    = if_else(is.na(YPD2_SnpWeight), 0, YPD2_SnpWeight),
      y_val    = if_else(is.na(YPD1_SnpWeight), 0, YPD1_SnpWeight),
      Category = factor(Category, levels = c("YPD1-ONLY", "YPD2-ONLY", "SHARED"))
    )
  ax_abs   <- max(abs(c(plot_data$x_val, plot_data$y_val)), na.rm = TRUE) * 1.05
  n_shared <- sum(plot_data$Category == "SHARED")
  n_ypd1   <- sum(plot_data$Category == "YPD1-ONLY")
  n_ypd2   <- sum(plot_data$Category == "YPD2-ONLY")

  ggplot(plot_data, aes(x = x_val, y = y_val, color = Category)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.7) +
    geom_abline(slope = -1, intercept = 0, linetype = "dotted", color = "grey80", linewidth = 0.5) +
    geom_hline(yintercept = 0, color = "grey85", linewidth = 0.4) +
    geom_vline(xintercept = 0, color = "grey85", linewidth = 0.4) +
    geom_point(data = plot_data %>% filter(Category != "SHARED"), alpha = 0.4, size = 1.8) +
    geom_point(data = plot_data %>% filter(Category == "SHARED"), alpha = 0.6, size = 2) +
    scale_color_manual(values = category_colors,
      labels = c("YPD1-ONLY" = paste0("YPD.1 only (n=", comma(n_ypd1), ")"),
                 "YPD2-ONLY" = paste0("YPD.2 only (n=", comma(n_ypd2), ")"),
                 "SHARED"    = paste0("Shared (n=", comma(n_shared), ")"))) +
    coord_fixed(ratio = 1, xlim = c(-ax_abs, ax_abs), ylim = c(-ax_abs, ax_abs)) +
    labs(title = variant_label,
         x = expression(bold("YPD.2  SnpWeight") ~ (beta)),
         y = expression(bold("YPD.1  SnpWeight") ~ (beta)), color = NULL) +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
          axis.text = element_text(size = 11, color = "black"),
          legend.position = "bottom", legend.text = element_text(size = 9),
          plot.margin = margin(10, 12, 10, 10))
}

scatter_effect_list <- list(
  make_scatter_effect(unified_snp, "SNP"), make_scatter_effect(unified_cnv, "CNV"),
  make_scatter_effect(unified_indel, "Indel"), make_scatter_effect(unified_all, "All")
)
fig_scatter_effect <- wrap_plots(scatter_effect_list, nrow = 1)
print(fig_scatter_effect)
save_tiff(fig_scatter_effect, "figS_YPD1_vs_YPD2_scatter_effect_by_variant.tiff", width = 15, height = 4.5)
```

## 8. Sign concordance of SNP effect sizes (YPD.1-significant eQTLs looked up in YPD.2)

Raw per-gene FastLMM association files (`<gene>.first_assoc.txt`) are only
available for SNPs in the paths supplied, so the sign-concordance /
permutation analysis below is restricted to SNP eQTLs. Extending it to
CNV/Indel would need the equivalent per-gene raw-association directories for
YPD.2.

```{r sign-concordance-lookup}
YPD1_SNP_DIR <- file.path(BASE, "YPD.1/fastlmm/results/results_SNPs_968_with_double_id")
YPD2_SNP_DIR <- file.path(BASE, "YPD.2/fastlmm/results/results_SNPs_968_with_double_id")

ypd1_genes <- unique(all_snps_eqtl_YPD1$Pheno)
cat("Unique genes with YPD.1-significant SNP eQTLs:", length(ypd1_genes), "\n")

ypd2_lookup <- lapply(seq_along(ypd1_genes), function(i) {
  gene <- ypd1_genes[i]
  path <- file.path(YPD2_SNP_DIR, gene, paste0(gene, ".first_assoc.txt"))
  if (!file.exists(path)) return(NULL)

  ypd1_snps <- all_snps_eqtl_YPD1 %>%
    filter(Pheno == gene) %>%
    select(SNP, Beta_YPD1 = SnpWeight, P_YPD1 = PValue)

  ypd2 <- read.table(path, header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote = "") %>%
    select(SNP, Beta_YPD2 = SnpWeight, P_YPD2 = PValue)

  inner_join(ypd1_snps, ypd2, by = "SNP") %>% mutate(Pheno = gene)
}) %>% bind_rows()

cat("Total SNP x Pheno pairs retrieved:", nrow(ypd2_lookup), "\n")
cat("Genes with YPD.2 data found:      ", n_distinct(ypd2_lookup$Pheno), "\n")

write.table(ypd2_lookup, file.path(OUT, "ypd2_lookup_YPD1_signif_SNPs.txt"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r fig-sign-concordance}
ypd2_lookup <- ypd2_lookup %>%
  mutate(direction = ifelse(sign(Beta_YPD1) == sign(Beta_YPD2), "Concordant", "Discordant"))

beta_cor         <- cor(ypd2_lookup$Beta_YPD1, ypd2_lookup$Beta_YPD2, method = "pearson", use = "complete.obs")
sign_concordance <- mean(sign(ypd2_lookup$Beta_YPD1) == sign(ypd2_lookup$Beta_YPD2), na.rm = TRUE)

cat(sprintf("Beta correlation (Pearson r): %.3f\n", beta_cor))
cat(sprintf("Sign concordance:             %.1f%%\n", sign_concordance * 100))

fig_effect_concordance <- ggplot(ypd2_lookup, aes(x = Beta_YPD1, y = Beta_YPD2, color = direction)) +
  geom_point(size = 1.5, alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_smooth(method = "lm", color = "grey30", se = TRUE, linewidth = 0.8) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_color_manual(values = c("Concordant" = "#27AE60", "Discordant" = "#E74C3C")) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = sprintf("Pearson r = %.3f\nSign concordance = %.1f%%\nn = %d eQTLs",
                            beta_cor, sign_concordance * 100, nrow(ypd2_lookup)),
           size = 4, fontface = "bold") +
  labs(x = "Effect size Beta (YPD.1)", y = "Effect size Beta (YPD.2)", color = "Direction") +
  theme_classic(base_size = 13) +
  theme(legend.position = "bottom")

print(fig_effect_concordance)
save_tiff(fig_effect_concordance, "figS_YPD1_signif_SNP_effect_size_concordance.tiff", width = 7, height = 6.5)
```

## 9. Statistical significance of the sign concordance (binomial + permutation test)

```{r sign-concordance-stats}
n_total       <- nrow(ypd2_lookup)
n_concordant  <- sum(sign(ypd2_lookup$Beta_YPD1) == sign(ypd2_lookup$Beta_YPD2), na.rm = TRUE)
obs_sign_conc <- n_concordant / n_total
obs_r         <- cor(ypd2_lookup$Beta_YPD1, ypd2_lookup$Beta_YPD2, method = "pearson", use = "complete.obs")

cat(sprintf("Observed sign concordance : %.3f (%d / %d)\n", obs_sign_conc, n_concordant, n_total))
cat(sprintf("Observed Pearson r        : %.3f\n\n", obs_r))

binom_result <- binom.test(n_concordant, n_total, p = 0.5, alternative = "greater")
cat("Binomial test (H0: sign concordance = 0.50):\n")
cat(sprintf("  p-value : %.3e\n", binom_result$p.value))
cat(sprintf("  95%% CI  : [%.3f, %.3f]\n\n", binom_result$conf.int[1], binom_result$conf.int[2]))

set.seed(42)
n_perm    <- 5000
perm_sign <- numeric(n_perm)
perm_r    <- numeric(n_perm)

complete <- complete.cases(ypd2_lookup$Beta_YPD1, ypd2_lookup$Beta_YPD2)
b1 <- ypd2_lookup$Beta_YPD1[complete]
b2 <- ypd2_lookup$Beta_YPD2[complete]

for (i in seq_len(n_perm)) {
  b2_shuf      <- sample(b2)
  perm_sign[i] <- mean(sign(b1) == sign(b2_shuf))
  perm_r[i]    <- cor(b1, b2_shuf, method = "pearson")
}

perm_p_sign <- mean(perm_sign >= obs_sign_conc)
perm_p_r    <- mean(perm_r    >= obs_r)

cat("Permutation test (n = 5000, beta-shuffling):\n")
cat(sprintf("  Sign concordance: obs = %.3f | null mean = %.3f | p = %.4f\n", obs_sign_conc, mean(perm_sign), perm_p_sign))
cat(sprintf("  Pearson r       : obs = %.3f | null mean = %.3f | p = %.4f\n", obs_r, mean(perm_r), perm_p_r))
```

```{r fig-null-dist}
perm_df <- data.frame(sign_conc = perm_sign, pearson_r = perm_r)

p_null_sign <- ggplot(perm_df, aes(x = sign_conc)) +
  geom_histogram(bins = 60, fill = "grey70", color = "white", linewidth = 0.2) +
  geom_vline(xintercept = obs_sign_conc, color = "#E74C3C", linewidth = 1) +
  geom_vline(xintercept = 0.5, color = "black", linewidth = 0.6, linetype = "dashed") +
  annotate("text", x = obs_sign_conc, y = Inf, label = sprintf("Observed\n%.3f", obs_sign_conc),
           hjust = -0.1, vjust = 1.5, color = "#E74C3C", size = 3.5, fontface = "bold") +
  annotate("text", x = Inf, y = Inf,
           label = sprintf("Binomial p = %.2e\nPerm p = %.4f", binom_result$p.value, perm_p_sign),
           hjust = 1.1, vjust = 1.5, size = 3.5, fontface = "bold") +
  labs(x = "Sign concordance (fraction same direction)", y = "Count") +
  theme_classic(base_size = 12)

p_null_r <- ggplot(perm_df, aes(x = pearson_r)) +
  geom_histogram(bins = 60, fill = "grey70", color = "white", linewidth = 0.2) +
  geom_vline(xintercept = obs_r, color = "#27AE60", linewidth = 1) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.6, linetype = "dashed") +
  annotate("text", x = obs_r, y = Inf, label = sprintf("Observed\nr = %.3f", obs_r),
           hjust = -0.1, vjust = 1.5, color = "#27AE60", size = 3.5, fontface = "bold") +
  annotate("text", x = Inf, y = Inf, label = sprintf("Perm p = %.4f", perm_p_r),
           hjust = 1.1, vjust = 1.5, size = 3.5, fontface = "bold") +
  labs(x = "Pearson r (Beta YPD.1 vs Beta YPD.2)", y = "Count") +
  theme_classic(base_size = 12)

fig_null_dist <- p_null_sign + p_null_r
print(fig_null_dist)
save_tiff(fig_null_dist, "figS_YPD1_vs_YPD2_permutation_null_distributions.tiff", width = 11, height = 5)
```

## Session info

```{r session-info}
sessionInfo()
```
