---
title: "Fig.3-Supplementary-panel-A-cis_trans_eqtl_mapping_flu-sc2glu"
author: "Akriti Agrawal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: true
    number_section: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../results/") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 10)
```

```{r libs, include=FALSE}
library(tidyverse)
library(data.table)
library(GenomicRanges)
select <- dplyr::select
filter <- dplyr::filter
```
0. Config — set your four .clumped/assoc paths here; the indel switch is the confirmed REF_ALT. for fluconazole
```{r config}
RES  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results"
CLMP <- file.path(RES)
# All NEW outputs (annotated tables + figures) go here, NOT into RES:
OUT  <- "../results"
ANNOT_CSV <- "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"
GENE_LOC  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/data/pheno_data/gene_location.txt"
CIS_WINDOW_BP <- 2000

variant_cfg <- list(
  SNP   = list(clumped = file.path(CLMP, "clumped_results_snp_82.clumped"), #it is in results folder only not in a subfolder
               assoc   = file.path(RES,  "all_signif_SNPs_flu_with_threshold.txt")),
  CNV   = list(clumped = file.path(CLMP, "clumped_results_CNVs.clumped"),
               assoc   = file.path(RES,  "all_signif_CNVs_flu_with_threshold.txt")),
  Indel = list(clumped = file.path(CLMP, "clumped_results_Indels.clumped"),
               assoc   = file.path(RES,  "all_signif_Indels_flu_with_threshold.txt")),
  SV    = list(clumped = file.path(CLMP, "clumped_results_SVs.clumped"),
               assoc   = file.path(RES,  "all_signif_SVs_flu_with_threshold.txt"))
)
```

```{}
#fluconazole plink files:
variant_types = {
        'SNPs': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/SNPs/SNPs_change_names_in_fam_968/SNPs_82fluc/full969Matrix.SNPs.Biallelic.Var.DP10.99percNonMissing.MAF15.NoSubTelo20kb.plink_YPD.1_82fluc_strains',
            'outdir': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/results_SNPs'
        },
        'CNVs': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/CNVs/CNVs_82fluc/CNV_969strains_gain_maf15_N400_20221121.plink_82fluc_strains',
            'outdir': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/results_CNVs'
        },
        'Indels': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SNPs_indels/modified_snp_indel_plink_for_96strains/Indels_only_fluc_81strains/indels._fluc_81strains',
            'outdir': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/results_Indels'
        },
        'SVs': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SVs_from_SV.1087/SV.1087samples_updated_fam_files/SVs_for_fluc_81strains/SV.1087samples.SVs_to_fluc_81strains',
            'outdir': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/results_SVs'
        }
    }
```

0. Config — set your four .clumped/assoc paths here; the indel switch is the confirmed REF_ALT. for sc2glu
```{r config}
RES  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results"
CLMP <- file.path(RES, "plink_without_double_id")
# All NEW outputs (annotated tables + figures) go here, NOT into RES:
OUT  <- "../results"
ANNOT_CSV <- "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"
GENE_LOC  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/data/pheno_data/gene_location.txt"
CIS_WINDOW_BP <- 2000

variant_cfg <- list(
  SNP   = list(clumped = file.path(CLMP, "clumped_results_snp.clumped"),
               assoc   = file.path(RES,  "all_signif_SNP_71_with_threshold.txt")),
  CNV   = list(clumped = file.path(CLMP, "clumped_results_CNVs.clumped"),
               assoc   = file.path(RES,  "all_signif_CNV_71_with_threshold.txt")),
  Indel = list(clumped = file.path(CLMP, "clumped_results_Indels.clumped"),
               assoc   = file.path(RES,  "all_signif_Indels_71_with_threshold.txt")),
  SV    = list(clumped = file.path(CLMP, "clumped_results_SVs.clumped"),
               assoc   = file.path(RES,  "all_signif_SVs_71_with_threshold.txt"))
)
```


```{}
#sc2glucose variant file information - the only difference is here I have filtered for 71 strains
variant_types = {
        'SNPs': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/SNPs/SNPs_change_names_in_fam_968/SNPs_71/full969Matrix.SNPs.Biallelic.Var.DP10.99percNonMissing.MAF15.NoSubTelo20kb.plink_71strains',
            'outdir': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/results_SNPs_71'
        },
        'CNVs': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/CNVs/CNVs_71/CNV_969strains_gain_maf15_N400_20221121.plink_71strains',
            'outdir': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/results_CNVs_71'
        },
        'CNVs.2_cov': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/structural_variant_vcf/SVcalls/CNVs_for_71_strains/full1086.CNVs.MAC1.plink_to_71strains',
            'outdir': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/results_CNVs_71_cov'
        },
        'Indels': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SNPs_indels/modified_snp_indel_plink_for_96strains/Indels_only_71/indels.71samples',
            'outdir': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/results_Indels_71'
        },
        'SVs': {
            'genotype': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SVs_from_SV.1087/SV.1087samples_updated_fam_files/SVs_for_71_strains/SV.1087samples.SVs_to_71strains',
            'outdir': '/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/results_SVs_71'
        }
    }
```

1. Shared reference objects (chromosome coords, gene annotations, phenotype coords)
```{r reference}
ann <- read.csv(ANNOT_CSV)
annotations_gr <- GRanges(ann$seqnames,
                          IRanges(ann$start, ann$end),
                          strand = ann$strand,
                          type = ann$type, gene_id = ann$gene_id, gene_name = ann$gene_name)

roman16 <- c("I","II","III","IV","V","VI","VII","VIII","IX","X","XI","XII","XIII","XIV","XV","XVI")
num_to_roman <- function(x) roman16[x]

chr_lengths <- c(I=230218, II=813184, III=316620, IV=1531933, V=576874, VI=270161,
                 VII=1090940, VIII=562643, IX=439888, X=745751, XI=666816,
                 XII=1078177, XIII=924431, XIV=784333, XV=1091291, XVI=948066)
chr_offsets <- c(0, cumsum(chr_lengths[-length(chr_lengths)])); names(chr_offsets) <- roman16
chr_mids <- chr_offsets + chr_lengths / 2

gene_location_fixed <- read.table(GENE_LOC, header = TRUE, stringsAsFactors = FALSE) %>%
  select(gene_id, gene_name, chr, left, right) %>%
  mutate(right = right + 2,                                  # gene_end is start of stop codon
         gene_id = str_replace(gene_id, "-", "."))

feature_priority <- c(promoter=1, three_prime_UTR=2, five_prime_UTR=3,
                      start_codon=4, stop_codon=5, CDS=6)
prio <- function(f) { p <- feature_priority[f]; p[is.na(p)] <- 99; p }
```
Chunk 4 — reader (your process_clumped_variants, condensed)
```{r}
read_clumped <- function(clumped, assoc) {
  ids <- read.table(clumped, header = TRUE)$SNP
  read.table(assoc, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    filter(SNP %in% ids)
}
```


Chunk 5 — point annotation (SNP = priority, CNV = first overlap)

```{r annot-point}
annotate_point <- function(df, pick = c("priority","first")) {
  pick <- match.arg(pick)
  df$var_chr_roman <- num_to_roman(df$Chr)
  gr <- GRanges(df$var_chr_roman, IRanges(df$ChrPos, df$ChrPos))
  df$causal_gene <- NA_character_; df$causal_gene_feature <- NA_character_

  ov <- findOverlaps(gr, annotations_gr, type = "any")
  for (i in unique(queryHits(ov))) {
    subj <- subjectHits(ov)[queryHits(ov) == i]
    k <- if (pick == "priority") subj[which.min(prio(annotations_gr$type[subj]))] else subj[1]
    df$causal_gene[i] <- annotations_gr$gene_id[k]
    df$causal_gene_feature[i] <- annotations_gr$type[k]
  }
  # intergenic: nearest gene within 2 kb, else intergenic_distant
  m <- which(is.na(df$causal_gene))
  if (length(m)) {
    ng <- nearest(gr[m], annotations_gr); ok <- !is.na(ng)
    d <- rep(NA_real_, length(m)); d[ok] <- distance(gr[m][ok], annotations_gr[ng[ok]])
    near <- !is.na(d) & d <= CIS_WINDOW_BP
    df$causal_gene[m[near]]          <- annotations_gr$gene_id[ng[near]]
    df$causal_gene_feature[m[near]]  <- "intergenic_nearby"
    df$causal_gene_feature[m[!near]] <- "intergenic_distant"
  }
  df
}
```


Chunk 6 — range annotation (Indel = no nearest, SV = nearest)

```{r annot-sv}
annotate_range <- function(df, range_end, use_nearest) {
  df$var_chr_roman <- num_to_roman(df$Chr)
  gr <- GRanges(df$var_chr_roman, IRanges(df$ChrPos, range_end))
  df$causal_gene <- NA_character_; df$causal_gene_feature <- NA_character_

  ov <- findOverlaps(gr, annotations_gr, type = "any")
  for (i in unique(queryHits(ov))) {
    subj <- subjectHits(ov)[queryHits(ov) == i]
    df$causal_gene[i]         <- paste(unique(annotations_gr$gene_id[subj]), collapse = "; ")
    df$causal_gene_feature[i] <- paste(unique(annotations_gr$type[subj]),   collapse = "; ")
  }
  if (use_nearest) {                              # SV
    m <- which(is.na(df$causal_gene))
    if (length(m)) {
      ng <- nearest(gr[m], annotations_gr); ok <- !is.na(ng)
      d <- rep(NA_real_, length(m)); d[ok] <- distance(gr[m][ok], annotations_gr[ng[ok]])
      near <- !is.na(d) & d <= CIS_WINDOW_BP
      df$causal_gene[m[near]]          <- annotations_gr$gene_id[ng[near]]
      df$causal_gene_feature[m[near]]  <- "intergenic_nearby"
      df$causal_gene_feature[m[!near]] <- "intergenic_distant"
    }
  } else {                                        # Indel: original had nearest commented out
    df$causal_gene_feature[is.na(df$causal_gene)] <- "intergenic_distant"
  }
  df
}
```
Chunk 7 — per-type dispatch (builds each type's ranges the way your originals do)
```{r annot-dispatch}
annotate_variant <- function(df, vt) {
  if (vt == "SNP") return(annotate_point(df, "priority"))
  if (vt == "CNV") return(annotate_point(df, "first"))

  if (vt == "Indel") {
    p <- strsplit(df$SNP, "_")                    # CHR_POS_REF_ALT  (A1=ALT, A2=REF)
    ref_len <- vapply(p, function(x) nchar(x[3]), 1L)
    alt_len <- vapply(p, function(x) nchar(x[4]), 1L)
    df$Indel_Size <- alt_len - ref_len            # +ve = insertion, -ve = deletion
    range_end <- ifelse(df$Indel_Size < 0, df$ChrPos + ref_len - 1, df$ChrPos)
    return(annotate_range(df, range_end, use_nearest = FALSE))
  }
  if (vt == "SV") {
    sub <- vapply(strsplit(df$SNP, "_"), `[`, "", 2)
    sz  <- vapply(strsplit(df$SNP, "_"), function(x){ i <- which(x=="len"); if(length(i)) as.numeric(x[i+1]) else NA_real_ }, 0)
    range_end <- ifelse(sub %in% c("INS","TRA"), df$ChrPos, df$ChrPos + abs(sz))
    range_end[is.na(range_end)] <- df$ChrPos[is.na(range_end)]
    return(annotate_range(df, range_end, use_nearest = TRUE))
  }
}
```
Chunk 8 — cis/trans + harmonise (identical rule across all four originals)
```{r harmonise}
add_cis_trans <- function(df, vt) {
  df %>%
    left_join(gene_location_fixed, by = c("Pheno" = "gene_id")) %>%
    mutate(Pheno_chr = chr, Pheno_start = left, Pheno_end = right,
           dist_to_pheno = case_when(
             var_chr_roman != Pheno_chr ~ NA_real_,
             ChrPos < Pheno_start ~ Pheno_start - ChrPos,
             ChrPos > Pheno_end   ~ ChrPos - Pheno_end,
             TRUE ~ 0),
           cis_trans = case_when(
             var_chr_roman != Pheno_chr ~ "Trans",
             dist_to_pheno == 0    ~ "Cis",
             dist_to_pheno <= 2000 ~ "Cis",
             causal_gene == Pheno  ~ "Cis",
             TRUE                  ~ "Trans"),
           variant_type = vt,
           abs_effect_size = abs(SnpWeight)) %>%
    select(variant_type, SNP, Pheno, var_chr_roman, ChrPos,
           Pheno_chr, Pheno_start, Pheno_end, causal_gene,
           cis_trans, abs_effect_size, PValue)
}
```

Chunk 9 — build combined table
```{r build-table}
combined <- lapply(names(variant_cfg), function(vt) {
  cfg <- variant_cfg[[vt]]
  message("annotating ", vt, " ...")
  read_clumped(cfg$clumped, cfg$assoc) %>% annotate_variant(vt) %>% add_cis_trans(vt)
}) %>% bind_rows() %>%
  mutate(var_chr_roman = factor(var_chr_roman, levels = roman16),
         Pheno_chr     = factor(Pheno_chr,     levels = roman16),
         variant_type  = factor(variant_type,  levels = c("SNP","CNV","Indel","SV")),
         var_genome    = chr_offsets[as.character(var_chr_roman)] + ChrPos,
         pheno_genome  = chr_offsets[as.character(Pheno_chr)] + (Pheno_start + Pheno_end)/2) %>%
  filter(!is.na(var_genome), !is.na(pheno_genome), !is.na(cis_trans))

table(combined$variant_type, combined$cis_trans)
```
Chunk 10 — Panel A: 2×2 cis/trans facet
```{r panelA, fig.width=11, fig.height=11}
panelA <- ggplot(combined, aes(var_genome, pheno_genome, color = cis_trans, size = abs_effect_size)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ variant_type, ncol = 2) +
  scale_color_manual(name = "eQTL type", values = c(Cis = "#d73027", Trans = "#4575b4")) +
  scale_size_continuous(name = "Effect size (|β|)", range = c(0.4, 2.6), trans = "log10") +
  scale_x_continuous(breaks = chr_mids, labels = names(chr_mids), expand = c(0.01, 0)) +
  scale_y_continuous(breaks = chr_mids, labels = names(chr_mids), expand = c(0.01, 0)) +
  labs(x = "Variant position (chromosome)", y = "Gene position (chromosome)") +
  theme_bw(base_size = 12) +
  theme(strip.text = element_text(face = "bold"),
        axis.text = element_text(size = 8),
        legend.position = "bottom")

ggsave(file.path(OUT, "Fig3_panelA_cis_trans_2x2.tiff"), panelA,
       width = 11, height = 10, units = "in", dpi = 600, compression = "lzw")
ggsave(file.path(OUT, "Fig3_panelA_cis_trans_2x2.png"),  panelA,
       width = 11, height = 10, units = "in", dpi = 300)
panelA
```
```{r}
library(data.table)

# --- per-type annotated tables (full columns, before harmonise) ---
annotated_list <- lapply(names(variant_cfg), function(vt) {
  cfg <- variant_cfg[[vt]]
  read_clumped(cfg$clumped, cfg$assoc) %>% annotate_variant(vt) %>% add_cis_trans(vt)
})
names(annotated_list) <- names(variant_cfg)

out_names <- c(SNP  = "snp_data_annotated_cis_trans.txt",
               CNV  = "cnv_data_annotated_cis_trans.txt",
               Indel= "indel_data_annotated_cis_trans.txt",
               SV   = "sv_data_annotated_cis_trans.txt")

for (vt in names(annotated_list)) {
  f <- file.path(OUT, out_names[[vt]])
  fwrite(annotated_list[[vt]], f, sep = "\t", quote = FALSE,
         row.names = FALSE, na = "NA")
  cat("Wrote", out_names[[vt]], "—", nrow(annotated_list[[vt]]), "rows\n")
}

# --- combined harmonised table used by the plot ---
fwrite(combined, file.path(OUT, "all_variants_cis_trans_combined.txt"),
       sep = "\t", quote = FALSE, row.names = FALSE, na = "NA")
cat("Wrote all_variants_cis_trans_combined.txt —", nrow(combined), "rows\n")
```

```{r save-table}
fwrite(combined, "../results/Fig3_panelA_combined_cis_trans_YPD.txt",
       sep = "\t", quote = FALSE, na = "NA")
```