---
title: "Fig.3-Supplementary — cis/trans eQTL mapping (Fluconazole & sc2glu), all variant types"
author: "Akriti Agrawal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: true
    number_section: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../results/") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 11, fig.height = 11)
```

```{r libs, include=FALSE}
library(tidyverse)
library(data.table)
library(GenomicRanges)
select <- dplyr::select
filter <- dplyr::filter
```

# Config — both conditions in one list

Each condition carries its own `RES`, clumped-file directory (`CLMP`), gene-location
file, and the four `.clumped`/assoc paths. The indel switch is the confirmed
`REF_ALT` convention (A1 = ALT, A2 = REF). All NEW outputs go to `OUT`, never into `RES`.

Note the per-condition differences:

  * **Fluconazole** — SNP clumped file is `clumped_results_snp_82.clumped` and all four
    clumped files sit directly in `RES` (no `plink_without_double_id` subfolder).
  * **sc2glu** — clumped files sit in the `plink_without_double_id` subfolder; assoc files
    use the `_71` strain-count suffix.

```{r config}
OUT           <- "../results"
ANNOT_CSV     <- "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"
CIS_WINDOW_BP <- 2000

# ---- Fluconazole ----------------------------------------------------------
flu_RES  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results"
flu_CLMP <- flu_RES                                    # clumped files are in RES directly

# ---- sc2glu ---------------------------------------------------------------
sc2_RES  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results"
sc2_CLMP <- file.path(sc2_RES, "plink")

CONDITIONS <- list(
  Fluconazole = list(
    GENE_LOC = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/data/pheno_data/gene_location.txt",
    variant_cfg = list(
      SNP   = list(clumped = file.path(flu_CLMP, "clumped_results_snp_82.clumped"),
                   assoc   = file.path(flu_RES,  "all_signif_SNPs_flu_with_threshold.txt")),
      CNV   = list(clumped = file.path(flu_CLMP, "clumped_results_CNVs.clumped"),
                   assoc   = file.path(flu_RES,  "all_signif_CNVs_flu_with_threshold.txt")),
      Indel = list(clumped = file.path(flu_CLMP, "clumped_results_Indels.clumped"),
                   assoc   = file.path(flu_RES,  "all_signif_Indels_flu_with_threshold.txt")),
      SV    = list(clumped = file.path(flu_CLMP, "clumped_results_SVs.clumped"),
                   assoc   = file.path(flu_RES,  "all_signif_SVs_flu_with_threshold.txt"))
    )
  ),
  sc2glu = list(
    GENE_LOC = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/data/pheno_data/gene_location.txt",
    variant_cfg = list(
      SNP   = list(clumped = file.path(sc2_CLMP, "clumped_results_snp.clumped"),
                   assoc   = file.path(sc2_RES,  "all_signif_SNP_71_with_threshold.txt")),
      CNV   = list(clumped = file.path(sc2_CLMP, "clumped_results_CNVs.clumped"),
                   assoc   = file.path(sc2_RES,  "all_signif_CNV_71_with_threshold.txt")),
      Indel = list(clumped = file.path(sc2_CLMP, "clumped_results_Indels.clumped"),
                   assoc   = file.path(sc2_RES,  "all_signif_Indels_71_with_threshold.txt")),
      SV    = list(clumped = file.path(sc2_CLMP, "clumped_results_SVs.clumped"),
                   assoc   = file.path(sc2_RES,  "all_signif_SVs_71_with_threshold.txt"))
    )
  )
)
```

# Shared reference objects (annotations, chromosome coordinates)

These are the same across both conditions (one genome, R64) so they are built once.
The per-condition `gene_location.txt` is rebuilt inside the loop.

```{r reference}
ann <- read.csv(ANNOT_CSV)
annotations_gr <- GRanges(ann$seqnames,
                          IRanges(ann$start, ann$end),
                          strand = ann$strand,
                          type = ann$type, gene_id = ann$gene_id, gene_name = ann$gene_name)

roman16 <- c("I","II","III","IV","V","VI","VII","VIII","IX","X","XI","XII","XIII","XIV","XV","XVI")
num_to_roman <- function(x) roman16[x]

chr_lengths <- c(I=230218, II=813184, III=316620, IV=1531933, V=576874, VI=270161,
                 VII=1090940, VIII=562643, IX=439888, X=745751, XI=666816,
                 XII=1078177, XIII=924431, XIV=784333, XV=1091291, XVI=948066)
chr_offsets <- c(0, cumsum(chr_lengths[-length(chr_lengths)])); names(chr_offsets) <- roman16
chr_mids <- chr_offsets + chr_lengths / 2

feature_priority <- c(promoter=1, three_prime_UTR=2, five_prime_UTR=3,
                      start_codon=4, stop_codon=5, CDS=6)
prio <- function(f) { p <- feature_priority[f]; p[is.na(p)] <- 99; p }

# per-condition gene-location loader (called inside the loop)
build_gene_loc <- function(path) {
  read.table(path, header = TRUE, stringsAsFactors = FALSE) %>%
    select(gene_id, gene_name, chr, left, right) %>%
    mutate(right = right + 2,                           # gene_end is start of stop codon
           gene_id = str_replace(gene_id, "-", "."))
}
```

# Annotation functions (your originals, unchanged)

```{r reader}
read_clumped <- function(clumped, assoc) {
  ids <- read.table(clumped, header = TRUE)$SNP
  read.table(assoc, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    filter(SNP %in% ids)
}
```

```{r annot-point}
# point annotation (SNP = priority feature, CNV = first overlap)
annotate_point <- function(df, pick = c("priority","first")) {
  pick <- match.arg(pick)
  df$var_chr_roman <- num_to_roman(df$Chr)
  gr <- GRanges(df$var_chr_roman, IRanges(df$ChrPos, df$ChrPos))
  df$causal_gene <- NA_character_; df$causal_gene_feature <- NA_character_

  ov <- findOverlaps(gr, annotations_gr, type = "any")
  for (i in unique(queryHits(ov))) {
    subj <- subjectHits(ov)[queryHits(ov) == i]
    k <- if (pick == "priority") subj[which.min(prio(annotations_gr$type[subj]))] else subj[1]
    df$causal_gene[i] <- annotations_gr$gene_id[k]
    df$causal_gene_feature[i] <- annotations_gr$type[k]
  }
  # intergenic: nearest gene within 2 kb, else intergenic_distant
  m <- which(is.na(df$causal_gene))
  if (length(m)) {
    ng <- nearest(gr[m], annotations_gr); ok <- !is.na(ng)
    d <- rep(NA_real_, length(m)); d[ok] <- distance(gr[m][ok], annotations_gr[ng[ok]])
    near <- !is.na(d) & d <= CIS_WINDOW_BP
    df$causal_gene[m[near]]          <- annotations_gr$gene_id[ng[near]]
    df$causal_gene_feature[m[near]]  <- "intergenic_nearby"
    df$causal_gene_feature[m[!near]] <- "intergenic_distant"
  }
  df
}
```

```{r annot-range}
# range annotation (Indel = overlap only, SV = overlap + nearest)
annotate_range <- function(df, range_end, use_nearest) {
  df$var_chr_roman <- num_to_roman(df$Chr)
  gr <- GRanges(df$var_chr_roman, IRanges(df$ChrPos, range_end))
  df$causal_gene <- NA_character_; df$causal_gene_feature <- NA_character_

  ov <- findOverlaps(gr, annotations_gr, type = "any")
  for (i in unique(queryHits(ov))) {
    subj <- subjectHits(ov)[queryHits(ov) == i]
    df$causal_gene[i]         <- paste(unique(annotations_gr$gene_id[subj]), collapse = "; ")
    df$causal_gene_feature[i] <- paste(unique(annotations_gr$type[subj]),   collapse = "; ")
  }
  if (use_nearest) {                              # SV
    m <- which(is.na(df$causal_gene))
    if (length(m)) {
      ng <- nearest(gr[m], annotations_gr); ok <- !is.na(ng)
      d <- rep(NA_real_, length(m)); d[ok] <- distance(gr[m][ok], annotations_gr[ng[ok]])
      near <- !is.na(d) & d <= CIS_WINDOW_BP
      df$causal_gene[m[near]]          <- annotations_gr$gene_id[ng[near]]
      df$causal_gene_feature[m[near]]  <- "intergenic_nearby"
      df$causal_gene_feature[m[!near]] <- "intergenic_distant"
    }
  } else {                                        # Indel: original had nearest commented out
    df$causal_gene_feature[is.na(df$causal_gene)] <- "intergenic_distant"
  }
  df
}
```

```{r annot-dispatch}
# per-type dispatch (builds each type's ranges the way your originals do)
annotate_variant <- function(df, vt) {
  if (vt == "SNP") return(annotate_point(df, "priority"))
  if (vt == "CNV") return(annotate_point(df, "first"))

  if (vt == "Indel") {
    p <- strsplit(df$SNP, "_")                    # CHR_POS_REF_ALT  (A1=ALT, A2=REF)
    ref_len <- vapply(p, function(x) nchar(x[3]), 1L)
    alt_len <- vapply(p, function(x) nchar(x[4]), 1L)
    df$Indel_Size <- alt_len - ref_len            # +ve = insertion, -ve = deletion
    range_end <- ifelse(df$Indel_Size < 0, df$ChrPos + ref_len - 1, df$ChrPos)
    return(annotate_range(df, range_end, use_nearest = FALSE))
  }
  if (vt == "SV") {
    sub <- vapply(strsplit(df$SNP, "_"), `[`, "", 2)
    sz  <- vapply(strsplit(df$SNP, "_"), function(x){ i <- which(x=="len"); if(length(i)) as.numeric(x[i+1]) else NA_real_ }, 0)
    range_end <- ifelse(sub %in% c("INS","TRA"), df$ChrPos, df$ChrPos + abs(sz))
    range_end[is.na(range_end)] <- df$ChrPos[is.na(range_end)]
    return(annotate_range(df, range_end, use_nearest = TRUE))
  }
}
```

```{r harmonise}
# cis/trans + harmonise (identical rule across all four originals)
# reads gene_location_fixed from the global env (set per condition inside the loop)
add_cis_trans <- function(df, vt) {
  df %>%
    left_join(gene_location_fixed, by = c("Pheno" = "gene_id")) %>%
    mutate(Pheno_chr = chr, Pheno_start = left, Pheno_end = right,
           dist_to_pheno = case_when(
             var_chr_roman != Pheno_chr ~ NA_real_,
             ChrPos < Pheno_start ~ Pheno_start - ChrPos,
             ChrPos > Pheno_end   ~ ChrPos - Pheno_end,
             TRUE ~ 0),
           cis_trans = case_when(
             var_chr_roman != Pheno_chr ~ "Trans",
             dist_to_pheno == 0    ~ "Cis",
             dist_to_pheno <= 2000 ~ "Cis",
             causal_gene == Pheno  ~ "Cis",
             TRUE                  ~ "Trans"),
           variant_type = vt,
           abs_effect_size = abs(SnpWeight)) %>%
    select(variant_type, SNP, Pheno, var_chr_roman, ChrPos,
           Pheno_chr, Pheno_start, Pheno_end, causal_gene,
           cis_trans, abs_effect_size, PValue)
}
```

# Per-condition runner

For one condition this: (1) rebuilds `gene_location_fixed` from that condition's gene-location
file, (2) reads → annotates → harmonises each of the four variant types, (3) writes the four
per-type annotated tables and the combined table (condition-suffixed), and (4) builds and saves
the 2×2 cis/trans facet (one panel per variant type, Cis/Trans coloured).

```{r runner}
run_condition <- function(cond_label, cond) {
  message("==== ", cond_label, " ====")
  gene_location_fixed <<- build_gene_loc(cond$GENE_LOC)   # pipeline reads this global
  vcfg <- cond$variant_cfg

  # annotate each type once, keep the per-type tables
  per_type <- lapply(names(vcfg), function(vt) {
    cfg <- vcfg[[vt]]
    message("  annotating ", vt, " ...")
    read_clumped(cfg$clumped, cfg$assoc) %>% annotate_variant(vt) %>% add_cis_trans(vt)
  })
  names(per_type) <- names(vcfg)

  combined <- bind_rows(per_type) %>%
    mutate(var_chr_roman = factor(var_chr_roman, levels = roman16),
           Pheno_chr     = factor(Pheno_chr,     levels = roman16),
           variant_type  = factor(variant_type,  levels = c("SNP","CNV","Indel","SV")),
           var_genome    = chr_offsets[as.character(var_chr_roman)] + ChrPos,
           pheno_genome  = chr_offsets[as.character(Pheno_chr)] + (Pheno_start + Pheno_end)/2) %>%
    filter(!is.na(var_genome), !is.na(pheno_genome), !is.na(cis_trans))

  # ---- write per-type annotated tables + combined (condition-suffixed) ----
  out_stub <- c(SNP="snp", CNV="cnv", Indel="indel", SV="sv")
  for (vt in names(per_type)) {
    fwrite(per_type[[vt]],
           file.path(OUT, sprintf("%s_data_annotated_cis_trans_%s.txt", out_stub[[vt]], cond_label)),
           sep = "\t", quote = FALSE, row.names = FALSE, na = "NA")
  }
  fwrite(combined,
         file.path(OUT, sprintf("all_variants_cis_trans_combined_%s.txt", cond_label)),
         sep = "\t", quote = FALSE, row.names = FALSE, na = "NA")

  # ---- 2x2 cis/trans facet ----
  panelA <- ggplot(combined, aes(var_genome, pheno_genome,
                                 color = cis_trans, size = abs_effect_size)) +
    geom_point(alpha = 0.6) +
    facet_wrap(~ variant_type, ncol = 2) +
    scale_color_manual(name = "eQTL type", values = c(Cis = "#d73027", Trans = "#4575b4")) +
    scale_size_continuous(name = "Effect size (|\u03b2|)", range = c(0.4, 2.6), trans = "log10") +
    scale_x_continuous(breaks = chr_mids, labels = names(chr_mids), expand = c(0.01, 0)) +
    scale_y_continuous(breaks = chr_mids, labels = names(chr_mids), expand = c(0.01, 0)) +
    labs(x = "Variant position (chromosome)", y = "Gene position (chromosome)",
         title = cond_label) +
    theme_bw(base_size = 12) +
    theme(strip.text = element_text(face = "bold"),
          axis.text = element_text(size = 8),
          legend.position = "bottom")

  ggsave(file.path(OUT, sprintf("Fig3S_panelA_cis_trans_2x2_%s.tiff", cond_label)), panelA,
         width = 11, height = 10, units = "in", dpi = 600, device = "tiff", compression = "lzw")
  ggsave(file.path(OUT, sprintf("Fig3S_panelA_cis_trans_2x2_%s.png", cond_label)), panelA,
         width = 11, height = 10, units = "in", dpi = 300)

  cat("\n---", cond_label, "cis/trans counts ---\n")
  print(table(combined$variant_type, combined$cis_trans))

  list(combined = combined, panelA = panelA)
}
```

# Run both conditions

```{r run-all}
results <- lapply(names(CONDITIONS), function(cl) run_condition(cl, CONDITIONS[[cl]]))
names(results) <- names(CONDITIONS)
```

# Figures

## Fluconazole

```{r fig-flu, fig.width=11, fig.height=11}
results$Fluconazole$panelA
```

## sc2glu

```{r fig-sc2, fig.width=11, fig.height=11}
results$sc2glu$panelA
```

# Session info

```{r session-info}
sessionInfo()
```
