---
title: "Fig.1 Panel C & D — reproducibility, bulk agreement, and strain concordance"
output: html_document
---

Three ranked-concordance panels (YPD replicates, Flu replicates, scRNA-seq vs bulk)
Jackson et al. 2020 (eLife 51254) Figure 1A/1B.

Scales: 10x pseudobulk matrices are DESeq2 size-factor normalized; the bulk
pan-transcriptome enters the global comparison as TPM (length-corrected) and the
per-gene comparison as raw counts (length cancels there, and DESeq2 needs counts).


#Legend: Each point represents one gene. Expression values are DESeq2 size-factor normalized pseudobulk means (10x Chromium) across shared strains. Axes show gene expression rank. Spearman r: YPD replicates = X.XXX; Fluconazole replicates = X.XXX; sc 2% glucose vs. Caudal (2024) = X.XXX.

Panel C: ranked-concordance (YPD replicates, Flu replicates, scRNA-seq vs bulk).
Panel D: same-strain vs different-strain replicate concordance (YPD main, Flu supplementary).
Jackson et al. 2020 (eLife 51254) Figure 1A/1B.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE,
                      fig.width = 9, fig.height = 3.4, dpi = 300)
```

```{r}
library(tidyverse)
library(DESeq2)
library(patchwork)
```

# Configuration

```{r config}
PATHS <- list(
  YPD1       = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/data/pheno_data/pseudobulk_reliable_transcripts_YPD.txt",
  YPD2       = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.2/fastlmm/data/pheno_data/pseudobulk_reliable_transcripts_YPD.txt",
  FLU1       = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/Flu.1/fastlmm/data/pheno_data/pseudobulk_reliable_transcripts_Flu1.txt",
  FLU2       = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/Flu.2/fastlmm/data/pheno_data/pseudobulk_reliable_transcripts_Flu2.txt",
  SC         = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/data/pheno_data/pseudobulked_sc2glu_before_any_filtering.csv",

  BULK_TPM   = "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/pan_transcriptome/data/969_count_annotated_merged_tpm_before_any_subfiltering.txt",
  BULK_COUNT = "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/pan_transcriptome/data/969_count_annotated_merged_before_any_subfiltering.txt",

  ANNOT      = "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"
)

OUT_DIR <- "../results"
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

# Helpers

```{r helpers}
read_pseudobulk <- function(path, sep = "\t") {
  if (!file.exists(path)) {
    stop("File not found: ", path, "\n  (getwd(): ", getwd(), ")")
  }
  read.csv(path, sep = sep, row.names = 1, header = TRUE, check.names = FALSE)
}

deseq_normalize <- function(mat, label = "") {
  m   <- round(as.matrix(mat))
  dds <- DESeqDataSetFromMatrix(
    countData = m,
    colData   = data.frame(row.names = colnames(m),
                           condition = factor(colnames(m))),
    design    = ~ 1
  )
  dds <- tryCatch(
    estimateSizeFactors(dds),
    error = function(e) {
      message(label, ": standard size factors failed, using type = 'poscounts'")
      estimateSizeFactors(dds, type = "poscounts")
    }
  )
  n_used <- sum(rowSums(m == 0) == 0)
  message(label, ": size factors from ", n_used, " genes with no zeros")
  as.data.frame(counts(dds, normalized = TRUE))
}
```
#Load
```{r}
YPD1_pseudobulk <- read_pseudobulk(PATHS$YPD1)
YPD2_pseudobulk <- read_pseudobulk(PATHS$YPD2)
Flu1_pseudobulk <- read_pseudobulk(PATHS$FLU1)
Flu2_pseudobulk <- read_pseudobulk(PATHS$FLU2)
sc_pseudobulk   <- read_pseudobulk(PATHS$SC, sep = ",")

bulk_tpm   <- read_pseudobulk(PATHS$BULK_TPM)
bulk_count <- read_pseudobulk(PATHS$BULK_COUNT)

for (nm in c("YPD1_pseudobulk", "YPD2_pseudobulk", "Flu1_pseudobulk", "Flu2_pseudobulk",
             "sc_pseudobulk", "bulk_tpm", "bulk_count")) {
  cat(sprintf("%-18s %5d genes x %4d strains\n", nm, nrow(get(nm)), ncol(get(nm))))
}
```




```{r orf_conversion}
all_custom_gene_annotations <- read.csv(PATHS$ANNOT, stringsAsFactors = FALSE)

common_to_orf <- all_custom_gene_annotations %>%
  filter(!is.na(gene_name), !is.na(gene_id)) %>%
  distinct(gene_name, gene_id) %>%
  deframe()

expr_gene_ids <- rownames(sc_pseudobulk)
is_orf        <- grepl("^Y[A-Z]{2}\\d{3}[WC](-[A-Z])?$", expr_gene_ids)

expr_orf_ids <- ifelse(
  is_orf,
  expr_gene_ids,
  ifelse(expr_gene_ids %in% names(common_to_orf), common_to_orf[expr_gene_ids], expr_gene_ids)
)

dup <- duplicated(expr_orf_ids)
if (any(dup)) {
  cat("Dropping", sum(dup), "rows with duplicate ORF IDs after conversion\n")
  sc_pseudobulk <- sc_pseudobulk[!dup, , drop = FALSE]
  expr_orf_ids  <- expr_orf_ids[!dup]
}
rownames(sc_pseudobulk) <- expr_orf_ids

stopifnot(!any(duplicated(rownames(sc_pseudobulk))))
```

# Normalize

```{r normalize}
YPD1_norm <- deseq_normalize(YPD1_pseudobulk, "YPD.1")
YPD2_norm <- deseq_normalize(YPD2_pseudobulk, "YPD.2")
Flu1_norm <- deseq_normalize(Flu1_pseudobulk, "Flu.1")
Flu2_norm <- deseq_normalize(Flu2_pseudobulk, "Flu.2")
sc_norm   <- deseq_normalize(sc_pseudobulk,   "sc SC-2%glu")
# bulk_tpm is already per-strain normalized and length-corrected; leave it alone.

```

# Global ranked expression concordance

Subset to shared genes and strains, `rowMeans` per gene, Spearman on the means,
`rank()` for the plot axes.

```{r rank_concordance_fn}
rank_concordance <- function(mat_x, mat_y, label, xlab, ylab) {

  shared_genes   <- intersect(rownames(mat_x), rownames(mat_y))
  shared_strains <- intersect(colnames(mat_x), colnames(mat_y))

  cat("\n---", label, "---\n")
  cat("Shared strains:", length(shared_strains), "\n")
  cat("Shared genes:  ", length(shared_genes),   "\n")

  x_mean <- rowMeans(mat_x[shared_genes, shared_strains, drop = FALSE])
  y_mean <- rowMeans(mat_y[shared_genes, shared_strains, drop = FALSE])

  spearman_r <- cor(x_mean, y_mean, method = "spearman")
  cat("Spearman r:", round(spearman_r, 3), "\n")

  list(
    df    = data.frame(comparison = label, gene = shared_genes,
                       x = rank(x_mean), y = rank(y_mean)),
    stats = data.frame(comparison = label, xlab = xlab, ylab = ylab,
                       n_genes    = length(shared_genes),
                       n_strains  = length(shared_strains),
                       spearman_r = spearman_r)
  )
}
```

```{r run_rank_concordance}
res <- list(
  rank_concordance(YPD1_norm, YPD2_norm, "YPD",
                   "YPD.1", "YPD.2"),
  rank_concordance(Flu1_norm, Flu2_norm, "Fluconazole",
                   "Fluconazole.1", "Fluconazole.2"),
  rank_concordance(bulk_tpm, sc_norm, "Synthetic complete vs. Caudal (2024)",
                   "Caudal (2024)", "Synthetic complete")
)

lvl        <- c("YPD", "Fluconazole", "Synthetic complete vs. Caudal (2024)")
rank_df    <- bind_rows(lapply(res, `[[`, "df"))    %>% mutate(comparison = factor(comparison, levels = lvl))
rank_stats <- bind_rows(lapply(res, `[[`, "stats")) %>% mutate(comparison = factor(comparison, levels = lvl))
```

# Figure 1C — ranked concordance panels
```{r theme}
pub_theme <- theme_classic(base_size = 9, base_family = "Arial") +
  theme(
    axis.title = element_text(size = 12,  face = "bold", color = "black"),
    axis.text  = element_text(size = 11,  color = "black"),
    axis.line  = element_line(linewidth = 0.4, color = "black"),
    axis.ticks = element_line(linewidth = 0.35, color = "black"),
    panel.grid       = element_blank(),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

```{r fig1c}
# ── panel colours ────────────────────────────────────────────────────────────
panel_colors <- c(
  "YPD"                                  = "#4E79A7",
  "Fluconazole"                          = "#E15759",
  "Synthetic complete vs. Caudal (2024)" = "#59A14F"
)

global_range <- range(c(rank_df$x, rank_df$y), na.rm = TRUE)

make_panel <- function(cmp) {
  d <- filter(rank_df,    comparison == cmp)
  s <- filter(rank_stats, comparison == cmp)
  ggplot(d, aes(x = x, y = y, color = comparison)) +
    geom_point(alpha = 0.1, size = 0.3, shape = 16) +
    scale_color_manual(values = panel_colors, guide = "none") +
    annotate("text", x = Inf, y = -Inf,
             label = paste0("Spearman r = ", round(s$spearman_r, 3)),
             hjust = 1.1, vjust = -0.8, size = 4, fontface = "bold") +
    scale_x_continuous(limits = global_range, breaks = seq(0, global_range[2], by = 1000),
                       expand = expansion(mult = 0.02)) +
    scale_y_continuous(limits = global_range, breaks = seq(0, global_range[2], by = 1000),
                       expand = expansion(mult = 0.02)) +
    coord_fixed() +
    labs(x = s$xlab, y = s$ylab) +
    pub_theme
}

fig1c <- wrap_plots(lapply(lvl, make_panel), nrow = 1)
fig1c


```
Figure 1C. Gene expression rankings are highly concordant across independent replicates and datasets.

Scatter plots showing ranked gene expression concordance. For each condition, genes are ranked by their mean expression level across strains, and this ranking is compared between two datasets. Panel (i) shows YPD replicates with high concordance (Spearman r = 0.953), demonstrating reproducible gene expression rankings in standard growth conditions. Panel (ii) shows fluconazole-treated replicates (Spearman r = 0.908), indicating strong reproducibility under stress. Panel (iii) compares sc 2% glucose (single cell study) against published Caudal (2024, bulk RNAseq) data (Spearman r = 0.838), showing good concordance across experimental approaches.
```{r save_fig1c}
ggsave(file.path(OUT_DIR, "Fig1C_rank_concordance.pdf"), fig1c,
       width = 9, height = 3.4, units = "in")
ggsave(file.path(OUT_DIR, "Fig1C_rank_concordance.tiff"), fig1c,
       width = 9, height = 3.4, units = "in", dpi = 300,
       device = "tiff", compression = "lzw")s

write.csv(rank_stats %>% select(-xlab, -ylab),
          file.path(OUT_DIR, "Fig1C_summary_stats.csv"), row.names = FALSE)
```



# Panel D — same-strain vs different-strain replicate concordance

"How similar is this strain's overall gene expression profile between two independent experiments?" 

Strain A's gene expression profile is more similar to strain A in replicate 2 then to other strains

# ============================================================
# Figure 1D / Supplementary — same-strain vs different-strain
# replicate concordance
#
# H0: same-strain r = different-strain r (strain identity is not
#     reproducible across independent experiments)
# H1: same-strain r > different-strain r (a strain's expression
#     profile is reproducible, i.e. driven by strain identity
#     rather than shared technical noise)
# ============================================================

Is a strain's transcriptome reproducible across replicates?


Figure 1D - MOVE IT TO SUPPLEMENTARY. Gene expression profiles are reproducible within strains across independent replicates.

Density distributions of Spearman correlations (log₁p-transformed, replicate 1 vs 2) for the YPD condition. Each correlation represents genome-wide expression similarity between two replicate experiments. Same-strain comparisons (blue, n = 96 strains) measure expression reproducibility within individual strains across replicates and show a median r of 0.932. Cross-strain comparisons (red, n = 9,120 strain pairs) measure expression similarity between different strains across replicates and serve as a null expectation, with median r of 0.894. The marked separation between distributions demonstrates that a strain's transcriptional profile is highly reproducible and driven by intrinsic strain-specific factors rather than technical noise or shared environmental responses. Wilcoxon rank-sum test (one-sided, H₁: same-strain r > cross-strain r): p = 7.1 × 10⁻¹⁷.

```{r}
strain_concordance <- function(mat1, mat2, label) {
  shared_genes   <- intersect(rownames(mat1), rownames(mat2))
  shared_strains <- intersect(colnames(mat1), colnames(mat2))
  cat("\n---", label, "---\n")
  cat("Shared strains:", length(shared_strains), "\n")
  cat("Shared genes:  ", length(shared_genes),   "\n")
  mat1_sub <- mat1[shared_genes, shared_strains]
  mat2_sub <- mat2[shared_genes, shared_strains]
  # ── Same-strain r ─────────────────────────────────────────────
  same_strain_r <- sapply(shared_strains, function(s) {
    cor(log1p(as.numeric(mat1_sub[, s])),
        log1p(as.numeric(mat2_sub[, s])),
        method = "spearman", use = "complete.obs")
  })
  # ── Cross-strain r (null) ──────────────────────────────────────
  all_diff_r <- c()
  for (i in 1:length(shared_strains)) {
    for (j in 1:length(shared_strains)) {
      if (shared_strains[i] != shared_strains[j]) {
        r <- cor(log1p(as.numeric(mat1_sub[, shared_strains[i]])),
                 log1p(as.numeric(mat2_sub[, shared_strains[j]])),
                 method = "spearman", use = "complete.obs")
        all_diff_r <- c(all_diff_r, r)
      }
    }
  }
  cat("Same-strain pairs:", length(same_strain_r), "\n")
  cat("Cross-strain pairs:", length(all_diff_r), "\n")
  wt <- wilcox.test(same_strain_r, all_diff_r, alternative = "greater")
  cat("Wilcoxon p-value:", format(wt$p.value, scientific = TRUE), "\n")
  df <- bind_rows(
    data.frame(r = same_strain_r, group = "Same strain"),
    data.frame(r = all_diff_r,    group = "Cross-strain")
  ) %>%
    mutate(group = factor(group, levels = c("Cross-strain", "Same strain")))
  stats <- data.frame(
    comparison    = label,
    n_genes       = length(shared_genes),
    n_strains     = length(shared_strains),
    n_same_pairs  = length(same_strain_r),
    n_diff_pairs  = length(all_diff_r),
    median_same_r = median(same_strain_r),
    median_diff_r = median(all_diff_r),
    wilcox_p      = wt$p.value
  )
  list(df = df, stats = stats)
}
```

```{r}
library(ggrepel)

concordance_colors <- c(
  "Same strain" = "#2166AC",
  "Cross-strain" = "#D6604D"
)

make_concordance_panel <- function(res, condition_label) {
  medians <- res$df %>%
    group_by(group) %>%
    summarise(med = median(r), .groups = "drop")
  
  p_label <- format(res$stats$wilcox_p, digits = 2, scientific = TRUE)
  
  ggplot(res$df, aes(x = r, fill = group, color = group)) +
    geom_density(alpha = 0.5, linewidth = 0.5, color = NA) +
    geom_vline(data = medians, aes(xintercept = med, color = group),
               linetype = "dashed", linewidth = 0.6, show.legend = FALSE) +
    ggrepel::geom_text_repel(
      data = medians, 
      aes(x = med, y = Inf, label = round(med, 3), color = group),
      size = 4, fontface = "bold", show.legend = FALSE,
      direction = "y",
      nudge_y = -0.5,
      segment.size = 0.3,
      force = 2
    ) +
    scale_fill_manual(values  = concordance_colors) +
    scale_color_manual(values = concordance_colors) +
    annotate("text", x = -Inf, y = Inf,
             label = paste0("Wilcoxon p = ", p_label),
             hjust = -0.1, vjust = 1.4, size = 4, fontface = "italic") +
    labs(
      x = "Spearman r (log1p, replicate 1 vs replicate 2)",
      y = "Density", fill = NULL, color = NULL
    ) +
    pub_theme +
    theme(
      legend.position = "bottom",
      legend.text     = element_text(size = 15)
    )
}
```

```{r}
res_YPD <- strain_concordance(YPD1_norm, YPD2_norm, "YPD")
res_Flu <- strain_concordance(Flu1_norm, Flu2_norm, "Fluconazole")

fig1d_YPD <- make_concordance_panel(res_YPD, "YPD")
figS_Flu  <- make_concordance_panel(res_Flu, "Fluconazole")

fig1d_YPD
figS_Flu

knitr::kable(
  bind_rows(res_YPD$stats, res_Flu$stats),
  digits = 3,
  caption = "Same-strain vs cross-strain replicate concordance"
)
```

```{r}
ggsave(file.path(OUT_DIR, "Fig1D_YPD_strain_concordance.pdf"), fig1d_YPD,
       width = 4.5, height = 3.5, units = "in")

ggsave(file.path(OUT_DIR, "Fig1D_YPD_strain_concordance.tiff"), fig1d_YPD,
       width = 4.5, height = 3.5, units = "in", dpi = 300)

ggsave(file.path(OUT_DIR, "FigS_Flu_strain_concordance.pdf"), figS_Flu,
       width = 4.5, height = 3.5, units = "in")
ggsave(file.path(OUT_DIR, "FigS_Flu_strain_concordance.tiff"), figS_Flu,
       width = 4.5, height = 3.5, units = "in", dpi = 300,
       device = "tiff", compression = "lzw")

write.csv(bind_rows(res_YPD$stats, res_Flu$stats),
          file.path(OUT_DIR, "Fig1D_strain_concordance_summary_stats.csv"),
          row.names = FALSE)
```

