---
title: "Cell-cycle expression across natural yeast isolates — harmonic model"
author: "Akriti"
output: html_document
---

# Overview

Cell-cycle genes are **periodic with a known period** (88 min). This version
replaces the LOESS smoother with a harmonic (Fourier) regression, which encodes
that periodicity directly instead of rediscovering it with ~13 free parameters
per strain.

What this buys:

| Problem in the LOESS version | Fix here |
|---|---|
| R² inflated in small strains (`cor(R², n_cells) = -0.63`) | 5–6 parameters instead of ~13, plus df-adjusted R² → measured at **+0.09** |
| Amplitude inflated in small strains | Analytic noise-variance subtraction (`amp1_db`) |
| Seam artifact at t = 0/88 | Harmonic basis is periodic by construction — no seam, no padding |
| `peak`/`amp` read off a per-strain grid `min(t)..max(t)` | Fixed 0–88 grid for every strain |
| Peak time needs circular statistics as a patch | Phase is an angle in the model itself |
| Fit used for stats ≠ fit used for plots | One fit, used for both |
| Depth differences invisible to the model | Optional per-cell log-depth covariate |

**Three questions, kept separate:**

- **Q1 (pooled, no strain split).** Does this gene cycle at all? One fit on all
  cells, judged on **effect size** (`r2_adj`), not on a p-value — see §4 for why
  the F-test is useless at n = 16,000.
- **Q2 (per strain).** Among genes that cycle, which are *consistent* across
  strains and which *vary*? Three phenotypes off one per-strain fit: `r2_adj`
  (structure), `amp1_db` (amplitude), `peak` (timing).
- **Q3 (replicates, §11).** Does any of it reproduce in an independent capture?
  This is the only thing that separates biology from technical artifact.

**On R².** R² is fine as a descriptive measure of how tightly cells track the
curve — it just was not comparable across strains of unequal size under LOESS.
The df adjustment restores that (§5). It is still not the whole story, which is
why amplitude and phase are carried alongside: a gene can have identical R² in
two strains while cresting 15 minutes apart.

**No detection filter.** `pct_strains_detected` is reported as a phenotype
rather than used as a cutoff.

---

# 1. Setup

1. GRID <- seq(0, WRAP, length.out = 89)
This creates a fixed sequence of 89 evenly spaced time points from 0 to 88 minutes: 0, 1, 2, 3, ... 88. Every strain uses this same grid when the fitted curve is evaluated.

2. 

```{r libraries, message=FALSE}
library(Seurat)
library(tidyverse)
library(RColorBrewer)
library(AnnotationDbi)
library(org.Sc.sgd.db)
library(ggrepel)

select <- dplyr::select
filter <- dplyr::filter

options(future.globals.maxSize = 9 * 1024^3)

# ── Analysis parameters ──────────────────────────────────────────────────────
WRAP          <- 88     # cell cycle period (min)
HARMONICS     <- 2      # 1 = pure sinusoid; 2 allows asymmetric peaks
INCLUDE_DEPTH <- TRUE   # centred log(nCount) as a within-strain covariate
GRID          <- seq(0, WRAP, length.out = 89)   # FIXED grid for every strain
MIN_CELLS     <- 25     # per-strain minimum
N_PERM        <- 500    # permutations for the null
SEED          <- 1

# Amplitude statistic used for all Q2 phenotypes.
# "amp1_db" = bias-corrected first-harmonic amplitude (recommended)
# "amp"     = peak-to-trough of the K-harmonic grid curve (size-biased)
AMP_STAT <- "amp1_db"

CC_LEVELS  <- c('M-G1', 'G1', 'S', 'G2', 'M')
PHASE_FILL <- c('M-G1'='#a8d5a2','G1'='#ffd6a5','S'='#ffadad',
                'G2'='#bde0fe','M'='#ffc8dd')

ROOT <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1"
```

## 1.1 Load counts and strain assignments

```{r load_data}
expression_data <- Read10X(
  data.dir = file.path(ROOT, "alignment/output_of_alignment_YPD1_with_clean_gtf",
                       "run_count/outs/filtered_feature_bc_matrix/")
)

Na_va_data <- CreateSeuratObject(
  counts = expression_data, project = "natural variants",
  min.cells = 3, min.features = 200
)
cat("Cells:", ncol(Na_va_data), "| Genes:", nrow(Na_va_data), "\n")

demuxlet <- read.delim(file.path(ROOT, "demuxlet/merged_demuxlet_output_YPD.tsv")) %>%
  mutate(across(c(BEST.LLK, NUM.READS, NUM.SNPS, NEXT.LLK, DIFF.LLK.BEST.NEXT,
                  BEST.POSTERIOR, SNG.POSTERIOR, SNG.BEST.LLK, SNG.NEXT.LLK,
                  SNG.ONLY.POSTERIOR, DBL.BEST.LLK, DIFF.LLK.SNG.DBL),
                as.numeric)) %>%
  na.omit()

Na_va_data <- AddMetaData(
  Na_va_data, metadata = demuxlet[match(colnames(Na_va_data), demuxlet$BARCODE), ]
)
cat("Strains:", n_distinct(Na_va_data@meta.data$SNG.BEST.GUESS), "\n")
```

## 1.2 Strain QC

The `too_few_timepoints` category in the old code fired on the *same* strains for
every gene — pseudotime is a per-cell property, so it was a strain filter wearing
a per-gene costume. It is applied once, here.

```{r strain_qc}
Na_va <- subset(
  Na_va_data,
  subset = SNG.BEST.GUESS != "DGY3" & !(DROPLET.TYPE %in% c("AMB", "DBL"))
)

keep_strains <- Na_va@meta.data %>%
  dplyr::count(strain = SNG.BEST.GUESS, name = "n") %>%
  dplyr::filter(n >=MIN_CELLS) %>% pull(strain)

Na_va <- subset(Na_va, subset = SNG.BEST.GUESS %in% keep_strains)

cat("Strains kept:", n_distinct(Na_va@meta.data$SNG.BEST.GUESS),
    "| cells:", ncol(Na_va), "\n")
print(summary(as.numeric(table(Na_va@meta.data$SNG.BEST.GUESS))))
```

## 1.3 Pseudotime

```{r add_pseudotime}
cc_time <- read.csv(
  file.path(ROOT, "continuous_time_trajectory/data/cc_time_all_genes_spellman_annotated.csv"),
  row.names = 1
)
cc_time$CC <- factor(cc_time$CC, levels = CC_LEVELS)

shared <- intersect(rownames(cc_time), colnames(Na_va))
cat("Cells with cc_time:", nrow(cc_time), "| shared:", length(shared), "\n")

Na_va <- AddMetaData(Na_va, cc_time[shared, c('CC', 'cc_time')])
Na_va@meta.data$CC <- factor(Na_va@meta.data$CC, levels = CC_LEVELS)

Na_va <- subset(Na_va, cells = colnames(Na_va)[!is.na(Na_va@meta.data$cc_time)])
cat("Cells with pseudotime:", ncol(Na_va), "\n")


# Check average time per phase for each program
ggplot(cc_time, aes(x = cc_time)) +
  geom_histogram(bins = 88, fill = "steelblue") +
  facet_wrap(~CC, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 88, 11)) +
  theme_classic()

p_cc_time_by_phase <- ggplot(cc_time, aes(x = cc_time)) +
  geom_histogram(bins = 88, fill = "steelblue") +
  facet_wrap(~CC, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 88, 11)) +
  labs(x = "Cell-cycle time (min)", y = "Cell count",
       title = "Assigned pseudotime distribution by phase") +
  theme_classic()

p_cc_time_by_phase

dir.create("../results", showWarnings = FALSE)
ggsave("../results/cc_time_by_phase.png", p_cc_time_by_phase,
       width = 6, height = 10, dpi = 300)
```

## 1.3b Pseudotime coverage per strain

A harmonic fit extrapolates smoothly across gaps, so a strain whose cells all
land in one arc will still report a confident peak somewhere it has no data.
Cell count alone does not catch this.

#basically strains with really large gap in its time extrapolation. i.e., if a strain has all its value of time in G1 phase and G2 phase but none in S phase, then it has a large gap. in this case code will give a peak in the unobserved region i.e., could be anywhere but it can't be trusted so peak time can't be trusted because its a mathematical extrapolation/ an artifact. But can trust amplitude and r2

22 minutes is one quarter of the 88-minute cycle.A gap smaller than a quarter cycle means the harmonic fit is mostly reasonable

```{r coverage_check}
circ_max_gap <- function(t, period = WRAP) {
  s <- sort(t %% period)
  max(c(diff(s), period - max(s) + min(s)))
}

gap_by_strain <- tapply(Na_va@meta.data$cc_time,
                        Na_va@meta.data$SNG.BEST.GUESS, circ_max_gap)
gap_by_strain <- gap_by_strain[!is.na(gap_by_strain)]

cat("Largest empty arc per strain (min):\n"); print(summary(gap_by_strain))
cat("Strains with a gap > 22 min (quarter cycle):",
    sum(gap_by_strain > 22), "\n")
print(sort(gap_by_strain, decreasing = TRUE)[1:10])

# flag, do not drop — used later when interpreting `peak`
POOR_COVERAGE <- names(gap_by_strain)[gap_by_strain > 22]
```

## 1.4 Normalize (PCA/UMAP are for display only)

```{r normalize}
Na_va <- Na_va %>%
  NormalizeData() %>%
  FindVariableFeatures(selection.method = "vst", nfeatures = 2000)
Na_va <- ScaleData(Na_va, features = VariableFeatures(Na_va))
Na_va <- Na_va %>% RunPCA(features = VariableFeatures(.)) %>% RunUMAP(dims = 1:5)
```

```{r umap_plots}
umap_df <- as.data.frame(Embeddings(Na_va, "umap"))
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$CC      <- Na_va@meta.data$CC
umap_df$cc_time <- Na_va@meta.data$cc_time

centroids <- umap_df %>% filter(!is.na(CC)) %>%
  group_by(CC) %>%
  summarise(umap_1 = median(umap_1), umap_2 = median(umap_2), .groups = "drop")

# pseudotime, with phase labels on top
p_umap_pseudotime <- ggplot(umap_df, aes(umap_1, umap_2)) +
  geom_point(aes(colour = cc_time), size = 0.3, alpha = 0.6) +
  scale_colour_viridis_c(option = "magma", direction = -1, name = "cc_time (min)") +
  geom_label_repel(data = centroids,
                   mapping = aes(x = umap_1, y = umap_2, label = CC),
                   inherit.aes = FALSE, size = 4.5, alpha = 0.85,
                   label.size = 0.2, box.padding = 0.5) +
  labs(title = "Cell-cycle time with phase labels") +
  theme_classic()

p_umap_pseudotime

dir.create("../results", showWarnings = FALSE)
ggsave("../results/umap_pseudotime_phase_labels.png", p_umap_pseudotime,
       width = 7, height = 6, dpi = 300)

# discrete phase, same labels
ggplot(umap_df %>% filter(!is.na(CC)), aes(umap_1, umap_2, colour = CC)) +
  geom_point(size = 0.3, alpha = 0.6) +
  scale_colour_manual(values = PHASE_FILL, name = "Phase") +
  geom_label_repel(data = centroids,
                   mapping = aes(x = umap_1, y = umap_2, label = CC),
                   inherit.aes = FALSE, size = 4.5, alpha = 0.85) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  labs(title = "Cell-cycle phase") +
  theme_classic()
```

## 1.5 ORF ↔ common name lookup

```{r orf_lookup}
CUSTOM_ANNOT_CSV <- "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"

build_orf_lookup <- function(seurat_obj, custom_csv = CUSTOM_ANNOT_CSV) {
  custom <- read.csv(custom_csv) %>%
    filter(!is.na(gene_id), !is.na(gene_name)) %>%
    distinct(gene_id, gene_name) %>% deframe()
  missing <- rownames(seurat_obj)[!rownames(seurat_obj) %in% names(custom)]
  sgd <- tryCatch(
    AnnotationDbi::select(org.Sc.sgd.db, keys = missing,
                          columns = "GENENAME", keytype = "ORF"),
    error = function(e) data.frame(ORF = character(), GENENAME = character())
  ) %>% filter(!is.na(GENENAME)) %>% distinct(ORF, GENENAME) %>% deframe()
  c(custom, sgd[!names(sgd) %in% names(custom)])
}

orf_lookup <- build_orf_lookup(Na_va)
to_name <- function(x, lk = orf_lookup) ifelse(x %in% names(lk), unname(lk[x]), x)
to_orf  <- function(x, lk = orf_lookup) {
  hit <- names(lk)[match(x, unname(lk))]; ifelse(is.na(hit), x, hit)
}
cat("YBR009C ->", to_name("YBR009C"), "| HHF1 ->", to_orf("HHF1"), "\n")
```

---

# 2. The harmonic engine

## 2.1 What the model is

For a cell at pseudotime `t`, with `K` harmonics:

```
expr ~ b0 + b1·cos(2πt/88) + b2·sin(2πt/88) + b3·cos(4πt/88) + b4·sin(4πt/88)
```

Everything is read off the coefficients:

- **`amp`** — peak-to-trough of the fitted curve on the fixed 0–88 grid.
  Descriptive and shape-aware, but a *range* is upward-biased by estimation
  noise, so small strains report larger amplitudes.
- **`amp1`** — first-harmonic amplitude, `2·sqrt(b1² + b2²)`. Same bias.
- **`amp1_db`** — the bias-corrected version. Each coefficient has sampling
  variance ≈ σ²·(XᵀX)⁻¹, so `E[b1² + b2²] = A²/4 + var(b1) + var(b2)`.
  Subtracting the noise term removes the 1/√n inflation analytically. **This is
  the amplitude phenotype used downstream.**
- **`peak`** — grid time where the curve crests. Circular by construction; no
  0/88 seam.
- **`r2_adj`** — df-adjusted, unbiased at any n.
- **`f_p`** — F-test of the harmonic terms. Reported, but see §4: at n = 16,000
  it is not a useful filter.

## 2.2 Design matrix - this gives us a matrix

```{r design}
harmonic_design <- function(t, K = HARMONICS, period = WRAP, depth = NULL) {
  X <- matrix(1, nrow = length(t), ncol = 1, dimnames = list(NULL, "intercept")) #this is intercept
  for (k in seq_len(K)) {
    X <- cbind(X, cos(2 * pi * k * t / period), sin(2 * pi * k * t / period))
    colnames(X)[(ncol(X) - 1):ncol(X)] <- paste0(c("cos", "sin"), k) #The term 2πt/88 is converting time into an angle in radians
  }
  if (!is.null(depth)) { X <- cbind(X, depth = depth) }   # centred, so grid uses 0
  X
}
```
It builds a table of predictor columns. Given t (the pseudotime of each cell), it returns a matrix X with one row per cell and one column per predictor:
#X will have 6 columns

intercept | cos1 | sin1 | cos2 | sin2 | depth
    1     | ...  | ...  | ...  | ...  | -0.3   ← shallowly sequenced cell
    1     | ...  | ...  | ...  | ...  |  0.1   ← average cell
    1     | ...  | ...  | ...  | ...  |  0.8   ← deeply sequenced cell

## 2.3 Block fit — all genes at once

The design matrix depends only on the *cells*, not the gene. So for one strain
we solve for every gene in a single QR decomposition. This is what makes 500
permutations feasible.

```{r block_fit}
fit_block <- function(Y, t, depth = NULL, K = HARMONICS, grid = GRID) {
  # Y: cells x genes (dense). Returns per-gene vectors.
  n <- nrow(Y) #n = number of cells in this strain
  X <- harmonic_design(t, K, WRAP, depth)
  p <- ncol(X) #number of columns in X 
  if (n < p + 3 || length(unique(t)) < p + 1) return(NULL) #The checks make sure you have enough cells and enough distinct pseudotime values to actually fit the model.

  qrX <- qr(X)
  if (qrX$rank < p) return(NULL)
  B <- qr.coef(qrX, Y); B[is.na(B)] <- 0     # p x genes

  resid <- Y - X %*% B
  RSS <- colSums(resid^2)
  TSS <- colSums(sweep(Y, 2, colMeans(Y), "-")^2)

  r2     <- ifelse(TSS > 0, 1 - RSS / TSS, NA_real_)
  r2_adj <- ifelse(TSS > 0, 1 - (RSS / (n - p)) / (TSS / (n - 1)), NA_real_)

  n_harm <- 2 * K
  Fstat  <- ((TSS - RSS) / n_harm) / (RSS / (n - p))
  f_p    <- ifelse(TSS > 0 & RSS > 0,
                   pf(Fstat, n_harm, n - p, lower.tail = FALSE), NA_real_)

  # curve on the FIXED grid, depth held at its centre (0)
  Gd <- harmonic_design(grid, K, WRAP,
                        if (is.null(depth)) NULL else rep(0, length(grid)))
  P  <- Gd %*% B                             # grid x genes
  amp  <- apply(P, 2, max) - apply(P, 2, min)
  peak <- grid[max.col(t(P), ties.method = "first")]

  amp1   <- 2 * sqrt(B["cos1", ]^2 + B["sin1", ]^2)
  phase1 <- (atan2(B["sin1", ], B["cos1", ]) %% (2 * pi)) / (2 * pi) * WRAP

  # ── bias-corrected first-harmonic amplitude ──────────────────────────────
  # E[b1^2 + b2^2] = A_true^2/4 + var(b1) + var(b2); subtract the noise part
  XtXinv <- chol2inv(qr.R(qrX))
  dimnames(XtXinv) <- list(colnames(X), colnames(X))
  sigma2  <- RSS / (n - p)
  v_noise <- XtXinv["cos1", "cos1"] + XtXinv["sin1", "sin1"]
  amp1_db <- 2 * sqrt(pmax(0, B["cos1", ]^2 + B["sin1", ]^2 - sigma2 * v_noise))

  bad <- is.na(TSS) | TSS == 0
  amp[bad] <- NA; peak[bad] <- NA; amp1[bad] <- NA
  phase1[bad] <- NA; amp1_db[bad] <- NA

  list(r2 = r2, r2_adj = r2_adj, f_p = f_p, amp = amp, peak = peak,
       amp1 = amp1, amp1_db = amp1_db, phase1 = phase1,
       constant = bad, n = n, p = p)
}
```

## 2.4 Fit every strain

```{r fit_all_strains}
FIT_STATS <- c("r2","r2_adj","amp","peak","amp1","amp1_db","phase1","f_p")

fit_by_group <- function(Ymat, tvec, group, depth = NULL, K = HARMONICS) {
  strains <- sort(unique(group))
  genes   <- colnames(Ymat)
  blank   <- matrix(NA_real_, length(genes), length(strains),
                    dimnames = list(genes, strains))
  out <- setNames(replicate(length(FIT_STATS), blank, simplify = FALSE), FIT_STATS)
  out$constant <- blank
  ncell <- setNames(integer(length(strains)), strains)

  for (j in seq_along(strains)) {
    idx <- which(group == strains[j])
    ncell[j] <- length(idx)
    fit <- fit_block(Ymat[idx, , drop = FALSE], tvec[idx],
                     if (is.null(depth)) NULL else scale(depth[idx], scale = FALSE)[, 1],
                     K = K)
    if (is.null(fit)) next
    for (nm in FIT_STATS) out[[nm]][, j] <- fit[[nm]]
    out$constant[, j] <- as.numeric(fit$constant)
  }
  out$n_cells <- ncell
  out
}
```

## 2.5 Circular summaries

Peak time lives on a circle of circumference 88, so minute 2 and minute 86 are
4 minutes apart. Each peak becomes a unit vector `(cos θ, sin θ)`; the direction
of the mean vector is the circular mean, its length `R ∈ [0,1]` the concentration.

`circ_sd = sqrt(-2·log R)` is unbounded — a huge value means "no concentration at
all", not "peaks are 60 minutes apart".

Its direction is the average peak time.
Its length measures how well the strains agree.

```{r circular}
circ_mean_rows <- function(M, period = WRAP) {
  th <- 2 * pi * M / period
  (atan2(rowMeans(sin(th), na.rm = TRUE),
         rowMeans(cos(th), na.rm = TRUE)) %% (2 * pi)) / (2 * pi) * period
}
circ_sd_rows <- function(M, period = WRAP) {
  th <- 2 * pi * M / period
  R  <- sqrt(rowMeans(cos(th), na.rm = TRUE)^2 + rowMeans(sin(th), na.rm = TRUE)^2)
  R  <- pmin(pmax(R, 1e-8), 1)
  sqrt(-2 * log(R)) / (2 * pi) * period
}
row_sd   <- function(M) apply(M, 1, sd,   na.rm = TRUE)
row_mean <- function(M) rowMeans(M, na.rm = TRUE)
row_n    <- function(M) rowSums(!is.na(M))
```

---

# 3. Build the matrices

```{r build_matrices}
norm_counts <- GetAssayData(Na_va, layer = "data")      # log-normalized

meta <- Na_va@meta.data %>%
  rownames_to_column("cell") %>%
  transmute(cell, strain = SNG.BEST.GUESS, pseudotime = cc_time,
            CC, nCount = nCount_RNA) %>%
  filter(!is.na(strain), !is.na(pseudotime))

spellman <- read.delim(file.path(ROOT, "cell_cycle_fastlmm/data/CellCycle_Genes.tsv")) %>%
  distinct(Gene) %>% pull(Gene)
spellman_in <- intersect(spellman, rownames(norm_counts))
cat("Spellman genes present:", length(spellman_in), "of", length(spellman), "\n")

cells   <- meta$cell
tvec    <- meta$pseudotime
gvec    <- meta$strain
depthv  <- log1p(meta$nCount)

Y_all <- t(as.matrix(norm_counts[, cells]))              # cells x all genes
Y_cc  <- Y_all[, spellman_in, drop = FALSE]              # cells x Spellman genes

cat("Matrix:", nrow(Y_cc), "cells x", ncol(Y_cc), "genes |",
    n_distinct(gvec), "strains\n")
cat("Pseudotime range:", round(range(tvec), 2), "\n")
```

---

# 4. Q1 — which genes cycle? (pooled, no strain split) does this gene cycle at all?

**The F-test is not usable here.** With ~16,000 cells and 4 harmonic terms, the
R² needed to clear p < 0.05 is about 0.0006 — a gene explaining 0.06% of its
variance "passes". That is why 93% of the transcriptome comes back significant.
Rank and threshold on **effect size** (`r2_adj`) instead.

Y_all is every gene = 5745 genes. One fit on all ~16000 cells

```{r q1}
pooled <- fit_block(Y_all, tvec,
                    if (INCLUDE_DEPTH) scale(depthv, scale = FALSE)[, 1] else NULL) #all cells and all pseudotime

q1 <- tibble(
  gene        = colnames(Y_all),
  name        = to_name(colnames(Y_all)),
  r2_pooled   = pooled$r2,
  r2_adj      = pooled$r2_adj,
  amp         = pooled$amp,
  amp1_db     = pooled$amp1_db,
  peak        = pooled$peak,
  p_periodic  = pooled$f_p,
  pct_nonzero = colMeans(Y_all > 0),
  is_spellman = colnames(Y_all) %in% spellman_in
) %>%
  mutate(fdr_periodic = p.adjust(p_periodic, method = "BH")) %>%
  arrange(desc(r2_adj))

n_pooled <- length(tvec)
p_pooled <- 1 + 2*HARMONICS + as.integer(INCLUDE_DEPTH)
r2_crit  <- (2*HARMONICS) * qf(0.95, 2*HARMONICS, n_pooled - p_pooled) /
            (n_pooled - p_pooled)

cat("Cells in pooled fit:", n_pooled, "\n")
cat("R2 needed for p<0.05 at this n:", signif(r2_crit, 4),
    " <- why", sum(q1$fdr_periodic < 0.05, na.rm = TRUE), "genes 'pass'\n\n")

cat("Distribution of pooled adjusted R2:\n")
print(round(quantile(q1$r2_adj, c(.5,.75,.9,.95,.99,1), na.rm = TRUE), 3))

cat("\nGenes above an effect-size threshold:\n")
for (th in c(0.02, 0.05, 0.10, 0.20, 0.30))
  cat(sprintf("  r2_adj > %.2f : %4d total | %3d Spellman\n", th,
              sum(q1$r2_adj > th, na.rm = TRUE),
              sum(q1$r2_adj > th & q1$is_spellman, na.rm = TRUE))) #100% means the maximum observed R² = 0.715

# main sanity check: is the Spellman set enriched?
print(q1 %>% group_by(is_spellman) %>%
        summarise(n = n(),
                  median_r2adj  = round(median(r2_adj, na.rm = TRUE), 4),
                  pct_above_0.1 = round(mean(r2_adj > 0.1, na.rm = TRUE), 3)))

R2_CYCLING <- 0.1        # adjust after reading the numbers above
cycling <- q1 %>% filter(r2_adj > R2_CYCLING)
cat("\nCalling", nrow(cycling), "genes cycling at r2_adj >", R2_CYCLING, "\n")

q1 %>% select(name, gene, r2_adj, amp1_db, peak, is_spellman) %>%
  head(25) %>% print(n = 25)


q1 %>% filter(r2_adj > R2_CYCLING, !is_spellman) %>%
  select(name, gene, r2_adj, amp1_db, peak, pct_nonzero) %>%
  head(30) %>% print(n = 30)
```

```{r q1_peak_distribution}
q1 %>% filter(r2_adj > R2_CYCLING) %>%
  ggplot(aes(peak)) +
  geom_histogram(binwidth = 4, fill = "grey30") +
  scale_x_continuous(breaks = seq(0, WRAP, 11), limits = c(0, WRAP)) +
  labs(x = "Peak time (min)", y = "Genes",
       title = "When cycling genes crest",
       subtitle = "Two waves is the expected biology — a strong sanity check") +
  theme_classic()
```

---

# 5. Per gene- per strain fit

Only Spellman 478 genes were fit within each 82 strains

```{r q2_fits}
set.seed(SEED)
obs <- fit_by_group(Y_cc, tvec, gvec,
                    depth = if (INCLUDE_DEPTH) depthv else NULL)
cat("Fitted", nrow(obs$r2), "genes x", ncol(obs$r2), "strains\n")
n_by_strain <- obs$n_cells

range(obs$r2_adj, na.rm = TRUE)                              # true min, max
quantile(obs$r2_adj, c(0.25, 0.5, 0.75), na.rm = TRUE)         # IQR + median
```


# Diagnostics - 1. size dependance 

```{r size_dependence}
#correlation per strain, can correlate anything across sytrains
cor_with_n <- function(M) {
  apply(M, 1, function(v) {
    ok <- !is.na(v); if (sum(ok) < 10) return(NA_real_)
    cor(v[ok], n_by_strain[ok], method = "spearman")
  })
}

cat("Median per-gene correlation with cells per strain:\n")
cat("  raw R2                   :", round(median(cor_with_n(obs$r2),      na.rm = TRUE), 3), "\n")
cat("  adjusted R2              :", round(median(cor_with_n(obs$r2_adj),  na.rm = TRUE), 3),
    "  <- LOESS version was -0.629\n")
cat("  grid amplitude           :", round(median(cor_with_n(obs$amp),     na.rm = TRUE), 3), "\n")
cat("  first-harmonic amplitude :", round(median(cor_with_n(obs$amp1),    na.rm = TRUE), 3), "\n")
cat("  bias-corrected amplitude :", round(median(cor_with_n(obs$amp1_db), na.rm = TRUE), 3),
    "  <- the one used downstream\n")
cat("  peak time                :", round(median(cor_with_n(obs$peak),    na.rm = TRUE), 3), "\n")

# does the correction change WHAT is measured, or only remove bias?
cat("\ncor(grid amp, debiased amp1) over all gene x strain:",
    round(cor(as.vector(obs$amp), as.vector(obs$amp1_db),
              use = "complete.obs", method = "spearman"), 3),
    "\n(high = same quantity, bias removed)\n")

tibble(n_cells      = n_by_strain,
       median_r2adj = apply(obs$r2_adj,  2, median, na.rm = TRUE),
       median_amp   = apply(obs$amp1_db, 2, median, na.rm = TRUE)) %>%
  pivot_longer(-n_cells) %>%
  ggplot(aes(n_cells, value)) +
  geom_point() + geom_smooth(method = "loess", se = TRUE, colour = "firebrick") +
  facet_wrap(~name, scales = "free_y") +
  scale_x_log10() +
  labs(x = "Cells per strain (log10)", y = "Median across genes",
       title = "Residual size dependence after correction") +
  theme_classic()


histones <- c("HTB2"="YBL002W","HHT2"="YNL031C","HHF1"="YBR009C","HHF2"="YNL030W")
per_gene_agree <- sapply(rownames(obs$amp), function(g) {
  a <- obs$amp[g, ]; b <- obs$amp1_db[g, ]; ok <- !is.na(a) & !is.na(b)
  if (sum(ok) < 20) NA_real_ else cor(a[ok], b[ok], method = "spearman")
})
cat("Per-gene agreement:\n"); print(summary(per_gene_agree))
cat("\nHistones:\n")
print(round(per_gene_agree[intersect(histones, names(per_gene_agree))], 3))
cat("\nWorst 10 (2nd harmonic matters most here):\n")
print(round(sort(per_gene_agree)[1:10], 3))
```

## 5.1 Is amplitude just expression level?

Measured answer: no. Mean expression is nearly free of size bias (~-0.07), and
dividing amplitude by it does not help — the amplitude bias comes from
estimation noise in a *range*, not from expression level. Kept as a diagnostic.

```{r amp_vs_expression}
strains   <- colnames(obs$amp1_db)
mean_expr <- sapply(strains, function(s) colMeans(Y_cc[gvec == s, , drop = FALSE]))
dimnames(mean_expr) <- dimnames(obs$amp1_db)

cat("  mean expression vs n_cells:",
    round(median(cor_with_n(mean_expr), na.rm = TRUE), 3), "\n")

cv_amp_obs   <- row_sd(obs[[AMP_STAT]]) / row_mean(obs[[AMP_STAT]])
cv_mean_expr <- row_sd(mean_expr) / row_mean(mean_expr)

cat("cor(cv_amp, cv of mean expression):",
    round(cor(cv_amp_obs, cv_mean_expr, use = "complete.obs",
              method = "spearman"), 3),
    "\n(high = amplitude spread is really expression-level spread)\n")
```

---

# 6. Q2 — per-gene summaries across strains

```{r pheno_summary}
AMP <- obs[[AMP_STAT]]      # amplitude matrix used for all Q2 phenotypes

pheno <- tibble(
  gene   = rownames(obs$r2_adj),
  name   = to_name(rownames(obs$r2_adj)),

  # structure
  n_strains_fit = row_n(obs$r2_adj),
  mean_r2   = row_mean(obs$r2_adj),
  median_r2 = apply(obs$r2_adj, 1, median, na.rm = TRUE),
  sd_r2     = row_sd(obs$r2_adj),

  # amplitude (bias-corrected)
  mean_amp  = row_mean(AMP),
  sd_amp    = row_sd(AMP),
  cv_amp    = row_sd(AMP) / row_mean(AMP),
  amp_range = apply(AMP, 1, function(v) diff(range(v, na.rm = TRUE))),

  # timing
  mean_peak = circ_mean_rows(obs$peak),
  sd_peak   = circ_sd_rows(obs$peak),

  # detection — reported, NOT filtered on
  n_strains_total = ncol(obs$r2_adj),
  n_not_detected  = rowSums(obs$constant == 1, na.rm = TRUE),
  pct_strains_detected = 1 - rowSums(obs$constant == 1, na.rm = TRUE) /
                             ncol(obs$constant),
  mean_pct_expressed   = colMeans(Y_cc > 0)[rownames(obs$r2_adj)]
)

cat("Genes summarised:", nrow(pheno), "\n")
cat("Detected in every strain:", sum(pheno$pct_strains_detected == 1), "\n")
cat("Detected in <50% of strains:", sum(pheno$pct_strains_detected < 0.5), "\n\n")
print(summary(pheno$pct_strains_detected))

AMP_FLOOR <- quantile(pheno$mean_amp, 0.25, na.rm = TRUE)
cat("\nAmplitude floor for cv_amp ranking:", round(AMP_FLOOR, 3),
    "| genes below it:", sum(pheno$mean_amp < AMP_FLOOR, na.rm = TRUE), "\n")
```

```{r}
AMP <- obs[[AMP_STAT]]
amp_resid <- t(sapply(rownames(AMP), function(g) {
  y <- AMP[g, ]; x <- mean_expr[g, ]; ok <- !is.na(y) & !is.na(x)
  out <- rep(NA_real_, length(y)); names(out) <- names(y)
  if (sum(ok) >= 20) out[ok] <- residuals(lm(y[ok] ~ x[ok]))
  out
}))
resid_spread <- row_sd(amp_resid) / row_mean(AMP)

cat("Median cv_amp:", round(median(cv_amp_obs, na.rm = TRUE), 3),
    "| median after removing expression:", round(median(resid_spread, na.rm = TRUE), 3), "\n")

cat("\nHistones — how much survives:\n")
tibble(gene = intersect(histones, rownames(AMP))) %>%
  mutate(name = to_name(gene),
         cv_amp = cv_amp_obs[gene],
         resid  = resid_spread[gene],
         frac_explained = round(1 - resid / cv_amp, 2)) %>%
  print()
```



## 6.1 Detection as a phenotype

```{r detection_as_phenotype}
# patchy <- pheno %>%
#   filter(pct_strains_detected < 0.9, pct_strains_detected > 0.1) %>%
#   arrange(pct_strains_detected)
# 
# cat("Genes with patchy detection (10-90% of strains):", nrow(patchy), "\n")
# patchy %>% select(name, gene, pct_strains_detected, mean_pct_expressed,
#                   mean_r2, mean_amp) %>% head(20) %>% print(n = 20)
# 
# if (nrow(patchy) > 0) {
#   loss_by_strain <- colMeans(obs$constant[patchy$gene, , drop = FALSE] == 1,
#                              na.rm = TRUE)
#   print(head(sort(loss_by_strain, decreasing = TRUE), 10))
#   r_drop <- cor(loss_by_strain, n_by_strain, method = "spearman")
#   cat("cor(dropout rate, n_cells):", round(r_drop, 3), "\n")
#   cat(if (abs(r_drop) > 0.4)
#         "  STRONG -> dropout tracks cell count. Treat these as a sampling
#   artifact, not presence/absence variation, unless they survive §11.\n"
#       else
#         "  Weak -> dropout is not explained by cell count; worth following up.\n")
# }
```

---

# 7. Permutation null

Shuffling the cell → strain map preserves each strain's exact cell count and each
cell's `(expr, pseudotime)` pair, and breaks only the genotype link.

**Read the output of this section with §8.1 in mind.** With 74 strains and
16,000 cells, "significant" mostly tracks how precisely a gene is measured, not
how much it truly differs. Significance is a floor, not a ranking.

```{r permutation, cache=TRUE}
set.seed(SEED)
depth_arg <- if (INCLUDE_DEPTH) depthv else NULL

perm_stats <- function() {
  gperm <- sample(gvec)                 # preserves per-strain cell counts exactly
  f <- fit_by_group(Y_cc, tvec, gperm, depth = depth_arg)
  A <- f[[AMP_STAT]]
  cbind(sd_r2   = row_sd(f$r2_adj),
        cv_amp  = row_sd(A) / row_mean(A),
        sd_peak = circ_sd_rows(f$peak))
}

null_arr <- array(NA_real_, dim = c(ncol(Y_cc), 3, N_PERM),
                  dimnames = list(colnames(Y_cc),
                                  c("sd_r2","cv_amp","sd_peak"), NULL))

for (i in seq_len(N_PERM)) {
  null_arr[, , i] <- perm_stats()
  if (i %% 25 == 0) cat("permutation", i, "/", N_PERM, "\n")
}
```

```{r calibrate}
emp_p <- function(obs_v, null_mat) {
  (1 + rowSums(null_mat >= obs_v, na.rm = TRUE)) / (1 + rowSums(!is.na(null_mat)))
}
z_of <- function(obs_v, null_mat) {
  (obs_v - rowMeans(null_mat, na.rm = TRUE)) / apply(null_mat, 1, sd, na.rm = TRUE)
}

obs_vec <- list(sd_r2 = pheno$sd_r2, cv_amp = pheno$cv_amp, sd_peak = pheno$sd_peak)
for (nm in names(obs_vec)) names(obs_vec[[nm]]) <- pheno$gene

cal <- pheno
for (st in c("sd_r2", "cv_amp", "sd_peak")) {
  nm <- null_arr[pheno$gene, st, ]
  cal[[paste0(st, "_null_mean")]] <- rowMeans(nm, na.rm = TRUE)
  cal[[paste0(st, "_z")]]         <- z_of(obs_vec[[st]][pheno$gene], nm)
  cal[[paste0(st, "_p")]]         <- emp_p(obs_vec[[st]][pheno$gene], nm)
}
cal <- cal %>%
  mutate(across(ends_with("_p"), ~p.adjust(.x, method = "BH"), .names = "{.col}_fdr"))

cat("Significant at FDR < 0.05:\n")
cat("  structure (sd_r2) :", sum(cal$sd_r2_p_fdr   < 0.05, na.rm = TRUE), "\n")
cat("  amplitude (cv_amp):", sum(cal$cv_amp_p_fdr  < 0.05, na.rm = TRUE), "\n")
cat("  timing (sd_peak)  :", sum(cal$sd_peak_p_fdr < 0.05, na.rm = TRUE), "\n")

#how many distinct genes showed significant between-strain variation in any of the three phenotypes
n_any_sig <- sum(cal$cv_amp_p_fdr < 0.05 | cal$sd_peak_p_fdr < 0.05 | cal$sd_r2_p_fdr < 0.05, na.rm = TRUE)
cat("Genes significant in at least one phenotype:", n_any_sig, "\n")
```



---

# 8. The three answers

## 8.1 Why significance is not the ranking

If `pct_sig` climbs monotonically with `mean_r2`, the permutation test is
detecting *measurement precision*, not biological difference — well-measured
genes have the power to reject, poorly-measured ones do not. That is exactly why
the earlier `mean_r2 > 0.2` AND `not significant` filter returned zero genes:
those two conditions are nearly mutually exclusive by construction.

```{r why_significance_tracks_precision}
cat("Genes with mean_r2 > 0.15:", sum(cal$mean_r2 > 0.15, na.rm = TRUE),
    "of", nrow(cal), "\n\n")

cal %>%
  mutate(bin = cut(mean_r2, c(-Inf, 0.02, 0.05, 0.10, 0.20, 0.40, Inf))) %>%
  group_by(bin) %>%
  summarise(n = n(),
            sig_any = sum(sd_r2_p_fdr < 0.05 | cv_amp_p_fdr < 0.05 |
                          sd_peak_p_fdr < 0.05, na.rm = TRUE),
            pct_sig = round(sig_any / n, 2), .groups = "drop") %>%
  print()
```
```{r}
# ── Where does pct_sig stop climbing steeply? ────────────────────────────
inflection_check <- audit %>%
  filter(cycles) %>%
  mutate(sig_any = sd_r2_p_fdr < 0.05 | cv_amp_p_fdr < 0.05 | sd_peak_p_fdr < 0.05)

# finer bins than before
fine_bins <- inflection_check %>%
  mutate(bin = cut(mean_r2, breaks = seq(0, 0.5, by = 0.025))) %>%
  group_by(bin) %>%
  summarise(n = n(), pct_sig = mean(sig_any, na.rm = TRUE), .groups = "drop") %>%
  filter(n >= 5)   # drop bins too sparse to trust

print(fine_bins, n = 30)

# rate of change between consecutive bins — this IS the inflection
fine_bins <- fine_bins %>%
  mutate(bin_mid = as.numeric(str_extract(bin, "(?<=,)[0-9.]+")) - 0.0125,
         slope = c(NA, diff(pct_sig)) / c(NA, diff(bin_mid)))

print(fine_bins %>% select(bin, n, pct_sig, slope), n = 30)

# alternative, smoother view: fit a loess and find where its derivative flattens
lo <- loess(sig_any ~ mean_r2, data = inflection_check, span = 0.5)
grid_r2 <- seq(0.02, 0.5, by = 0.005)
pred    <- predict(lo, newdata = data.frame(mean_r2 = grid_r2))
deriv   <- diff(pred) / diff(grid_r2)

tibble(mean_r2 = grid_r2[-1], pct_sig_smoothed = pred[-1], slope = deriv) %>%
  filter(mean_r2 >= 0.05) %>%
  ggplot(aes(mean_r2, slope)) +
  geom_line() + geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Rate of change in pct_sig — inflection is where this flattens toward 0",
       x = "mean_r2", y = "d(pct_sig)/d(mean_r2)") +
  theme_classic()

audit %>% filter(cycles) %>%
  mutate(bin = cut(mean_r2, breaks = seq(0, 0.5, by = 0.025))) %>%
  group_by(bin) %>%
  summarise(n = n(),
            median_peak_excess = median(sd_peak / sd_peak_null_mean, na.rm = TRUE),
            median_amp_excess  = median(cv_amp / cv_amp_null_mean, na.rm = TRUE),
            .groups = "drop") %>%
  filter(n >= 5) %>%
  print(n = 20)
```
```{r}
n_measurable_and_sig <- sum(cal$mean_r2 > 0.15 &
                            (cal$cv_amp_p_fdr < 0.05 | cal$sd_peak_p_fdr < 0.05 | cal$sd_r2_p_fdr < 0.05),
                            na.rm = TRUE)
cat("Well-measured genes (mean_r2 > 0.15) significant in at least one phenotype:",
    n_measurable_and_sig, "of", sum(cal$mean_r2 > 0.15, na.rm = TRUE), "\n")

cal %>% filter(mean_r2 > 0.15) %>%
  summarise(sig_amp  = sum(cv_amp_p_fdr  < 0.05, na.rm = TRUE),
            sig_time = sum(sd_peak_p_fdr < 0.05, na.rm = TRUE),
            sig_str  = sum(sd_r2_p_fdr   < 0.05, na.rm = TRUE),
            sig_any  = sum(cv_amp_p_fdr < 0.05 | sd_peak_p_fdr < 0.05 | sd_r2_p_fdr < 0.05, na.rm = TRUE),
            n_total  = n())
```


## 8.2 Three-way accounting of every gene

Every Spellman gene lands in exactly one class. The third class — genes with too
little cell-cycle signal to say anything about — was missing from the earlier
version, which is why the numbers did not add up.

```{r audit}
audit <- cal %>%
  left_join(q1 %>% select(gene, r2_adj_pooled = r2_adj), by = "gene") %>%
  mutate(
    cycles  = r2_adj_pooled > R2_CYCLING,
    sig_any = (sd_r2_p_fdr < 0.05 | cv_amp_p_fdr < 0.05 | sd_peak_p_fdr < 0.05),
    class = case_when(
      !cycles           ~ "not measurably cycling",
      cycles &  sig_any ~ "cycling, varies between strains",
      cycles & !sig_any ~ "cycling, consistent between strains"))

#print(count(audit, class))
str(audit)
```

desc(consistency_score) = most variable

rank() gives 1 to the smallest value. So:

Low sd_peak + low cv_amp → low ranks → low score → most consistent
High sd_peak + high cv_amp → high ranks → high score → most variable

```{r}
amp_resid <- t(sapply(rownames(AMP), function(g) {
  y <- AMP[g, ]; x <- mean_expr[g, ]; ok <- !is.na(y) & !is.na(x)
  out <- rep(NA_real_, length(y)); names(out) <- names(y)
  if (sum(ok) >= 20) out[ok] <- residuals(lm(y[ok] ~ x[ok]))
  out
}))

resid_spread <- row_sd(amp_resid) / row_mean(AMP)   # spread not explained by expression

audit %>%
  mutate(resid_spread = resid_spread[gene],
         frac_explained = 1 - resid_spread / cv_amp) %>%
  filter(cycles, cv_amp_p_fdr < 0.05) %>%
  arrange(desc(frac_explained)) %>%
  select(name, gene, mean_r2, cv_amp, resid_spread, frac_explained) %>%
  print(n = 25)
```


## 8.3 The variable set, ranked by effect size in real units

A 10-minute peak shift is a claim you can defend at the bench. A z of 18 is not.
Structure

mean_r2 — not "spread," and not across cells. It's the average of HHF1's 74 per-strain adjusted R² values. It answers "how strongly does this gene cycle, typically, in a strain." Used as a measurability filter, not as a variation measure.

sd_r2 — the standard deviation of those same 74 values. This is the structure variation.

Timing

sd_peak — circular SD of the 74 peak times, in minutes. How scattered the crest is across strains.

sd_peak_null_mean — the average sd_peak over the 500 permutations. It is not "how much it varies from its null." It is the null itself: how much scatter this gene would show if strain labels meant nothing. HHF1's is 1.43 minutes.

peak_excess = sd_peak / sd_peak_null_mean — a ratio, not minutes. HHT2's 3.18 means "3.18 times more timing scatter than chance," not 3.18 minutes and not 3.18 minutes ahead. It's unitless by design, so genes with different noise levels can be ranked together.

peak_diff = sd_peak - sd_peak_null_mean — the same comparison in minutes. HHT2: 5.18 − 1.63 = 3.55 minutes of excess scatter. This is the number for a talk; peak_excess is the number for ranking.

The outlier diagnostics

sd_trim — recompute the spread after dropping the 10% of strains furthest from the consensus. Nothing is removed from your actual analysis; this is a check. If sd_peak is 12 but sd_trim is 3, a few strains were doing all the work.

n_outlier — how many of the 74 strains sit more than 10 minutes from the circular mean. Also diagnostic only. Not trimmed anywhere.

Read them together. MRH1: sd_peak 12.46, sd_trim 9.17, n_outlier 19 — broad, not outlier-driven, but noisy. HHF1: sd_peak 4.35, sd_trim 1.86, n_outlier 2 — compact and tight.

Amplitude

mean_amp — average of the 74 per-strain amplitudes. Typical size of the wave. Each of those 74 came from that strain's cells, but the mean is across strains.

cv_amp — sd(74 amplitudes) / mean(74 amplitudes). Relative spread, unitless.

cv_amp_null_mean — the average cv_amp across the 500 permutations. Again, the null itself.

amp_excess = cv_amp / cv_amp_null_mean — a ratio, not a difference between extreme strains. HHF1: 0.193 / 0.0745 = 2.60.

amp_spread = amp_range = max(74 amplitudes) − min(74 amplitudes) — this is the extreme-to-extreme difference you were describing, but it's amp_spread, not amp_excess. In absolute log-normalized units.

cv_amp vs amp_spread

Different in two ways.

Scale. cv_amp divides by the mean, so it's relative. amp_spread doesn't, so it's absolute. A gene with amplitudes 1.0–1.5 and one with 10–15 have identical cv_amp but amp_spread of 0.5 versus 5.

Robustness. amp_spread uses only the two most extreme strains, so one bad strain sets it. cv_amp uses all 74.

```{r variable_effect_size}
# ── Per-gene timing structure: how many strains shift, which way, to where ──
peak_shape <- map_dfr(rownames(obs$peak), function(g) {
  p <- obs$peak[g, ]; p <- p[!is.na(p)]
  if (length(p) < 20)
    return(tibble(gene = g, n_fit = NA_integer_, sd_trim = NA_real_,
                  n_outlier = NA_integer_, n_early = NA_integer_, n_late = NA_integer_,
                  shift_early = NA_real_, shift_late = NA_real_,
                  peak_early = NA_real_, peak_late = NA_real_,
                  early_strains = NA_character_, late_strains = NA_character_))

  ctr <- circ_mean_rows(matrix(p, nrow = 1))     # consensus peak, minutes
  d   <- ((p - ctr + WRAP/2) %% WRAP) - WRAP/2   # signed shift: - earlier, + later

  early <- d[d < -10]
  late  <- d[d >  10]

  tibble(
    gene        = g,
    n_fit       = length(p),
    sd_trim     = round(sd(d[abs(d) <= quantile(abs(d), 0.9)]), 2),
    n_outlier   = length(early) + length(late),
    n_early     = length(early),
    n_late      = length(late),
    shift_early = if (length(early)) round(median(early), 1) else NA_real_,
    shift_late  = if (length(late))  round(median(late),  1) else NA_real_,
    peak_early  = if (length(early)) round((ctr + median(early)) %% WRAP, 1) else NA_real_,
    peak_late   = if (length(late))  round((ctr + median(late))  %% WRAP, 1) else NA_real_,
    early_strains = if (length(early))
      paste(names(sort(early))[1:min(5, length(early))], collapse = ",") else NA_character_,
    late_strains  = if (length(late))
      paste(names(sort(late, decreasing = TRUE))[1:min(5, length(late))], collapse = ",") else NA_character_
  )
})

stopifnot(intersect(names(audit), names(peak_shape)) == "gene")

# ── Why raw spread cannot be the ranking ────────────────────────────────────
cat("Correlation of raw spread statistics with mean_r2 (cycling genes):\n")
audit %>% filter(cycles) %>%
  summarise(sd_peak   = cor(mean_r2, sd_peak,   method = "spearman", use = "complete.obs"),
            amp_range = cor(mean_r2, amp_range, method = "spearman", use = "complete.obs"),
            cv_amp    = cor(mean_r2, cv_amp,    method = "spearman", use = "complete.obs")) %>%
  round(3) %>% print()
cat("  -> all three track how strongly a gene cycles, so rank on excess over each\n",
    "     gene's own permutation null instead.\n\n")
```
peak_shape, amp_shape, and str_shape are the fix for that. For each gene, they:

Find the middle (the median) of the 74 values.
Find how far each strain's value is from that middle.
Drop the 10% of strains that are furthest away — the most extreme ones.
Recompute the spread using only the remaining 90%.

robustness: If a gene's raw spread and trimmed spread are almost the same, robustness is close to 1 (100%) — dropping the extremes barely mattered, so the variation is spread broadly across many strains, not concentrated in a few.

```{r}
# ── Trimmed spread for ALL THREE phenotypes, computed the same way ──────────

trimmed_sd <- function(v, trim = 0.10) {
  v <- v[is.finite(v)]
  if (length(v) < 10) return(NA_real_)
  d <- abs(v - median(v))
  keep <- d <= quantile(d, probs = 1 - trim, na.rm = TRUE)
  if (sum(keep) < 2) return(NA_real_)
  sd(v[keep])
}

trimmed_cv <- function(v, trim = 0.10) {
  v <- v[is.finite(v)]
  if (length(v) < 10) return(NA_real_)
  d <- abs(v - median(v))
  keep <- d <= quantile(d, probs = 1 - trim, na.rm = TRUE)
  v_trim <- v[keep]
  if (length(v_trim) < 2 || mean(v_trim) <= 0) return(NA_real_)
  sd(v_trim) / mean(v_trim)
}

amp_shape <- map_dfr(rownames(AMP), function(g) {
  tibble(gene = g, cv_trim_amp = trimmed_cv(AMP[g, ]))
})

str_shape <- map_dfr(rownames(obs$r2_adj), function(g) {
  tibble(gene = g, sd_trim_str = trimmed_sd(obs$r2_adj[g, ]))
})

# ── frac_explained: how much of amplitude spread is really expression level ──
# (from your amp_resid chunk — resid_spread and cv_amp_obs are already gene-named)
frac_explained_tbl <- tibble(
  gene = names(resid_spread),
  frac_explained = 1 - resid_spread / cv_amp_obs[names(resid_spread)]
)



# ── One table, three parallel trimmed-excess columns ─────────────────────────
gene_summary <- audit %>%
  filter(cycles, mean_r2 > 0.15) %>%
  left_join(peak_shape,         by = "gene") %>%
  left_join(amp_shape,          by = "gene") %>%
  left_join(str_shape,          by = "gene") %>%
  left_join(frac_explained_tbl, by = "gene") %>%
  mutate(
    n_strains_total = ncol(AMP),        # 74 — same for every gene, for reference
    n_fit_struct    = n_strains_fit,    # already in audit/pheno
    n_fit_amp       = n_strains_fit,    # amp and structure are fit together per strain
    n_fit_peak      = n_fit,            # from peak_shape — can be lower (needs >=20 non-NA)

    # raw excess (all strains) — kept for the outlier-diagnostic comparison
    peak_excess  = sd_peak / sd_peak_null_mean,
    amp_excess   = cv_amp  / cv_amp_null_mean,
    str_excess   = sd_r2   / sd_r2_null_mean,

    # trimmed excess (drop worst 10%) — THIS is what ranking uses, for all three
    peak_trim_excess = sd_trim     / sd_peak_null_mean,
    amp_trim_excess = cv_trim_amp / cv_amp_null_mean,   # now both sides are CV
    str_trim_excess  = sd_trim_str / sd_r2_null_mean,

    # ── inside the mutate(), after the trim_excess block, add: ──────────────────
    peak_robustness = sd_trim     / sd_peak,
    amp_robustness  = cv_trim_amp / cv_amp,
    str_robustness  = sd_trim_str / sd_r2,

    frac_outlier_peak = n_outlier / n_fit,
    peak_diff  = sd_peak - sd_peak_null_mean,
    amp_spread = amp_range,

    # gate 1: exclude amplitude candidates where expression level explains
    # most (>50%) of the between-strain amplitude spread
    expr_gated = frac_explained > 0.5,

    amp_gated = expr_gated,
    amp_trim_excess_final = ifelse(amp_gated, NA_real_, amp_trim_excess),

    # ── replace the variability line with: ───────────────────────────────────────
    rank_peak = percent_rank(peak_trim_excess),
    rank_str  = percent_rank(str_trim_excess),
    rank_amp  = percent_rank(amp_trim_excess_final),
    n_phenotypes_available = rowSums(!is.na(cbind(rank_peak, rank_str, rank_amp))),
    variability          = rowMeans(cbind(rank_peak, rank_str, rank_amp), na.rm = TRUE),
    variability_strict   = ifelse(n_phenotypes_available == 3, variability, NA_real_),

    which_phenotype = paste0(
      ifelse(sd_r2_p_fdr   < 0.05, "structure ", ""),
      ifelse(cv_amp_p_fdr  < 0.05 & mean_amp >= AMP_FLOOR & !amp_gated, "amplitude ", ""),
      ifelse(sd_peak_p_fdr < 0.05, "timing", "")) %>% str_trim()
  ) %>%
  arrange(desc(variability))
```




```{r}
cat("Well-measured cycling genes:", nrow(gene_summary),
    " (pooled r2_adj >", R2_CYCLING, "and mean per-strain r2 > 0.1)\n\n")
cat("Amplitude candidates excluded as expression-driven (frac_explained > 0.5):",
    sum(gene_summary$amp_gated, na.rm = TRUE), "of", nrow(gene_summary), "\n\n")

# ── Outlier diagnostic, now for all three ────────────────────────────────────
cat("Is the spread broad or outlier-driven? (raw excess vs trimmed excess)\n")
cat("  A large gap = a few strains carrying the signal.\n\n")
gene_summary %>%
  mutate(gap_peak = round(peak_excess - peak_trim_excess, 2),
         gap_amp  = round(amp_excess  - amp_trim_excess,  2),
         gap_str  = round(str_excess  - str_trim_excess,  2)) %>%
  arrange(desc(gap_peak + gap_amp + gap_str)) %>%
  select(name, gene, peak_excess, peak_trim_excess, gap_peak,
         amp_excess, amp_trim_excess, gap_amp,
         str_excess, str_trim_excess, gap_str) %>%
  head(20) %>% print(n = 20, width = Inf)

# ── OVERALL: most variable (combined) and most consistent (combined) ────────
cat("\nMOST VARIABLE — combined ranking across structure, amplitude, timing:\n")
gene_summary %>% head(15) %>%
  select(name, gene, which_phenotype, mean_r2,
         str_trim_excess, amp_trim_excess_final, peak_trim_excess, variability) %>%
  print(n = 15)

cat("\nMOST CONSISTENT — combined ranking, low end:\n")
gene_summary %>% tail(15) %>%
  select(name, gene, mean_r2,
         str_trim_excess, amp_trim_excess_final, peak_trim_excess, variability) %>%
  print(n = 15)
```

## gene significant varying in each phenotype

amp_gated means "NOT gated" — i.e., this gene passed that check, so its amplitude signal is trustworthy enough to report. Filtering !amp_gated is how you exclude the expression-confounded genes from sig_amplitude.
```{r}
# ── Canonical per-phenotype significant-gene tables ─────────────────────────
# Each contains ONLY genes passing that phenotype's own FDR < 0.05, sorted by
# trimmed excess (largest first). These are what all downstream figures and
# text should reference — not raw gene_summary %>% arrange() %>% head(), which
# does not require significance.

sig_structure <- gene_summary %>%
  filter(sd_r2_p_fdr < 0.05) %>%
  arrange(desc(str_trim_excess)) %>%
  select(name, gene, mean_r2, sd_r2, sd_r2_null_mean, str_trim_excess,
         str_robustness, sd_r2_p_fdr, which_phenotype, early_strains, late_strains)

sig_structure

sig_amplitude <- gene_summary %>%
  filter(cv_amp_p_fdr < 0.05, !amp_gated) %>%
  arrange(desc(amp_trim_excess_final)) %>%
  select(name, gene, mean_r2, cv_amp, amp_trim_excess_final,
         amp_robustness, frac_explained, cv_amp_p_fdr, which_phenotype,
         early_strains, late_strains)
sig_amplitude

sig_timing <- gene_summary %>%
  filter(sd_peak_p_fdr < 0.05) %>%
  arrange(desc(peak_trim_excess)) %>%
  select(name, gene, mean_r2, mean_peak, sd_peak, peak_trim_excess,
         peak_robustness, n_early, n_late, sd_peak_p_fdr, which_phenotype,
         early_strains, late_strains)
sig_timing
cat("Significant genes — structure:", nrow(sig_structure),
    "| amplitude:", nrow(sig_amplitude),
    "| timing:", nrow(sig_timing), "\n")
```

#histone variability is the most surprising output so let's look at that:
```{r}
gene_summary %>%
  filter(gene %in% c("YDR033W", "YBR067C", "YBR009C", "YNL030W", "YNL031C", "YBL002W")) %>%
  select(name, gene, str_trim_excess, str_robustness,
         amp_trim_excess_final, amp_robustness,
         peak_trim_excess, peak_robustness)
#gene summary of all top hit genes
gene_summary %>% filter(gene %in% c("YDR033W","YBR067C","YBR009C","YNL030W","YNL031C","YBL002W")) %>%
  select(name, early_strains, late_strains)
#gene summary of all top histone genes
gene_summary %>% filter(gene %in% c("YBR009C","YNL030W","YNL031C","YBL002W")) %>%
  select(name, early_strains, late_strains)

# for HHT2, was BDL/APQ actually part of the 10% trimmed out, or did they survive?
p <- obs$peak["YNL031C", ]  # HHT2's peak times across all strains
p <- p[!is.na(p)]
ctr <- circ_mean_rows(matrix(p, nrow = 1))
d <- ((p - ctr + WRAP/2) %% WRAP) - WRAP/2
d_sorted <- sort(abs(d), decreasing = TRUE)

# which strains are in the trimmed-OUT top 10%?
n_trim <- ceiling(0.10 * length(d))
trimmed_out <- names(sort(abs(d), decreasing = TRUE))[1:n_trim]
cat("Strains trimmed out for HHT2:", paste(trimmed_out, collapse = ", "), "\n")
cat("Was BDL trimmed out?", "BDL" %in% trimmed_out, "\n")
cat("Was APQ trimmed out?", "APQ" %in% trimmed_out, "\n")

# HHT2 peak spread, excluding APQ entirely (not just relative-trim)
p_no_apq <- obs$peak["YNL031C", setdiff(colnames(obs$peak), "APQ")]
p_no_apq <- p_no_apq[!is.na(p_no_apq)]
ctr_no_apq <- circ_mean_rows(matrix(p_no_apq, nrow = 1))
sd_no_apq <- circ_sd_rows(matrix(p_no_apq, nrow = 1))

cat("HHT2 sd_peak with APQ:", cal$sd_peak[cal$gene == "YNL031C"], "\n")
cat("HHT2 sd_peak without APQ:", sd_no_apq, "\n")

for (g in c("YBR009C", "YBL002W", "YNL030W")) {  # HHF1, HTB2, HHF2
  p_all <- obs$peak[g, ]; p_all <- p_all[!is.na(p_all)]
  p_no_apq <- obs$peak[g, setdiff(colnames(obs$peak), "APQ")]
  p_no_apq <- p_no_apq[!is.na(p_no_apq)]
  cat(g, "| with APQ:", round(circ_sd_rows(matrix(p_all, nrow=1)), 2),
      "| without APQ:", round(circ_sd_rows(matrix(p_no_apq, nrow=1)), 2), "\n")
}
```





```{r phenotype_overlap}
# ── shared axis limits so the two panels are visually comparable ──────────
lims <- gene_summary %>%
  summarise(x_max = max(c(amp_excess, amp_trim_excess), na.rm = TRUE),
            y_max = max(c(peak_excess, peak_trim_excess), na.rm = TRUE))

make_overlap_plot <- function(df, x, y, title) {
  df %>%
    mutate(sig_status = case_when(
      gene %in% sig_amplitude$gene & gene %in% sig_timing$gene ~ "both significant",
      gene %in% sig_amplitude$gene                              ~ "amplitude only",
      gene %in% sig_timing$gene                                 ~ "timing only",
      TRUE                                                      ~ "neither"
    )) %>%
    ggplot(aes(.data[[x]], .data[[y]])) +
    geom_hline(yintercept = 1, linetype = 2, colour = "grey70") +
    geom_vline(xintercept = 1, linetype = 2, colour = "grey70") +
    geom_point(aes(size = mean_r2, colour = sig_status), alpha = 0.6) +
    scale_colour_manual(values = c("both significant" = "firebrick",
                                   "amplitude only"   = "darkorange",
                                   "timing only"       = "steelblue",
                                   "neither"           = "grey70"),
                        name = "Significance (FDR < 0.05)") +
    ggrepel::geom_text_repel(
      data = . %>% filter(sig_status != "neither"),
      aes(label = name), size = 2.5, max.overlaps = 20) +
    coord_cartesian(xlim = c(0, lims$x_max), ylim = c(0, lims$y_max)) +
    labs(x = "Amplitude excess over null", y = "Timing excess over null",
         title = title) +
    theme_classic()
}

p_overlap_raw     <- make_overlap_plot(gene_summary, "amp_excess", "peak_excess",
                                       "Untrimmed (all strains)")
p_overlap_trimmed <- make_overlap_plot(gene_summary, "amp_trim_excess_final", "peak_trim_excess",
                                       "Trimmed (worst 10% dropped)")

print(p_overlap_raw + p_overlap_trimmed)

dir.create("../results", showWarnings = FALSE)
ggsave("../results/phenotype_overlap_amp_vs_timing.png",
       p_overlap_raw + p_overlap_trimmed, width = 12, height = 6, dpi = 300)
```

```{r save}
dir.create("../results", showWarnings = FALSE)

# ── the expensive objects — save as .rds so you never need to rerun ────────
saveRDS(obs,      "../results/harmonic_obs.rds")       # per-strain fits (all 74 x 478)
saveRDS(null_arr, "../results/harmonic_null_arr.rds")   # 500 permutations, raw
saveRDS(pooled,   "../results/harmonic_pooled.rds")     # Q1 pooled fit

cat("null_arr size:", format(object.size(null_arr), units = "MB"), "\n")

# ── summary tables — .csv for readability / sharing outside R ──────────────
write.csv(q1,            "../results/harmonic_q1_pooled.csv",           row.names = FALSE)
write.csv(cal,            "../results/harmonic_pheno_calibrated.csv",   row.names = FALSE)
write.csv(audit,          "../results/harmonic_gene_audit.csv",         row.names = FALSE)
write.csv(gene_summary,   "../results/harmonic_gene_summary.csv",       row.names = FALSE)

# ── the six canonical per-phenotype gene lists ──────────────────────────────
write.csv(sig_structure,        "../results/sig_structure.csv",        row.names = FALSE)
write.csv(sig_amplitude,        "../results/sig_amplitude.csv",        row.names = FALSE)
write.csv(sig_timing,           "../results/sig_timing.csv",           row.names = FALSE)
write.csv(consistent_structure, "../results/consistent_structure.csv", row.names = FALSE)
write.csv(consistent_amplitude, "../results/consistent_amplitude.csv", row.names = FALSE)
write.csv(consistent_timing,    "../results/consistent_timing.csv",    row.names = FALSE)

long_pheno <- map_dfr(c("r2_adj","amp","amp1_db","peak","phase1"), function(st)
  as.data.frame(obs[[st]]) %>% rownames_to_column("gene") %>%
    pivot_longer(-gene, names_to = "strain", values_to = "value") %>%
    mutate(stat = st))
write.csv(long_pheno, "../results/harmonic_strain_phenotypes_long.csv",
          row.names = FALSE)

#to reload in the future:
# obs      <- readRDS("../results/harmonic_obs.rds")
# null_arr <- readRDS("../results/harmonic_null_arr.rds")
# pooled   <- readRDS("../results/harmonic_pooled.rds")
# sig_structure        <- read.csv("../results/sig_structure.csv")
# sig_amplitude        <- read.csv("../results/sig_amplitude.csv")
# sig_timing           <- read.csv("../results/sig_timing.csv")
# consistent_structure <- read.csv("../results/consistent_structure.csv")
# consistent_amplitude <- read.csv("../results/consistent_amplitude.csv")
# consistent_timing    <- read.csv("../results/consistent_timing.csv")
```

---

# 9. Plotting

The curve plotted is exactly the curve that was measured — no padding, no seam
handling, no second smoother.

```{r plot_toolkit}
PHASE_BOUNDS <- tibble(
  xmin  = c(0, 7, 22.5, 39.5, 56.5, 77.5),
  xmax  = c(7, 22.5, 39.5, 56.5, 77.5, WRAP),
  phase = factor(c('M','M-G1','G1','S','G2','M'), levels = CC_LEVELS)
)

harmonic_curve <- function(gene, strain = NULL, K = HARMONICS) {
  idx <- if (is.null(strain)) seq_along(tvec) else which(gvec == strain)
  y <- Y_cc[idx, gene]
  if (length(idx) < 2 * K + 4 || sd(y) == 0) return(NULL)
  X <- harmonic_design(tvec[idx], K)
  b <- qr.coef(qr(X), y); b[is.na(b)] <- 0
  tibble(time = GRID, y = as.numeric(harmonic_design(GRID, K) %*% b),
         strain = strain %||% "global")
}

cc_panel <- function(genes, per_strain = TRUE, ncol = 3, title = NULL) {
  orfs <- to_orf(genes)
  orfs <- orfs[orfs %in% colnames(Y_cc)]
  if (length(orfs) == 0) { message("cc_panel: no genes to plot"); return(invisible(NULL)) }

  global <- map_dfr(orfs, ~ harmonic_curve(.x) %>% mutate(gene = .x))
  strains <- names(n_by_strain)
  per <- if (per_strain)
    map_dfr(orfs, function(g)
      map_dfr(strains, ~ harmonic_curve(g, .x) %>% mutate(gene = g))) else NULL

  binned <- tibble(gene = rep(orfs, each = length(tvec)),
                   t = rep(round(tvec) %% WRAP, times = length(orfs)),
                   y = as.vector(Y_cc[, orfs])) %>%
    group_by(gene, t) %>% summarise(y = mean(y), .groups = "drop")

  p <- ggplot() +
    geom_rect(data = PHASE_BOUNDS,
              aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = phase),
              alpha = 0.20, inherit.aes = FALSE) +
    scale_fill_manual(values = PHASE_FILL, name = "Phase")
  if (per_strain)
    p <- p + geom_line(data = per %>% mutate(label = to_name(gene)),
                       aes(time, y, group = strain),
                       colour = "grey45", alpha = 0.4, linewidth = 0.3)
  p +
    geom_point(data = binned %>% mutate(label = to_name(gene)),
               aes(t, y), size = 1, alpha = 0.6, colour = "grey20") +
    geom_line(data = global %>% mutate(label = to_name(gene)),
              aes(time, y), colour = "black", linewidth = 1) +
    scale_x_continuous(breaks = seq(0, WRAP, 11)) +
    coord_cartesian(xlim = c(0, WRAP)) +
    facet_wrap(~label, scales = "free_y", ncol = ncol) +
    labs(x = "Time in cell cycle (min)", y = "Normalized expression", title = title) +
    theme_classic() + theme(legend.position = "bottom")
}

```

```{r}
dir.create("../results/cc_panels", showWarnings = FALSE, recursive = TRUE)

save_cc <- function(plot, filename, width = 10, height = 7) {
  ggsave(file.path("../results/cc_panels", filename), plot,
         width = width, height = height, dpi = 300)
}

# ── MARKERS ──────────────────────────────────────────────────────────────
p_markers <- cc_panel(c("HHF1", "CTS1"), ncol = 2,
                      title = "Markers — global (black) + per strain (grey)")
print(p_markers)
save_cc(p_markers, "markers_HHF1_CTS1.png", width = 7, height = 4)

# ── STRUCTURE ─────────────────────────────────────────────────────────────
p_structure_varies <- cc_panel(head(sig_structure$gene, 6),
             title = "Largest SIGNIFICANT structure (R2) differences —\ncurves should differ in tightness")
print(p_structure_varies)
save_cc(p_structure_varies, "structure_varies.png")

# p_structure_consistent <- cc_panel(head(consistent_structure$gene, 6),
#              title = "Most consistent structure (not significant, FDR >= 0.05) —\ncurves should overlap tightly")
# print(p_structure_consistent)
# save_cc(p_structure_consistent, "structure_consistent.png")

# ── AMPLITUDE (gated, trimmed, FDR-significant) ────────────────────────────
p_amplitude_varies <- cc_panel(head(sig_amplitude$gene, 6),
             title = "Largest SIGNIFICANT amplitude differences\n(outlier-trimmed, expression-gated) — same shape, different heights")
print(p_amplitude_varies)
save_cc(p_amplitude_varies, "amplitude_varies.png")

# p_amplitude_consistent <- cc_panel(head(consistent_amplitude$gene, 6),
#              title = "Most consistent amplitude (not significant, FDR >= 0.05) —\ncurves should reach similar heights")
# print(p_amplitude_consistent)
# save_cc(p_amplitude_consistent, "amplitude_consistent.png")

cat("Amplitude genes excluded from these plots (expression-driven):",
    sum(gene_summary$amp_gated, na.rm = TRUE), "\n")

# ── TIMING (trimmed, FDR-significant) ──────────────────────────────────────
p_timing_varies <- cc_panel(head(sig_timing$gene, 6),
             title = "Largest SIGNIFICANT timing shifts (outlier-trimmed) —\ncurves should be offset sideways")
print(p_timing_varies)
save_cc(p_timing_varies, "timing_varies.png")

# p_timing_consistent <- cc_panel(head(consistent_timing$gene, 6),
#              title = "Most consistent timing (not significant, FDR >= 0.05) —\ncurves should crest together")
# print(p_timing_consistent)
# save_cc(p_timing_consistent, "timing_consistent.png")

# ── SIGNIFICANT IN BOTH amplitude AND timing ───────────────────────────────
both_sig <- intersect(sig_amplitude$gene, sig_timing$gene)
cat("\nGenes significant in BOTH amplitude and timing:", length(both_sig), "|",
    paste(to_name(head(both_sig, 10)), collapse = ", "), "\n")

p_both_amp_timing <- cc_panel(head(both_sig, 6), title = "Significant in both timing and amplitude")
print(p_both_amp_timing)
save_cc(p_both_amp_timing, "both_amplitude_and_timing.png")
```

```{r}
strain_scatter <- function(gene_ids, title, subtitle = NULL) {
  map_dfr(gene_ids, ~ tibble(gene = .x, strain = colnames(AMP),
                             amp = AMP[.x, ], peak = obs$peak[.x, ],
                             n_cells = n_by_strain,
                             poor_cov = colnames(AMP) %in% POOR_COVERAGE)) %>%
    filter(!is.na(amp)) %>%
    mutate(name = to_name(gene)) %>%
    ggplot(aes(peak, amp)) +
    geom_point(aes(size = n_cells, colour = poor_cov), alpha = 0.6) +
    scale_colour_manual(values = c(`FALSE` = "black", `TRUE` = "firebrick"),
                        name = "gap > 22 min") +
    ggrepel::geom_text_repel(
      data = . %>% group_by(name) %>%
        mutate(ctr = circ_mean_rows(matrix(peak, nrow = 1)),
               d   = ((peak - ctr + WRAP/2) %% WRAP) - WRAP/2) %>%
        filter(abs(d) > 10) %>% ungroup(),
      aes(label = strain), size = 2.5, max.overlaps = 15) +
    facet_wrap(~name, scales = "free_y") +
    scale_x_continuous(limits = c(0, WRAP), breaks = seq(0, WRAP, 22)) +
    labs(x = "Peak time (min)", y = "Amplitude (bias-corrected)",
         title = title, subtitle = subtitle) +
    theme_classic()
}

strain_scatter_structure <- function(gene_ids, title, subtitle = NULL) {
  map_dfr(gene_ids, ~ tibble(gene = .x, strain = colnames(obs$r2_adj),
                             r2 = obs$r2_adj[.x, ],
                             n_cells = n_by_strain,
                             poor_cov = colnames(obs$r2_adj) %in% POOR_COVERAGE)) %>%
    filter(!is.na(r2)) %>%
    mutate(name = to_name(gene)) %>%
    ggplot(aes(reorder(strain, r2), r2)) +
    geom_point(aes(size = n_cells, colour = poor_cov), alpha = 0.6) +
    scale_colour_manual(values = c(`FALSE` = "black", `TRUE` = "firebrick"),
                        name = "gap > 22 min") +
    facet_wrap(~name, scales = "free_x") +
    labs(x = "Strain (sorted by R2)", y = "Adjusted R2 (fit quality)",
         title = title, subtitle = subtitle) +
    theme_classic() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
}

dir.create("../results/strain_scatter", showWarnings = FALSE, recursive = TRUE)

save_scatter <- function(plot, filename, width = 10, height = 7) {
  ggsave(file.path("../results/strain_scatter", filename), plot,
         width = width, height = height, dpi = 300)
}

# ── STRUCTURE ────────────────────────────────────────────────────────────
p_struct_sig <- strain_scatter_structure(head(sig_structure$gene, 6),
                                         "Significant structure variation (FDR < 0.05)")
print(p_struct_sig)
save_scatter(p_struct_sig, "structure_highest.png")

# ── AMPLITUDE (gated genes excluded) ──────────────────────────────────────
p_amp_sig <- strain_scatter(head(sig_amplitude$gene, 6),
                            "Significant amplitude variation (FDR < 0.05)",
                            "(expression-driven genes excluded)")
print(p_amp_sig)
save_scatter(p_amp_sig, "amplitude_highest.png")

# ── TIMING ───────────────────────────────────────────────────────────────
p_timing_sig <- strain_scatter(head(sig_timing$gene, 6),
                               "Significant timing variation (FDR < 0.05)")
print(p_timing_sig)
save_scatter(p_timing_sig, "timing_highest.png")
```



---

# 10. Is a strain's deviation gene-specific or global?

A strain with shallow coverage looks "different" for every gene at once.
Subtract each strain's global tendency before claiming a gene-specific effect.

```{r strain_offset}
zrow <- function(M) t(scale(t(M)))

well   <- pheno$gene[pheno$mean_r2 > 0.3 & pheno$pct_strains_detected > 0.9]
Z      <- zrow(obs$r2_adj[well, , drop = FALSE])
offset <- colMeans(Z, na.rm = TRUE)

cat("Strain offset range:", round(range(offset, na.rm = TRUE), 2), "\n")
cat("cor(offset, n_cells):",
    round(cor(offset, n_by_strain, method = "spearman", use = "complete.obs"), 3),
    "  <- non-zero means part of the strain signal is cell number\n")

resid_mat <- sweep(zrow(obs$r2_adj), 2, offset, "-")

histones <- c("HTB2"="YBL002W","HHT2"="YNL031C","HHF1"="YBR009C","HHF2"="YNL030W")
h <- intersect(histones, rownames(resid_mat))

if (length(h) >= 2) {
  set.seed(SEED)
  rand <- sample(setdiff(well, histones), min(12, length(setdiff(well, histones))))
  cm <- cor(t(resid_mat[c(h, rand), , drop = FALSE]), use = "pairwise.complete.obs")
  cat("\nAfter removing the strain offset:\n")
  cat("  histone-histone median r:", round(median(cm[h, h][upper.tri(cm[h, h])]), 3), "\n")
  cat("  random-random   median r:", round(median(cm[rand, rand][upper.tri(cm[rand, rand])]), 3), "\n")
  cat("  histone-random  median r:", round(median(cm[h, rand]), 3), "\n")
  cat("  (histone-histone >> random-random = one shared regulatory cause)\n")
}
```
```{r}
# Q1: how many of 478 cycled at all
sum(q1$r2_adj > R2_CYCLING & q1$is_spellman, na.rm = TRUE)

# how many retained after the mean_r2 > 0.15 measurability floor
nrow(gene_summary)   # once you've set the filter to 0.15

# significant per phenotype
sum(gene_summary$sd_r2_p_fdr < 0.05, na.rm = TRUE)      # structure
sum(gene_summary$cv_amp_p_fdr < 0.05 & !gene_summary$amp_gated, na.rm = TRUE)  # amplitude
sum(gene_summary$sd_peak_p_fdr < 0.05, na.rm = TRUE)     # timing

# overlap
gene_summary %>% mutate(n_sig = (sd_r2_p_fdr<0.05) + (cv_amp_p_fdr<0.05 & !amp_gated) + (sd_peak_p_fdr<0.05)) %>%
  count(n_sig)
```

