---
title: "Fig.2-panel-A-B-C-D-gene-expression-comparison"
output: html_document
---

#I am only showing YPD.1 data in the main paper.

#Panel A: Heatmap of average gene expression within strains - does it cluster by clade?
#Panel B:Dimension reduction plot of gene expression within strains - does it cluster by clade?
#Panel C:Focused examples of differential gene expression - GAL genes?, NCR genes?
#Panel D:Is within strain gene expression correlation/anticorrelation conserved among strains?

The code for pseudobulking step is in figure 3
```{r}
# ============================================================
# Figure 2 — Multiplexed scRNA-seq provides an efficient means
# of differential gene expression analysis (YPD.1 only)
#
# Panel A: strain-level DEG heatmap — does it cluster by clade?
# Panel B: dimension reduction on DEGs — does it cluster by clade?
# Panel C: focused DGE examples (GAL / NCR), GO-ORA-selected
# Panel D: gene-gene correlation/anticorrelation, conserved across strains?
# ============================================================

library(DESeq2)
library(tidyverse)
library(pheatmap)
library(circlize)
library(RColorBrewer)
library(clusterProfiler)
library(org.Sc.sgd.db)
library(AnnotationDbi)
library(Rtsne)
library(vegan)
library(cowplot)
library(patchwork)
library(ComplexHeatmap)
library(circlize)


select <- dplyr::select
filter <- dplyr::filter
first  <- dplyr::first

BASE    <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all"
OUT_DIR <- "../results"
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
set.seed(42)

pub_theme <- theme_classic(base_size = 9) +
  theme(
    axis.title = element_text(size = 9,  face = "bold", color = "black"),
    axis.text  = element_text(size = 8,  color = "black"),
    axis.line  = element_line(linewidth = 0.4, color = "black"),
    axis.ticks = element_line(linewidth = 0.35, color = "black"),
    plot.title = element_text(size = 9,  face = "bold", hjust = 0.5),
    panel.grid       = element_blank(),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```


# ─────────────────────────────────────────────────────────────────────
# Input preparation: pseudobulk counts, gene ID harmonization, metadata
#
# 1. Load YPD.1 pseudobulk counts (genes x strains; counts already summed
#    across cells within each strain, so each column is one bulk sample).
#
# 2. Harmonize gene IDs to systematic ORF names. Rows already matching the
#    ORF pattern (Y__###W/C) are kept as-is; common names are mapped via the
#    custom annotation table. Rows that are neither are dropped (not_converted).
#
# 3. Restrict to nuclear standard chromosomes — mitochondrial and Marker
#    features are removed — then round to integers for DESeq2.
#
# 4. Build an ORF -> common-name lookup (orf_to_name) for figure labels,
#    using the custom annotation first and falling back to SGD (org.Sc.sgd.db)
#    for any ORFs it doesn't cover. Wrapped in tryCatch so a failed SGD query
#    degrades to ORF-only labels rather than stopping the script.
#
# 5. Build sample_meta (colData): one row per strain with its clade, filtered
#    to strains present in the count matrix and REORDERED to match column
#    order — DESeq2 pairs colData rows to count columns by position, not by
#    name, so this ordering is load-bearing. stopifnot() guards it.
#
# 6. Define clade_colors, a named palette (Paired, interpolated to the number
#    of clades) so clade colors stay consistent across all Figure 2 panels.
#
# Outputs used downstream: raw_counts_mat, orf_to_name, sample_meta, clade_colors
# ─────────────────────────────────────────────────────────────────────
```{r}
# ── Load, ORF-convert, filter to nuclear standard chromosomes ─────────
raw_counts <- read.table(
  file.path(BASE, "YPD.1/fastlmm/data/pheno_data/pseudobulk_reliable_transcripts_YPD.txt"),
  header = TRUE, sep = "\t", row.names = 1, check.names = FALSE, quote = ""
)
cat("Loaded:", nrow(raw_counts), "genes x", ncol(raw_counts), "strains\n")

all_custom_gene_annotations <- read.csv(
  "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"
)

common_to_orf <- all_custom_gene_annotations %>%
  filter(!is.na(gene_name), !is.na(gene_id)) %>%
  distinct(gene_name, gene_id) %>%
  deframe()

expr_gene_ids <- rownames(raw_counts)
is_orf <- grepl("^Y[A-Z]{2}\\d{3}[WC](-[A-Z])?$", expr_gene_ids)
expr_orf_ids <- ifelse(
  is_orf, expr_gene_ids,
  ifelse(expr_gene_ids %in% names(common_to_orf), common_to_orf[expr_gene_ids], expr_gene_ids)
)
not_converted <- expr_gene_ids[!is_orf & !expr_gene_ids %in% names(common_to_orf)]
rownames(raw_counts) <- expr_orf_ids
raw_counts <- raw_counts[!rownames(raw_counts) %in% not_converted, ]

gene_location <- all_custom_gene_annotations %>%
  filter(type == "CDS", !is.na(gene_id)) %>%
  group_by(gene_id) %>%
  summarise(gene_name = first(gene_name), chr = first(seqnames), .groups = "drop")

standard_orfs <- gene_location %>% filter(!chr %in% c("Mito", "Marker")) %>% pull(gene_id)
raw_counts    <- raw_counts[rownames(raw_counts) %in% standard_orfs, ]
raw_counts_mat <- round(as.matrix(raw_counts))
cat("After ORF conversion + chromosome filter:", nrow(raw_counts_mat), "genes\n")

# ORF -> common name lookup (custom annotation, SGD fallback)
orf_to_name_custom <- all_custom_gene_annotations %>%
  filter(!is.na(gene_id), !is.na(gene_name)) %>%
  distinct(gene_id, gene_name) %>%
  deframe()

orfs_missing_name <- rownames(raw_counts_mat)[!rownames(raw_counts_mat) %in% names(orf_to_name_custom)]
sgd_lookup <- tryCatch(
  AnnotationDbi::select(org.Sc.sgd.db, keys = orfs_missing_name,
                        columns = "GENENAME", keytype = "ORF"),
  error = function(e) data.frame(ORF = character(), GENENAME = character())
)
orf_to_name_sgd <- sgd_lookup %>% filter(!is.na(GENENAME)) %>% distinct(ORF, GENENAME) %>% deframe()
orf_to_name <- c(orf_to_name_custom,
                 orf_to_name_sgd[!names(orf_to_name_sgd) %in% names(orf_to_name_custom)])

# ── Clade metadata ──────────────────────────────────────────────────
strains_sc <- read.csv(file.path(BASE, "StrainsforSC-RNAseq.csv"), header = TRUE)

sample_meta <- strains_sc %>%
  select(strain = Standardized.name, clade = Clades) %>%
  mutate(clade = str_trim(clade)) %>%
  filter(strain %in% colnames(raw_counts_mat)) %>%
  distinct(strain, clade) %>%
  arrange(match(strain, colnames(raw_counts_mat))) %>%
  column_to_rownames("strain")



# sample_meta <- strains_sc %>%
#   select(strain = Standardized.name, clade = Clades) %>%
#   mutate(clade = str_trim(clade)) %>%
#   # ── Collapse singletons into broader clades ──────────────────────
#   mutate(clade = case_when(
#     clade %in% c("1. Wine/European", "1. Wine/European (subclade 1)") ~ "Wine/European",
#     clade %in% c("16. CHNI", "15. CHNII", "14. CHNIII", "17. Taiwanese")       ~ "CHNT",
#     TRUE ~ clade
#   )) %>%
#   filter(strain %in% colnames(raw_counts_mat)) %>%
#   distinct(strain, clade) %>%
#   arrange(match(strain, colnames(raw_counts_mat))) %>%
#   column_to_rownames("strain")

# ── Coverage check: does every count-matrix strain have a clade? ──────
mat_strains  <- colnames(raw_counts_mat)
meta_strains <- rownames(sample_meta)

missing_meta <- setdiff(mat_strains, meta_strains)   # in matrix, absent from CSV
na_clade     <- meta_strains[is.na(sample_meta$clade) | sample_meta$clade == ""]
extra_csv    <- setdiff(strains_sc$Standardized.name, mat_strains)  # in CSV, not sequenced

cat("Strains in raw_counts_mat:      ", length(mat_strains), "\n")
cat("Strains with clade annotation:  ", length(meta_strains), "\n")
cat("Missing from strain sheet:      ", length(missing_meta), "\n")
cat("Present but clade is NA/blank:  ", length(na_clade), "\n")
cat("In sheet but not in matrix:     ", length(extra_csv), "\n")
cat("Distinct clades:                ", length(unique(na.omit(sample_meta$clade))), "\n")

if (length(missing_meta) > 0)
  cat("  -> no metadata row:", paste(missing_meta, collapse = ", "), "\n")
if (length(na_clade) > 0)
  cat("  -> clade missing:  ", paste(na_clade, collapse = ", "), "\n")

# Strict check — identical() catches length mismatch AND misordering.
# all(==) alone recycles on unequal lengths and can pass while wrong.
stopifnot(identical(meta_strains, mat_strains))

# ── Clade palette ─────────────────────────────────────────────────────
all_clade_levels <- sort(unique(sample_meta$clade))
all_clade_levels <- all_clade_levels[!is.na(all_clade_levels)]
clade_colors <- setNames(
  colorRampPalette(brewer.pal(12, "Paired"))(length(all_clade_levels)),
  all_clade_levels
)
```

#create a deseq dataset and use vst for normalization

# ─────────────────────────────────────────────────────────────────────
# Variance-stabilizing transform (VST) for visualization only.
#
# Builds an intercept-only DESeqDataSet (design = ~1) purely as a container
# for the transform — no model is fit and no testing happens here. DE testing
# is done separately per clade in run_clade_dge() with design = ~ group,
# operating on RAW counts with DESeq2's own internal normalization.
#
# vst() stabilizes the mean-variance relationship so highly- and lowly-
# expressed genes contribute comparably to distance metrics, and normalizes
# for library size. blind = TRUE estimates dispersions without reference to
# clade, keeping the transform independent of the labels the downstream
# clustering is being asked to recover.
#
# vst_mat is used for: Panel A heatmap (z-scored), Panel B t-SNE input,
# Panel C GAL/NCR boxplots, and Panel D gene selection.
# ─────────────────────────────────────────────────────────────────────
```{r}
# ── VST on the full matrix (used by Panels A, B, D) ────────────────
dds_full <- DESeqDataSetFromMatrix(countData = raw_counts_mat, colData = sample_meta, design = ~ 1) #it's a container only, to be used for vst_mat and that's it
vst_mat  <- SummarizedExperiment::assay(DESeq2::vst(dds_full, blind = TRUE))
cat("VST matrix:", nrow(vst_mat), "genes x", ncol(vst_mat), "strains\n")
```




```{r}
# ── Per-clade DGE (design ~ group), clades with >=2 strains only ──────
run_clade_dge <- function(clade_of_interest, counts_mat, meta,
                          padj_thresh = 0.05, lfc_thresh = 0.3) {
  n_focal <- sum(meta$clade == clade_of_interest, na.rm = TRUE)
  if (n_focal < 2) return(NULL)

  meta_de <- meta %>%
    mutate(group = factor(ifelse(clade == clade_of_interest, "focal", "rest"),
                          levels = c("rest", "focal")))
  dds <- DESeqDataSetFromMatrix(countData = counts_mat, colData = meta_de, design = ~ group)
  dds <- DESeq(dds, quiet = TRUE)
  res <- results(dds, contrast = c("group", "focal", "rest"), alpha = padj_thresh)

  as.data.frame(res) %>%
    rownames_to_column("gene") %>%
    filter(!is.na(padj)) %>%
    mutate(
      gene_name = ifelse(gene %in% names(orf_to_name), orf_to_name[gene], gene),
      direction = case_when(
        padj < padj_thresh & log2FoldChange >  lfc_thresh ~ "Upregulated",
        padj < padj_thresh & log2FoldChange < -lfc_thresh ~ "Downregulated",
        TRUE ~ "NS"),
      clade = clade_of_interest, n_focal = n_focal)
}

all_clades  <- sort(unique(sample_meta$clade)); all_clades <- all_clades[!is.na(all_clades)]

# ── Which clades are testable? ────────────────────────────────────────
clade_sizes <- sample_meta %>%
  rownames_to_column("strain") %>%
  filter(!is.na(clade)) %>%
  dplyr::count(clade, name = "n_strains")

singleton_clades <- clade_sizes %>% filter(n_strains < 2)
testable_clades  <- clade_sizes %>% filter(n_strains >= 2)
n_na_clade       <- sum(is.na(sample_meta$clade))

cat("Clades with clade annotation:   ", nrow(clade_sizes), "\n")
cat("  testable (>= 2 strains):      ", nrow(testable_clades), "\n")
cat("  skipped (only 1 strain):      ", nrow(singleton_clades),
    "  covering", sum(singleton_clades$n_strains), "strains\n")
cat("Strains with NA clade:          ", n_na_clade,
    "  (these fall into 'rest' in every contrast)\n")

if (nrow(singleton_clades) > 0) {
  cat("\nSingleton clades not tested:\n")
  print(singleton_clades, n = Inf)
}

dge_results <- lapply(all_clades, run_clade_dge, counts_mat = raw_counts_mat, meta = sample_meta)
names(dge_results) <- all_clades
dge_results <- Filter(Negate(is.null), dge_results)
dge_combined <- bind_rows(dge_results)

sig_degs <- dge_combined %>% filter(direction != "NS") %>% pull(gene) %>% unique()
degs_in_matrix <- intersect(sig_degs, rownames(vst_mat))

cat("\nClades tested:", length(dge_results),
    "| unique DEGs across clades:", length(degs_in_matrix), "\n")

# DEG count per clade — shows which contrasts carry the union
#summarizes how many DEGs each clade has, split by direction
dge_combined %>%
  filter(direction != "NS") %>%
  dplyr::count(clade, n_focal, direction) %>%
  pivot_wider(names_from = direction, values_from = n, values_fill = 0) %>%
  arrange(desc(Upregulated + Downregulated)) %>%
  print(n = Inf)


sign_deg <- dge_combined %>%
  filter(direction != "NS")

```




```{r}
# ── Panel A: strain-level DEG heatmap using ComplexHeatmap ───────
library(ComplexHeatmap)
library(circlize)

vst_degs_strain <- vst_mat[degs_in_matrix, , drop = FALSE]
scaled_A <- t(scale(t(vst_degs_strain)))
scaled_A[scaled_A >  3] <-  3
scaled_A[scaled_A < -3] <- -3

# ── Remove singleton clades not tested in DE analysis ─────────────
tested_clades <- dge_combined %>%
  distinct(clade, n_focal) %>%
  filter(n_focal > 1) %>%
  pull(clade)

cat("Tested clades (n_focal > 1):", length(tested_clades), "\n")
cat("Clades included:", paste(sort(tested_clades), collapse = ", "), "\n\n")

all_clades    <- unique(sample_meta$clade)
excluded_clades <- setdiff(all_clades, tested_clades)
cat("Excluded singleton clades:", length(excluded_clades), "\n")
cat("Clades excluded:", paste(sort(excluded_clades), collapse = ", "), "\n\n")

strains_to_keep  <- rownames(sample_meta)[sample_meta$clade %in% tested_clades]
strains_in_matrix <- intersect(strains_to_keep, colnames(scaled_A))

cat("Strains present in scaled matrix:  ", ncol(scaled_A), "\n")
cat("Strains after filtering:           ", length(strains_in_matrix), "\n")
cat("Strains removed:                   ", ncol(scaled_A) - length(strains_in_matrix), "\n\n")

scaled_A_filtered <- scaled_A[, strains_in_matrix, drop = FALSE]
cat("Final filtered matrix dimensions:", nrow(scaled_A_filtered), "genes x",
    ncol(scaled_A_filtered), "strains\n")

# ── Column annotation: clade colors (on filtered matrix) ──────────
col_ann_A <- HeatmapAnnotation(
  clade = sample_meta[colnames(scaled_A_filtered), "clade"],
  col = list(clade = clade_colors),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    clade = list(
      title = "Clade",
      title_gp = gpar(fontsize = 10, fontface = "bold"),
      labels_gp = gpar(fontsize = 8),
      ncol = 1
    )
  ),
  which = "column"
)

# Color function
col_fun_A <- colorRamp2(
  c(-3, -1.5, 0, 1.5, 3),
  c("#2166AC", "#92C5DE", "white", "#F4A582", "#B2182B")
)

# ── Main heatmap (scaled_A_filtered) ──────────────────────────────
fig2A <- Heatmap(
  scaled_A_filtered,           # <-- filtered matrix
  
  col  = col_fun_A,
  name = "z-score",
  
  cluster_rows    = TRUE,
  cluster_columns = TRUE,
  clustering_method_rows    = "ward.D2",
  clustering_method_columns = "ward.D2",
  
  top_annotation = col_ann_A,
  
  show_row_names   = FALSE,
  show_column_names = TRUE,
  column_names_rot = 90,       # 90 degrees as requested
  column_names_gp  = gpar(fontsize = 8),
  
  heatmap_legend_param = list(
    title      = "DEG expression",
    title_gp   = gpar(fontsize = 10, fontface = "bold"),
    labels_gp  = gpar(fontsize = 8),
    at         = c(-3, -1.5, 0, 1.5, 3),
    labels     = c("-3 (low)", "-1.5", "0 (mean)", "1.5", "+3 (high)"),
    legend_height = unit(4, "cm")
  ),
  
  # Title now uses filtered dimensions
  column_title = paste0("Strain-level DEG expression (YPD.1)\n",
                        nrow(scaled_A_filtered), " genes × ",
                        ncol(scaled_A_filtered), " strains"),
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  
  border = FALSE
)

# ── Draw and save ─────────────────────────────────────────────────
draw(fig2A,
     heatmap_legend_side      = "right",
     annotation_legend_side   = "right",
     merge_legend             = TRUE)

tiff(file.path(OUT_DIR, "Fig2A_strain_DEG_heatmap.tiff"),
     width = 10, height = 7, units = "in", res = 600, compression = "lzw")
draw(fig2A,
     heatmap_legend_side    = "right",
     annotation_legend_side = "right",
     merge_legend           = TRUE)
dev.off()
```


```{r}
# ── Panel A: strain-level DEG heatmap using ComplexHeatmap ───────
library(ComplexHeatmap)
library(circlize)

vst_degs_strain <- vst_mat[degs_in_matrix, , drop = FALSE]
scaled_A <- t(scale(t(vst_degs_strain)))
scaled_A[scaled_A >  3] <-  3
scaled_A[scaled_A < -3] <- -3

# ── Remove singleton clades not tested in DE analysis ─────────────
tested_clades <- dge_combined %>%
  distinct(clade, n_focal) %>%
  filter(n_focal > 1) %>%
  pull(clade)

cat("Tested clades (n_focal > 1):", length(tested_clades), "\n")
cat("Clades included:", paste(sort(tested_clades), collapse = ", "), "\n\n")

all_clades      <- unique(sample_meta$clade)
excluded_clades <- setdiff(all_clades, tested_clades)
cat("Excluded singleton clades:", length(excluded_clades), "\n")
cat("Clades excluded:", paste(sort(excluded_clades), collapse = ", "), "\n\n")

strains_to_keep   <- rownames(sample_meta)[sample_meta$clade %in% tested_clades]
strains_in_matrix <- intersect(strains_to_keep, colnames(scaled_A))

cat("Strains present in scaled matrix:  ", ncol(scaled_A), "\n")
cat("Strains after filtering:           ", length(strains_in_matrix), "\n")
cat("Strains removed:                   ", ncol(scaled_A) - length(strains_in_matrix), "\n\n")

scaled_A_filtered <- scaled_A[, strains_in_matrix, drop = FALSE]
cat("Final filtered matrix dimensions:", nrow(scaled_A_filtered), "genes x",
    ncol(scaled_A_filtered), "strains\n")

# ── Column annotation with clade colors AND clade name labels ─────
clade_labels <- as.character(sample_meta[colnames(scaled_A_filtered), "clade"])

col_ann_A <- HeatmapAnnotation(
  clade = clade_labels,
  col   = list(clade = clade_colors),
  show_annotation_name = FALSE,
  clade_name = anno_text(
    clade_labels,
    rot      = 90,
    gp       = gpar(fontsize = 7),
    location = 1,
    just     = "left",
    height   = max_text_height(clade_labels,
                               gp = gpar(fontsize = 7)) + unit(2, "mm")
  ),
  annotation_legend_param = list(
    clade = list(
      title      = "Clade",
      title_gp   = gpar(fontsize = 10, fontface = "bold"),
      labels_gp  = gpar(fontsize = 8),
      ncol       = 1
    )
  ),
  which = "column"
)

# ── Color function ────────────────────────────────────────────────
col_fun_A <- colorRamp2(
  c(-3, -1.5, 0, 1.5, 3),
  c("#2166AC", "#92C5DE", "white", "#F4A582", "#B2182B")
)

# ── Main heatmap ──────────────────────────────────────────────────
fig2A <- Heatmap(
  scaled_A_filtered,

  col  = col_fun_A,
  name = "z-score",

  cluster_rows              = TRUE,
  cluster_columns           = TRUE,
  clustering_method_rows    = "ward.D2",
  clustering_method_columns = "ward.D2",

  top_annotation = col_ann_A,

  show_row_names    = FALSE,
  show_column_names = TRUE,
  column_names_rot  = 90,
  column_names_side = "bottom",
  column_names_gp   = gpar(fontsize = 8),

  heatmap_legend_param = list(
    title         = "DEG expression",
    title_gp      = gpar(fontsize = 10, fontface = "bold"),
    labels_gp     = gpar(fontsize = 8),
    at            = c(-3, -1.5, 0, 1.5, 3),
    labels        = c("-3 (low)", "-1.5", "0 (mean)", "1.5", "+3 (high)"),
    legend_height = unit(4, "cm")
  ),

  column_title = paste0("Strain-level DEG expression (YPD.1)\n",
                        nrow(scaled_A_filtered), " genes × ",
                        ncol(scaled_A_filtered), " strains"),
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),

  border = FALSE
)

fig2A

#── Save ──────────────────────────────────────────────────────────
tiff(file.path(OUT_DIR, "Fig2B_strain_DEG_heatmap.tiff"),
     width = 10, height = 7, units = "in", res = 600, compression = "lzw")
draw(fig2A,
     heatmap_legend_side    = "right",
     annotation_legend_side = "right",
     merge_legend           = TRUE)
dev.off()
```
#looking into why some strain/clade have a very different degree, some strains within a clade are also different from each other
```{r}
# ── Investigate suspicious strains ────────────────────────────────
suspicious_strains <- c("CDI", "CDF", "BTG", "YDE", "ANH", "AVI")

# Load cell counts directly from the saved Seurat object
ypd1_obj <- readRDS(file.path(BASE, 
  "YPD.1/gene_expression_analysis/output_from_pseudobulk_fastlmm/Na_va_data_subsetted.rds"))

cell_counts <- ypd1_obj@meta.data %>%
  dplyr::count(strain = SNG.BEST.GUESS, name = "n_cells") %>%
  filter(strain %in% suspicious_strains) %>%
  arrange(n_cells)

cat("Cell counts for suspicious strains:\n")
print(cell_counts)
rm(ypd1_obj); gc(verbose = FALSE)   # drop the object immediately to free memory

# 2. Within-clade Spearman correlation on DEG expression
clade_pairs <- list(
  "22. Far East Russian"   = c("CDI", "CDF"),
  "4. Mediterranean oak"   = c("BTG"),
  "24. Asian islands"      = c("YDE"),
  "23. North American oak" = c("ANH", "AVI")
)

for (clade_name in names(clade_pairs)) {
  all_clade_strains <- rownames(sample_meta)[sample_meta$clade == clade_name]
  
  cat("\n---", clade_name, "---\n")
  cat("All strains in clade:", paste(all_clade_strains, collapse = ", "), "\n")
  
  strains_available <- intersect(all_clade_strains, colnames(vst_mat))
  
  if (length(strains_available) >= 2) {
    expr_sub <- vst_mat[degs_in_matrix, strains_available, drop = FALSE]
    cor_mat  <- cor(expr_sub, method = "spearman")
    cat("Within-clade Spearman correlations:\n")
    print(round(cor_mat, 3))
  } else {
    cat("Only", length(strains_available), "strain(s) available in matrix — cannot compute pairwise correlation\n")
  }
}

# 3. Compare suspicious strains' expression to their clade mean
cat("\n--- Expression deviation from clade mean ---\n")
for (s in suspicious_strains) {
  clade_s <- sample_meta[s, "clade"]
  if (is.na(clade_s)) next
  
  clade_strains <- rownames(sample_meta)[sample_meta$clade == clade_s]
  clade_strains <- intersect(clade_strains, colnames(vst_mat))
  
  if (length(clade_strains) < 2) next
  
  clade_mean <- rowMeans(vst_mat[degs_in_matrix, clade_strains, drop = FALSE])
  strain_expr <- vst_mat[degs_in_matrix, s]
  
  deviation <- mean(abs(strain_expr - clade_mean))
  cat(s, "(", clade_s, ") — mean absolute deviation from clade mean:", round(deviation, 3), "\n")
}
```

```{r}
# How many of the 2310 DEGs were actually significant for these clades?
dge_combined %>%
  filter(clade %in% c("22. Far East Russian", "4. Mediterranean oak",
                      "24. Asian islands", "23. North American oak"),
         direction != "NS") %>%
  dplyr::count(clade, direction) %>%
  pivot_wider(names_from = direction, values_from = n, values_fill = 0) %>%
  mutate(total_own_degs = Downregulated + Upregulated,
         pct_of_union   = round(100 * total_own_degs / length(degs_in_matrix), 1))
```



```{r}
# ── GO ORA per clade — used to select Panel C's gene set ──────────────
run_ora <- function(genes, universe, keytype = "ORF") {
  if (length(genes) < 5) return(NULL)
  enrichGO(gene = genes, universe = universe, OrgDb = org.Sc.sgd.db,
           keyType = keytype, ont = "BP", pAdjustMethod = "BH",
           pvalueCutoff = 0.05, qvalueCutoff = 0.2, readable = FALSE)
}

universe_genes <- unique(dge_combined$gene)
ora_results <- lapply(names(dge_results), function(cl) {
  list(up   = run_ora(dge_results[[cl]]$gene[dge_results[[cl]]$direction == "Upregulated"],   universe_genes),
       down = run_ora(dge_results[[cl]]$gene[dge_results[[cl]]$direction == "Downregulated"], universe_genes))
})
names(ora_results) <- names(dge_results)

ora_combined <- lapply(names(ora_results), function(cl) {
  bind_rows(
    if (!is.null(ora_results[[cl]]$up))   as.data.frame(ora_results[[cl]]$up)   %>% mutate(direction = "Upregulated",   clade = cl),
    if (!is.null(ora_results[[cl]]$down)) as.data.frame(ora_results[[cl]]$down) %>% mutate(direction = "Downregulated", clade = cl)
  )
}) %>% bind_rows()

write_csv(ora_combined, file.path(OUT_DIR, "Fig2_clade_GO_ORA_all.csv"))

# Search enriched BP terms for galactose- / nitrogen-catabolite-repression-related
# descriptions. Only the regulon(s) that actually show enrichment get plotted.
gal_hits <- ora_combined %>% filter(str_detect(tolower(Description), "galactose"))
ncr_hits <- ora_combined %>% filter(str_detect(tolower(Description),
                          "nitrogen catabolite|allantoin|urea catabolic|nitrogen compound catabolic"))

cat("GO terms matching GAL regulon:", nrow(gal_hits), "\n")
cat("GO terms matching NCR regulon:", nrow(ncr_hits), "\n")

# Canonical regulon members, as fallback if ORA finds nothing for a regulon
# (e.g. too few DEGs survive FDR in any clade) — intersected with detected genes.
gal_canonical_names <- c("GAL1","GAL2","GAL3","GAL4","GAL7","GAL10","GAL80")
ncr_canonical_names <- c("MEP1","MEP2","GAP1","DAL5","DUR1,2","DUR3","PUT1","PUT4","GLN1","GLN3","GAT1")

name_to_orf <- setNames(names(orf_to_name), orf_to_name)  # invert lookup

resolve_focus_genes <- function(go_hits, canonical_names) {
  from_go <- if (nrow(go_hits) > 0) unique(unlist(str_split(go_hits$geneID, "/"))) else character(0)
  from_canonical <- intersect(name_to_orf[canonical_names], rownames(vst_mat))
  genes <- union(intersect(from_go, rownames(vst_mat)), from_canonical)
  genes
}

gal_focus_genes <- resolve_focus_genes(gal_hits, gal_canonical_names)
ncr_focus_genes <- resolve_focus_genes(ncr_hits, ncr_canonical_names)
cat("GAL focus genes resolved:", length(gal_focus_genes), "\n")
cat("NCR focus genes resolved:", length(ncr_focus_genes), "\n")
```




```{r}
# ── Panel A: strain-level DEG heatmap, clade strip, legend fixed ───────
# vst_degs_strain <- vst_mat[degs_in_matrix, , drop = FALSE]
# scaled_A <- t(scale(t(vst_degs_strain)))
# scaled_A[scaled_A >  3] <-  3
# scaled_A[scaled_A < -3] <- -3
# 
# hm_A   <- plot_heatmap_gg(scaled_A, xlab = "Strain", ylab = "Gene",
#                           show_colnames = TRUE, show_rownames = FALSE,
#                           limits = c(-3, 3), x_font_size = 6)
# strip_A <- make_clade_strip(hm_A$col_order, sample_meta, clade_colors)

# guides = "collect" pulls both plots' legends (clade fill + z-score fill)
# into one column on the right, instead of each plot fighting for its own
# legend space — this is what was pushing the clade legend off-panel before.
# fig2A <- (strip_A / hm_A$plot) +
#   plot_layout(heights = c(0.05, 1), guides = "collect") &
#   theme(legend.position = "right")
# 
# print(fig2A)
# ggsave(file.path(OUT_DIR, "Fig2A_strain_DEG_heatmap.tiff"), fig2A, width = 10, height = 7,
#        units = "in", dpi = 600, device = "tiff", compression = "lzw")
```


```{r}
# ── t-SNE panel, reusable for either gene set ─────────────────────────
plot_tsne_geneset <- function(genes, label, vst_mat, sample_meta,
                              clade_colors, seed = 42,
                              label_centroids = TRUE) {
  genes <- intersect(genes, rownames(vst_mat))
  if (length(genes) < 3) return(NULL)

  tsne_input <- t(vst_mat[genes, , drop = FALSE])
  set.seed(seed)
  tsne_res <- Rtsne(tsne_input, dims = 2,
                    perplexity = min(15, floor((nrow(tsne_input) - 1) / 3)),
                    max_iter = 1000, check_duplicates = FALSE, verbose = FALSE)

  tsne_df <- data.frame(tSNE1 = tsne_res$Y[, 1], tSNE2 = tsne_res$Y[, 2],
                        strain = rownames(tsne_input)) %>%
    left_join(sample_meta %>% rownames_to_column("strain"), by = "strain")

  centroids <- tsne_df %>% filter(!is.na(clade)) %>%
    group_by(clade) %>%
    summarise(tSNE1 = mean(tSNE1), tSNE2 = mean(tSNE2), n = n(), .groups = "drop") %>%
    filter(n >= 2)

  p <- ggplot() +
    geom_point(data = filter(tsne_df, is.na(clade)), aes(tSNE1, tSNE2),
               shape = 1, size = 2, color = "grey50") +
    geom_point(data = filter(tsne_df, !is.na(clade)), aes(tSNE1, tSNE2, color = clade),
               size = 2.3, alpha = 0.85)

  # Centroid labels only where clades actually cluster — labelling a
  # structureless embedding invites the reader to see grouping that isn't there.
  if (label_centroids) {
    p <- p + ggrepel::geom_text_repel(
      data = centroids, aes(tSNE1, tSNE2, label = clade),
      size = 2.4, fontface = "bold", color = "black",
      max.overlaps = 20, segment.size = 0.3)
  }

  p +
    scale_color_manual(values = clade_colors, na.value = "grey50", name = "Clade") +
    guides(color = guide_legend(override.aes = list(size = 3, alpha = 1), ncol = 1)) +
    labs(x = "t-SNE 1", y = "t-SNE 2",
         title = paste0(label, " (", length(genes), " genes)")) +
    pub_theme +
    theme(legend.position = "right",
          legend.key.size = unit(0.35, "cm"),
          legend.text     = element_text(size = 7),
          legend.title    = element_text(size = 8, face = "bold"))
}

fig2B_coexpr <- plot_tsne_geneset(coexpr_genes, "Co-expression",
                                  vst_mat, sample_meta, clade_colors,
                                  label_centroids = FALSE)

degs_clean <- setdiff(degs_in_matrix, coexpr_genes)
fig2B_deg  <- plot_tsne_geneset(degs_clean, "Differential expression",
                                vst_mat, sample_meta, clade_colors,
                                label_centroids = TRUE)

print(fig2B_coexpr)
print(fig2B_deg)


# Exclude co-expression genes from the DEG set (as in Caudal et al.)
degs_clean <- setdiff(degs_in_matrix, coexpr_genes)
cat("DEGs after removing co-expression overlap:", length(degs_clean),
    "(removed", length(degs_in_matrix) - length(degs_clean), ")\n")

# Which clades contributed the overlapping genes?
dge_combined %>%
  filter(gene %in% intersect(degs_in_matrix, coexpr_genes), direction != "NS") %>%
  dplyr::count(clade, sort = TRUE) %>%
  print(n = Inf)

fig2B_deg <- plot_tsne_geneset(degs_clean, "Differential expression",
                               vst_mat, sample_meta, clade_colors)

# ggsave(file.path(OUT_DIR, "Fig2B_tsne_coexpression.tiff"), fig2B_coexpr,
#        width = 6.2, height = 4, units = "in", dpi = 600,
#        device = "tiff", compression = "lzw")
# ggsave(file.path(OUT_DIR, "Fig2B_tsne_DEG.tiff"), fig2B_deg,
#        width = 6.2, height = 4, units = "in", dpi = 600,
#        device = "tiff", compression = "lzw")

cat("Overlap between co-expression and DEG sets:",
    length(intersect(coexpr_genes, degs_in_matrix)), "genes\n")
```
```{r}
fig2B_deg 
```




```{r}
# ── Clade cohesion test: are strains from the same clade closer than chance? ──
library(Rtsne)

clade_cohesion <- function(coords, meta, n_perm = 999, seed = 42) {
  coords <- as.matrix(coords)
  strains <- rownames(coords)

  # match strains to clades, keeping only those with annotation
  clade_vec <- meta[strains, "clade"]
  ok        <- !is.na(clade_vec)
  coords    <- coords[ok, , drop = FALSE]
  clade_vec <- clade_vec[ok]

  # keep clades with >= 2 strains (singletons have no within-clade distance)
  tab  <- table(clade_vec)
  keep <- clade_vec %in% names(tab)[tab >= 2]
  coords    <- coords[keep, , drop = FALSE]
  clade_vec <- clade_vec[keep]

  # Euclidean distance in whatever dimensionality was passed
  d    <- as.matrix(dist(coords))
  same <- outer(clade_vec, clade_vec, "==")
  diag(same) <- NA
  obs <- mean(d[which(same)], na.rm = TRUE)

  set.seed(seed)
  null <- replicate(n_perm, {
    shuf <- sample(clade_vec)
    s <- outer(shuf, shuf, "=="); diag(s) <- NA
    mean(d[which(s)], na.rm = TRUE)
  })

  list(observed_within = obs,
       null_mean       = mean(null),
       ratio           = obs / mean(null),
       p_value         = (sum(null <= obs) + 1) / (n_perm + 1),
       n_strains       = nrow(coords),
       n_clades        = length(unique(clade_vec)))
}

# ── t-SNE coordinates for a gene set ──────────────────────────────────
get_tsne_coords <- function(genes, vst_mat, seed = 42) {
  genes <- intersect(genes, rownames(vst_mat))
  input <- t(vst_mat[genes, , drop = FALSE])
  set.seed(seed)
  r <- Rtsne(input, dims = 2,
             perplexity = min(15, floor((nrow(input) - 1) / 3)),
             max_iter = 1000, check_duplicates = FALSE, verbose = FALSE)
  coords <- r$Y
  rownames(coords) <- rownames(input)
  coords
}

# ── Run across several seeds — t-SNE layout varies, so check stability ──
seeds <- c(1, 42, 123, 2024, 777)

cohesion_results <- lapply(seeds, function(s) {
  cx <- clade_cohesion(get_tsne_coords(coexpr_genes,   vst_mat, seed = s),
                       sample_meta, seed = s)
  dg <- clade_cohesion(get_tsne_coords(degs_in_matrix, vst_mat, seed = s),
                       sample_meta, seed = s)
  bind_rows(
    tibble(seed = s, gene_set = "Co-expression",
           n_genes = length(intersect(coexpr_genes, rownames(vst_mat))),
           ratio = cx$ratio, p_value = cx$p_value,
           observed = cx$observed_within, null_mean = cx$null_mean),
    tibble(seed = s, gene_set = "Differential expression",
           n_genes = length(intersect(degs_in_matrix, rownames(vst_mat))),
           ratio = dg$ratio, p_value = dg$p_value,
           observed = dg$observed_within, null_mean = dg$null_mean)
  )
}) %>% bind_rows()

cat("\n=== Clade cohesion in t-SNE space, across seeds ===\n")
print(cohesion_results, n = Inf)

cat("\n=== Summary across seeds ===\n")
cohesion_results %>%
  group_by(gene_set) %>%
  summarise(n_genes    = first(n_genes),
            mean_ratio = mean(ratio),
            min_ratio  = min(ratio),
            max_ratio  = max(ratio),
            n_sig      = sum(p_value < 0.05),
            .groups = "drop") %>%
  print()

# ── Same test on full expression space (no embedding) ──────────────────
# t-SNE distances are a 2-D distortion of the real geometry. This version
# uses Euclidean distance across all genes in the set, so it is deterministic
# and free of embedding artifacts — the more defensible number of the two.
cohesion_full_space <- function(genes, vst_mat, meta, n_perm = 999) {
  genes  <- intersect(genes, rownames(vst_mat))
  coords <- t(vst_mat[genes, , drop = FALSE])   # strains x genes
  clade_cohesion(coords, meta, n_perm = n_perm)
}

cat("\n=== Clade cohesion in FULL expression space (deterministic) ===\n")
full_space <- bind_rows(
  tibble(gene_set = "Co-expression",
         !!!cohesion_full_space(coexpr_genes,   vst_mat, sample_meta)[
             c("ratio", "p_value", "n_strains", "n_clades")]),
  tibble(gene_set = "Differential expression",
         !!!cohesion_full_space(degs_in_matrix, vst_mat, sample_meta)[
             c("ratio", "p_value", "n_strains", "n_clades")])
)
print(full_space)

#coexpression only
coexpr_only <- setdiff(coexpr_genes, degs_in_matrix)
cat("Co-expression genes not overlapping DEGs:", length(coexpr_only), "\n")
clade_cohesion(get_tsne_coords(coexpr_only, vst_mat), sample_meta)
```
```{r}
#random gene set
set.seed(42)
random_genes <- sample(setdiff(rownames(vst_mat), degs_in_matrix), length(coexpr_only))
clade_cohesion(get_tsne_coords(random_genes, vst_mat), sample_meta)
```



```{r}
# ── Panel B: t-SNE, no title label, no gene-count annotation ───────────
tsne_input <- t(vst_mat[degs_in_matrix, , drop = FALSE])
set.seed(42)
tsne_res <- Rtsne(tsne_input, dims = 2,
                  perplexity = min(15, floor((nrow(tsne_input) - 1) / 3)),
                  max_iter = 1000, check_duplicates = FALSE, verbose = FALSE)

tsne_df <- data.frame(tSNE1 = tsne_res$Y[, 1], tSNE2 = tsne_res$Y[, 2],
                      strain = rownames(tsne_input)) %>%
  left_join(sample_meta %>% rownames_to_column("strain"), by = "strain")

centroids_B <- tsne_df %>% filter(!is.na(clade)) %>%
  group_by(clade) %>% summarise(tSNE1 = mean(tSNE1), tSNE2 = mean(tSNE2), n = n(), .groups = "drop") %>%
  filter(n >= 2)

fig2B <- ggplot() +
  geom_point(data = filter(tsne_df, is.na(clade)), aes(tSNE1, tSNE2),
            shape = 1, size = 2, color = "grey50") +
  geom_point(data = filter(tsne_df, !is.na(clade)), aes(tSNE1, tSNE2, color = clade),
            size = 2.3, alpha = 0.85) +
  ggrepel::geom_text_repel(data = centroids_B, aes(tSNE1, tSNE2, label = clade),
                           size = 2.4, fontface = "bold", color = "black",
                           max.overlaps = 20, segment.size = 0.3) +
  scale_color_manual(values = clade_colors, na.value = "grey50", name = "Clade",
                     breaks = names(clade_colors)) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1), ncol = 1)) +
  labs(x = "t-SNE 1", y = "t-SNE 2") +
  pub_theme +
  theme(legend.position = "right",
        legend.key.size = unit(0.35, "cm"),
        legend.text  = element_text(size = 7),
        legend.title = element_text(size = 8, face = "bold"))

print(fig2B)
# wider canvas so the added legend doesn't compress the plotting area
ggsave(file.path(OUT_DIR, "Fig2B_tsne_clade.tiff"), fig2B, width = 6.2, height = 4,
       units = "in", dpi = 600, device = "tiff", compression = "lzw")
```

```{r}
# ── Panel C1: GAL and NCR as two SEPARATE figures, 3 facets/line, ──────
# shared y-axis within each figure (fixed scale, not free_y).

plot_focus_regulon <- function(genes, regulon_label) {
  if (length(genes) == 0) return(NULL)

  df <- as.data.frame(t(vst_mat[genes, , drop = FALSE])) %>%
    rownames_to_column("strain") %>%
    pivot_longer(-strain, names_to = "gene", values_to = "expr") %>%
    mutate(gene_name = ifelse(gene %in% names(orf_to_name), orf_to_name[gene], gene)) %>%
    left_join(sample_meta %>% rownames_to_column("strain"), by = "strain") %>%
    filter(!is.na(clade))

  ggplot(df, aes(x = clade, y = expr, fill = clade)) +
    geom_boxplot(outlier.size = 0.5, linewidth = 0.3) +
    facet_wrap(~ gene_name, ncol = 3) +               # fixed shared y-axis, 3 per row
    scale_fill_manual(values = clade_colors, guide = "none") +
    labs(title = paste0(regulon_label, " (YPD.1)"), x = NULL, y = "VST expression") +
    pub_theme +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7))
}

fig2C_GAL <- plot_focus_regulon(gal_focus_genes, "GAL regulon")
fig2C_NCR <- plot_focus_regulon(ncr_focus_genes, "NCR regulon (nitrogen catabolite repression)")

print(fig2C_GAL)
print(fig2C_NCR)

n_row_GAL <- ceiling(length(gal_focus_genes) / 3)
n_row_NCR <- ceiling(length(ncr_focus_genes) / 3)

ggsave(file.path(OUT_DIR, "Fig2C1_GAL_regulon.tiff"), fig2C_GAL, width = 7.5, height = 2.6 * n_row_GAL,
       units = "in", dpi = 600, device = "tiff", compression = "lzw")
ggsave(file.path(OUT_DIR, "Fig2C1_NCR_regulon.tiff"), fig2C_NCR, width = 7.5, height = 2.6 * n_row_NCR,
       units = "in", dpi = 600, device = "tiff", compression = "lzw")
```
```{r}
# ── Test normality for GAL genes ────

# 1. Shapiro-Wilk test (statistical test for normality)
cat("\n=== Shapiro-Wilk normality test (p > 0.05 = normal) ===\n")

shapiro_results <- lapply(gal_focus_genes, function(g) {
  gene_name <- ifelse(g %in% names(orf_to_name), orf_to_name[g], g)
  expr_vals <- as.numeric(vst_mat[g, ])
  test <- shapiro.test(expr_vals)
  
  tibble(
    gene = g,
    gene_name = gene_name,
    p.value = test$p.value,
    normal = test$p.value > 0.05
  )
}) %>% bind_rows()

print(shapiro_results)
cat("Genes that appear normal:", sum(shapiro_results$normal), "/", nrow(shapiro_results), "\n")

# 2. Visual check with Q-Q plots
library(ggplot2)

qq_plots <- lapply(gal_focus_genes[1:3], function(g) {  # just first 3 to avoid too many plots
  gene_name <- ifelse(g %in% names(orf_to_name), orf_to_name[g], g)
  expr_vals <- as.numeric(vst_mat[g, ])
  
  df_qq <- data.frame(
    sample = qqnorm(expr_vals, plot.it = FALSE)$x,
    theoretical = qqnorm(expr_vals, plot.it = FALSE)$y
  )
  
  ggplot(df_qq, aes(sample = expr_vals)) +
    stat_qq() +
    stat_qq_line() +
    labs(title = gene_name, x = "Theoretical", y = "Sample") +
    theme_minimal()
})

print(qq_plots)
```



```{r}
# ── Investigate GAL1 in detail ────
gal1_data <- as.data.frame(t(vst_mat["YBR020W", , drop = FALSE])) %>%
  rownames_to_column("strain") %>%
  pivot_longer(-strain, names_to = "gene", values_to = "expr") %>%
  left_join(sample_meta %>% rownames_to_column("strain"), by = "strain") %>%
  filter(!is.na(clade))

# Breakdown by clade
cat("\n=== GAL1 expression by clade ===\n")
gal1_summary <- gal1_data %>%
  group_by(clade) %>%
  summarise(
    n = n(),
    mean = mean(expr),
    median = median(expr),
    sd = sd(expr),
    min = min(expr),
    max = max(expr),
    .groups = "drop"
  ) %>%
  arrange(desc(mean))

print(gal1_summary)

# Manual comparison: French dairy vs all others
cat("\n=== French dairy vs Others (Wilcoxon test) ===\n")
gal1_data_grouped <- gal1_data %>%
  mutate(group = ifelse(clade == "5. French dairy", "French dairy", "Others"))

french_vs_others <- wilcox.test(expr ~ group, data = gal1_data_grouped)
cat("p-value:", french_vs_others$p.value, "\n")
cat("Mean French dairy:", mean(gal1_data_grouped$expr[gal1_data_grouped$group == "French dairy"]), "\n")
cat("Mean Others:", mean(gal1_data_grouped$expr[gal1_data_grouped$group == "Others"]), "\n")
```
```{r}
# # ── French dairy vs Others for all significant GAL genes ────
# gal_sig_genes <- filter(gal_overall, significant == TRUE)$gene
# 
# results_french_vs_others <- lapply(gal_sig_genes, function(g) {
#   gene_name <- filter(gal_overall, gene == g)$gene_name[1]
#   
#   df <- as.data.frame(t(vst_mat[g, , drop = FALSE])) %>%
#     rownames_to_column("strain") %>%
#     pivot_longer(-strain, names_to = "gene", values_to = "expr") %>%
#     left_join(sample_meta %>% rownames_to_column("strain"), by = "strain") %>%
#     filter(!is.na(clade)) %>%
#     mutate(group = ifelse(clade == "5. French dairy", "French dairy", "Others"))
#   
#   test <- wilcox.test(expr ~ group, data = df)
#   
#   tibble(
#     gene = g,
#     gene_name = gene_name,
#     p.value = test$p.value,
#     mean_french = mean(df$expr[df$group == "French dairy"]),
#     mean_others = mean(df$expr[df$group == "Others"]),
#     diff = mean_french - mean_others
#   )
# }) %>% bind_rows()
# 
# cat("\n=== GAL genes: French dairy vs Others ===\n")
# print(results_french_vs_others)
# 
# write_csv(results_french_vs_others, file.path(OUT_DIR, "GAL_French_dairy_comparison.csv"))
```




#function created
```{r}
# ── Publication-ready volcano plot ────────────────────────────────
plot_volcano <- function(res_df, clade_label, padj_thresh = 0.05,
                         lfc_thresh = 0.3, top_n_labels = 15) {
  sig       <- res_df %>% filter(direction != "NS")
  top_genes <- sig %>% arrange(padj) %>% slice_head(n = top_n_labels)
  n_up      <- sum(sig$direction == "Upregulated")
  n_down    <- sum(sig$direction == "Downregulated")

  ggplot(res_df, aes(log2FoldChange, -log10(padj),
                     color = direction, size = -log10(padj))) +
    # NS points first (bottom layer)
    geom_point(data = filter(res_df, direction == "NS"),
               alpha = 0.3, size = 0.8, color = "grey80") +
    # Significant points on top
    geom_point(data = filter(res_df, direction != "NS"),
               alpha = 0.7) +
    geom_hline(yintercept = -log10(padj_thresh),
               linetype   = "dashed", color = "grey50", linewidth = 0.35) +
    geom_vline(xintercept = c(-lfc_thresh, lfc_thresh),
               linetype   = "dashed", color = "grey50", linewidth = 0.35) +
    ggrepel::geom_text_repel(
      data          = top_genes,
      aes(label     = gene_name),
      size          = 2.8,
      color         = "black",
      fontface      = "italic",
      max.overlaps  = 25,
      segment.color = "grey60",
      segment.size  = 0.3,
      box.padding   = 0.4,
      point.padding = 0.3,
      force         = 2
    ) +
    scale_color_manual(
      values = c(Upregulated   = "#C0392B",
                 Downregulated = "#2471A3",
                 NS            = "grey80"),
      labels = c(Upregulated   = paste0("Up (", n_up, ")"),
                 Downregulated = paste0("Down (", n_down, ")"),
                 NS            = "NS"),
      breaks = c("Upregulated", "Downregulated", "NS")  # legend order
    ) +
    scale_size_continuous(range = c(0.5, 3.5), guide = "none") +
    labs(
      title    = clade_label,
      subtitle = paste0(unique(res_df$n_focal),
                        " strains vs. rest  |  FDR < ",
                        padj_thresh, ", |log₂FC| > ", lfc_thresh),
      x        = expression(log[2]~"fold change (focal vs. rest)"),
      y        = expression(-log[10]~"(adjusted"~italic(p)~")"),
      color    = NULL
    ) +
    theme_classic(base_size = 10, base_family = "Arial") +
    theme(
      # Title
      plot.title         = element_text(size = 11, face = "bold",
                                        color = "black", hjust = 0),
      plot.subtitle      = element_text(size = 8,  color = "grey40",
                                        hjust = 0, margin = margin(b = 6)),
      # Axes
      axis.title         = element_text(size = 10, face = "bold", color = "black"),
      axis.text          = element_text(size = 9,  color = "black"),
      axis.line          = element_line(linewidth = 0.4, color = "black"),
      axis.ticks         = element_line(linewidth = 0.35, color = "black"),
      # Legend
      legend.position    = "bottom",
      legend.text        = element_text(size = 9),
      legend.key.size    = unit(0.4, "cm"),
      legend.spacing.x   = unit(0.3, "cm"),
      # Panel
      panel.grid         = element_blank(),
      plot.background    = element_rect(fill = "white", color = NA),
      panel.background   = element_rect(fill = "white", color = NA),
      plot.margin        = margin(8, 10, 6, 8)
    )
}

# ── Publication-ready GO dot plot ─────────────────────────────────
plot_go_dotplot <- function(ora_up, ora_down, clade_label, top_n = 8) {
  df_up   <- if (!is.null(ora_up)   && nrow(as.data.frame(ora_up))   > 0)
               as.data.frame(ora_up)   %>% arrange(p.adjust) %>%
               slice_head(n = top_n)  %>% mutate(direction = "Upregulated")
             else NULL
  df_down <- if (!is.null(ora_down) && nrow(as.data.frame(ora_down)) > 0)
               as.data.frame(ora_down) %>% arrange(p.adjust) %>%
               slice_head(n = top_n)  %>% mutate(direction = "Downregulated")
             else NULL

  df <- bind_rows(df_up, df_down)
  if (is.null(df) || nrow(df) == 0) return(NULL)

  df <- df %>%
    mutate(
      GeneRatio_num = sapply(GeneRatio, function(x) {
        p <- strsplit(x, "/")[[1]]
        as.numeric(p[1]) / as.numeric(p[2])
      }),
      Description = str_wrap(Description, width = 40),
      direction   = factor(direction, levels = c("Upregulated", "Downregulated")),
      Description = reorder(Description, GeneRatio_num)
    )

  p <- ggplot(df, aes(GeneRatio_num, Description,
                      color = p.adjust, size = Count)) +
    geom_point() +
    facet_grid(direction ~ ., scales = "free_y", space = "free_y") +
    scale_color_gradient(
      low    = "#C0392B",
      high   = "#2980B9",
      name   = "FDR",
      limits = c(0, 0.05),
      oob    = scales::squish,
      labels = scales::label_scientific(digits = 1)
    ) +
    scale_size_continuous(
      name   = "Gene count",
      range  = c(2, 7),
      breaks = scales::pretty_breaks(n = 4)
    ) +
    labs(
      title    = clade_label,
      subtitle = "GO Biological Process (ORA)",
      x        = "Gene ratio",
      y        = NULL
    ) +
    theme_bw(base_size = 10, base_family = "Arial") +
    theme(
      # Title
      plot.title       = element_text(size = 11, face = "bold",
                                      color = "black", hjust = 0),
      plot.subtitle    = element_text(size = 8,  color = "grey40", hjust = 0),
      # Axes
      axis.title.x     = element_text(size = 10, face = "bold", color = "black"),
      axis.text.x      = element_text(size = 9,  color = "black"),
      axis.text.y      = element_text(size = 8,  color = "black"),
      axis.ticks       = element_line(linewidth = 0.35, color = "black"),
      # Strip
      strip.text.y     = element_text(size = 9, face = "bold",
                                      color = "white", angle = 90),
      strip.background = element_rect(fill = "grey30", color = NA),
      # Panel
      panel.grid.major.x = element_line(linewidth = 0.3, color = "grey90"),
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      panel.border       = element_rect(linewidth = 0.4, color = "grey40"),
      # Legend
      legend.position  = "right",
      legend.text      = element_text(size = 8),
      legend.title     = element_text(size = 9, face = "bold"),
      legend.key.size  = unit(0.4, "cm"),
      # Background
      plot.background  = element_rect(fill = "white", color = NA),
      plot.margin      = margin(8, 10, 6, 8)
    )

  # ── Color facet strips by direction ──────────────────────────────
  g             <- ggplotGrob(p)
  strip_idx     <- which(grepl("strip-r", g$layout$name))
  strip_fill    <- c("Upregulated" = "#C0392B", "Downregulated" = "#2471A3")
  present_dirs  <- levels(droplevels(df$direction))

  for (i in seq_along(strip_idx)) {
    fill_col      <- strip_fill[present_dirs[i]]
    rect_grob_idx <- which(sapply(
      g$grobs[[strip_idx[i]]]$grobs[[1]]$children,
      function(ch) inherits(ch, "rect")
    ))
    g$grobs[[strip_idx[i]]]$grobs[[1]]$children[[rect_grob_idx]]$gp$fill <- fill_col
  }

  g
}
```

```{r}
# ── Run: one file per clade, named after the clade ────────────────
clade_deg_counts <- dge_combined %>%
  filter(direction != "NS") %>%
  dplyr::count(clade, name = "n_deg") %>%
  arrange(desc(n_deg))

clades_with_degs <- clade_deg_counts %>% filter(n_deg >= 5) %>% pull(clade)

cat("Saving plots for", length(clades_with_degs), "clades\n\n")

# Clean clade name for use in filename (remove special characters)
clean_filename <- function(clade_name) {
  clade_name %>%
    str_replace_all("[^A-Za-z0-9_]", "_") %>%  # replace non-alphanumeric with _
    str_replace_all("_+", "_") %>%              # collapse multiple underscores
    str_remove("^_|_$")                         # remove leading/trailing _
}

for (cl in clades_with_degs) {
  fn <- clean_filename(cl)
  cat("Processing:", cl, "->", fn, "\n")
  
  # ── Volcano ───────────────────────────────────────────────────
  v <- plot_volcano(dge_results[[cl]], clade_label = cl)
  print(v)
  ggsave(
    file.path(OUT_DIR, paste0("volcano_", fn, ".tiff")),
    v, width = 6, height = 5, units = "in",
    dpi = 600, device = "tiff", compression = "lzw"
  )

  # ── GO dot plot ───────────────────────────────────────────────
  g <- plot_go_dotplot(ora_results[[cl]]$up, ora_results[[cl]]$down,
                       clade_label = cl)
  if (!is.null(g)) {
    grid::grid.newpage()
    grid::grid.draw(g)
    ggsave(
      file.path(OUT_DIR, paste0("GO_dotplot_", fn, ".tiff")),
      g, width = 6, height = 5, units = "in",
      dpi = 600, device = "tiff", compression = "lzw"
    )
  } else {
    cat("  No significant GO terms for:", cl, "\n")
  }
}

cat("\nDone. Files saved to:", OUT_DIR, "\n")
```

```{r}
# ── GSEA per clade (YPD.1) — complement to the ORA above ──────────────
# ORA tests over-representation within the thresholded DEG list; GSEA ranks
# ALL tested genes by effect size and asks whether a GO term's members
# cluster at either end. Coordinated modest shifts that never clear
# padj < 0.05 are invisible to ORA but detectable here.

library(clusterProfiler)

run_gsea <- function(res_df, min_genes = 100) {
  if (nrow(res_df) < min_genes) return(NULL)

  # Rank by signed significance: direction from LFC, magnitude from p-value.
  # Ranking on LFC alone lets a noisy high-fold-change gene outrank a
  # well-estimated moderate one.
  ranks <- res_df %>%
    filter(!is.na(padj), !is.na(log2FoldChange), padj > 0) %>%
    mutate(stat_rank = sign(log2FoldChange) * -log10(padj)) %>%
    arrange(desc(stat_rank))

  vec <- setNames(ranks$stat_rank, ranks$gene)
  vec <- vec[is.finite(vec)]
  if (length(vec) < min_genes) return(NULL)

  tryCatch(
    gseGO(geneList     = vec,
          OrgDb        = org.Sc.sgd.db,
          keyType      = "ORF",
          ont          = "BP",
          minGSSize    = 10,
          maxGSSize    = 500,
          pvalueCutoff = 0.05,
          pAdjustMethod = "BH",
          eps          = 0,        # allow p-values below 1e-10
          verbose      = FALSE),
    error = function(e) { message("GSEA failed for a clade: ",
                                  conditionMessage(e)); NULL }
  )
}

gsea_results <- lapply(names(dge_results), function(cl) run_gsea(dge_results[[cl]]))
names(gsea_results) <- names(dge_results)
gsea_results <- Filter(function(x) !is.null(x) && nrow(as.data.frame(x)) > 0, gsea_results)

cat("Clades with GSEA hits:", length(gsea_results), "of", length(dge_results), "\n")

gsea_combined <- lapply(names(gsea_results), function(cl) {
  as.data.frame(gsea_results[[cl]]) %>%
    mutate(clade = cl,
           direction = ifelse(NES > 0, "Upregulated", "Downregulated"))
}) %>% bind_rows()

write_csv(gsea_combined, file.path(OUT_DIR, "Fig2_clade_GSEA_YPD1.csv"))

cat("\nGSEA terms per clade:\n")
gsea_combined %>% dplyr::count(clade, direction) %>%
  pivot_wider(names_from = direction, values_from = n, values_fill = 0) %>%
  arrange(desc(Upregulated + Downregulated)) %>% print(n = Inf)
```


```{r}
# ── Per-clade volcano highlighting GSEA leading-edge genes ────────────
# GSEA's core_enrichment field lists the genes driving each term's score.
# Plotting them on the volcano shows *where* they sit — leading-edge genes
# below the DEG thresholds are exactly what ORA could not have seen.

# Extract leading-edge genes per clade (optionally per term)
gsea_leading_edge <- gsea_combined %>%
  select(clade, ID, Description, NES, p.adjust, direction, core_enrichment) %>%
  separate_rows(core_enrichment, sep = "/") %>%
  rename(gene = core_enrichment)

cat("Leading-edge gene-term pairs:", nrow(gsea_leading_edge), "\n")
cat("Unique leading-edge genes:", n_distinct(gsea_leading_edge$gene), "\n")

# How many leading-edge genes were NOT called DEGs?
le_summary <- gsea_leading_edge %>%
  distinct(clade, gene) %>%
  left_join(dge_combined %>% select(clade, gene, direction_deg = direction),
            by = c("clade", "gene")) %>%
  mutate(is_deg = direction_deg != "NS") %>%
  dplyr::count(clade, is_deg) %>%
  pivot_wider(names_from = is_deg, values_from = n, values_fill = 0) %>%
  rename(leading_edge_NS = `FALSE`, leading_edge_DEG = `TRUE`) %>%
  mutate(pct_missed_by_ORA = round(100 * leading_edge_NS /
                                     (leading_edge_NS + leading_edge_DEG), 1)) %>%
  arrange(desc(pct_missed_by_ORA))

cat("\nLeading-edge genes NOT passing DEG thresholds (invisible to ORA):\n")
print(le_summary, n = Inf)

# ── Volcano with leading-edge genes highlighted ───────────────────────
plot_volcano_gsea <- function(res_df, clade_label, le_genes,
                              padj_thresh = 0.05, lfc_thresh = 0.3,
                              top_n_labels = 12) {

  df <- res_df %>%
    mutate(in_gsea = gene %in% le_genes,
           class = case_when(
             in_gsea & direction != "NS" ~ "GSEA leading edge (also DEG)",
             in_gsea & direction == "NS" ~ "GSEA leading edge (not DEG)",
             direction != "NS"           ~ "DEG only",
             TRUE                        ~ "NS"))

  n_le_deg <- sum(df$class == "GSEA leading edge (also DEG)")
  n_le_ns  <- sum(df$class == "GSEA leading edge (not DEG)")

  labs_df <- df %>% filter(in_gsea) %>% arrange(padj) %>% slice_head(n = top_n_labels)

  ggplot(df, aes(log2FoldChange, -log10(padj))) +
    geom_point(data = filter(df, class == "NS"),
               color = "grey88", size = 0.7, alpha = 0.4) +
    geom_point(data = filter(df, class == "DEG only"),
               color = "grey55", size = 1, alpha = 0.6) +
    geom_point(data = filter(df, in_gsea), aes(color = class),
               size = 1.8, alpha = 0.9) +
    geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed",
               color = "grey50", linewidth = 0.35) +
    geom_vline(xintercept = c(-lfc_thresh, lfc_thresh), linetype = "dashed",
               color = "grey50", linewidth = 0.35) +
    ggrepel::geom_text_repel(data = labs_df, aes(label = gene_name),
                             size = 2.4, fontface = "italic", color = "black",
                             max.overlaps = 20, segment.size = 0.25,
                             box.padding = 0.35) +
    scale_color_manual(values = c("GSEA leading edge (also DEG)" = "#C0392B",
                                  "GSEA leading edge (not DEG)"  = "#E67E22"),
                       name = NULL) +
    labs(title = clade_label,
         subtitle = sprintf("GSEA leading edge: %d also DEG, %d below DEG thresholds",
                            n_le_deg, n_le_ns),
         x = expression(log[2]~"fold change (focal vs. rest)"),
         y = expression(-log[10]~"(adjusted"~italic(p)~")")) +
    theme_classic(base_size = 10) +
    theme(plot.title    = element_text(size = 11, face = "bold"),
          plot.subtitle = element_text(size = 8, color = "grey40"),
          legend.position = "bottom",
          legend.text     = element_text(size = 8))
}

#── One volcano per clade with GSEA hits ──────────────────────────────
GSEA_DIR <- file.path(OUT_DIR, "gsea_volcano")
dir.create(GSEA_DIR, showWarnings = FALSE, recursive = TRUE)

for (cl in names(gsea_results)) {
  le <- gsea_leading_edge %>% filter(clade == cl) %>% pull(gene) %>% unique()
  if (length(le) == 0) next

  v <- plot_volcano_gsea(dge_results[[cl]], clade_label = cl, le_genes = le)
  print(v)

#   ggsave(file.path(GSEA_DIR, paste0("gsea_volcano_", clean_filename(cl), ".tiff")),
#          v, width = 6, height = 5.5, units = "in", dpi = 600,
#          device = "tiff", compression = "lzw")
 }
```

```{r}
# ── Faceted grid: all clades with GSEA hits, 4 per row ────────────────
gsea_clades <- names(gsea_results)

grid_df <- lapply(gsea_clades, function(cl) {
  le <- gsea_leading_edge %>% filter(clade == cl) %>% pull(gene) %>% unique()
  dge_results[[cl]] %>%
    mutate(in_gsea = gene %in% le,
           class = case_when(
             in_gsea & direction != "NS" ~ "Leading edge (DEG)",
             in_gsea & direction == "NS" ~ "Leading edge (not DEG)",
             TRUE                        ~ "Other"))
}) %>% bind_rows()

panel_lab <- grid_df %>% filter(in_gsea) %>%
  dplyr::count(clade, class) %>%
  pivot_wider(names_from = class, values_from = n, values_fill = 0) %>%
  { if (!"Leading edge (not DEG)" %in% names(.))
      mutate(., `Leading edge (not DEG)` = 0L) else . } %>%
  mutate(panel = sprintf("%s\n(%d LE | %d not DEG)", clade,
                         `Leading edge (DEG)` + `Leading edge (not DEG)`,
                         `Leading edge (not DEG)`))

grid_df <- left_join(grid_df, select(panel_lab, clade, panel), by = "clade")

labs_df <- grid_df %>% filter(in_gsea) %>%
  group_by(clade) %>% slice_min(padj, n = 5) %>% ungroup()

gsea_volcano_grid <- ggplot(grid_df, aes(log2FoldChange, -log10(padj))) +
  geom_point(data = filter(grid_df, class == "Other"),
             color = "grey88", size = 0.4, alpha = 0.35) +
  geom_point(data = filter(grid_df, in_gsea), aes(color = class),
             size = 0.9, alpha = 0.85) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             color = "grey50", linewidth = 0.25) +
  geom_vline(xintercept = c(-0.3, 0.3), linetype = "dashed",
             color = "grey50", linewidth = 0.25) +
  ggrepel::geom_text_repel(data = labs_df, aes(label = gene_name),
                           size = 1.8, fontface = "italic",
                           max.overlaps = 10, segment.size = 0.2) +
  scale_color_manual(values = c("Leading edge (DEG)"     = "#C0392B",
                                "Leading edge (not DEG)" = "#E67E22"),
                     name = NULL) +
  facet_wrap(~ panel, ncol = 4, scales = "fixed") +
  labs(title = "GSEA leading-edge genes on clade-vs-rest volcanoes (YPD.1)",
       x = expression(log[2]~"fold change"),
       y = expression(-log[10]~"(adjusted"~italic(p)~")")) +
  theme_bw(base_size = 8) +
  theme(strip.text       = element_text(size = 6.5, face = "bold", lineheight = 0.9),
        strip.background = element_rect(fill = "grey92", color = NA),
        panel.grid       = element_blank(),
        legend.position  = "bottom",
        plot.title       = element_text(face = "bold", size = 10))

print(gsea_volcano_grid)

# ggsave(file.path(OUT_DIR, "Fig2_GSEA_leadingedge_volcano_GRID.tiff"),
#        gsea_volcano_grid, width = 11,
#        height = 2.6 * ceiling(length(gsea_clades) / 4),
#        units = "in", dpi = 600, device = "tiff", compression = "lzw",
#        limitsize = FALSE)
```

