---
title: "05_custom_gtf_annotations"
author: "Akriti Agrawal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: true
    number_section: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../results/") })

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.width = 10, 
                      fig.height = 6)
```

# Trans-eQTL Analysis with Allele Frequency Filtering

This analysis processes eQTL data by filtering for SNPs with allele frequency greater than 5%. The workflow consists of two main steps:

1. Process the VCF file to identify SNPs with allele frequency > 5%
2. Filter each bootstrap file to keep only significant eQTLs with these common SNPs

## Load Required Libraries

```{r libraries}
# Load necessary libraries
library(data.table)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(org.Sc.sgd.db)
# Be explicit if needed
select <- dplyr::select
filter <- dplyr::filter
```

---------------------------------------------------------
# STEP 0: Load SNP and gene location data
---------------------------------------------------------

```{r load-snp-data}
# 2. Read SNP matrix
snp_matrix_df <- read.table("/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast/yeast.1/code/snp_matrix.txt", 
                           header = TRUE, 
                           sep = "\t", 
                           stringsAsFactors = FALSE)

# Convert back to matrix with rownames
snp_matrix <- as.matrix(snp_matrix_df[, -1])  # Remove the SNP_ID column
rownames(snp_matrix) <- snp_matrix_df$SNP_ID

# 3. Read SNP location
snp_location <- read.table("/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast/yeast.1/code/snp_location.txt", 
                          header = TRUE, 
                          sep = "\t", 
                          stringsAsFactors = FALSE)

dim(snp_matrix)
dim(snp_location)


# Read gene matrix file
gene_matrix_df <- read.table("../results/pseudobulk_and_eqtl/gene_matrix.txt", 
                            header = TRUE, 
                            sep = "\t", 
                            stringsAsFactors = FALSE)

# Convert back to matrix with row names
gene_ids <- gene_matrix_df$Gene_ID
gene_matrix <- as.matrix(gene_matrix_df[, -1])  # Remove Gene_ID column
rownames(gene_matrix) <- gene_ids

# Check dimensions
cat("Gene matrix dimensions:", nrow(gene_matrix), "genes by", ncol(gene_matrix), "samples\n")

# Read gene location file
gene_location <- read.table("../results/pseudobulk_and_eqtl/gene_location.txt", 
                           header = TRUE,
                           stringsAsFactors = FALSE)

# Check dimensions
cat("Gene location file:", nrow(gene_location), "rows\n")

# Verify that the files were read correctly
head(gene_matrix[, 1:5])  # Show first 5 columns of gene matrix
head(gene_location)

# Check for NA values in the position columns
sum(is.na(gene_location$left))  # Count NA values in start position
sum(is.na(gene_location$right)) # Count NA values in end position

# Option 1: Remove genes with NA positions
gene_location_fixed <- gene_location[!is.na(gene_location$left) & !is.na(gene_location$right), ]
gene_location_fixed <- gene_location_fixed[, c("geneid", "chr", "left", "right")]
names(gene_location_fixed) <- c("geneid", "chr", "start", "end")

length(unique(gene_location$chr))
```


#Create custom annotations

Look at the features

```{r}
library(rtracklayer)
# Process the GTF annotation file
gtf_file <- "/scratch/aa9679/SingleCell_cerevisiae/Saccharomyces_cerevisiae.R64-1-1.CLEAN.gtf.gz"
yeast_annotations <- import(gtf_file)
gene_annotations <- yeast_annotations[yeast_annotations$type == "gene"]

# Explore different biotypes in yeast annotations
library(dplyr)

# Create the gene_type dataframe
gene_type <- data.frame(
  geneid = yeast_annotations$gene_id,
  type = yeast_annotations$type,
  gene_biotype = yeast_annotations$gene_biotype,
  transcript_biotype = yeast_annotations$transcript_biotype,
  stringsAsFactors = FALSE
)

cat("=== OVERVIEW OF ANNOTATION TYPES ===\n")
cat("Total annotations:", nrow(gene_type), "\n\n")

# 1. Explore the 'type' column (gene, transcript, exon, etc.)
cat("=== ANNOTATION TYPES (type column) ===\n")
type_counts <- table(gene_type$type, useNA = "ifany")
print(type_counts)
cat("\n")

# 2. Explore gene_biotype
cat("=== GENE BIOTYPES ===\n")
gene_biotype_counts <- table(gene_type$gene_biotype, useNA = "ifany")
print(gene_biotype_counts)
cat("\nUnique gene biotypes:\n")
unique_gene_biotypes <- unique(gene_type$gene_biotype)
print(unique_gene_biotypes[!is.na(unique_gene_biotypes)])
cat("\n")

# 3. Explore transcript_biotype
cat("=== TRANSCRIPT BIOTYPES ===\n")
transcript_biotype_counts <- table(gene_type$transcript_biotype, useNA = "ifany")
print(transcript_biotype_counts)
cat("\nUnique transcript biotypes:\n")
unique_transcript_biotypes <- unique(gene_type$transcript_biotype)
print(unique_transcript_biotypes[!is.na(unique_transcript_biotypes)])
cat("\n")


# Step 1: Convert gene annotations to data frame
gene_annotations_df <- as.data.frame(yeast_annotations)
```

#### Step 6b. Create custom annotation file for SNP annotations
```{r}
# Step 1: Convert gene annotations to data frame
# Process the GTF annotation file
gtf_file <- "/scratch/aa9679/SingleCell_cerevisiae/Saccharomyces_cerevisiae.R64-1-1.CLEAN.gtf.gz"
yeast_annotations <- import(gtf_file)
gene_annotations_df <- as.data.frame(yeast_annotations)


# Create custom gene annotations for ALL genes
# Processes both + and - strand genes

library(dplyr)

# Get all unique gene IDs
all_gene_ids <- unique(gene_annotations_df$gene_id)
all_gene_ids <- all_gene_ids[!is.na(all_gene_ids)]

cat("Processing", length(all_gene_ids), "genes...\n")

# Create a new dataframe for all custom annotations
all_custom_gene_annotations <- data.frame()

# Counter for progress tracking
processed_count <- 0
failed_count <- 0

# Process each gene
for(gene_id in all_gene_ids) {
  
  processed_count <- processed_count + 1
  
  # Progress update every 500 genes
  if(processed_count %% 500 == 0) {
    cat("Processed", processed_count, "of", length(all_gene_ids), "genes...\n")
  }
  
  tryCatch({
    
    # Filter annotations for this gene
    gene_features <- gene_annotations_df[gene_annotations_df$gene_id == gene_id, ]
    
    if(nrow(gene_features) == 0) {
      failed_count <- failed_count + 1
      next
    }
    
    # Get gene info from first row
    gene_info <- gene_features[1, ]
    strand <- gene_info$strand
    seqname <- gene_info$seqnames
    source <- gene_info$source
    
    # Get CDS coordinates
    cds_features <- gene_features[gene_features$type == "CDS", ]
    if(nrow(cds_features) == 0) {
      # Skip genes without CDS (like ncRNA, etc.)
      next
    }
    
    cds_start <- min(cds_features$start)
    cds_end <- max(cds_features$end)
    
    # Get transcript coordinates
    transcript_features <- gene_features[gene_features$type == "transcript", ]
    if(nrow(transcript_features) > 0) {
      transcript_start <- min(transcript_features$start)
      transcript_end <- max(transcript_features$end)
    } else {
      # Use gene coordinates if transcript not available
      gene_row <- gene_features[gene_features$type == "gene", ]
      if(nrow(gene_row) > 0) {
        transcript_start <- gene_row$start[1]
        transcript_end <- gene_row$end[1]
      } else {
        # Fallback: use min/max of all features
        transcript_start <- min(gene_features$start)
        transcript_end <- max(gene_features$end)
      }
    }
    
    # Add existing CDS annotations
    all_custom_gene_annotations <- rbind(all_custom_gene_annotations, cds_features)
    
    # Add start_codon and stop_codon if they exist
    start_codon_features <- gene_features[gene_features$type == "start_codon", ]
    if(nrow(start_codon_features) > 0) {
      all_custom_gene_annotations <- rbind(all_custom_gene_annotations, start_codon_features)
    }
    
    stop_codon_features <- gene_features[gene_features$type == "stop_codon", ]
    if(nrow(stop_codon_features) > 0) {
      all_custom_gene_annotations <- rbind(all_custom_gene_annotations, stop_codon_features)
    }
    
    # Create UTR annotations based on strand
    if(strand == "+") {
      
      # For positive strand:
      # 5' UTR is from transcript start to CDS start
      # 3' UTR is from CDS end to transcript end
      
      if(transcript_start < cds_start) {
        # 5' UTR
        five_prime_utr <- gene_info
        five_prime_utr$start <- transcript_start
        five_prime_utr$end <- cds_start - 1
        five_prime_utr$width <- (cds_start - 1) - transcript_start + 1
        five_prime_utr$type <- "five_prime_UTR"
        five_prime_utr$score <- NA
        five_prime_utr$phase <- NA
        
        all_custom_gene_annotations <- rbind(all_custom_gene_annotations, five_prime_utr)
      }
      
      if(cds_end < transcript_end) {
        # 3' UTR
        three_prime_utr <- gene_info
        three_prime_utr$start <- cds_end + 1
        three_prime_utr$end <- transcript_end
        three_prime_utr$width <- transcript_end - (cds_end + 1) + 1
        three_prime_utr$type <- "three_prime_UTR"
        three_prime_utr$score <- NA
        three_prime_utr$phase <- NA
        
        all_custom_gene_annotations <- rbind(all_custom_gene_annotations, three_prime_utr)
        
        # Terminator region (15bp downstream of 3' UTR)
        terminator <- gene_info
        terminator$start <- transcript_end + 1
        terminator$end <- transcript_end + 15
        terminator$width <- 15
        terminator$type <- "terminator"
        terminator$score <- NA
        terminator$phase <- NA
        
        all_custom_gene_annotations <- rbind(all_custom_gene_annotations, terminator)
      }
      
      # Promoter region (200bp upstream of transcription start site)
      promoter <- gene_info
      promoter$start <- transcript_start - 200
      promoter$end <- transcript_start - 1
      promoter$width <- 200
      promoter$type <- "promoter"
      promoter$score <- NA
      promoter$phase <- NA
      
      all_custom_gene_annotations <- rbind(all_custom_gene_annotations, promoter)
      
    } else if(strand == "-") {
      
      # For negative strand:
      # 5' UTR is from CDS end to transcript end (3' in genomic coordinates)
      # 3' UTR is from transcript start to CDS start (5' in genomic coordinates)
      
      if(transcript_start < cds_start) {
        # 3' UTR (5' in genomic coordinates, but 3' relative to gene direction)
        three_prime_utr <- gene_info
        three_prime_utr$start <- transcript_start
        three_prime_utr$end <- cds_start - 1
        three_prime_utr$width <- (cds_start - 1) - transcript_start + 1
        three_prime_utr$type <- "three_prime_UTR"
        three_prime_utr$score <- NA
        three_prime_utr$phase <- NA
        
        all_custom_gene_annotations <- rbind(all_custom_gene_annotations, three_prime_utr)
        
        # Terminator region (15bp upstream of 3' UTR in genomic coordinates)
        terminator <- gene_info
        terminator$start <- transcript_start - 15
        terminator$end <- transcript_start - 1
        terminator$width <- 15
        terminator$type <- "terminator"
        terminator$score <- NA
        terminator$phase <- NA
        
        all_custom_gene_annotations <- rbind(all_custom_gene_annotations, terminator)
      }
      
      if(cds_end < transcript_end) {
        # 5' UTR (3' in genomic coordinates, but 5' relative to gene direction)
        five_prime_utr <- gene_info
        five_prime_utr$start <- cds_end + 1
        five_prime_utr$end <- transcript_end
        five_prime_utr$width <- transcript_end - (cds_end + 1) + 1
        five_prime_utr$type <- "five_prime_UTR"
        five_prime_utr$score <- NA
        five_prime_utr$phase <- NA
        
        all_custom_gene_annotations <- rbind(all_custom_gene_annotations, five_prime_utr)
      }
      
      # Promoter region (200bp upstream of gene start in genomic coordinates)
      # For negative strand, upstream means higher coordinates
      promoter <- gene_info
      promoter$start <- transcript_end + 1
      promoter$end <- transcript_end + 200
      promoter$width <- 200
      promoter$type <- "promoter"
      promoter$score <- NA
      promoter$phase <- NA
      
      all_custom_gene_annotations <- rbind(all_custom_gene_annotations, promoter)
    }
    
  }, error = function(e) {
    failed_count <<- failed_count + 1
    cat("Error processing gene", gene_id, ":", e$message, "\n")
  })
}

# Sort all annotations by chromosome and start position
all_custom_gene_annotations <- all_custom_gene_annotations[order(all_custom_gene_annotations$seqnames, 
                                                                 all_custom_gene_annotations$start), ]

# Reset row names
rownames(all_custom_gene_annotations) <- NULL

# Final summary
cat("\n=== PROCESSING COMPLETE ===\n")
cat("Total genes processed:", processed_count, "\n")
cat("Genes failed:", failed_count, "\n")
cat("Total annotations created:", nrow(all_custom_gene_annotations), "\n")

cat("\nSummary of annotation types:\n")
print(table(all_custom_gene_annotations$type))

cat("\nAnnotations per chromosome:\n")
print(table(all_custom_gene_annotations$seqnames))

# Display first few rows
cat("\nFirst few annotations:\n")
print(head(all_custom_gene_annotations[, c("seqnames", "type", "start", "end", "width", "strand", "gene_id")]))

# Save to file (optional)
write.csv(all_custom_gene_annotations, "../data/yeast_custom_annotations.csv", row.names = FALSE)

```




