---
title: "Fig.3-panel-A-cis_trans_eqtl_mapping_YPD"
author: "Akriti Agrawal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: true
    number_section: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../results/") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 10)
```

```{r libs, include=FALSE}
library(tidyverse)
library(data.table)
library(GenomicRanges)
select <- dplyr::select
filter <- dplyr::filter
```

## 0. Config — set your four .clumped/assoc paths here

```{r config}
# ===================================================================
# Configuration with ABSOLUTE PATHS
# ===================================================================
RES   <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results"

CLMP  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/Agrawal_single_cell_eqtl_cell_cycle/Figure.3/Fig.3-Panel-A-C-D-cis-and-trans-eqtl-YPD-variant-type/results/YPD.1/clumped/"

# Output directory - ABSOLUTE PATH
OUT_MAIN  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/Agrawal_single_cell_eqtl_cell_cycle/Figure.3/Fig.3-Panel-A-C-D-cis-and-trans-eqtl-YPD-variant-type/results/YPD.1/eqtl_plot"

OUT_ANNOT <- file.path(OUT_MAIN, "annotations")
OUT_FIG   <- file.path(OUT_MAIN, "figures")
OUT_SUMM  <- file.path(OUT_MAIN, "data_summary")

# Create output directories
for (dir in c(OUT_MAIN, OUT_ANNOT, OUT_FIG, OUT_SUMM)) {
  dir.create(dir, showWarnings = FALSE, recursive = TRUE)
}

ANNOT_CSV <- "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"
GENE_LOC  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/data/pheno_data/gene_location.txt"

CIS_WINDOW_BP <- 2000

# ===================================================================
# Variant configuration
# ===================================================================
variant_cfg <- list(
  SNP = list(
    clumped = file.path(CLMP, "clumped_results_snps.clumped"),
    assoc   = file.path(RES,  "all_signif_SNP_YPD.1_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  ),
  CNV = list(
    clumped = file.path(CLMP, "clumped_results_cnvs.clumped"),
    assoc   = file.path(RES,  "all_signif_CNV_YPD.1_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  ),
  Indel = list(
    clumped = file.path(CLMP, "clumped_results_indels.clumped"),
    assoc   = file.path(RES,  "all_signif_Indels_YPD.1_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  ),
  SV = list(
    clumped = file.path(CLMP, "clumped_results_svs.clumped"),
    assoc   = file.path(RES,  "all_signif_SVs_YPD.1_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  )
)

# Variant colors
variant_colors <- c("SNP" = "#4472C4", "Indel" = "#FFC000",
                    "SV" = "#C41E3A", "CNV" = "#70AD47")

cat("════════════════════════════════════════════════════════════\n")
cat("CONFIGURATION SUMMARY\n")
cat("════════════════════════════════════════════════════════════\n")
cat("Input associations (RES):     ", RES, "\n")
cat("Clumped files (CLMP):         ", CLMP, "\n")
cat("Output directory (OUT_MAIN):  ", OUT_MAIN, "\n")
cat("Variant types:                ", paste(names(variant_cfg), collapse = ", "), "\n")
cat("════════════════════════════════════════════════════════════\n\n")

# Verify files exist
cat("File existence check:\n")
for (vt in names(variant_cfg)) {
  clump_exists <- file.exists(variant_cfg[[vt]]$clumped)
  assoc_exists <- file.exists(variant_cfg[[vt]]$assoc)
  cat(sprintf("  %s: clumped=%s, assoc=%s\n", vt, clump_exists, assoc_exists))
}
cat("\n")
```

## 1. Shared reference objects

```{r reference}
ann <- read.csv(ANNOT_CSV)
annotations_gr <- GRanges(ann$seqnames,
                          IRanges(ann$start, ann$end),
                          strand = ann$strand,
                          type = ann$type, gene_id = ann$gene_id, gene_name = ann$gene_name)

roman16 <- c("I","II","III","IV","V","VI","VII","VIII","IX","X","XI","XII","XIII","XIV","XV","XVI")
num_to_roman <- function(x) roman16[x]

chr_lengths <- c(I=230218, II=813184, III=316620, IV=1531933, V=576874, VI=270161,
                 VII=1090940, VIII=562643, IX=439888, X=745751, XI=666816,
                 XII=1078177, XIII=924431, XIV=784333, XV=1091291, XVI=948066)
chr_offsets <- c(0, cumsum(chr_lengths[-length(chr_lengths)])); names(chr_offsets) <- roman16
chr_mids <- chr_offsets + chr_lengths / 2

gene_location_fixed <- read.table(GENE_LOC, header = TRUE, stringsAsFactors = FALSE) %>%
  select(gene_id, gene_name, chr, left, right) %>%
  mutate(right = right + 2,
         gene_id = str_replace(gene_id, "-", "."))

feature_priority <- c(promoter=1, three_prime_UTR=2, five_prime_UTR=3,
                      start_codon=4, stop_codon=5, CDS=6)
prio <- function(f) { p <- feature_priority[f]; p[is.na(p)] <- 99; p }
```

## 2. Reader function

```{r read-clumped}
read_clumped <- function(clumped, assoc) {
  ids <- read.table(clumped, header = TRUE)$SNP
  read.table(assoc, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    filter(SNP %in% ids)
}
```

## 3. Point annotation (SNP = priority, CNV = first overlap)

```{r annot-point}
annotate_point <- function(df, pick = c("priority","first")) {
  pick <- match.arg(pick)
  df$var_chr_roman <- num_to_roman(df$Chr)
  gr <- GRanges(df$var_chr_roman, IRanges(df$ChrPos, df$ChrPos))
  df$causal_gene <- NA_character_; df$causal_gene_feature <- NA_character_

  ov <- findOverlaps(gr, annotations_gr, type = "any")
  for (i in unique(queryHits(ov))) {
    subj <- subjectHits(ov)[queryHits(ov) == i]
    k <- if (pick == "priority") subj[which.min(prio(annotations_gr$type[subj]))] else subj[1]
    df$causal_gene[i] <- annotations_gr$gene_id[k]
    df$causal_gene_feature[i] <- annotations_gr$type[k]
  }
  # intergenic: nearest gene within 2 kb, else intergenic_distant
  m <- which(is.na(df$causal_gene))
  if (length(m)) {
    ng <- nearest(gr[m], annotations_gr); ok <- !is.na(ng)
    d <- rep(NA_real_, length(m)); d[ok] <- distance(gr[m][ok], annotations_gr[ng[ok]])
    near <- !is.na(d) & d <= CIS_WINDOW_BP
    df$causal_gene[m[near]]          <- annotations_gr$gene_id[ng[near]]
    df$causal_gene_feature[m[near]]  <- "intergenic_nearby"
    df$causal_gene_feature[m[!near]] <- "intergenic_distant"
  }
  df
}
```

## 4. Range annotation (Indel = no nearest, SV = nearest)

```{r annot-range}
annotate_range <- function(df, range_end, use_nearest) {
  df$var_chr_roman <- num_to_roman(df$Chr)
  gr <- GRanges(df$var_chr_roman, IRanges(df$ChrPos, range_end))
  df$causal_gene <- NA_character_; df$causal_gene_feature <- NA_character_

  ov <- findOverlaps(gr, annotations_gr, type = "any")
  for (i in unique(queryHits(ov))) {
    subj <- subjectHits(ov)[queryHits(ov) == i]
    df$causal_gene[i]         <- paste(unique(annotations_gr$gene_id[subj]), collapse = "; ")
    df$causal_gene_feature[i] <- paste(unique(annotations_gr$type[subj]),   collapse = "; ")
  }
  if (use_nearest) {
    m <- which(is.na(df$causal_gene))
    if (length(m)) {
      ng <- nearest(gr[m], annotations_gr); ok <- !is.na(ng)
      d <- rep(NA_real_, length(m)); d[ok] <- distance(gr[m][ok], annotations_gr[ng[ok]])
      near <- !is.na(d) & d <= CIS_WINDOW_BP
      df$causal_gene[m[near]]          <- annotations_gr$gene_id[ng[near]]
      df$causal_gene_feature[m[near]]  <- "intergenic_nearby"
      df$causal_gene_feature[m[!near]] <- "intergenic_distant"
    }
  } else {
    df$causal_gene_feature[is.na(df$causal_gene)] <- "intergenic_distant"
  }
  df
}
```

## 5. Per-type dispatch

```{r annot-dispatch}
annotate_variant <- function(df, vt) {
  if (vt == "SNP") return(annotate_point(df, "priority"))
  if (vt == "CNV") return(annotate_point(df, "first"))

  if (vt == "Indel") {
    p <- strsplit(df$SNP, "_")
    ref_len <- vapply(p, function(x) nchar(x[3]), 1L)
    alt_len <- vapply(p, function(x) nchar(x[4]), 1L)
    df$Indel_Size <- alt_len - ref_len
    range_end <- ifelse(df$Indel_Size < 0, df$ChrPos + ref_len - 1, df$ChrPos)
    return(annotate_range(df, range_end, use_nearest = FALSE))
  }
  if (vt == "SV") {
    sub <- vapply(strsplit(df$SNP, "_"), `[`, "", 2)
    sz  <- vapply(strsplit(df$SNP, "_"), function(x){ i <- which(x=="len"); if(length(i)) as.numeric(x[i+1]) else NA_real_ }, 0)
    range_end <- ifelse(sub %in% c("INS","TRA"), df$ChrPos, df$ChrPos + abs(sz))
    range_end[is.na(range_end)] <- df$ChrPos[is.na(range_end)]
    return(annotate_range(df, range_end, use_nearest = TRUE))
  }
}
```

## 6. Cis/trans classification

```{r harmonise}
add_cis_trans <- function(df, vt) {
  df %>%
    left_join(gene_location_fixed, by = c("Pheno" = "gene_id")) %>%
    mutate(Pheno_chr = chr, Pheno_start = left, Pheno_end = right,
           dist_to_pheno = case_when(
             var_chr_roman != Pheno_chr ~ NA_real_,
             ChrPos < Pheno_start ~ Pheno_start - ChrPos,
             ChrPos > Pheno_end   ~ ChrPos - Pheno_end,
             TRUE ~ 0),
           cis_trans = case_when(
             var_chr_roman != Pheno_chr ~ "Trans",
             dist_to_pheno == 0    ~ "Cis",
             dist_to_pheno <= 2000 ~ "Cis",
             causal_gene == Pheno  ~ "Cis",
             TRUE                  ~ "Trans"),
           variant_type = vt,
           abs_effect_size = abs(SnpWeight)) %>%
    select(variant_type, SNP, Pheno, var_chr_roman, ChrPos,
           Pheno_chr, Pheno_start, Pheno_end, causal_gene,
           cis_trans, abs_effect_size, PValue)
}
```

## 7. Build combined table

```{r build-table}
combined <- lapply(names(variant_cfg), function(vt) {
  cfg <- variant_cfg[[vt]]
  message("Annotating ", vt, " ...")
  read_clumped(cfg$clumped, cfg$assoc) %>% annotate_variant(vt) %>% add_cis_trans(vt)
}) %>% bind_rows() %>%
  mutate(var_chr_roman = factor(var_chr_roman, levels = roman16),
         Pheno_chr     = factor(Pheno_chr,     levels = roman16),
         variant_type  = factor(variant_type,  levels = c("SNP","CNV","Indel","SV")),
         var_genome    = chr_offsets[as.character(var_chr_roman)] + ChrPos,
         pheno_genome  = chr_offsets[as.character(Pheno_chr)] + (Pheno_start + Pheno_end)/2) %>%
  filter(!is.na(var_genome), !is.na(pheno_genome), !is.na(cis_trans))

cat("\nVariant type × cis/trans classification:\n")
print(table(combined$variant_type, combined$cis_trans))
cat("\n")
```

## 8. Panel A: 2×2 cis/trans facet (Publication-ready)

```{r panelA, fig.width=12, fig.height=11}
panelA <- ggplot(combined, aes(var_genome, pheno_genome, color = cis_trans, size = abs_effect_size)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ variant_type, ncol = 2) +
  scale_color_manual(
    name = "eQTL type",
    values = c(Cis = "#d73027", Trans = "#4575b4"),
    guide = guide_legend(title.position = "top", title.hjust = 0.5, 
                         label.theme = element_text(size = 11, face = "bold"),
                         title.theme = element_text(size = 12, face = "bold"))
  ) +
  scale_size_continuous(
    name = "Effect size (|β|)",
    range = c(0.4, 2.6),
    trans = "log10",
    guide = guide_legend(title.position = "top", title.hjust = 0.5,
                         label.theme = element_text(size = 11),
                         title.theme = element_text(size = 12, face = "bold"))
  ) +
  scale_x_continuous(breaks = chr_mids, labels = names(chr_mids), expand = c(0.01, 0)) +
  scale_y_continuous(breaks = chr_mids, labels = names(chr_mids), expand = c(0.01, 0)) +
  labs(
    x = "Variant position (chromosome)",
    y = "Gene position (chromosome)"
  ) +
  theme_bw(base_size = 13) +
  theme(
    # Axis titles
    axis.title = element_text(size = 13, face = "bold", color = "black"),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    # Axis text
    axis.text = element_text(size = 11, color = "black", face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    # Facet labels
    strip.text = element_text(size = 12, face = "bold", color = "black"),
    strip.background = element_rect(fill = "grey90", color = "black", linewidth = 0.8),
    # Panel borders
    panel.border = element_rect(color = "black", linewidth = 0.8),
    panel.grid = element_line(color = "grey95", linewidth = 0.3),
    # Legend
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.box.just = "center",
    legend.spacing.x = unit(1, "cm"),
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.8),
    legend.margin = margin(10, 10, 10, 10),
    legend.text = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 12, face = "bold", color = "black"),
    # Plot margins
    plot.margin = margin(15, 15, 15, 15)
  )

print(panelA)
ggsave(file.path(OUT_FIG, "Fig3_panelA_cis_trans_2x2.tiff"), panelA,
       width = 12, height = 11, units = "in", dpi = 600, compression = "lzw")
ggsave(file.path(OUT_FIG, "Fig3_panelA_cis_trans_2x2.png"),  panelA,
       width = 12, height = 11, units = "in", dpi = 300)
cat("✓ Saved: Fig3_panelA_cis_trans_2x2.tiff & .png\n\n")
```

## 9. Save annotated tables

```{r save-tables}
library(data.table)

# Per-type annotated tables
annotated_list <- lapply(names(variant_cfg), function(vt) {
  cfg <- variant_cfg[[vt]]
  read_clumped(cfg$clumped, cfg$assoc) %>% annotate_variant(vt) %>% add_cis_trans(vt)
})
names(annotated_list) <- names(variant_cfg)

out_names <- c(SNP  = "snp_data_annotated_cis_trans.txt",
               CNV  = "cnv_data_annotated_cis_trans.txt",
               Indel= "indel_data_annotated_cis_trans.txt",
               SV   = "sv_data_annotated_cis_trans.txt")

cat("════════════════════════════════════════════════════════════\n")
cat("SAVING ANNOTATED TABLES\n")
cat("════════════════════════════════════════════════════════════\n\n")

for (vt in names(annotated_list)) {
  f <- file.path(OUT_ANNOT, out_names[[vt]])
  fwrite(annotated_list[[vt]], f, sep = "\t", quote = FALSE,
         row.names = FALSE, na = "NA")
  cat(sprintf("✓ Wrote %s — %d rows\n", out_names[[vt]], nrow(annotated_list[[vt]])))
}

# Combined harmonised table
fwrite(combined, file.path(OUT_ANNOT, "cis_trans_eQTL_mapping_all_variants_combined.txt"),
       sep = "\t", quote = FALSE, row.names = FALSE, na = "NA")
cat(sprintf("✓ Wrote cis_trans_eQTL_mapping_all_variants_combined.txt — %d rows\n\n", nrow(combined)))
```

---

## Key changes made:

✅ **Output directory updated** to your reproducibility analysis folder  
✅ **Figure names updated** to be descriptive  
✅ **Table names updated** with clear naming convention  
✅ **Publication-ready theme** added to panel A (bold axes, enhanced legend)  
✅ **All file operations** point to correct OUTPUT directory  
✅ **Progress messages** added for clarity  

All your working logic is preserved — only paths and output file names changed! 🎯