---
title: "eQTL publication figures — YPD.1 (SNP / Indel / SV / CNV)"
author: "Akriti Agrawal"
output: html_document
---

<!--
Self-contained pipeline:
  PLINK clumping  ->  per-type annotation (inline)  ->  Fig 1-6
Every figure reads ONE standardized table (`all_clumped`) with these columns:
  Genetic_variants, SNP, Variant_Chr, ChrPos, Pheno, Pheno_chr,
  PValue, SnpWeight, EffectSize, Causal_Gene, Causal_Gene_feature,
  Variant_subtype, dist_to_pheno, cis_trans, Location, abs_SnpWeight
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 0. Packages
```{r libs}
library(tidyverse)
library(data.table)
library(GenomicRanges)
library(ggpubr)
library(ggsignif)

select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
mutate <- dplyr::mutate
```

# 1. Configuration (paths + parameters)
```{r config}
RES   <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/results"
CLMP  <- file.path(RES, "plink_without_double_id")
# All NEW outputs (annotated tables + figures) go here, NOT into RES:
OUT   <- "../results"
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)

ANNOT_CSV <- "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"
GENE_LOC  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/YPD.1/fastlmm/data/pheno_data/gene_location.txt"

CIS_WINDOW_BP <- 2000   # gene-assignment + cis/trans window

# per-type clumped + association tables (from your results dir)
variant_cfg <- list(
  SNPs   = list(clumped = file.path(CLMP, "clumped_results_snp.clumped"),
                assoc   = file.path(RES,  "all_signif_SNP_YPD.1_with_double_id_with_threshold.txt")),
  CNVs   = list(clumped = file.path(CLMP, "clumped_results_CNVs.clumped"),
                assoc   = file.path(RES,  "all_signif_CNV_YPD.1_with_double_id_with_threshold.txt")),
  InDels = list(clumped = file.path(CLMP, "clumped_results_Indels.clumped"),
                assoc   = file.path(RES,  "all_signif_Indels_YPD.1_with_double_id_with_threshold.txt")),
  SVs    = list(clumped = file.path(CLMP, "clumped_results_SVs.clumped"),
                assoc   = file.path(RES,  "all_signif_SVs_YPD.1_with_double_id_with_threshold.txt"))
)

# genotype matrices for clumping (Section 2)
geno_matrix <- list(
  SNPs   = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/SNPs/SNPs_no_change_name_and_no_strain_filtering_only_MAF15_filtered/full969Matrix.SNPs.Biallelic.Var.DP10.99percNonMissing.MAF15.NoSubTelo20kb.plink",
  CNVs   = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/CNVs/CNVs_no_change_name_and_no_strain_filtering_only_MAF15_filtered/CNV_969strains_gain_maf15_N400_20221121.plink",
  InDels = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SNPs_indels/modified_snp_indel_plink_for_96strains/Indels_only_96/indels.96samples",
  SVs    = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SVs_from_SV.1087/SV.1087samples_updated_fam_files/SV.1087samples.SVs"
)

# .bim files used ONLY for total variant counts (donut + odds ratio backgrounds)
bim_paths <- list(
  SNPs   = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/SNPs/SNPs_change_names_in_fam_968/SNPs_85/full969Matrix.SNPs.Biallelic.Var.DP10.99percNonMissing.MAF15.NoSubTelo20kb.plink_YPD.1_85strains.bim",
  InDels = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SNPs_indels/modified_snp_indel_plink_for_96strains/Indels_only/indels.85samples.bim",
  SVs    = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/vcf_files_sent_by_Anne/SVs_from_SV.1087/SV.1087samples_updated_fam_files/SVs_for_84_strains/SV.1087samples.SVs_to_84strains.bim",
  CNVs   = "/projects/rps/cgsb/gresham/Akriti/wildyeast1/the_vcfs/pan_transcriptome_paper/CNVs/CNVs_85/CNV_969strains_gain_maf15_N400_20221121.plink_YPD.1_85strains.bim"
)

variant_colors <- c("SNPs" = "#4472C4", "InDels" = "#FFC000",
                    "SVs" = "#C41E3A", "CNVs" = "#70AD47")
```
# 2. PLINK clumping  (run on the cluster; not evaluated in R)

Prepare the `--clump` input (SNP CHR BP P) from each association table, then clump.
`--clump-kb 2` is the window derived from the SNP LD-decay curve and applied to all four types.

```{r prep-clump-input, eval=FALSE}
prepare_plink_file <- function(assoc, out) {
  read.table(assoc, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    dplyr::select(SNP = SNP, CHR = Chr, BP = ChrPos, P = PValue) %>%
    write.table(out, row.names = FALSE, quote = FALSE, sep = "\t")
}
for (vt in names(variant_cfg))
  prepare_plink_file(variant_cfg[[vt]]$assoc,
                     file.path(CLMP, paste0("plink_", tolower(vt), ".txt")))
```

```{bash clump, eval=FALSE}
# one call per variant type (edit --bfile / --clump / --out per type)
plink --bfile <geno_matrix> \
      --clump  <plink_input.txt> \
      --clump-p1 0.05 \
      --clump-r2 0.6 \
      --clump-kb 2 \
      --out    <clumped_results_TYPE>
```

# 3. Shared annotation reference
```{r reference}
ann <- read.csv(ANNOT_CSV)
annotations_gr <- GRanges(ann$seqnames,
                          IRanges(ann$start, ann$end),
                          strand  = ann$strand,
                          type    = ann$type,
                          gene_id = ann$gene_id,
                          gene_name = ann$gene_name)

roman16 <- c("I","II","III","IV","V","VI","VII","VIII",
             "IX","X","XI","XII","XIII","XIV","XV","XVI")
num_to_roman <- function(x) roman16[x]

chr_lengths <- c(I=230218, II=813184, III=316620, IV=1531933, V=576874, VI=270161,
                 VII=1090940, VIII=562643, IX=439888, X=745751, XI=666816,
                 XII=1078177, XIII=924431, XIV=784333, XV=1091291, XVI=948066)

gene_location_fixed <- read.table(GENE_LOC, header = TRUE, stringsAsFactors = FALSE) %>%
  dplyr::select(gene_id, gene_name, chr, left, right) %>%
  mutate(right = right + 2,                              # gene end = start of stop codon
         gene_id = str_replace(gene_id, "-", "."))

# feature priority for the representative-feature pick (promoter first ... CDS last)
feature_priority <- c(promoter = 1, three_prime_UTR = 2, five_prime_UTR = 3,
                      start_codon = 4, stop_codon = 5, CDS = 6)
prio <- function(f) { p <- feature_priority[f]; p[is.na(p)] <- 99; p }
```

# 4. Annotation functions (each type keeps its own logic)

* **SNP**  — point overlap → highest-priority feature → nearest gene ≤2 kb
* **CNV**  — causal gene read directly from the ID (`rsYOR390W → YOR390W`); feature from point overlap
* **Indel** — range overlap (paste all features) → nearest gene ≤2 kb (NA-guarded)
* **SV**   — range overlap (paste all features) → nearest gene ≤2 kb (NA-guarded)

```{r annot-helpers}
read_clumped <- function(clumped, assoc) {
  ids <- read.table(clumped, header = TRUE)$SNP
  read.table(assoc, header = TRUE, sep = "\t", quote = "", stringsAsFactors = FALSE) %>%
    filter(SNP %in% ids)
}

# nearest-gene ≤ window, NA-guarded. Returns updated causal_gene / feature vectors.
assign_nearest <- function(df, gr) {
  m <- which(is.na(df$Causal_Gene))
  if (length(m)) {
    ng <- nearest(gr[m], annotations_gr); ok <- !is.na(ng)
    d  <- rep(NA_real_, length(m))
    d[ok] <- distance(gr[m][ok], annotations_gr[ng[ok]])
    near <- !is.na(d) & d <= CIS_WINDOW_BP
    df$Causal_Gene[m[near]]          <- annotations_gr$gene_id[ng[near]]
    df$Causal_Gene_feature[m[near]]  <- "intergenic_nearby"
    df$Causal_Gene_feature[m[!near]] <- "intergenic_distant"   # >2kb OR no reachable gene
  }
  df
}

annotate_point <- function(df, pick = c("priority","first")) {
  pick <- match.arg(pick)
  df$Variant_Chr <- num_to_roman(df$Chr)
  gr <- GRanges(df$Variant_Chr, IRanges(df$ChrPos, df$ChrPos))
  df$Causal_Gene <- NA_character_; df$Causal_Gene_feature <- NA_character_
  ov <- findOverlaps(gr, annotations_gr, type = "any")
  for (i in unique(queryHits(ov))) {
    subj <- subjectHits(ov)[queryHits(ov) == i]
    k <- if (pick == "priority") subj[which.min(prio(annotations_gr$type[subj]))] else subj[1]
    df$Causal_Gene[i]         <- annotations_gr$gene_id[k]
    df$Causal_Gene_feature[i] <- annotations_gr$type[k]
  }
  assign_nearest(df, gr)
}

annotate_range <- function(df, range_end, use_nearest) {
  df$Variant_Chr <- num_to_roman(df$Chr)
  gr <- GRanges(df$Variant_Chr, IRanges(pmin(df$ChrPos, range_end), pmax(df$ChrPos, range_end)))
  df$Causal_Gene <- NA_character_; df$Causal_Gene_feature <- NA_character_
  ov <- findOverlaps(gr, annotations_gr, type = "any")
  for (i in unique(queryHits(ov))) {
    subj <- subjectHits(ov)[queryHits(ov) == i]
    df$Causal_Gene[i]         <- paste(unique(annotations_gr$gene_id[subj]), collapse = "; ")
    df$Causal_Gene_feature[i] <- paste(unique(annotations_gr$type[subj]),   collapse = "; ")
  }
  if (use_nearest) {                              # SV: assign nearest gene <= window (NA-guarded)
    df <- assign_nearest(df, gr)
  } else {                                        # Indel: original had nearest commented out
    df$Causal_Gene_feature[is.na(df$Causal_Gene)] <- "intergenic_distant"
  }
  df
}

# per-type dispatch: also sets Variant_subtype used by Fig.3
annotate_variant <- function(df, vt) {
  if (vt == "SNPs") {
    df <- annotate_point(df, "priority")
    df$Variant_subtype <- "SNP"
    return(df)
  }
  if (vt == "CNVs") {
    df <- annotate_point(df, "first")        # gives a feature label at the point
    gene_from_id <- sub("^rs", "", df$SNP)   # ID encodes the ORF -> causal gene
    df$Causal_Gene <- ifelse(gene_from_id != "" & !is.na(gene_from_id),
                             gene_from_id, df$Causal_Gene)
    df$Variant_subtype <- "CNV"
    return(df)
  }
  if (vt == "InDels") {
    p <- strsplit(df$SNP, "_")               # CHR_POS_REF_ALT  (A1=ALT, A2=REF)
    ref_len <- vapply(p, function(x) nchar(x[3]), 1L)
    alt_len <- vapply(p, function(x) nchar(x[4]), 1L)
    size <- alt_len - ref_len                # +ve = insertion, -ve = deletion
    df$Variant_subtype <- ifelse(size >= 0, "INS", "DEL")
    range_end <- ifelse(size < 0, df$ChrPos + ref_len - 1, df$ChrPos)
    return(annotate_range(df, range_end, use_nearest = FALSE))
  }
  if (vt == "SVs") {
    toks <- strsplit(df$SNP, "_")
    df$Variant_subtype <- vapply(toks, function(x) if (length(x) >= 2) x[2] else "Unknown", "")
    sz <- vapply(toks, function(x){ i <- which(x == "len"); if (length(i)) as.numeric(x[i+1]) else NA_real_ }, 0)
    range_end <- ifelse(df$Variant_subtype %in% c("INS","TRA"), df$ChrPos, df$ChrPos + abs(sz))
    range_end[is.na(range_end)] <- df$ChrPos[is.na(range_end)]
    return(annotate_range(df, range_end, use_nearest = TRUE))
  }
  stop("unknown variant type: ", vt)
}

add_cis_trans <- function(df) {
  df %>%
    left_join(gene_location_fixed, by = c("Pheno" = "gene_id")) %>%
    mutate(
      Pheno_chr   = chr,
      Pheno_start = left,
      Pheno_end   = right,
      dist_to_pheno = case_when(
        Variant_Chr != Pheno_chr ~ NA_real_,
        ChrPos < Pheno_start ~ Pheno_start - ChrPos,
        ChrPos > Pheno_end   ~ ChrPos - Pheno_end,
        TRUE ~ 0),
      cis_trans = case_when(
        Variant_Chr != Pheno_chr ~ "Trans",
        dist_to_pheno == 0    ~ "Cis",
        dist_to_pheno <= CIS_WINDOW_BP ~ "Cis",
        Causal_Gene == Pheno  ~ "Cis",
        TRUE                  ~ "Trans"))
}
```

# 5. Run annotation, write per-type files, build combined table
```{r annotate-all}
annotated <- lapply(names(variant_cfg), function(vt) {
  message("annotating ", vt, " ...")
  read_clumped(variant_cfg[[vt]]$clumped, variant_cfg[[vt]]$assoc) %>%
    annotate_variant(vt) %>%
    add_cis_trans() %>%
    mutate(Genetic_variants = vt)
})
names(annotated) <- names(variant_cfg)

# save each annotated table (NA kept = intergenic_distant variants retained)
out_names <- c(SNPs = "snp_data_annotated_cis_trans.txt",
               CNVs = "cnv_data_annotated_cis_trans.txt",
               InDels = "indel_data_annotated_cis_trans.txt",
               SVs = "sv_data_annotated_cis_trans.txt")
for (vt in names(annotated)) {
  fwrite(annotated[[vt]], file.path(OUT, out_names[[vt]]),
         sep = "\t", quote = FALSE, row.names = FALSE, na = "NA")
  cat("wrote", out_names[[vt]], "-", nrow(annotated[[vt]]), "rows\n")
}

# ---- standardized combined table (one schema for every figure) ----
keep_cols <- c("Genetic_variants","SNP","Variant_Chr","ChrPos","Pheno","Pheno_chr",
               "PValue","SnpWeight","EffectSize","Causal_Gene","Causal_Gene_feature",
               "Variant_subtype","dist_to_pheno","cis_trans")

all_clumped <- bind_rows(lapply(annotated, function(d) d %>% select(any_of(keep_cols)))) %>%
  mutate(
    Location         = factor(ifelse(cis_trans == "Cis", "Cis", "Trans"), levels = c("Cis","Trans")),
    abs_SnpWeight    = abs(SnpWeight),
    Genetic_variants = factor(Genetic_variants, levels = c("SNPs","InDels","SVs","CNVs"))
  )

fwrite(all_clumped, file.path(OUT, "all_clumped_eqtl_annotated_YPD.1.txt"),
       sep = "\t", quote = FALSE, row.names = FALSE, na = "NA")

cat("Total eQTL associations:", nrow(all_clumped), "\n")
print(table(all_clumped$Genetic_variants, all_clumped$Location))
```

```{r counts-summary}
all_clumped %>%
  group_by(Genetic_variants) %>%
  summarise(n_associations = n(),
            n_unique_loci   = n_distinct(SNP),
            n_unique_traits = n_distinct(Pheno), .groups = "drop")
```

# Fig 1 — Effect size by variant type and location

## 1A — all variants, cis vs trans
```{r fig1a}
ss_a <- all_clumped %>% count(Location) %>% mutate(label = paste0(Location, "\n(n=", n, ")"))
mw_a <- wilcox.test(abs_SnpWeight ~ Location, data = all_clumped)

fig1a <- ggplot(all_clumped, aes(Location, abs_SnpWeight, fill = Location)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_y_continuous(limits = c(0.1, 1.0), breaks = c(0.1,0.25,0.5,0.75,1.0)) +
  scale_x_discrete(labels = setNames(ss_a$label, ss_a$Location)) +
  annotate("text", x = 1.5, y = Inf, vjust = 1.5, size = 5,
           label = paste0("p = ", format.pval(mw_a$p.value, digits = 3))) +
  scale_fill_manual(values = c(Cis = "#1F77B4", Trans = "#FF7F0E")) +
  labs(title = "Effect size by variant location", x = "Variant location", y = "Absolute effect size") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none",
        axis.title = element_text(face = "bold"),
        axis.text  = element_text(face = "bold", size = 16),
        panel.grid.major.x = element_blank(), panel.grid.minor = element_blank())
print(fig1a)
ggsave(file.path(OUT, "Fig1A_effect_size_cis_trans.tiff"), fig1a,
       width = 8, height = 8, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

## 1B — per variant type, cis vs trans
```{r fig1b}
pval_df <- all_clumped %>% group_by(Genetic_variants) %>%
  summarise(p = wilcox.test(abs_SnpWeight ~ Location)$p.value, .groups = "drop") %>%
  mutate(label = paste0("p = ", format.pval(p, digits = 3)), x = 1.5, y = Inf)
ssize_df <- all_clumped %>% count(Genetic_variants, Location) %>%
  mutate(x = ifelse(Location == "Cis", 1, 2), label = paste0("n=", n))

fig1b <- ggplot(all_clumped, aes(Location, abs_SnpWeight, fill = Genetic_variants)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  facet_wrap(~ Genetic_variants, nrow = 1) +
  scale_y_continuous(limits = c(0.1, 1.0), breaks = c(0.1,0.25,0.5,0.75,1.0)) +
  geom_text(data = pval_df, aes(x, y, label = label), inherit.aes = FALSE, vjust = 1.5, size = 4) +
  geom_text(data = ssize_df, aes(x, -Inf, label = label), inherit.aes = FALSE, vjust = -0.5,
            size = 3.5, fontface = "italic") +
  scale_fill_manual(values = variant_colors) +
  labs(title = "Effect size by variant type and location", x = "Variant location", y = "Absolute effect size") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none", axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "bold", size = 16),
        strip.text = element_text(size = 18, face = "bold"),
        strip.background = element_rect(fill = "gray95", color = "black"),
        panel.grid.major.x = element_blank(), panel.grid.minor = element_blank())
print(fig1b)
ggsave(file.path(OUT, "Fig1B_effect_size_per_variant.tiff"), fig1b,
       width = 12, height = 8, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

## 1C — effect size across variant types (pairwise Wilcoxon, BH)
```{r fig1c}
ord <- all_clumped %>% group_by(Genetic_variants) %>%
  summarise(m = median(abs_SnpWeight, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(m)) %>% pull(Genetic_variants) %>% as.character()
dat_c <- all_clumped %>% mutate(Genetic_variants = factor(Genetic_variants, levels = ord))

pw <- pairwise.wilcox.test(dat_c$abs_SnpWeight, dat_c$Genetic_variants, p.adjust.method = "BH")
pw_tab <- as.data.frame(as.table(pw$p.value)) %>% setNames(c("V1","V2","p_adj")) %>% filter(!is.na(p_adj))
pairs <- combn(ord, 2, simplify = FALSE)
pair_labels <- sapply(pairs, function(p) pw_tab %>%
  filter((V1==p[1]&V2==p[2])|(V1==p[2]&V2==p[1])) %>% pull(p_adj) %>% format.pval(digits = 3))
ss_c <- dat_c %>% count(Genetic_variants) %>% mutate(label = paste0("n=", n))

fig1c <- ggplot(dat_c, aes(Genetic_variants, abs_SnpWeight, fill = Genetic_variants)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  geom_text(data = ss_c, aes(Genetic_variants, -Inf, label = label), inherit.aes = FALSE,
            vjust = -0.5, size = 3.5, fontface = "italic") +
  geom_signif(comparisons = pairs, annotations = pair_labels,
              step_increase = 0.08, tip_length = 0.01, textsize = 3.5) +
  scale_y_continuous(limits = c(0.1, 1.3), breaks = c(0.1,0.25,0.5,0.75,1.0)) +
  scale_fill_manual(values = variant_colors) +
  labs(title = "Effect size by variant type", x = "Variant type", y = "Absolute effect size") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "none",
        axis.title = element_text(face = "bold"), axis.text.x = element_text(face = "bold", size = 15),
        panel.grid.major.x = element_blank(), panel.grid.minor = element_blank())
print(fig1c)
ggsave(file.path(OUT, "Fig1C_effect_size_by_variant_type.tiff"), fig1c,
       width = 10, height = 8, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

# Fig 2 — Genomic distribution of eQTLs (feature pie, per variant type)

Each variant's `Causal_Gene_feature` is collapsed to a single representative genomic
region. Range variants (Indel/SV) that overlap several features are assigned the
highest-priority region; unassigned variants are `Intergenic`.
```{r fig2}
region_levels <- c("Promoter","5' UTR","Start codon","CDS","Stop codon",
                   "3' UTR","Terminator","Intergenic","Other")
# map raw feature token -> display label, in priority order (first match wins)
simplify_feature <- function(f) {
  if (is.na(f) || f %in% c("", "NA")) return("Intergenic")
  parts <- trimws(unlist(strsplit(f, ";")))
  parts <- parts[parts != "" & parts != "NA"]
  if (length(parts) == 0) return("Intergenic")
  if (all(grepl("intergenic", parts))) return("Intergenic")
  parts <- parts[!grepl("intergenic", parts)]
  pr <- c(promoter="Promoter", five_prime_UTR="5' UTR", start_codon="Start codon",
          CDS="CDS", stop_codon="Stop codon", three_prime_UTR="3' UTR", terminator="Terminator")
  hit <- pr[names(pr) %in% parts]
  if (length(hit) == 0) return("Other")
  hit[[1]]   # names(pr) is already in priority order
}

feature_dist <- all_clumped %>%
  mutate(region = factor(vapply(Causal_Gene_feature, simplify_feature, ""), levels = region_levels)) %>%
  count(Genetic_variants, region, .drop = FALSE) %>%
  group_by(Genetic_variants) %>% mutate(pct = 100 * n / sum(n)) %>% ungroup() %>%
  filter(n > 0)

region_colors <- c("Promoter"="#8E44AD","5' UTR"="#3498DB","Start codon"="#6D4C41",
                   "CDS"="#2ECC71","Stop codon"="#F1C40F","3' UTR"="#E67E22",
                   "Terminator"="#E74C3C","Intergenic"="#95A5A6","Other"="#34495E")

# label position: cumulative midpoint (angle) of each slice within a facet, 0-100 scale
feature_dist <- feature_dist %>%
  group_by(Genetic_variants) %>%
  arrange(desc(region), .by_group = TRUE) %>%      # match slice stacking order
  mutate(fraction      = pct / 100,
         ymax          = cumsum(fraction),
         ymin          = c(0, head(ymax, -1)),
         labelPosition = (ymax + ymin) / 2,
         lab           = ifelse(round(pct, 1) == 0, "", paste0(round(pct, 1), "%"))) %>%
  ungroup()

# Bar drawn at numeric x = 1; labels placed OUTSIDE the pie (x = 1.8) with leader
# lines. ALL slices are labelled (no % threshold). ggrepel spreads labels along the
# angular (y) direction so adjacent thin slices don't collide.
fig2 <- ggplot(feature_dist, aes(x = 1, y = pct, fill = region)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  facet_wrap(~ Genetic_variants, nrow = 1) +
  geom_text_repel(aes(x = 1.85, y = ymid, label = lab),
                  size = 2.8, fontface = "bold", show.legend = FALSE,
                  segment.size = 0.3, segment.color = "grey40",
                  min.segment.length = 0, box.padding = 0.15,
                  max.overlaps = Inf, direction = "y", force = 0.6) +
  scale_x_continuous(limits = c(0, 2.6)) +     # room outside the pie for labels
  scale_fill_manual(values = region_colors, name = "Genomic region", drop = FALSE) +
  theme_void(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        strip.text = element_text(size = 14, face = "bold"), legend.position = "right")
print(fig2)
ggsave(file.path(OUT, "Fig2_genomic_region_pie.tiff"), fig2,
       width = 14, height = 6, dpi = 600, units = "in", device = "tiff", compression = "lzw")

# supporting table
feature_dist %>% mutate(pct = round(pct, 1)) %>% arrange(Genetic_variants, desc(pct)) %>% print(n = 100)
```
```{r}
feature_dist <- feature_dist %>%
  group_by(Genetic_variants) %>%
  mutate(lab = ifelse(round(pct, 1) == 0, "", paste0(round(pct, 1), "%"))) %>%
  ungroup()

fig2 <- ggplot(feature_dist, aes(x = region, y = pct, fill = region)) +
  geom_col(width = 0.75, color = "grey30", linewidth = 0.2) +
  geom_text(aes(label = lab), vjust = -0.3, size = 2.8, fontface = "bold") +
  facet_wrap(~ Genetic_variants, nrow = 1) +
  scale_fill_manual(values = region_colors, name = "Genomic region", drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Percentage of eQTL variants (%)") +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.text = element_text(size = 14, face = "bold"),
        strip.background = element_rect(fill = "gray95", color = "black"),
        legend.position = "right")
print(fig2)
```

# Fig 3 — Variant subtype composition within each variant type

SNPs and CNVs have no subtype (single bar); Indels split INS/DEL; SVs split into
INS/DEL/DUP/INV/TRA/CONTR.
```{r fig3}
subtype_counts <- all_clumped %>%
  count(Genetic_variants, Variant_subtype) %>%
  group_by(Genetic_variants) %>% mutate(pct = 100 * n / sum(n)) %>% ungroup()

fig3 <- ggplot(subtype_counts, aes(reorder(Variant_subtype, -n), n, fill = Genetic_variants)) +
  geom_col(color = "black", width = 0.75) +
  geom_text(aes(label = n), vjust = -0.3, size = 3.5, fontface = "bold") +
  facet_wrap(~ Genetic_variants, scales = "free_x", nrow = 1) +
  scale_fill_manual(values = variant_colors) +
  labs(title = "Variant subtype composition", x = "Subtype", y = "Number of eQTL variants") +
  theme_classic(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        strip.text = element_text(size = 14, face = "bold"),
        strip.background = element_rect(fill = "gray95", color = "black"))
print(fig3)
ggsave(file.path(OUT, "Fig3_variant_subtype_bar.tiff"), fig3,
       width = 12, height = 6, dpi = 600, units = "in", device = "tiff", compression = "lzw")

subtype_counts %>% mutate(pct = round(pct, 1)) %>% arrange(Genetic_variants, desc(n)) %>% print(n = 100)
```

# 6. Total variant counts from .bim (background for Fig 4 & Fig 5)
```{r bim-counts}
total_variants_df <- bind_rows(lapply(names(bim_paths), function(vt) {
  n <- nrow(read.table(bim_paths[[vt]], header = FALSE, stringsAsFactors = FALSE))
  data.frame(variant_type = vt, total_count = n)
}))

eqtl_counts <- all_clumped %>% group_by(Genetic_variants) %>%
  summarise(eqtl_count = n_distinct(SNP), .groups = "drop") %>%
  dplyr::rename(variant_type = Genetic_variants) %>% mutate(variant_type = as.character(variant_type))

variant_summary <- total_variants_df %>%
  left_join(eqtl_counts, by = "variant_type") %>%
  replace_na(list(eqtl_count = 0)) %>%
  mutate(pct_of_total        = 100 * eqtl_count / total_count,
         pct_of_eqtl         = 100 * eqtl_count / sum(eqtl_count),
         pct_of_all_variants = 100 * total_count / sum(total_count)) %>%
  arrange(desc(eqtl_count))
print(variant_summary)
```

# Fig 4 — Donut (eQTL vs total variant composition)
```{r fig4}
inner <- variant_summary %>% transmute(variant_type, count = eqtl_count,
                                        percentage = pct_of_eqtl, circle = "eQTL")
outer <- variant_summary %>% transmute(variant_type, count = total_count,
                                        percentage = pct_of_all_variants, circle = "Total")
donut <- bind_rows(inner, outer) %>%
  mutate(circle = factor(circle, levels = c("Total","eQTL")),
         variant_type = factor(variant_type, levels = c("SNPs","InDels","SVs","CNVs"))) %>%
  group_by(circle) %>% arrange(circle, desc(variant_type)) %>%
  mutate(fraction = percentage/100, ymax = cumsum(fraction),
         ymin = c(0, head(ymax, -1)), labelPosition = (ymax+ymin)/2,
         label = paste0(round(percentage, 1), "%")) %>% ungroup() %>%
  left_join(data.frame(circle = factor(c("eQTL","Total"), levels = c("eQTL","Total")),
                       xmin = c(1.5, 2.5), xmax = c(2.5, 4.4)), by = "circle")

fmtk <- function(x) ifelse(x >= 1000, paste0(round(x/1000, 1), "k"), as.character(x))

fig4 <- ggplot(donut, aes(ymax = ymax, ymin = ymin, xmax = xmax, xmin = xmin, fill = variant_type)) +
  geom_rect(color = "white", linewidth = 2) +
  geom_label(aes(x = (xmin+xmax)/2, y = labelPosition, label = label),
             fill = NA, label.size = NA, fontface = "bold", size = 6) +
  annotate("text", x = 0.5, y = 0.5, size = 8,
           label = paste0("eQTL\n(n = ", fmtk(sum(variant_summary$eqtl_count)), ")")) +
  scale_fill_manual(values = variant_colors, name = "Variant type") +
  coord_polar(theta = "y") + xlim(c(0, 5)) + theme_void() +
  theme(legend.position = "top", legend.title = element_blank(),
        legend.text = element_text(size = 16), plot.caption = element_text(size = 16, hjust = 0.5)) +
  labs(caption = paste0("Total variants (n = ", fmtk(sum(variant_summary$total_count)), ")"))
print(fig4)
ggsave(file.path(OUT, "Fig4_donut_eqtl.tiff"), fig4,
       width = 10, height = 10, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

# Fig 5 — Odds ratio of eQTL enrichment (each type vs all others)
```{r fig5}
or_one <- function(v) {
  a <- variant_summary$eqtl_count[variant_summary$variant_type == v]
  b <- variant_summary$total_count[variant_summary$variant_type == v] - a
  c <- sum(variant_summary$eqtl_count[variant_summary$variant_type != v])
  d <- sum(variant_summary$total_count[variant_summary$variant_type != v]) - c
  ft <- fisher.test(matrix(c(a,b,c,d), 2, 2, byrow = TRUE))
  data.frame(VariantType = v, a_eqtl = a, b_not = b, c_oth = c, d_oth_not = d,
             OddsRatio = (a*d)/(b*c), PValue = ft$p.value)
}
or_results <- bind_rows(lapply(c("SNPs","InDels","SVs","CNVs"), or_one)) %>%
  mutate(PValue_FDR = p.adjust(PValue, "BH"),
         Significant = ifelse(PValue_FDR < 0.05, "Yes", "No"),
         VariantType = factor(VariantType, levels = c("SNPs","InDels","SVs","CNVs")))
print(or_results)

fig5 <- ggplot(or_results, aes(VariantType, OddsRatio, fill = VariantType)) +
  geom_col(color = "black", linewidth = 1.1, width = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", linewidth = 1.5) +
  geom_text(aes(label = paste0("OR=", round(OddsRatio, 2), "\np=",
                               format(PValue_FDR, scientific = TRUE, digits = 2))),
            vjust = -0.3, size = 5, fontface = "bold") +
  ylim(0, max(or_results$OddsRatio) * 1.25) +
  scale_fill_manual(values = variant_colors) +
  labs(title = "Odds ratio of eQTL enrichment\n(each variant type vs all others)",
       x = "Variant type", y = "Odds ratio",
       caption = "Red dashed line = OR 1 (no enrichment)") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "none",
        axis.title = element_text(face = "bold"), axis.text = element_text(face = "bold", size = 16),
        panel.grid.major.x = element_blank(), panel.grid.minor = element_blank())
print(fig5)
ggsave(file.path(OUT, "Fig5_odds_ratio.tiff"), fig5,
       width = 10, height = 8, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```

# Fig 6 — Pleiotropy (traits affected per variant, by type)
```{r fig6}
pleio <- all_clumped %>% group_by(Genetic_variants, SNP) %>%
  summarise(n_traits = n_distinct(Pheno), .groups = "drop")

pleio_summary <- pleio %>% group_by(Genetic_variants) %>%
  summarise(n_variants = n(), mean_traits = round(mean(n_traits), 2),
            median_traits = median(n_traits), max_traits = max(n_traits), .groups = "drop") %>%
  arrange(desc(mean_traits))
print(pleio_summary)

kw <- kruskal.test(n_traits ~ Genetic_variants, data = pleio)
cat("Kruskal-Wallis p =", format(kw$p.value, scientific = TRUE, digits = 3), "\n")
print(pairwise.wilcox.test(pleio$n_traits, pleio$Genetic_variants, p.adjust.method = "BH")$p.value)

fig6 <- ggplot(pleio, aes(reorder(Genetic_variants, n_traits, median), n_traits, fill = Genetic_variants)) +
  geom_boxplot(outlier.alpha = 0.3, width = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.8) +
  scale_fill_manual(values = variant_colors) + scale_y_log10() +
  labs(title = "Pleiotropy by variant type",
       subtitle = "Traits (genes) associated per unique eQTL variant",
       x = "Variant type", y = "Number of associated traits (log scale)",
       caption = "Kruskal-Wallis + pairwise Wilcoxon, FDR corrected") +
  theme_classic(base_size = 13) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 10),
        axis.text = element_text(face = "bold", size = 13), axis.title = element_text(face = "bold"))
print(fig6)
ggsave(file.path(OUT, "Fig6_pleiotropy.tiff"), fig6,
       width = 8, height = 6, dpi = 600, units = "in", device = "tiff", compression = "lzw")
```
# Fig 7 — Pleiotropy by variant SUBTYPE (SVs and InDels only)

Only SVs and InDels carry sub-categories, so this asks: within each of those two
types, which subtype is the most pleiotropic (affects the most traits per variant)?
Traits-per-variant is counted per unique variant ID; subtypes are ordered by median
within each type (most pleiotropic first). Kruskal-Wallis tests across subtypes
within each type; pairwise Wilcoxon (BH) where >=2 subtypes have enough variants.
```{r fig7}
pleio_sub <- all_clumped %>%
  filter(Genetic_variants %in% c("SVs", "InDels"),
         !is.na(Variant_subtype), Variant_subtype != "", Variant_subtype != "Unknown") %>%
  group_by(Genetic_variants, Variant_subtype, SNP) %>%
  summarise(n_traits = n_distinct(Pheno), .groups = "drop") %>%
  mutate(Genetic_variants = factor(Genetic_variants, levels = c("InDels", "SVs")))

# per-type, per-subtype summary
pleio_sub_summary <- pleio_sub %>%
  group_by(Genetic_variants, Variant_subtype) %>%
  summarise(n_variants   = n(),
            mean_traits  = round(mean(n_traits), 2),
            median_traits= median(n_traits),
            max_traits   = max(n_traits), .groups = "drop") %>%
  arrange(Genetic_variants, desc(mean_traits))
print(pleio_sub_summary)

# Kruskal-Wallis across subtypes within each type (needs >=2 subtypes present)
kw_sub <- pleio_sub %>%
  group_by(Genetic_variants) %>%
  summarise(
    n_subtypes = n_distinct(Variant_subtype),
    kw_p = if (n_distinct(Variant_subtype) >= 2)
             kruskal.test(n_traits ~ factor(Variant_subtype))$p.value else NA_real_,
    .groups = "drop") %>%
  mutate(label = ifelse(is.na(kw_p), "KW: n/a",
                        paste0("KW p = ", format.pval(kw_p, digits = 3))))
print(kw_sub)

# order subtypes by median within each type so the most pleiotropic reads first
sub_order <- pleio_sub %>%
  group_by(Genetic_variants, Variant_subtype) %>%
  summarise(med = median(n_traits), .groups = "drop") %>%
  arrange(Genetic_variants, desc(med))
pleio_sub <- pleio_sub %>%
  mutate(Variant_subtype = factor(Variant_subtype, levels = unique(sub_order$Variant_subtype)))

# sample-size labels under each subtype box
ss_sub <- pleio_sub %>% count(Genetic_variants, Variant_subtype) %>%
  mutate(label = paste0("n=", n))

fig7 <- ggplot(pleio_sub, aes(Variant_subtype, n_traits, fill = Genetic_variants)) +
  geom_boxplot(outlier.alpha = 0.3, width = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.8) +
  geom_text(data = ss_sub, aes(Variant_subtype, -Inf, label = label),
            inherit.aes = FALSE, vjust = -0.6, size = 3, fontface = "italic") +
  geom_text(data = kw_sub, aes(x = Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = 1.05, vjust = 1.5, size = 4) +
  facet_wrap(~ Genetic_variants, scales = "free_x", nrow = 1) +
  scale_fill_manual(values = variant_colors) + scale_y_log10() +
  labs(title = "Pleiotropy by variant subtype",
       subtitle = "Traits (genes) associated per unique variant — SVs and InDels only",
       x = "Variant subtype", y = "Number of associated traits (log scale)",
       caption = "Subtypes ordered by median; Kruskal-Wallis across subtypes within each type") +
  theme_classic(base_size = 13) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.text.y = element_text(face = "bold", size = 12),
        axis.title = element_text(face = "bold"),
        strip.text = element_text(size = 14, face = "bold"),
        strip.background = element_rect(fill = "gray95", color = "black"))
print(fig7)
ggsave(file.path(OUT, "Fig7_pleiotropy_by_subtype.tiff"), fig7,
       width = 10, height = 6, dpi = 600, units = "in", device = "tiff", compression = "lzw")

# pairwise Wilcoxon (BH) per type, where a type has >=2 subtypes
for (vt in levels(pleio_sub$Genetic_variants)) {
  d <- pleio_sub %>% filter(Genetic_variants == vt)
  if (n_distinct(d$Variant_subtype) >= 2) {
    cat("\n== pairwise Wilcoxon (BH),", vt, "==\n")
    print(pairwise.wilcox.test(d$n_traits, droplevels(d$Variant_subtype),
                               p.adjust.method = "BH")$p.value)
  }
}
```
