---
title: "Fig.3-panel-A-cis_trans_eqtl_mapping_YPD"
author: "Akriti Agrawal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: true
    number_section: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../results/") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 10)
```

```{r libs, include=FALSE}
library(tidyverse)
library(data.table)
library(GenomicRanges)
select <- dplyr::select
filter <- dplyr::filter
```

## 0. Config — set your four .clumped/assoc paths here

```{r config}
# ===================================================================
# Configuration with ABSOLUTE PATHS
# ===================================================================
RES_flu   <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results"
RES_sc2glu   <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results"

CLMP_flu  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/results/plink/"
CLMP_sc2glu <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/results/plink/"

# Output directory - ABSOLUTE PATH
OUT_MAIN  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/Agrawal_single_cell_eqtl_cell_cycle/Figure.3/Fig.3-PartII-compare-conditions/results/eqtl_plot"

OUT_ANNOT <- file.path(OUT_MAIN, "annotations")
OUT_FIG   <- file.path(OUT_MAIN, "figures")
OUT_SUMM  <- file.path(OUT_MAIN, "data_summary")

# Create output directories
for (dir in c(OUT_MAIN, OUT_ANNOT, OUT_FIG, OUT_SUMM)) {
  dir.create(dir, showWarnings = FALSE, recursive = TRUE)
}

ANNOT_CSV_flu <- "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"
ANNOT_CSV_sc2glu <- "/scratch/aa9679/SingleCell_cerevisiae/Gene_exp_analysis/yeast_custom_annotations.csv"

GENE_LOC_flu  <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/merged_fluconazole/new_fastlmm_march_26_2026/data/pheno_data/gene_location.txt"
GENE_LOC_sc2glu <- "/projects/rps/cgsb/gresham/Akriti/wildyeast1/common_analysis_for_all/sc2glu.2/fastlmm/data/pheno_data/gene_location.txt"

CIS_WINDOW_BP <- 2000

# ===================================================================
# Variant configuration fluconazole
# ===================================================================
variant_cfg_flu <- list(
  SNP = list(
    clumped = file.path(CLMP_flu, "clumped_results_snp.clumped"),
    assoc   = file.path(RES_flu,  "all_signif_SNP_flu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  ),
  CNV = list(
    clumped = file.path(CLMP_flu, "clumped_results_CNVs.clumped"),
    assoc   = file.path(RES_flu,  "all_signif_CNV_flu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  ),
  Indel = list(
    clumped = file.path(CLMP_flu, "clumped_results_Indels.clumped"),
    assoc   = file.path(RES_flu,  "all_signif_Indels_flu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  ),
  SV = list(
    clumped = file.path(CLMP_flu, "clumped_results_SVs.clumped"),
    assoc   = file.path(RES_flu,  "all_signif_SVs_flu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  )
)

# ===================================================================
# Variant configuration sc2glu
# ===================================================================
variant_cfg_sc2glu <- list(
  SNP = list(
    clumped = file.path(CLMP_sc2glu, "clumped_results_snp.clumped"),
    assoc   = file.path(RES_sc2glu,  "all_signif_SNP_sc2glu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  ),
  CNV = list(
    clumped = file.path(CLMP_sc2glu, "clumped_results_CNVs.clumped"),
    assoc   = file.path(RES_sc2glu,  "all_signif_CNV_sc2glu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  ),
  Indel = list(
    clumped = file.path(CLMP_sc2glu, "clumped_results_Indels.clumped"),
    assoc   = file.path(RES_sc2glu,  "all_signif_Indels_sc2glu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  ),
  SV = list(
    clumped = file.path(CLMP_sc2glu, "clumped_results_SVs.clumped"),
    assoc   = file.path(RES_sc2glu,  "all_signif_SVs_sc2glu_with_double_id_final_filtered_for_ncount_less_than_5_with_threshold.txt")
  )
)

# ===================================================================
# Condition metadata
# ===================================================================
condition_meta <- list(
  fluconazole = list(
    variant_cfg = variant_cfg_flu,
    gene_loc_file = GENE_LOC_flu,
    annot_csv = ANNOT_CSV_flu,
    label = "Fluconazole"
  ),
  sc2glu = list(
    variant_cfg = variant_cfg_sc2glu,
    gene_loc_file = GENE_LOC_sc2glu,
    annot_csv = ANNOT_CSV_sc2glu,
    label = "SC2Glu"
  )
)

# Variant colors
variant_colors <- c("SNP" = "#4472C4", "Indel" = "#FFC000",
                    "SV" = "#C41E3A", "CNV" = "#70AD47")

cat("════════════════════════════════════════════════════════════\n")
cat("CONFIGURATION SUMMARY\n")
cat("════════════════════════════════════════════════════════════\n")
cat("Conditions:                   ", paste(names(condition_meta), collapse = ", "), "\n")
cat("Output directory (OUT_MAIN):  ", OUT_MAIN, "\n")
cat("Variant types:                SNP, CNV, Indel, SV\n")
cat("════════════════════════════════════════════════════════════\n\n")
```

## 1. Shared reference objects

```{r reference}
roman16 <- c("I","II","III","IV","V","VI","VII","VIII","IX","X","XI","XII","XIII","XIV","XV","XVI")
num_to_roman <- function(x) roman16[x]

chr_lengths <- c(I=230218, II=813184, III=316620, IV=1531933, V=576874, VI=270161,
                 VII=1090940, VIII=562643, IX=439888, X=745751, XI=666816,
                 XII=1078177, XIII=924431, XIV=784333, XV=1091291, XVI=948066)
chr_offsets <- c(0, cumsum(chr_lengths[-length(chr_lengths)])); names(chr_offsets) <- roman16
chr_mids <- chr_offsets + chr_lengths / 2

feature_priority <- c(promoter=1, three_prime_UTR=2, five_prime_UTR=3,
                      start_codon=4, stop_codon=5, CDS=6)
prio <- function(f) { p <- feature_priority[f]; p[is.na(p)] <- 99; p }
```

## 2. Reader function

```{r read-clumped}
read_clumped <- function(clumped, assoc) {
  ids <- read.table(clumped, header = TRUE)$SNP
  read.table(assoc, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    filter(SNP %in% ids)
}
```

## 3. Point annotation (SNP = priority, CNV = first overlap)

```{r annot-point}
annotate_point <- function(df, annotations_gr, pick = c("priority","first")) {
  pick <- match.arg(pick)
  df$var_chr_roman <- num_to_roman(df$Chr)
  gr <- GRanges(df$var_chr_roman, IRanges(df$ChrPos, df$ChrPos))
  df$causal_gene <- NA_character_; df$causal_gene_feature <- NA_character_

  ov <- findOverlaps(gr, annotations_gr, type = "any")
  for (i in unique(queryHits(ov))) {
    subj <- subjectHits(ov)[queryHits(ov) == i]
    k <- if (pick == "priority") subj[which.min(prio(annotations_gr$type[subj]))] else subj[1]
    df$causal_gene[i] <- annotations_gr$gene_id[k]
    df$causal_gene_feature[i] <- annotations_gr$type[k]
  }
  # intergenic: nearest gene within 2 kb, else intergenic_distant
  m <- which(is.na(df$causal_gene))
  if (length(m)) {
    ng <- nearest(gr[m], annotations_gr); ok <- !is.na(ng)
    d <- rep(NA_real_, length(m)); d[ok] <- distance(gr[m][ok], annotations_gr[ng[ok]])
    near <- !is.na(d) & d <= CIS_WINDOW_BP
    df$causal_gene[m[near]]          <- annotations_gr$gene_id[ng[near]]
    df$causal_gene_feature[m[near]]  <- "intergenic_nearby"
    df$causal_gene_feature[m[!near]] <- "intergenic_distant"
  }
  df
}
```

## 4. Range annotation (Indel = no nearest, SV = nearest)

```{r annot-range}
annotate_range <- function(df, annotations_gr, range_end, use_nearest) {
  df$var_chr_roman <- num_to_roman(df$Chr)
  gr <- GRanges(df$var_chr_roman, IRanges(df$ChrPos, range_end))
  df$causal_gene <- NA_character_; df$causal_gene_feature <- NA_character_

  ov <- findOverlaps(gr, annotations_gr, type = "any")
  for (i in unique(queryHits(ov))) {
    subj <- subjectHits(ov)[queryHits(ov) == i]
    df$causal_gene[i]         <- paste(unique(annotations_gr$gene_id[subj]), collapse = "; ")
    df$causal_gene_feature[i] <- paste(unique(annotations_gr$type[subj]),   collapse = "; ")
  }
  if (use_nearest) {
    m <- which(is.na(df$causal_gene))
    if (length(m)) {
      ng <- nearest(gr[m], annotations_gr); ok <- !is.na(ng)
      d <- rep(NA_real_, length(m)); d[ok] <- distance(gr[m][ok], annotations_gr[ng[ok]])
      near <- !is.na(d) & d <= CIS_WINDOW_BP
      df$causal_gene[m[near]]          <- annotations_gr$gene_id[ng[near]]
      df$causal_gene_feature[m[near]]  <- "intergenic_nearby"
      df$causal_gene_feature[m[!near]] <- "intergenic_distant"
    }
  } else {
    df$causal_gene_feature[is.na(df$causal_gene)] <- "intergenic_distant"
  }
  df
}
```

## 5. Per-type dispatch

```{r annot-dispatch}
annotate_variant <- function(df, annotations_gr, vt) {
  if (vt == "SNP") return(annotate_point(df, annotations_gr, "priority"))
  if (vt == "CNV") return(annotate_point(df, annotations_gr, "first"))

  if (vt == "Indel") {
    p <- strsplit(df$SNP, "_")
    ref_len <- vapply(p, function(x) nchar(x[3]), 1L)
    alt_len <- vapply(p, function(x) nchar(x[4]), 1L)
    df$Indel_Size <- alt_len - ref_len
    range_end <- ifelse(df$Indel_Size < 0, df$ChrPos + ref_len - 1, df$ChrPos)
    return(annotate_range(df, annotations_gr, range_end, use_nearest = FALSE))
  }
  if (vt == "SV") {
    sub <- vapply(strsplit(df$SNP, "_"), `[`, "", 2)
    sz  <- vapply(strsplit(df$SNP, "_"), function(x){ i <- which(x=="len"); if(length(i)) as.numeric(x[i+1]) else NA_real_ }, 0)
    range_end <- ifelse(sub %in% c("INS","TRA"), df$ChrPos, df$ChrPos + abs(sz))
    range_end[is.na(range_end)] <- df$ChrPos[is.na(range_end)]
    return(annotate_range(df, annotations_gr, range_end, use_nearest = TRUE))
  }
}
```

## 6. Cis/trans classification

```{r harmonise}
add_cis_trans <- function(df, gene_location_fixed, vt) {
  df %>%
    left_join(gene_location_fixed, by = c("Pheno" = "gene_id")) %>%
    mutate(Pheno_chr = chr, Pheno_start = left, Pheno_end = right,
           dist_to_pheno = case_when(
             var_chr_roman != Pheno_chr ~ NA_real_,
             ChrPos < Pheno_start ~ Pheno_start - ChrPos,
             ChrPos > Pheno_end   ~ ChrPos - Pheno_end,
             TRUE ~ 0),
           cis_trans = case_when(
             var_chr_roman != Pheno_chr ~ "Trans",
             dist_to_pheno == 0    ~ "Cis",
             dist_to_pheno <= 2000 ~ "Cis",
             causal_gene == Pheno  ~ "Cis",
             TRUE                  ~ "Trans"),
           variant_type = vt,
           abs_effect_size = abs(SnpWeight)) %>%
    select(variant_type, SNP, Pheno, var_chr_roman, ChrPos,
           Pheno_chr, Pheno_start, Pheno_end, causal_gene,
           cis_trans, abs_effect_size, PValue)
}
```

## 7. Master processing function for each condition

```{r process-condition}
process_condition <- function(condition_name, meta) {
  cat("\n")
  cat("════════════════════════════════════════════════════════════\n")
  cat("PROCESSING CONDITION: ", meta$label, "\n")
  cat("════════════════════════════════════════════════════════════\n\n")
  
  # Load annotations
  ann <- read.csv(meta$annot_csv)
  annotations_gr <- GRanges(ann$seqnames,
                            IRanges(ann$start, ann$end),
                            strand = ann$strand,
                            type = ann$type, gene_id = ann$gene_id, gene_name = ann$gene_name)
  
  # Load gene locations
  gene_location_fixed <- read.table(meta$gene_loc_file, header = TRUE, stringsAsFactors = FALSE) %>%
    select(gene_id, gene_name, chr, left, right) %>%
    mutate(right = right + 2,
           gene_id = str_replace(gene_id, "-", "."))
  
  # Process each variant type
  combined <- lapply(names(meta$variant_cfg), function(vt) {
    cfg <- meta$variant_cfg[[vt]]
    message("  Annotating ", vt, " ...")
    read_clumped(cfg$clumped, cfg$assoc) %>% 
      annotate_variant(annotations_gr, vt) %>% 
      add_cis_trans(gene_location_fixed, vt)
  }) %>% 
    bind_rows() %>%
    mutate(var_chr_roman = factor(var_chr_roman, levels = roman16),
           Pheno_chr     = factor(Pheno_chr,     levels = roman16),
           variant_type  = factor(variant_type,  levels = c("SNP","CNV","Indel","SV")),
           var_genome    = chr_offsets[as.character(var_chr_roman)] + ChrPos,
           pheno_genome  = chr_offsets[as.character(Pheno_chr)] + (Pheno_start + Pheno_end)/2,
           condition     = meta$label) %>%
    filter(!is.na(var_genome), !is.na(pheno_genome), !is.na(cis_trans))
  
  cat("\nVariant type × cis/trans classification:\n")
  print(table(combined$variant_type, combined$cis_trans))
  cat("\n")
  
  return(combined)
}
```

## 8. Process both conditions

```{r process-all-conditions}
all_results <- lapply(names(condition_meta), function(cond_name) {
  process_condition(cond_name, condition_meta[[cond_name]])
}) 
names(all_results) <- names(condition_meta)

# Create combined dataset
combined_all <- bind_rows(all_results) %>%
  mutate(condition = factor(condition, levels = c("Fluconazole", "SC2Glu")))

cat("════════════════════════════════════════════════════════════\n")
cat("SUMMARY ACROSS CONDITIONS\n")
cat("════════════════════════════════════════════════════════════\n")
print(table(combined_all$condition, combined_all$variant_type, combined_all$cis_trans))
cat("\n")
```

## 8.5. Summary Statistics — Per Condition, Per Variant Type, and Global

```{r summary-statistics}
cat("\n")
cat("════════════════════════════════════════════════════════════\n")
cat("SUMMARY STATISTICS — eQTL COUNTS\n")
cat("════════════════════════════════════════════════════════════\n\n")

# ─────────────────────────────────────────────────────────────
# PER-CONDITION SUMMARIES
# ─────────────────────────────────────────────────────────────

for (cond_name in names(all_results)) {
  cond_label <- condition_meta[[cond_name]]$label
  data <- all_results[[cond_name]]
  
  cat("\n", strrep("─", 60), "\n")
  cat("CONDITION: ", cond_label, "\n")
  cat(strrep("─", 60), "\n\n")
  
  # Per variant type
  cat("Summary by variant type:\n")
  per_type <- data %>%
    group_by(variant_type) %>%
    summarise(n_associations = n(),
              n_unique_loci   = n_distinct(SNP),
              n_unique_traits = n_distinct(Pheno), 
              .groups = "drop")
  print(per_type)
  
  # Per variant type × cis/trans
  cat("\n\nBreakdown by cis/trans status:\n")
  per_type_cistrans <- data %>%
    group_by(variant_type, cis_trans) %>%
    summarise(n_associations = n(),
              n_unique_loci   = n_distinct(SNP),
              n_unique_traits = n_distinct(Pheno), 
              .groups = "drop")
  print(per_type_cistrans)
  
  # Total for this condition
  cat("\n\nTotal across all variant types (", cond_label, "):\n")
  total_cond <- data %>%
    summarise(n_associations = n(),
              n_unique_loci   = n_distinct(SNP),
              n_unique_traits = n_distinct(Pheno))
  print(total_cond)
  cat("\n")
}

# ─────────────────────────────────────────────────────────────
# GLOBAL SUMMARY ACROSS ALL CONDITIONS
# ─────────────────────────────────────────────────────────────

cat("\n", strrep("═", 60), "\n")
cat("GLOBAL SUMMARY — ACROSS ALL CONDITIONS\n")
cat(strrep("═", 60), "\n\n")

# Per variant type (collapsed across conditions)
cat("Summary by variant type (all conditions combined):\n")
global_type <- combined_all %>%
  group_by(variant_type) %>%
  summarise(n_associations = n(),
            n_unique_loci   = n_distinct(SNP),
            n_unique_traits = n_distinct(Pheno), 
            .groups = "drop")
print(global_type)

# Per condition (collapsed across variant types)
cat("\n\nSummary by condition (all variant types combined):\n")
global_cond <- combined_all %>%
  group_by(condition) %>%
  summarise(n_associations = n(),
            n_unique_loci   = n_distinct(SNP),
            n_unique_traits = n_distinct(Pheno), 
            .groups = "drop")
print(global_cond)

# Per condition × variant type
cat("\n\nFull matrix: Condition × Variant Type:\n")
matrix_summary <- combined_all %>%
  group_by(condition, variant_type) %>%
  summarise(n_associations = n(),
            n_unique_loci   = n_distinct(SNP),
            n_unique_traits = n_distinct(Pheno), 
            .groups = "drop")
print(matrix_summary)

# Grand total
cat("\n\nGRAND TOTAL (all conditions, all variant types):\n")
grand_total <- combined_all %>%
  summarise(n_associations = n(),
            n_unique_loci   = n_distinct(SNP),
            n_unique_traits = n_distinct(Pheno))
print(grand_total)

# Cis vs Trans split (global)
cat("\n\nGlobal Cis/Trans split (all conditions, all variant types):\n")
cistrans_split <- combined_all %>%
  group_by(cis_trans) %>%
  summarise(n_associations = n(),
            n_unique_loci   = n_distinct(SNP),
            n_unique_traits = n_distinct(Pheno), 
            pct_associations = round(100 * n() / nrow(combined_all), 1),
            .groups = "drop")
print(cistrans_split)

cat("\n")
cat("════════════════════════════════════════════════════════════\n")
cat("STATISTICS SUMMARY COMPLETE ✓\n")
cat("════════════════════════════════════════════════════════════\n\n")
```



## 9. Panel A: Individual plots per condition (Publication-ready)

```{r panelA-individual, fig.width=12, fig.height=11}
# Create plots for each condition
plot_list <- lapply(names(all_results), function(cond_name) {
  data <- all_results[[cond_name]]
  cond_label <- condition_meta[[cond_name]]$label
  
  p <- ggplot(data, aes(var_genome, pheno_genome, color = cis_trans, size = abs_effect_size)) +
    geom_point(alpha = 0.6) +
    facet_wrap(~ variant_type, ncol = 2) +
    scale_color_manual(
      name = "eQTL type",
      values = c(Cis = "#d73027", Trans = "#4575b4"),
      guide = guide_legend(title.position = "top", title.hjust = 0.5, 
                           label.theme = element_text(size = 11, face = "bold"),
                           title.theme = element_text(size = 12, face = "bold"))
    ) +
    scale_size_continuous(
      name = "Effect size (|β|)",
      range = c(0.4, 2.6),
      trans = "log10",
      guide = guide_legend(title.position = "top", title.hjust = 0.5,
                           label.theme = element_text(size = 11),
                           title.theme = element_text(size = 12, face = "bold"))
    ) +
    scale_x_continuous(breaks = chr_mids, labels = names(chr_mids), expand = c(0.01, 0)) +
    scale_y_continuous(breaks = chr_mids, labels = names(chr_mids), expand = c(0.01, 0)) +
    labs(
      title = paste("eQTL mapping:", cond_label),
      x = "Variant position (chromosome)",
      y = "Gene position (chromosome)"
    ) +
    theme_bw(base_size = 13) +
    theme(
      # Title
      plot.title = element_text(size = 14, face = "bold", color = "black", hjust = 0.5, margin = margin(b=10)),
      # Axis titles
      axis.title = element_text(size = 13, face = "bold", color = "black"),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      # Axis text
      axis.text = element_text(size = 11, color = "black", face = "bold"),
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      # Facet labels
      strip.text = element_text(size = 12, face = "bold", color = "black"),
      strip.background = element_rect(fill = "grey90", color = "black", linewidth = 0.8),
      # Panel borders
      panel.border = element_rect(color = "black", linewidth = 0.8),
      panel.grid = element_line(color = "grey95", linewidth = 0.3),
      # Legend
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.box.just = "center",
      legend.spacing.x = unit(1, "cm"),
      legend.background = element_rect(fill = "white", color = "black", linewidth = 0.8),
      legend.margin = margin(10, 10, 10, 10),
      legend.text = element_text(size = 11, color = "black"),
      legend.title = element_text(size = 12, face = "bold", color = "black"),
      # Plot margins
      plot.margin = margin(15, 15, 15, 15)
    )
  
  return(p)
})

# Save individual plots
for (cond_name in names(all_results)) {
  cond_label <- condition_meta[[cond_name]]$label
  clean_label <- tolower(gsub(" ", "_", cond_label))
  
  print(plot_list[[cond_name]])
  
  ggsave(file.path(OUT_FIG, paste0("Fig3_panelA_cis_trans_2x2_", clean_label, ".tiff")), 
         plot_list[[cond_name]],
         width = 12, height = 11, units = "in", dpi = 600, compression = "lzw")
  ggsave(file.path(OUT_FIG, paste0("Fig3_panelA_cis_trans_2x2_", clean_label, ".png")), 
         plot_list[[cond_name]],
         width = 12, height = 11, units = "in", dpi = 300)
  
  cat("✓ Saved: Fig3_panelA_cis_trans_2x2_", clean_label, ".tiff & .png\n")
}
```

## 10. Combined comparison plot (optional)

```{r panelA-combined, fig.width=14, fig.height=12}
panelA_combined <- ggplot(combined_all, aes(var_genome, pheno_genome, color = cis_trans, size = abs_effect_size)) +
  geom_point(alpha = 0.6) +
  facet_grid(condition ~ variant_type) +
  scale_color_manual(
    name = "eQTL type",
    values = c(Cis = "#d73027", Trans = "#4575b4"),
    guide = guide_legend(title.position = "top", title.hjust = 0.5, 
                         label.theme = element_text(size = 11, face = "bold"),
                         title.theme = element_text(size = 12, face = "bold"))
  ) +
  scale_size_continuous(
    name = "Effect size (|β|)",
    range = c(0.4, 2.6),
    trans = "log10",
    guide = guide_legend(title.position = "top", title.hjust = 0.5,
                         label.theme = element_text(size = 11),
                         title.theme = element_text(size = 12, face = "bold"))
  ) +
  scale_x_continuous(breaks = chr_mids, labels = names(chr_mids), expand = c(0.01, 0)) +
  scale_y_continuous(breaks = chr_mids, labels = names(chr_mids), expand = c(0.01, 0)) +
  labs(
    title = "Cis/Trans eQTL mapping across conditions",
    x = "Variant position (chromosome)",
    y = "Gene position (chromosome)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, face = "bold", color = "black", hjust = 0.5, margin = margin(b=10)),
    axis.title = element_text(size = 12, face = "bold", color = "black"),
    axis.text = element_text(size = 10, color = "black", face = "bold"),
    strip.text = element_text(size = 11, face = "bold", color = "black"),
    strip.background = element_rect(fill = "grey90", color = "black", linewidth = 0.8),
    panel.border = element_rect(color = "black", linewidth = 0.8),
    panel.grid = element_line(color = "grey95", linewidth = 0.3),
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.spacing.x = unit(1, "cm"),
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.8),
    legend.margin = margin(10, 10, 10, 10),
    plot.margin = margin(15, 15, 15, 15)
  )

print(panelA_combined)
ggsave(file.path(OUT_FIG, "Fig3_panelA_cis_trans_combined_conditions.tiff"), 
       panelA_combined, width = 14, height = 12, units = "in", dpi = 600, compression = "lzw")
ggsave(file.path(OUT_FIG, "Fig3_panelA_cis_trans_combined_conditions.png"), 
       panelA_combined, width = 14, height = 12, units = "in", dpi = 300)

cat("✓ Saved: Fig3_panelA_cis_trans_combined_conditions.tiff & .png\n\n")
```

## 11. Save annotated tables for each condition

```{r save-tables}
library(data.table)

cat("════════════════════════════════════════════════════════════\n")
cat("SAVING ANNOTATED TABLES\n")
cat("════════════════════════════════════════════════════════════\n\n")

# Save per-condition tables
for (cond_name in names(all_results)) {
  cond_label <- condition_meta[[cond_name]]$label
  clean_label <- tolower(gsub(" ", "_", cond_label))
  combined_cond <- all_results[[cond_name]]
  
  # Save combined harmonized table
  out_file_combined <- file.path(OUT_ANNOT, 
                                 paste0("cis_trans_eQTL_mapping_all_variants_combined_", clean_label, ".txt"))
  fwrite(combined_cond, out_file_combined, sep = "\t", quote = FALSE, row.names = FALSE, na = "NA")
  cat(sprintf("✓ Wrote cis_trans_eQTL_mapping_all_variants_combined_%s.txt — %d rows\n", 
              clean_label, nrow(combined_cond)))
  
  # Save per-variant-type tables
  for (vt in c("SNP", "CNV", "Indel", "SV")) {
    vt_data <- combined_cond %>% filter(variant_type == vt)
    if (nrow(vt_data) > 0) {
      out_file_vt <- file.path(OUT_ANNOT, 
                               paste0(tolower(vt), "_data_annotated_cis_trans_", clean_label, ".txt"))
      fwrite(vt_data, out_file_vt, sep = "\t", quote = FALSE, row.names = FALSE, na = "NA")
      cat(sprintf("  → %s: %d rows\n", basename(out_file_vt), nrow(vt_data)))
    }
  }
  cat("\n")
}

# Save combined across-condition table
fwrite(combined_all, file.path(OUT_ANNOT, "cis_trans_eQTL_mapping_all_variants_all_conditions.txt"),
       sep = "\t", quote = FALSE, row.names = FALSE, na = "NA")
cat(sprintf("✓ Wrote cis_trans_eQTL_mapping_all_variants_all_conditions.txt — %d rows\n\n", 
            nrow(combined_all)))

cat("════════════════════════════════════════════════════════════\n")
cat("ALL PROCESSING COMPLETE ✓\n")
cat("════════════════════════════════════════════════════════════\n")
```

---

## **Key improvements:**

✅ **Single unified codebase** — processes both fluconazole & sc2glu simultaneously  
✅ **Modular design** — `process_condition()` function handles each condition independently  
✅ **Per-condition plots** — separate 2×2 faceted plots for each condition  
✅ **Combined comparison plot** — 2×4 grid view (condition × variant type)  
✅ **Organized outputs** — all tables labeled by condition  
✅ **All variant types** — SNP, CNV, Indel, SV for both conditions  
✅ **Publication-ready** — bold axes, clear legends, high-res output  

Just copy the entire chunk and run! 🎯